\chapter{An Overview of Iridian}\label{ch:overview}

\section{The Iridian language}

Iridian (\ird{ircevná malno}) is a language isolate spoken by approximately 8
million people in the Iridian Republic (\ird{Ircevná respublika}), a small
nation in Central Europe bordering the Czech Republic, Slovakia, Poland, and
Ukraine. It is the sole member of the Iridian language family and is not
demonstrably related to any other known language, although extensive contact
with the surrounding Slavic languages has left a significant mark on its
vocabulary and certain areas of its grammar.

\section{Typological overview}

Iridian is a predominantly head-final, agglutinative language whose morphology
is concentrated chiefly on the verb. Its most characteristic properties are not
independent quirks but parts of a coherent grammatical profile: pervasive
head-finality, a clause structure organised around topic and comment, and a
close interaction between verbal aspect, alignment, and discourse prominence.
The following features are especially central:

\begin{enumerate}[wide]
	\item \textit{Split-ergativity conditioned by aspect domain.} Iridian shows
	a split-ergative alignment conditioned by a binary aspectual opposition.
	The five aspectual categories of the finite verb fall into two higher-order
	domains: the \textit{perfective domain} (perfective and retrospective) and
	the \textit{imperfective domain} (continuous, progressive, and
	contemplative). In the perfective domain the language follows an
	ergative-absolutive pattern, with the patient as absolutive pivot; in the
	imperfective domain it follows a nominative-oblique pattern, with the agent
	as absolutive pivot. This alternation is not merely morphological: it is
	bound up with the language's information structure, since the direct-case
	argument is typically also the clause topic. Chapter~\ref{ch:verbs}
	describes the morphology of the system, and Chapter~\ref{ch:simple} its
	discourse consequences.

	\item \textit{Strong head-finality.} The finite predicate is clause-final in
	the unmarked clause; nominal modifiers precede the noun; adpositions are
	postpositions; and subordinate clauses ordinarily precede the matrix predicate
	that selects them. This same head-final organisation is visible in the noun
	phrase, the verbal complex, and clause combining.

	\item \textit{Topic-comment organisation.} The basic clause is organised as
	topic plus comment. The topic, normally the direct-case argument, occupies the
	left edge of the clause; the comment contains the predicate and associated
	operators. Much of Iridian word order is best understood through this
	information-structural organisation rather than through configurational
	subjecthood alone. The clause's most stable privilege belongs to the
	absolutive pivot, not to a language-wide subject relation independent of
	alignment.

	\item \textit{A layered particle system.} A small but important class of
	preverbal particles marks phasal contour, interrogation, polarity,
	evidentiality, and speaker stance. Their rigid ordering reflects a structured
	clause spine: particles with wider scope occur farther from the predicate,
	while narrowly predicational particles occur closest to it.

	\item \textit{No grammatical gender.} Iridian nouns are not marked for
	gender. Animacy plays a limited role in the pronoun system (see
	Chapter~\ref{ch:nouns}).

	\item \textit{Morphological number.} Iridian encodes singular and plural
	through fused case-number endings. Each declension class provides six
	distinct exponents (three cases $\times$ two numbers); there is no
	separate number particle. Plural forms are obligatory when number is
	grammatically relevant, though bare forms may be used generically or
	when number is discourse-inferrable (see Chapter~\ref{ch:nouns}).

	\item \textit{Pro-drop.} Iridian is a consistently pro-drop language.
	Pronominal arguments are typically omitted and are recoverable from the
	discourse alone; since the verb carries no person morphology, recoverability
	rests entirely on pragmatic and topical factors.

	\item \textit{Aspect as the primary verbal category.} Iridian verbs do not
	mark tense. The primary categories encoded on the verb are aspect and
	alignment, which are fused into a single morpheme, the aspect-alignment
	suffix (Asp.Align). The five aspectual categories of the finite verb are
	organised into a \textit{perfective domain} and an \textit{imperfective
	domain}; these domains differ both in their temporal-semantic profile and
	in the alignment pattern they impose. Temporal reference is established
	through adverbials and context.

		\item \textit{Distinct predication types and a unified negation system.}
		Iridian sharply distinguishes verbal predication, zero-copula equative
		predication, and existential predication. Their negative forms are
		superficially different --- verbal \ird{zá-}, equative \ird{staly}, and
		existential \ird{zad-} --- but the current analysis treats them as different
		phonological realisations of one structural negation domain; see
		Figure~\ref{fig:neg0-comparative} in Chapter~\ref{ch:verbs} for the
		comparative structural summary.
\end{enumerate}

\section{Grammatical architecture}\label{sec:grammatical-architecture}
\label{sec:th-clause-structure}% cross-reference compatibility

At a broad level, the Iridian clause is built from the lexical predicate
outward through a sequence of functional layers. The lowest layers belong to
the predicate itself and to phasal and aspectual operators; above these stand
negation, inflection, focus, interrogation, topic, evidentiality, and
speech-act marking. Because the language is head-final, material from the
highest layers appears farthest to the left in the surface string, while the
predicate closes the clause.

This architecture is especially visible in the particle system of
Chapter~\ref{ch:function-words}. Particles nearest the predicate express
phasal meanings such as \ird{v} \trsl{already} and \ird{s} \trsl{still / not
yet}; above these stand comment-domain particles such as \ird{och}
\trsl{also} and \ird{moc} \trsl{indeed}; above these in turn stands
the interrogative clause-typing particle \ird{by}, which is the overt
exponent of the Pol\textsuperscript{0} head and thus structurally above the
comment domain; and farther left again stand evidential and stance
particles such as \ird{oče}, \ird{hlavdy}, and \ird{ktady}. The linear order
of particles is therefore not arbitrary but a surface reflex of scope and
clause structure. The particle system thus partly reflects the clause spine,
though the key analytical point is that \ird{by} is not merely another
comment-domain operator but the spell-out of a dedicated interrogative
clause-typing projection (\synhead{PolP}) structurally above \synhead{FocP} and below \synhead{TopP}.

The same architecture also helps explain several otherwise disconnected facts.
The final position of the finite verb, the preverbal placement of complements,
the order of subordinate and matrix clauses, and the clause-final position of
\ird{staly} in zero-copula negation all follow from the same head-final
design. Likewise, the alternation between ergative and nominative-style
alignment is not treated here as a purely mechanical case pattern, but as part
of a grammar in which the direct-case argument is normally the one made
prominent as topic.

Table~\ref{tab:th-clause-spine} sets this out formally. It serves as the
standing reference for the formal analyses throughout the grammar; readers
working through Chapters~\ref{ch:verbs}--\ref{ch:complex} will find it useful
to return to.

\begin{table}[ht]
	\footnotesize
	\caption{The Iridian clause spine (innermost to outermost).}\label{tab:th-clause-spine}
	\medskip
	\begin{tabularx}{\textwidth}{l l L}
		\toprule
		{\sc level} & {\sc projection} & {\sc content} \\
		\midrule
		1 (lowest) & \synhead{PhasP} & Phasal contour: attained state (\ird{v}) vs.\ ongoing interval (\ird{s}) \\
		2 & \synhead{PredP} & Lexical predicate; root + aspect-alignment morphology \\
		3 & \synhead{AspP} & Comment-domain operators: additive, exclusive, affirmative,
		                     interim-sequential \\
		4 & \synhead{NegP} & Clausal negation: \ird{zá-} (prefix, verbal predicates);
		                     \ird{staly} (free morpheme, equative/zero-copula predicates);
		                     suppletive root \ird{zad-} (existential predicates, Neg\textsuperscript{0}
		                     absorbed into root) \\
		5 & \synhead{IP} & Inflection: finiteness, mood \\
		6 & \synhead{FocP} & Focus position; optionally hosts one fronted wh-phrase or other focused constituent \\
		7 & \synhead{PolP} & Interrogative clause typing: \ird{by} (overt) / $\varnothing$ (null) \\
		8 & \synhead{TopP} & Topic: typically the absolutive argument \\
		9 & \synhead{EvidP} & Evidential and epistemic operators:
		                      \ird{oče}, \ird{hlavdy}, \ird{ktady}, \ird{či} \\
		10 (highest) & \synhead{SpeechActP} & Illocutionary and discourse-relational
		                      framing: \ird{nadě}, \ird{mlada}, \ird{jse} \\
		\bottomrule
	\end{tabularx}
\end{table}

\noindent Because Iridian is rigorously head-final, specifiers and complements
both precede their heads. The predicate therefore surfaces last; projections
that scope over it surface progressively earlier in the string. Reading the
clause from left to right is thus equivalent to reading the functional sequence
from the highest (\synhead{SpeechActP}) to the lowest (\synhead{PhasP})
projection. This principle provides the primary explanatory tool for
understanding both particle ordering and clause-combining word order; it is
developed in detail in §\,\ref{sec:th-head-finality} of
Chapter~\ref{ch:simple}.

One projection requires special comment. \synhead{NegP} (level~4) has a single
structural position in all Iridian clause types, but its phonological
realisation varies by predicate type. In verbal clauses,
Neg\textsuperscript{0} is spelled out as the procliticized prefix \ird{zá-},
attaching to the left edge of the clause-final verbal complex. In zero-copula
equative clauses, where there is no verbal host, Neg\textsuperscript{0} is the
free morpheme \ird{staly}, surfacing in the clause-final position expected of a
head-final \synhead{NegP} head. In existential clauses,
Neg\textsuperscript{0} triggers suppletive root replacement: the existential
root \ird{jec-} is replaced by \ird{zad-}, and Neg\textsuperscript{0} itself
has no independent PF exponent. These are three phonological reflexes of one
syntactic head, not three independent negation strategies; the structural
position of \synhead{NegP} is the same in all three cases.

The clause fields are schematized in Figure~\ref{fig:th-clause-fields};
Figure~\ref{fig:th-left-periphery} shows the detailed hierarchical structure of
the left periphery.

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}[font=\footnotesize]
		\node[syn band, minimum width=18mm] (top)  {Topic\\\synhead{TopP}};
		\node[syn band, minimum width=42mm, right=0pt of top]  (op)
			{Speech-act \& evidential field\\\synhead{SpeechActP} / \synhead{EvidP}};
		\node[syn band, minimum width=16mm, right=0pt of op]   (pol)
			{Clause typing\\\synhead{PolP}};
		\node[syn band, minimum width=34mm, right=0pt of pol]  (tp)
			{Comment domain\\\synhead{IP} / \synhead{FocP}};
		\node[syn band, minimum width=20mm, right=0pt of tp]   (asp)
			{Phasal\\\synhead{PhasP}};
		\node[syn band, minimum width=14mm, right=0pt of asp]  (pred)
			{Pred.\\\synhead{PredP}};
	\end{tikzpicture}
	\caption{Clause fields of the Iridian clause.}\label{fig:th-clause-fields}
\end{figure}

\begin{figure}[ht]
	\footnotesize
	\caption{Hierarchical structure of the Iridian left periphery (\synhead{SpeechActP}
	to \synhead{IP}).}\label{fig:th-left-periphery}
	\medskip
	\centering
	\begin{forest}
		syntax tree
		[\synhead{SpeechActP}
			[{SpeechAct\\\ird{nadě} / \ird{mlada} / \ird{jse}}]
			[\synhead{EvidP}
				[{Evid\\\ird{oče} / \ird{hlavdy} / \ird{ktady} / \ird{či}}]
				[\synhead{TopP}
					[{\synhead{DP}\\{\footnotesize topic}}]
					[{Top$'$}
						[Top]
						[\synhead{PolP}
							[{Pol\\\ird{by} / $\varnothing$}]
							[{Pol$'$}
								[Pol]
								[\synhead{FocP}
									[{XP\\{\footnotesize\textit{wh / focus}}}]
									[{Foc$'$}
										[Foc]
										[\synhead{IP} [{\ldots}]]
									]
								]
							]
						]
					]
				]
			]
		]
	\end{forest}
\end{figure}

\section{Areal context}

Although Iridian is a language isolate, it has been in sustained contact with
the Slavic languages of Central Europe for over a millennium, and this contact
has left significant traces. Iridian has borrowed extensively from Czech,
Slovak, Polish, and Ukrainian (and, in earlier periods, from German and Latin),
and many structural features of Iridian show the influence of this contact: the
grammaticalisation of the reflexive clitic \ird{se} (a Slavic borrowing),
initial-stress phonology, and a number of phonological processes parallel to
those found in neighboring languages.

Conversely, Iridian has influenced its neighbors, particularly in the
development of certain Slavic phonological and lexical features in the border
dialects. The Iridian Republic's multilingual border regions have historically
produced significant code-switching and calquing in both directions.

\section{Word classes}

Traditional Iridian grammar recognises four main word classes:
\irdp{kaděc}{nouns}, \irdp{hloutěc}{verbs}, \irdp{priděc}{modifiers}, and
\irdp{zvuk}{function words}. This grammar follows the same classification, with
terminological updates where the traditional framework requires clarification.
The modifier class includes demonstratives, quantifiers, and a very small
closed set of lexical \textit{property roots} used in adnominal description.
These roots are not the language's productive strategy for property
predication: that role belongs instead to derived stative verbs, discussed in
Chapter~\ref{ch:verbs}.

Within this traditional four-way classification, the present grammar adopts two
analytical refinements that recur throughout later chapters. First, the verbal
complex is treated as an articulated sequence of functional positions rather
than as an undifferentiated word: stem, voice or applicative morphology,
causative, aspect-alignment, mood, and evidentiality each correspond
to distinct layers of structure. Second, the noun phrase is analysed as an
extended projection whose suffix \ird{-to} is best understood not simply as
`definiteness' in the traditional sense but as an \term{anchoring} marker in
the nominal D-domain. Chapter~\ref{ch:nouns} develops this analysis in detail.

The verb (Chapter~\ref{ch:verbs}) and the noun phrase (Chapter~\ref{ch:nouns})
therefore receive the most detailed treatment, while Chapters~\ref{ch:function-words},
\ref{ch:simple}, and \ref{ch:complex} show how their internal structure
extends outward into the organisation of the clause and of larger discourse.

\section{Conventions used in this grammar}

Iridian words are cited in italics. Glosses follow the Leipzig Glossing Rules; a
full list of abbreviations used in this grammar is provided in Appendix~A.
English translations of Iridian sentences and phrases appear in single quotes.
Reconstructed or hypothetical forms are preceded by an asterisk. Ungrammatical
forms are preceded by an asterisk in parentheses, \ird{*}; forms of questionable
grammaticality are preceded by a question mark, \ird{?}.

In interlinear examples, the case of each noun phrase is shown using the labels
from the Iridian grammatical tradition: \Dir{} (direct case), \Trs{} (transitive
case), \Ind{} (indirect case), and \Voc{} (vocative). These correspond, in
typological terms, to the absolutive, ergative, and oblique cases of a
split-ergative system; that equivalence is noted where relevant in the body of
the grammar.
