\chapter{Phonology}\label{ch:phon}

\section{Introduction}

This chapter describes the phonology of standard Iridian as spoken in Roubže and the
surrounding region. The language shows a distinctly West Slavic profile: an
opposition between hard and soft consonant stems, productive coronal softening,
velar fronting, final devoicing, cluster voicing assimilation, and a dense system
of morphophonemic stem alternations in which the behaviour of weak \ird{e} is
central.

Three distinct phonological and morphophonological domains must be kept separate
from the outset, and the rest of this chapter is organised around that separation.
The first domain is the \term{phonemic consonant inventory}: the set of contrastive
segments that make up underlying representations, including the independent
postalveolar phonemes \ird{č\,š\,ž} and the affricate \ird{c}. The second domain is
\term{soft-stem selection}: the productive process by which coronal obstruents and
the lateral are replaced by their soft counterparts---\ird{ť\,ď\,ň\,ľ}---when
morphology carrying a front-vocalic trigger is affixed. The third domain is
\term{velar fronting and affrication}: a partially distinct process in which /k/
is fronted to /c/ or, before a more restricted class of suffixes, to /č/. These
three domains are not to be conflated. In particular, the postalveolars
\ird{č\,š\,ž} are not the default outputs of productive softening; they are
independent phonemes. The alternation \ird{k} $\rightarrow$ \ird{c / č} is velar
affrication, not coronal softening. And the consonant change associated with the
antipassive suffix \ird{-aš-} is neither of those: it is a separate morphologised
\term{sibilant replacement}, treated in full in Section~\ref{sec:sibilant-replacement}.

The analysis is synchronic throughout. Although several alternations are
historically motivated---the West Slavic areal pressure behind the coronal system
in particular---the goal here is a descriptively adequate synchronic account, not a
historical derivation. A brief note on the historical background is reserved for
Section~\ref{sec:historical-note}.


\section{Vowels}

\subsection{Vowel inventory}

Iridian has six short-long vowel pairs in careful standard speech: /a\,aː/,
/ɛ\,eː/, /ɪ\,iː/, /ɨ\,ɨː/, /ɔ\,oː/, /ʊ\,uː/. Long vowels are tenser than their
short counterparts except for /a\,aː/. Orthographic \orth{i, í} represent the
palatalizing front series /ɪ,\,iː/, while \orth{y, ý} represent the non-palatalizing
central series /ɨ,\,ɨː/.

\begin{table}[ht]
	\footnotesize
	\caption{Vowel inventory of standard Iridian.}\label{tab:vowels}
	\medskip
	\begin{tabularx}{0.9\textwidth}{L C C}
		\toprule
		& {\sc front / central} & {\sc back} \\
		\midrule
		High short & /ɪ/ \ird{i}, /ɨ/ \ird{y} & /ʊ/ \ird{u} \\
		High long & /iː/ \ird{í}, /ɨː/ \ird{ý} & /uː/ \ird{ú} \\
		Mid short & /ɛ/ \ird{e} & /ɔ/ \ird{o} \\
		Mid long & /eː/ \ird{é} & /oː/ \ird{ó} \\
		Low short / long & /a/, /aː/ \ird{a, á} & --- \\
		\bottomrule
	\end{tabularx}
\end{table}

Short \ird{y} /ɨ/ ranges from a near-front lax realisation [ɪ] or [ɪ̞] to a more
central [ɘ̟] or [ɘ], depending on speaker and environment. Its long counterpart
\ird{ý} /ɨː/ is narrower in realisation and most naturally surfaces as [ɨː], with
secondary [ɘː] in some dialects or phonetic environments. In more colloquial speech,
short /ɨ/ remains distinct but tends toward [ɪ \textasciitilde{} ɘ], while long
/ɨː/ is often realised as [iː]; as a result, older \ird{ý} may merge with \ird{í}
in casual urban pronunciation even when short \ird{y} remains distinct from \ird{i}.

\subsection{Minimal pairs}

Table~\ref{tab:vowel-minimal} gives constructed minimal pairs illustrating the vowel
contrasts. The lexemes are invented for expository purposes, but they follow ordinary
native word shapes.

\begin{table}[ht]
	\footnotesize
	\caption{Constructed minimal pairs for the vowel system.}\label{tab:vowel-minimal}
	\medskip
	\begin{tabularx}{\textwidth}{L L L}
		\toprule
		{\sc contrast} & {\sc lexeme 1} & {\sc lexeme 2} \\
		\midrule
		/a \textasciitilde{} aː/ & \phon{sam}{sam}{stake} & \phon{sám}{saːm}{soot} \\
		/ɛ \textasciitilde{} eː/ & \phon{sel}{sɛl}{ledge} & \phon{sél}{seːl}{sap} \\
		/ɪ \textasciitilde{} iː/ & \phon{bit}{bʲɪt}{marker} & \phon{bít}{bʲiːt}{lash} \\
		/ɨ \textasciitilde{} ɨː/ & \phon{byt}{bɘt}{ridge} & \phon{být}{bɨːt}{hut} \\
		/ɔ \textasciitilde{} oː/ & \phon{sot}{sɔt}{crumb} & \phon{sót}{soːt}{orchard} \\
		/ʊ \textasciitilde{} uː/ & \phon{mur}{mʊr}{silt} & \phon{múr}{muːr}{wall} \\
		/ɪ \textasciitilde{} ɨ/ & \phon{bil}{bʲɪl}{hook} & \phon{byl}{bɘl}{ash} \\
		/iː \textasciitilde{} ɨː/ & \phon{bíl}{bʲiːl}{whip} & \phon{býl}{bɨːl}{stump} \\
		\bottomrule
	\end{tabularx}
\end{table}

\subsection{Vowel reduction}

The short vowels /ɛ/ and /ɪ/ reduce toward centralised [ɨ]-like qualities in
unstressed positions, and in rapid speech this can partially overlap with the
ordinary phonetic range of stressed \ird{y}. Final /ɛ/ is not reduced before a
pause. Unstressed /a/ is centralised to [ɐ], and between soft consonants it may
further front.

\subsection{Diphthongs}

Iridian has three phonemic diphthongs: \ird{au}\,/au̯/, \ird{ei}\,/eɪ̯/, and
\ird{ou}\,/ou̯/. In older descriptions /eɪ̯/ stands closer to long \ird{é}; in the
modern language the contrast is maintained chiefly in careful speech and in
morphophonemically conservative items.


\section{Consonants}

\subsection{Consonant inventory}\label{sec:consonant-inventory}

The primary consonant phonemes of standard Iridian are set out in
Table~\ref{tab:consonants}. The inventory presents a clean West Slavic profile.
Surface [g] is not a primary phoneme: it appears in careful loan pronunciation and
as a narrow phonetic realisation of /k/ under lenition (see
Section~\ref{sec:lenition}), but native phonology otherwise treats this series as
/k/.

\begin{table}[ht]
	\footnotesize
	\caption{Primary consonant inventory of standard Iridian.}\label{tab:consonants}
	\medskip
	\begin{tabularx}{0.95\textwidth}{L l l l l l}
		\toprule
		& {\sc labial} & {\sc alveolar} & {\sc postalveolar} & {\sc palatal} & {\sc velar} \\
		\midrule
		Plosive & p, b & t, d & & & k \\
		Nasal & m & n & & ň & \\
		Liquid & & r, l & & & \\
		Fricative & v & s, z & š, ž & & h \\
		Affricate & & c & č & & \\
		Glide & & & & j & \\
		\bottomrule
	\end{tabularx}
\end{table}

The orthographic values of the affricate and postalveolar series must be stated
explicitly, since they differ from the conventions of several neighboring
languages. The letter \orth{c} always represents the alveolar affricate /ts/. It
is never to be reanalysed as a plain plosive or as a generic `soft \ird{t}'. The
letters \orth{č}, \orth{š}, and \orth{ž} represent the postalveolar phonemes
/tʃ/, /ʃ/, and /ʒ/ respectively. These four symbols---\orth{c\,č\,š\,ž}---are
ordinary members of the consonant inventory, not derived alternants. In particular,
\orth{č}, \orth{š}, and \orth{ž} are \emph{not} the default outputs of productive
front-vocalic softening; they are independent phonemes whose distribution in native
vocabulary is not primarily conditioned by adjacent front vowels.

The palatal nasal \ird{ň} /ɲ/ is a full phonemic member of the inventory. It occurs
most robustly before front vowels and in morphologically soft stems, but it also
appears in inherited lexical items before back vowels and in word-final position,
which secures its phonemic status. Unlike \ird{ť\,ď\,ľ}---the other members of the
soft coronal series---\ird{ň} is not merely a predictable morphophonemic variant of
/n/ in standard Iridian. It is therefore listed in the basic inventory rather than
being deferred to the morphophonemic section.

\subsection{Lenition and voicing}\label{sec:lenition}

The velar stop /k/ is frequently realised as [k͡x] after a pause and lenites to
[x] or [ɣ] intervocalically and in certain clusters. This distribution accounts
for most surface voiced velars in native vocabulary. Loanwords may preserve [g] in
careful pronunciation, but in ordinary speech they are commonly adapted with
\orth{h} or with the native lenited series.

Obstruent voicing is phonemic, but clusters obey regressive voicing assimilation,
while liquids and nasals do not themselves voice a preceding obstruent. Word-final
obstruents are devoiced. The same voicing rules apply across morpheme boundaries
and therefore form part of the regular morphophonemic surface cleanup described
in Section~\ref{sec:cluster-repair}.

\subsection{The soft coronal series}\label{sec:soft-coronals}

Standard Iridian distinguishes hard and soft members of four coronal consonant
classes. The four productive soft--hard correspondences are:

\begin{center}
	t $\leftrightarrow$ ť\quad
	d $\leftrightarrow$ ď\quad
	n $\leftrightarrow$ ň\quad
	l $\leftrightarrow$ ľ
\end{center}

These are the only members of the \term{soft coronal series} in the modern standard.
The coronal sibilants /s/ and /z/ do not participate: there is no productive
soft-coronal alternation yielding \ird{ś} or \ird{ź} from underlying /s/ and /z/.
Surface \ird{š} and \ird{ž} are independent postalveolar phonemes (see
Section~\ref{sec:postalveolars}), not softened sibilants.

Soft-coronal softening is triggered by front-vocalic morphology: specifically, by
suffixes whose initial segment is \ird{i}, \ird{í}, \ird{ě} (when following an
unmarkedly plain consonant), or the linker \ird{-e}. In the noun system these are
the front-vocalic oblique endings; in the verb system they include the linker
and the simultaneous converb. The process is therefore best described as
\term{soft-stem selection}: a morphologically conditioned readjustment that replaces
the plain coronal with the corresponding soft variant at the point of Vocabulary
Insertion for the affixed form. Within a Distributed Morphology framework, soft-stem
selection corresponds to a readjustment rule applied to the Root node when a
\mbox{[+soft]} trigger is present in the local Spell-Out domain.

\begin{table}[ht]
	\footnotesize
	\caption{Productive soft coronal correspondences.}\label{tab:soft-coronals}
	\medskip
	\begin{tabularx}{0.9\textwidth}{L L L L}
		\toprule
		{\sc plain} & {\sc soft} & {\sc phonetic value} & {\sc example (plain $\rightarrow$ soft)} \\
		\midrule
		t & ť & /c/ or [tʲ] & \ird{rat} $\rightarrow$ \ird{raťe} \\
		d & ď & /ɟ/ or [dʲ] & \ird{rad} $\rightarrow$ \ird{raďe} \\
		n & ň & /ɲ/ & \ird{ran} $\rightarrow$ \ird{raňe} \\
		l & ľ & /ʎ/ or [lʲ] & \ird{ral} $\rightarrow$ \ird{raľe} \\
		\bottomrule
	\end{tabularx}
\end{table}

Soft \ird{ť\,ď\,ľ} are predictable in most front-vocalic environments, but softness
can also become lexicalised in certain stems or paradigmatic shapes where the
original conditioning environment has been obscured or lost. Where softness is
lexically fixed, the soft letter is retained in all forms of the stem regardless
of the following vowel.

\subsection{Postalveolars as independent phonemes}\label{sec:postalveolars}

The postalveolar series \ird{č\,š\,ž} /tʃ\,ʃ\,ʒ/ must be explicitly distinguished
from the soft coronals described in the preceding section. Three points of
separation are essential.

First, \ird{č\,š\,ž} are not derived. They are underlying phonemes that appear in
lexical items without any front-vocalic conditioning: \ird{šum} `noise', \ird{žen}
`wife (transitive)', \ird{čer} `border'. The soft coronals \ird{ť\,ď\,ň\,ľ}, by
contrast, alternate with their plain counterparts across morphological paradigms.

Second, the alternation \ird{t} $\rightarrow$ \ird{ť} and the identity
\ird{š} $\rightarrow$ \ird{š} are categorically different: the first is a morphophonemic
operation; the second is a lexical phoneme remaining stable. There is no productive
rule deriving \ird{š} from /s/ before front vowels in standard Iridian.

Third, sequences of postalveolar plus front vowel are entirely normal in native
vocabulary and do not trigger any further mutation. The word \ird{šira} `silver
(direct)' has /ʃi-/ at its surface; there is no deeper softening to be read
into the initial consonant.

Table~\ref{tab:soft-vs-postalv} makes the contrast explicit. The left-hand column
shows the productive soft-coronal alternation; the right-hand column shows the
independent postalveolars.

\begin{table}[ht]
	\footnotesize
	\caption{Soft coronals (morphophonemic) versus postalveolars (phonemic).}\label{tab:soft-vs-postalv}
	\medskip
	\begin{tabularx}{\textwidth}{L L L L}
		\toprule
		\multicolumn{2}{c}{\sc soft coronals (alternating)} & \multicolumn{2}{c}{\sc postalveolars (stable)} \\
		{\sc plain} & {\sc before front-vocalic suffix} & {\sc phoneme} & {\sc example} \\
		\midrule
		t & ť & š /ʃ/ & \ird{šum} `noise' \\
		d & ď & ž /ʒ/ & \ird{žen} `wife (\Trs)' \\
		l & ľ & č /tʃ/ & \ird{čer} `border' \\
		n & ň & & \\
		\bottomrule
	\end{tabularx}
\end{table}

For minimal contrast between the two series, compare \ird{tě} [cɛ] (the sequence
of soft \ird{ť} followed by \ird{e}, as in a softened nominal stem) with \ird{č}
[tʃ] (the postalveolar phoneme), or \ird{ľe} [ʎɛ] (soft lateral plus vowel) with
\ird{ž} [ʒɛ] (postalveolar fricative in the same vowel environment). The
orthographic and phonetic distinction is maintained consistently in the standard.

\subsection{Positional palatalisation of /s/, /z/, and /c/}\label{sec:sibilant-palatalisation}

The coronal sibilants /s/ and /z/ and the affricate /c/ undergo a surface phonetic
palatalisation immediately before the palatalizing high vowel series /i\,iː/
(orthographic \ird{i\,í}). In this environment /s/ is realised as [ś], /z/ as [ź],
and /c/ as [tɕ]. The rule is restricted to the high palatalizing series: it does not
apply before the non-palatalizing central series /ɨ\,ɨː/ (orthographic \ird{y\,ý}),
nor before mid front \ird{e} or \ird{ě}, where any softening effect is weak and
unpredictable.

This is a post-lexical phonetic rule, categorically distinct from the morphophonemic
soft-stem selection described in Section~\ref{sec:front-softening}. Three properties
mark the distinction. First, the rule is not morphologically conditioned: it applies
wherever the relevant segments immediately precede /i\,iː/, regardless of suffix
class membership. Second, the outputs [ś\,ź\,tɕ] are entirely predictable from
context and create no phonemic contrast with /s\,z\,ts/. Third, the rule is not
written: standard orthography spells the sequences as \ird{si}, \ird{zi}, \ird{ci}
without any diacritic, and the palatalisation is supplied automatically by the
reader from the following vowel letter.

The outputs [ś\,ź\,tɕ] are phonetically distinct from the postalveolars
[ʃ\,ʒ\,tʃ]: they are higher and more front in place of articulation, not retracted
to the postalveolar region. This distinction matters for the interpretation of
Section~\ref{sec:postalveolars} and of the sibilant restrictions in
Section~\ref{sec:front-softening}. Those sections bar a \emph{morphophonemic}
derivation of /ʃ\,ʒ/ from /s\,z/ in front-vocalic environments; they say nothing
about this surface phonetic effect. The two restrictions occupy different strata of
the grammar and do not interact. In terms of rule ordering, positional palatalisation
applies after all morphophonemic processes have completed, including voicing cleanup,
and never feeds or bleeds any morphological rule.


\section{Phonotactics}

\subsection{Syllable structure}

The basic syllable structure is CV(C). The maximum is CCCCVCCC, though most syllables
fall into one of CV, CVC, CCV, CCVC, or CVCC.

\subsection{Syllabic liquids}

Like its West Slavic neighbors, Iridian permits liquids to bear a syllable in
reduced environments. The rhotic and the lateral may surface as syllabic [r̩] and
[l̩] when weak \ird{e} is deleted and the language chooses not to restore a vowel.
This is especially natural at morpheme boundaries and in fast speech, though
orthography remains conservative and normally writes only the consonantal letters.

Syllabic liquids are not a free alternative to ordinary vocalism. They appear chiefly
when three conditions are met: the deleted vowel is weak \ird{e}; the resulting
cluster already contains a liquid; and no simpler repair such as straightforward
cluster retention is available. In this way the language accommodates dense consonant
groupings without abandoning its preference for preserving inherited stem shapes.

\subsection{Stress}

Iridian has initial stress. Stress falls on the first syllable of the phonological
word, including prefixes. Clitics are unstressed and form a phonological unit with
the following word.


\section{Suprasegmentals}

\subsection{Intonation}

Declaratives are characterised by a falling contour, yes-no questions by a final
rise, and focused contrastive phrases by an expanded pitch excursion on the focused
constituent. Detailed discourse intonation is treated in Chapter~\ref{ch:discourse}.


\section{Morphophonemic processes}\label{sec:morphophon}

Iridian morphology is surface-rich but structurally regular. The same rule system
applies across noun inflection, verb derivation, converbs, clause-linking forms,
and nominalisations, though nouns expose it most densely. Before proceeding to the
individual processes, it is essential to recall the three-way separation stated in
the introduction. Every consonant alternation in the grammar belongs to exactly one
of three categories:

\begin{enumerate}[noitemsep]
	\item \term{Soft-stem selection}: coronal softening t $\rightarrow$ ť, d $\rightarrow$ ď,
	n $\rightarrow$ ň, l $\rightarrow$ ľ, triggered by ordinary front-vocalic morphology.

	\item \term{Velar fronting / affrication}: the alternation k $\rightarrow$ c or
	k $\rightarrow$ č, a distinct process with its own two-class suffix conditioning
	(Section~\ref{sec:velar-fronting}).

	\item \term{Sibilant replacement}: the morphologised stem mutation triggered by
	the antipassive \ird{-aš-} and its derivatives, not conditioned by front vowels
	(Section~\ref{sec:sibilant-replacement}).
\end{enumerate}

A fourth process also operates on consonants but belongs to a different stratum of
the grammar: the positional palatalisation of /s\,z\,c/ before /i\,iː/, which is a
post-lexical phonetic rule and not a morphophonemic alternation
(Section~\ref{sec:sibilant-palatalisation}). It is listed separately because it
creates no orthographic contrast, no phonemic opposition, and no morphological
conditioning, and therefore cannot be confused with the three categories above.

In the modern language the morphophonemic processes apply in the following ordered
sequence:

\begin{center}
\textsc{suffix class selection} $\rightarrow$ \textsc{weak-e deletion or retention}
$\rightarrow$ \textsc{grade formation} $\rightarrow$ \textsc{soft-stem selection}
$\rightarrow$ \textsc{velar fronting / affrication} $\rightarrow$
\textsc{cluster repair / epenthesis} $\rightarrow$ \textsc{voicing cleanup}
$\rightarrow$ \textsc{orthographic normalisation} $\rightarrow$
\textsc{analogical levelling} $\rightarrow$ \textsc{positional sibilant palatalisation}
\end{center}

Sibilant replacement belongs to a separate derivational block and does not feed the
above chain in either direction, except that the output of replacement may
subsequently undergo voicing cleanup and cluster repair. The ordering matters. A
noun such as \ird{lobek} first selects the transitive ending of its declension
class, then deletes weak \ird{e}, then raises \ird{o} to \ird{u}, and only after
that reaches its final cluster shape. A noun of the type \ird{perek} first deletes
weak \ird{e}, then forms the liquid cluster \ird{prk-}, and only then undergoes
velar affrication before a front-vocalic indirect ending, yielding \ird{prce}. A
verb such as \ird{zděk-} behaves the same way before a palatalizing suffix, except
that the palatalised grade rather than a nominal oblique stem is the output.

\subsection{Suffix classes and their mutation triggers}\label{sec:suffix-classes}

The most useful synchronic generalisation is that suffixes belong to morphophonemic
trigger classes. A suffix does not merely add segmental material; it also selects a
particular type of stem. Table~\ref{tab:morph-trigger} reorganises suffix classes
by the type of stem mutation they impose. The first three classes are the primary
productive ones. The sibilant-replacing class is kept separate because it belongs
to a different derivational block.

\begin{table}[ht]
	\footnotesize
	\caption{Morphophonemic trigger classes organised by mutation type.}\label{tab:morph-trigger}
	\medskip
	\begin{tabularx}{\textwidth}{L L L}
		\toprule
		{\sc trigger class} & {\sc typical suffixes} & {\sc mutation imposed} \\
		\midrule
		Plain & direct zero, continuous \ird{-a}, retrospective \ird{-ac} & no stem mutation \\
		Zero-grade / grade-forming & noun \ird{-u}, \ird{-a}, \ird{-am} obliques; vowel-initial nominalizers; irrealis \ird{-ěv}; selected nominalizers & weak-\ird{e} deletion; optionally feeds vowel raising \\
		Ordinary softening & noun \ird{-í}, \ird{-e}, \ird{-ě}, \ird{-ení}; converb \ird{-ěc}; linker \ird{-e} & soft-stem selection (t $\rightarrow$ ť, d $\rightarrow$ ď, n $\rightarrow$ ň, l $\rightarrow$ ľ) and k $\rightarrow$ c \\
		Strong palatalizing & verbal \ird{-ěv}; a restricted nominal class & k $\rightarrow$ č (in addition to coronal softening) \\
		Sibilant-replacing & antipassive \ird{-aš-} & consonant replacement per Table~\ref{tab:sibilant-mutation}; not front-vocalic softening \\
		Leveled & high-frequency obliques, conservative lexical items & preserves an analogical output \\
		\bottomrule
	\end{tabularx}
\end{table}

The two front-vocalic rows---ordinary softening and strong palatalizing---together
constitute the broad front-vocalic domain of the grammar. They share soft-stem
selection for coronals. They differ in what they do to a velar: ordinary softening
yields \ird{c} (/ts/), while strong palatalizing yields \ird{č} (/tʃ/). This
difference is morphologised at the level of suffix membership, not derived from
the phonological content of the vowel alone.

The negative irrealis \ird{-ir} occupies an intermediate position: it belongs to
the ordinary softening class for velar stems (k $\rightarrow$ c), but it lacks
the vowel-raising and the strong-palatalizing properties of \ird{-ěv}. Any suffix
not listed here that appears to trigger an alternation should be cross-checked
against the relevant morphology chapter before a trigger class is assigned.

\subsection{Vowel-replacing nominal endings}

Not all nominal alternations are driven by deletion alone. A small but productive
set of noun classes selects endings that replace a final stem vowel rather than
simply attaching to it. This is most visible in inherited \ird{-o} stems, where
the transitive replaces the final vowel with \ird{-a} and the indirect adds a
nasal element that coalesces with the preceding back vowel:

\pex
\a \ird{bro} + \ird{-a} $\rightarrow$ \ird{bra}
\a \ird{bro} + \ird{-m} $\rightarrow$ \ird{brum}
\xe

Synchronically the useful generalisation is \ird{o + m > um}. The rule belongs
to the noun system rather than to the verbal domain and should be treated as a
class-specific back-vocalic contraction.

\subsection{Weak \ird{e} and zero-grade formation}\label{sec:vowel-zero}

The central morphophonemic process of modern Iridian is the behaviour of \term{weak}
\ird{e}. Synchronically this vowel is not confined to a few fossilised stems. It
remains highly productive in nouns and still active, though less pervasive, in
verbs and derived forms. Weak \ird{e} is deleted before many vowel-initial and
front-vocalic suffixes whenever the resulting cluster is phonotactically acceptable.

The core environment is a final syllable of the shape -C(\sx{j})eC, especially
before case endings and other suffixes that begin with a vowel. In structural
terms the process is comparable to the West Slavic type illustrated by
\foreign{pes} $\rightarrow$ \foreign{psa}: a weak mid vowel in the final stem
syllable is lost when a vowel-initial oblique ending follows and the resulting
consonant sequence is acceptable.

\pex
\a \ird{pes} + \ird{-a} $\rightarrow$ \ird{psa}
\a \ird{Janek} + \ird{-u} $\rightarrow$ \ird{Janku}
\a \ird{perek} + \ird{-u} $\rightarrow$ \ird{prku}
\a \ird{obel} + \ird{-u} $\rightarrow$ \ird{oblu}
\a \ird{Merek} + \ird{-í} $\rightarrow$ \ird{Mercí}
\xe

Deletion applies most readily when the resulting cluster belongs to one of the
cluster shapes already tolerated by Iridian phonotactics: obstruent + sonorant,
sibilant + obstruent, sonorant + obstruent, or a cluster that may be resolved with
syllabic \ird{r} or \ird{l}. Deletion is blocked in exactly three circumstances.
First, some declension classes and some individual stems are lexically marked as
non-deleting. Second, deletion is blocked when it would create an illicit cluster,
above all a dense all-obstruent cluster without a licensed sibilant edge or liquid
nucleus. Third, a small set of high-frequency forms preserves an analogically
levelled vowel despite the general rule.

The productivity of weak-\ird{e} deletion is greatest in noun oblique formation.
A particularly important subtype is the \term{CeLeC} pattern, where deletion
yields a stable \ird{CLC-} oblique stem: \ird{perek} $\rightarrow$ \ird{prku}
(transitive), \ird{prce} (front-vocalic indirect), \ird{prkam} (\ird{-am} class).
In verbal morphology the process is most visible before the irrealis, the
simultaneous converb, the linker \ird{-e}, and selected nominalizers:

\pex
\a \ird{zděk-} + \ird{-ěv} $\rightarrow$ \ird{zdíčev}
\a \ird{perek} + \ird{-ěc} $\rightarrow$ \ird{prčec}
\a \ird{obel} + \ird{-ou} $\rightarrow$ \ird{oblou}
\xe

\subsection{Grade formation and vowel raising}\label{sec:vowel-raising}

When weak-\ird{e} deletion creates a heavy cluster, the stem may enter a raised
grade. This is a morphologised grade alternation, not a mere phonetic tendency. Two
productive correspondences dominate the modern language:

\begin{itemize}
	\item front-vowel raising: \ird{e/ě/é} $\rightarrow$ \ird{i/í}
	\item back-vowel raising: \ird{o} $\rightarrow$ \ird{u}
\end{itemize}

In the noun system the raised grade is lexically tied to particular declension
classes. In the verb system it is much rarer but still productive before a narrow
set of front-vocalic suffixes and in some nominalizing formations. Because several
noun classes have undergone areal reshaping under prolonged Slavic contact, raising
may coexist with newer analogical endings such as \ird{-ení} or \ird{-ovi}; the
stem grade and the case ending must therefore be treated as partially independent
layers.

\pex
\a \ird{lobek} + \ird{-u} $\rightarrow$ \ird{lubku} \lingcontext{back-vowel raising after weak-\ird{e} deletion}
\a \ird{pěštel} + \ird{-í} $\rightarrow$ \ird{píštlí} \lingcontext{front-vowel raising in a noun oblique}
\a \ird{zděk-} + \ird{-ěv} $\rightarrow$ \ird{zdíčev} \lingcontext{front-vowel raising in the verbal domain}
\xe

Not every zero-grade form raises. Whether raising applies is determined first by
declension class or stem class, and only secondarily by phonological environment.
The important synchronic generalisation is not `delete weak \ird{e}, then
optionally raise' but `some classes select a raised oblique stem once weak
\ird{e} has been lost'.

\subsection{Soft-stem selection}\label{sec:front-softening}

Soft-stem selection is the productive process by which the coronal stops and lateral
/t\,d\,l/ are replaced by their soft counterparts \ird{ť\,ď\,ľ}, and underlying
/n/ surfaces as \ird{ň}, when a front-vocalic suffix is attached. Labial
consonants /p\,b\,m\,v/ and /r/ show secondary palatal articulation in the same
environments but are not replaced by distinct phonemes and are therefore not
represented with special letters.

The trigger class for ordinary softening was defined in
Section~\ref{sec:suffix-classes}. It includes noun \ird{-í}, \ird{-e}, \ird{-ě},
\ird{-ení}, the converb \ird{-ěc}, and the verbal linker \ird{-e}. The output of
soft-stem selection is therefore consistently written with the explicit soft letters
\ird{ť\,ď\,ľ\,ň} in the standard orthography (see Section~\ref{sec:orthography}).

\pex
\a \ird{rat} + \ird{-e} $\rightarrow$ \ird{raťe} \lingcontext{t $\rightarrow$ ť before linker}
\a \ird{rad} + \ird{-í} $\rightarrow$ \ird{raďí} \lingcontext{d $\rightarrow$ ď before nominal \ird{-í}}
\a \ird{ran} + \ird{-ě} $\rightarrow$ \ird{raňě} $\rightarrow$ \ird{raňe} \lingcontext{n $\rightarrow$ ň; ě normalised to e after ň}
\a \ird{ral} + \ird{-e} $\rightarrow$ \ird{raľe} \lingcontext{l $\rightarrow$ ľ before linker}
\xe

The coronal sibilants /s/ and /z/ and the affricate /c/ are not subject to
soft-stem selection. No morphophonemic rule derives \ird{š} from /s/ or \ird{ž}
from /z/ in front-vocalic environments: any surface \ird{š} or \ird{ž} before a
front vowel reflects either an underlying postalveolar phoneme or the output of
sibilant replacement (Section~\ref{sec:sibilant-replacement}). These segments do,
however, undergo a surface phonetic palatalisation before /i\,iː/, producing
[ś\,ź\,tɕ] respectively. That process is post-lexical, leaves the orthography
unchanged, and is described in Section~\ref{sec:sibilant-palatalisation}. The two
statements are compatible: soft-stem selection does not apply to /s\,z\,c/, and the
phonetic palatalisation is not soft-stem selection.

Soft-stem selection applies after zero-grade and grade formation. If a noun both
deletes weak \ird{e} and takes a front-vocalic ending, the language first
constructs the zero or raised stem and only then applies softening:

\pex
\a \ird{lobek} + \ird{-í} $\rightarrow$ \ird{lubcí} \lingcontext{zero-grade; back-vowel raising; then velar fronting}
\a \ird{Janek} + \ird{-í} $\rightarrow$ \ird{Jancí} \lingcontext{zero-grade; then velar fronting}
\a \ird{perek} + \ird{-e} $\rightarrow$ \ird{prce} \lingcontext{CeLeC zero-grade; then velar fronting}
\xe

Note that these examples involve velar fronting (\ird{k} $\rightarrow$ \ird{c}),
treated separately below, rather than soft-stem selection proper. They are included
here to make the ordering relation clear.

Two verbal suffixes qualify the generalisation. The conditional converb \ird{-ic}
is wholly exceptional: despite its front vowel, it does not trigger soft-stem
selection. The negative irrealis \ird{-ir} is only partially integrated into the
front-vocalic system: it triggers velar fronting (k $\rightarrow$ c) but lacks
the broader coronal softening and vowel raising associated with \ird{-ěv}. Both
patterns are lexicalised irregularities preserved in the modern language and should
be treated as exceptional allomorphy rather than productive rule interaction.

\subsection{Velar fronting and affrication}\label{sec:velar-fronting}

Velar fronting is a distinct process from coronal soft-stem selection. It applies
to underlying /k/ and converts it to an affricate: /c/ (= /ts/) in the ordinary
case, or /č/ (= /tʃ/) before a more restricted class of suffixes. The change is
genuinely phonological---a velar stop becomes an alveolar or postalveolar
affricate---and is not merely a graphic convention.

Since /g/ is absent from the native phoneme inventory, the productive alternation
is built around /k/ alone. The voiced counterpart of this alternation, which in
a full West Slavic series would pair /g/ with /ʒ/, is thus largely absent from
native vocabulary. For /h/, see the note at the end of this section.

\subsubsection{Two-class analysis}

The key generalisation is that front-vocalic suffixes form two classes with respect
to the velar, and that this distinction is morphologically specified rather than
phonologically derivable from the vowel alone:

\begin{enumerate}[noitemsep]
	\item \term{Ordinary softening suffixes}: these yield k $\rightarrow$ c. They
	constitute the majority of the front-vocalic class. Examples: nominal \ird{-í},
	nominal \ird{-e}, nominal \ird{-ě}, nominal \ird{-ení}, negative irrealis \ird{-ir},
	converb \ird{-ěc}.

	\item \term{Strong palatalizing suffixes}: these yield k $\rightarrow$ č. This
	is a smaller, morphologically specified class. In the current grammar the principal
	member is the verbal suffix \ird{-ěv} (irrealis / modal stem). These suffixes
	also trigger front-vowel raising and the full range of coronal softening.
\end{enumerate}

\begin{table}[ht]
	\footnotesize
	\caption{Velar fronting by suffix class.}\label{tab:velar-fronting}
	\medskip
	\begin{tabularx}{\textwidth}{L L L L}
		\toprule
		{\sc suffix class} & {\sc velar output} & {\sc example suffixes} & {\sc example derivation} \\
		\midrule
		Ordinary softening & k $\rightarrow$ c & \ird{-í}, \ird{-e}, \ird{-ě}, \ird{-ení}, \ird{-ir}, \ird{-ěc} & \ird{Marek} + \ird{-í} $\rightarrow$ \ird{Marcí} \\
		& & & \ird{ruk} + \ird{-ě} $\rightarrow$ \ird{ruce} \\
		& & & \ird{zděk-} + \ird{-ir} $\rightarrow$ \ird{zděcir} \\
		Strong palatalizing & k $\rightarrow$ č & \ird{-ěv} & \ird{zděk-} + \ird{-ěv} $\rightarrow$ \ird{zdíčev} \\
		& & & \ird{perek} + \ird{-ěc} $\rightarrow$ \ird{prčec} \\
		\bottomrule
	\end{tabularx}
\end{table}

The contrast between the two classes is most clearly seen by pairing examples with
the same stem but different suffixes. The stem \ird{zděk-} yields \ird{zděcir}
before ordinary \ird{-ir} but \ird{zdíčev} before strong \ird{-ěv}. This
difference is synchronically morphologised: both suffixes begin with a front vowel,
yet they select categorically different velar outputs. The reader who encounters
an unfamiliar suffix should assign it to one of these two classes based on the
attested velar outcome, not on the surface phonology of the suffix vowel alone.

\subsubsection{Treatment of h}

The fricative /h/ does not participate productively in velar fronting. In a
historically fuller Slavic-type system, /h/ would be expected to front to /š/ or /ś/
before front-vocalic morphology, parallel to the Czech and Slovak series. Standard
Iridian does not maintain this rule productively. Surface \ird{š} before a
front-vocalic ending involving an original /h/-final stem reflects either a
lexicalised historical residue or borrowing, not a productive synchronic rule. For
descriptive purposes, /h/ before front-vocalic suffixes simply retains its
underlying value; any exceptions are listed individually in the lexicon.

\subsection{Sibilant replacement}\label{sec:sibilant-replacement}

Sibilant replacement is a morphologised stem mutation triggered chiefly by the
antipassive \ird{-aš-}. It is emphatically \emph{not} ordinary front-vowel
softening. Its trigger is not a front vowel but a specific derivational morpheme;
its outputs include \ird{ž} and \ird{š}, which are postalveolar phonemes, not soft
coronals; and it affects the thematic consonant of the verb stem, not any consonant
adjacent to a front vowel. Every time a consonant alternation is encountered in an
antipassive derivative, the reader should refer to this section rather than to the
soft-stem selection or velar fronting sections.

The mutation targets the thematic consonant only and follows the classes in
Table~\ref{tab:sibilant-mutation}:

\begin{table}[ht]
	\footnotesize
	\caption{Sibilant replacement classes triggered by antipassive \ird{-aš-}.}\label{tab:sibilant-mutation}
	\medskip
	\begin{tabularx}{0.9\textwidth}{l L L L}
		\toprule
		{\sc class} & {\sc thematic consonants} & {\sc sibilant output} & {\sc example} \\
		\midrule
		I    & -t, -k, -c, -č & $\rightarrow$ -č   & \ird{piašt-} $\rightarrow$ \ird{piašč-} \\
		II   & -s, -š         & $\rightarrow$ -š   & \ird{tras-} $\rightarrow$ \ird{traš-} \\
		II-A & -l, -p         & + -š   & \ird{dal-} $\rightarrow$ \ird{dalš-} \\
		III  & -d, -h, -j, -z, -ž & $\rightarrow$ -ž & \ird{vad-} $\rightarrow$ \ird{váž-} \\
		III-A & all remaining & + -ž   & \ird{ščen-} $\rightarrow$ \ird{ščenž-} \\
		\bottomrule
	\end{tabularx}
\end{table}

In Classes I and III the thematic consonant is replaced; in Classes II-A and
III-A the sibilant is added after the preserved consonant. Surface [g]-bearing
loans, where they participate in this pattern at all, follow the voiced
velar/fricative series and are assigned analogically to Class~III. Once replacement
has applied, later voicing cleanup and cluster repair may still affect the output
if the following suffix requires it.

The wording `sibilant replacement' is used consistently in the rest of this
grammar when referring to the antipassive mutation, and `softening' is reserved
for the productive coronal process. These terms must not be used interchangeably.

\subsection{Cluster repair, epenthesis, and voicing cleanup}\label{sec:cluster-repair}

The productivity of weak-\ird{e} deletion would produce unwieldy strings if it
were not paired with strong surface-repair rules. Iridian is permissive about
clustering by European standards, but it still regularises derived clusters in a
principled way. The language prefers to keep the full consonant sequence when
possible, then to exploit syllabic liquids, then to simplify or assimilate, and
only last of all to restore a vowel.

\begin{table}[ht]
	\footnotesize
	\caption{Regularization hierarchy at morpheme boundaries.}\label{tab:cluster-repair}
	\medskip
	\begin{tabularx}{\textwidth}{l L L}
		\toprule
		{\sc rank} & {\sc strategy} & {\sc effect} \\
		\midrule
		1 & retain the cluster & keep the result of deletion if the cluster is otherwise licit \\
		2 & syllabic liquid formation & realise \ird{r} or \ird{l} as [r̩] or [l̩] in liquid-bearing clusters \\
		3 & assimilation or simplification & spread voicing, soften leftward locally, or delete the weakest coronal in an illicit all-obstruent group \\
		4 & epenthesis / vowel retention & restore a short vowel only when no consonantal repair yields a well-formed output \\
		\bottomrule
	\end{tabularx}
\end{table}

Four repairs are central in the modern language. First, obstruent clusters obey
regressive voicing assimilation, including across morpheme boundaries:

\pex
\a \ird{obez + -ba} $\rightarrow$ \ird{obezba}, phonetically [obezba]
\a \ird{otk + -da} $\rightarrow$ [ogda]
\a \ird{vrad + -ka} $\rightarrow$ [vratka]
\xe

Second, word-final obstruents are devoiced regardless of the underlying consonant
or the derivational history of the word. Third, when deletion produces a cluster
containing a liquid, Iridian often preserves the deletion and allows the liquid
to become syllabic rather than restoring the vowel:

\pex
\a \ird{berel} + \ird{-a} $\rightarrow$ [br̩la] or orthographic \ird{brla}
\a \ird{perek} + \ird{-am} $\rightarrow$ [pr̩kam] in the most reduced liquid-bearing type
\xe

Fourth, when deletion yields an illicit all-obstruent cluster, the language repairs
it by assimilation or simplification before resorting to epenthesis. Softness may
also spread leftward one segment inside a cluster, especially when a coronal
directly precedes a newly soft consonant. This spread is limited and not always
reflected orthographically. If no consonantal repair succeeds, a short epenthetic
vowel is inserted or the weak vowel is retained. Epenthesis is therefore a last
resort, not a general stem-building strategy.

\subsection{Compensatory lengthening}\label{sec:comp-lengthening}

When a thematic consonant is replaced rather than merely augmented, the preceding
vowel may lengthen if the replacement would otherwise reduce the moraic weight of
the stem. In the modern language this rule remains useful chiefly in older
Class~I and Class~III verbal stems under true sibilant replacement. It is therefore
not a general repair strategy, but a restricted morphologised contrast between
replacement stems and additive stems.

\pex
\a \ird{vad-} + \ird{-aš-} $\rightarrow$ \ird{váž-} \lingcontext{replacement class; vowel lengthens}
\a \ird{dal-} + \ird{-aš-} $\rightarrow$ \ird{dalš-} \lingcontext{additive class; no lengthening}
\xe

Orthographically the lengthening is written with the regular acute accent.

\subsection{Orthography and the soft consonant letters}\label{sec:orthography}

Standard Iridian orthography now writes the soft coronal consonants with their
explicit letters. This is a deliberate departure from an earlier conservative policy
that avoided háček letters for \ird{t\,d\,l} and relied on the following vowel to
signal softness indirectly. Under the current standard:

\begin{itemize}
	\item Soft \ird{ť}, \ird{ď}, \ird{ľ}, and \ird{ň} are written whenever the
	consonant is phonologically soft, whether the softness is predictable or
	lexicalised.
	\item \ird{č}, \ird{š}, \ird{ž} are reserved for the independent postalveolar
	phonemes. They carry no implication of derivation from a plain coronal.
	\item \ird{c} is always /ts/ and is never to be reinterpreted as a plosive or
	as a soft \ird{t}. The sequence \ird{c} + front vowel simply represents the
	affricate followed by that vowel; no softening is implied.
\end{itemize}

The practical effect is that orthography now directly reflects the phonological
distinction between soft coronals and postalveolars, rather than obscuring it
behind an ambiguous vowel letter.

\subsection{The letter ě and orthographic normalisation}\label{sec:e-hook}

The letter \ird{ě} has a specific and delimited function. It is written only
where it signals that an otherwise plain (i.e.\ unmarkedly hard) consonant is
soft in this form. The letter is \emph{not} written after consonants that are
already explicitly soft or already fronted, because those consonants already carry
softness information in their own letter and a redundant diacritic adds no
information and violates the consistency of the system.

The rule is:

\begin{quote}
\textit{Write \ird{ě} only after a plain hard consonant. After a consonant that is
already soft or postalveolar, write plain \ird{e}.}
\end{quote}

This yields the normalised spellings in Table~\ref{tab:e-norm}.

\begin{table}[ht]
	\footnotesize
	\caption{Orthographic normalisation: \ird{ě} versus \ird{e} by preceding consonant.}\label{tab:e-norm}
	\medskip
	\begin{tabularx}{0.9\textwidth}{L L L}
		\toprule
		{\sc preceding consonant} & {\sc spelling with front vowel} & {\sc rationale} \\
		\midrule
		Plain t, d, n, l, p, b, \ldots & tě, dě, ně, lě, pě, bě & \ird{ě} signals softening of an unmarked consonant \\
		Explicit soft ť, ď, ľ, ň & ťe, ďe, ľe, ňe & softness already marked; write plain \ird{e} \\
		Postalveolar č, š, ž & če, še, že & postalveolars do not softened further; write plain \ird{e} \\
		Affricate c & ce & /ts/ is already fronted; write plain \ird{e} \\
		Glide j & je & follows the same pattern \\
		\bottomrule
	\end{tabularx}
\end{table}

The sequences \ird{*cě}, \ird{*čě}, \ird{*šě}, \ird{*žě}, and \ird{*ňě} are thus
not used in standard native orthography. Where such sequences would arise in
derivation---for instance when the nominal ending \ird{-ě} is attached to a stem
ending in \ird{c}---orthographic normalisation replaces them with the plain-\ird{e}
form as the last step in the morphophonemic chain. Similarly, once \ird{ľ} is
written explicitly as the soft lateral, the form \ird{ľ + ě} is normalised to
\ird{ľe} rather than \ird{*ľě}; the redundant diacritic is dropped.

\subsection{The letter ř}\label{sec:r-hook}

The letter \ird{ř} belongs to the orthographic and learned stratum of the language
rather than to core productive phonology. It does not represent a value equivalent
to Czech /r̝/ (the raised rhotic fricative). Its value in Iridian is a simpler
one, broadly equivalent to a rhotic cluster or to the value of the corresponding
grapheme in Polish \foreign{rz}, which is phonetically [ʒ] in most environments.

Because \ird{ř} does not introduce a new phonemic distinction beyond what is
already expressible with \ird{rž} or simply with \ird{ž} in the relevant
environments, it is not listed in the phonemic inventory table
(Table~\ref{tab:consonants}). Its use is confined to learned vocabulary, etymological
spellings, and items from a stratum of cultural borrowing associated with the
historical Czech-Iridian contact zone (see Section~\ref{sec:historical-note}). In
native alternating paradigms, \ird{ř} does not participate as an input or output
of any productive rule. The reader who encounters \ird{ř} in a text should treat
it as a spelling variant of \ird{rž} and not attempt to assign it a distinct
phonemic status in native morphology.

\subsection{Analogical levelling and historical residue}

The modern morphophonemics contain a natural but limited cluster of irregularities.
These arise from four sources: analogical levelling, differential preservation of
weak-\ird{e} deletion, suffix-specific affrication spellouts, and areal reshaping.
Several noun classes have generalised longer oblique endings such as \ird{-ení}
and animate-like endings such as \ird{-ovi} under prolonged contact pressure from
neighboring Slavic systems. The stem grade and the case ending must therefore be
treated as partially independent layers in these classes. The strongest irregularity
is concentrated in high-frequency nouns, animate lexemes, clause-linking forms,
and a few older verbal derivatives. The core noun case system nevertheless remains
rule-driven once the declension class of the lexeme is known.

\subsection{Perfective suffix dissimilation}\label{sec:pf-dissim}

The bare intransitive perfective suffix \ird{-ek} raises to \ird{-ik} when the
stem's final syllable has \ird{ě} or \ird{e} as its nucleus, avoiding a CěCe or
CeCe sequence across a single intervening consonant. The raising does not apply
when a voice morpheme intervenes between the stem and the perfective suffix. This
alternation belongs to the same general tendency toward grade optimization in
front-vocalic environments, but is more narrowly delimited than nounal raising.

\subsection{Rule interactions}\label{sec:morphophon-interactions}

The alternations above often stack. The following derivations illustrate the
intended order and, where relevant, which of the three mutation domains applies.

\pex
\a \ird{lobek} + \ird{-u} $\rightarrow$ \ird{lubku}
\lingcontext{suffix class selection; weak-\ird{e} deletion; back-vowel raising; no front-vocalic mutation}
\a \ird{Janek} + \ird{-í} $\rightarrow$ \ird{Jancí}
\lingcontext{weak-\ird{e} deletion; then velar fronting k $\rightarrow$ c (ordinary softening class)}
\a \ird{perek} + \ird{-e} $\rightarrow$ \ird{prce}
\lingcontext{CeLeC stem loses weak \ird{e}; velar fronting k $\rightarrow$ c applies only after the liquid cluster is formed}
\a \ird{ruk} + \ird{-ě} $\rightarrow$ \ird{ruce}
\lingcontext{front-vocalic ordinary suffix; velar fronting k $\rightarrow$ c; ě normalised to e after c}
\a \ird{louž} + \ird{-ení} $\rightarrow$ \ird{loužení}
\lingcontext{expanded front-vocalic indirect ending; stem-final ž is a postalveolar phoneme, not a softened sibilant}
\a \ird{zděk-} + \ird{-ěv} $\rightarrow$ \ird{zdíčev}
\lingcontext{strong palatalizing suffix; weak-\ird{e} deletion; front-vowel raising; velar fronting k $\rightarrow$ č}
\a \ird{zděk-} + \ird{-ir} $\rightarrow$ \ird{zděcir}
\lingcontext{ordinary softening suffix; velar fronting k $\rightarrow$ c; no raising or strong palatalisation}
\a \ird{piašt-} + \ird{-aš-} + \ird{-ek} $\rightarrow$ \ird{piašček}
\lingcontext{sibilant replacement (Class I), a separate derivational block; ordinary suffixation follows}
\a \ird{piašček} + \ird{-e} $\rightarrow$ \ird{piaščice}
\lingcontext{linker \ird{-e} is an ordinary softening suffix; stem-final č is a postalveolar; no mutation applies to č itself}
\xe

These derivations are cited throughout Chapters~\ref{ch:verbs}, \ref{ch:nouns},
and~\ref{ch:complex}.

\section{Historical note}\label{sec:historical-note}

The Iridian consonant system reflects substantial West Slavic areal pressure,
concentrated especially in the orthographic conventions, the coronal soft series,
and the two-way velar affrication pattern. The use of háček letters (\ird{č\,š\,ž\,ň}
and now \ird{ť\,ď\,ľ} in the updated standard) reflects long exposure to Czech
graphic norms, partly mediated through centuries of Moravian political and cultural
influence. The optional use of \ird{ř} in learned or etymological spelling is the
most visible trace of this prestige relationship; its value is not the Czech /r̝/
but a simpler cluster-like or fricative-like realisation inherited through the same
cultural stratum.

The two-class velar alternation---yielding /ts/ before ordinary suffixes and /tʃ/
before a smaller strong class---mirrors a historical distinction between earlier
palatalisation environments that has been reanalysed morphologically in the modern
language. The coronal soft series t/d/n/l likewise reflects inherited West Slavic
morphophonological categories that have been stabilized and made more regular under
areal pressure.

These historical points do not govern the synchronic analysis. The alternations
described in this chapter are treated as synchronic readjustment rules conditioned
by morphological trigger classes, not as reflexes of historical sound changes that
happen to have survived. Where historical motivation conflicts with a synchronically
cleaner statement of the rule, the synchronic account takes precedence.

\section{Operational summary}

When encountering a consonant alternation in any paradigm or derived form, the
analyst should proceed as follows.

First, identify the trigger. If the alternation appears in an antipassive derivative
with \ird{-aš-}, it belongs to sibilant replacement and should be analysed using
Table~\ref{tab:sibilant-mutation}. It is not soft-stem selection or velar fronting.

Second, if the trigger is a front-vocalic suffix, identify the suffix's class. If
the affected consonant is a plain coronal (/t\,d\,n\,l/), the process is soft-stem
selection and the output is \ird{ť\,ď\,ň\,ľ}. If the affected consonant is /k/,
the process is velar fronting: \ird{c} for ordinary softening suffixes,
\ird{č} for strong palatalizing suffixes.

Third, check whether a surface \ird{š} or \ird{ž} is involved. If so, this
consonant is almost certainly an underlying postalveolar phoneme or a sibilant
replacement output, not the product of productive softening from /s/ or /z/. There
is no morphophonemic rule s $\rightarrow$ š or z $\rightarrow$ ž in the core system.

Fourth, if the form contains \ird{si}, \ird{zi}, or \ird{ci} before a high
palatalizing vowel, note that these sequences are phonetically [śi], [źi], [tɕi]
due to the positional palatalisation rule. No orthographic change is involved and
no morphophonemic alternation has occurred; the palatalisation is a surface phonetic
detail below the level of the writing system.

Fifth, apply orthographic normalisation last: replace any \ird{ě} that follows a
soft or fronted consonant with plain \ird{e}, and ensure that soft \ird{ť\,ď\,ľ\,ň}
are written explicitly where phonologically present rather than being signalled
only by the following vowel.

This operational summary bridges into the morphology chapters that follow.
Subsequent paradigms and derivations will invoke `soft-stem selection',
`velar fronting', and `sibilant replacement' as named processes without
restating the full theory each time. The reader who needs to verify how a
particular alternation is classified should return to this chapter, and in
particular to the three-class framing in Section~\ref{sec:morphophon}, the
positional palatalisation rule in Section~\ref{sec:sibilant-palatalisation}, and the
suffix-class table at Table~\ref{tab:morph-trigger}.
