\chapter{Verbs}\label{ch:verbs}

\section{Introduction}

The verb is the morphological nucleus of the Iridian clause, and upon it falls the principal burden of grammatical expression. While nouns admit of inflection for case alone and are in all other respects morphologically inert, the verb concentrates within itself all information pertaining to the nature of the event described, the relation of its participants, and the speaker's orientation toward what is asserted. Every finite verb bears the aspect-alignment suffix, which is the sole obligatory element of the verbal complex; beyond this the verb may carry markers of voice, mood, and evidentiality, in an order fixed by the morphological template set out in the following section. Tense, gender, number agreement, and pronominal cross-reference find no expression on the Iridian verb.

\section{The morphological template}\label{sec:morph-template}

The structure of a finite verb in Iridian is given by the following template, wherein parentheses enclose optional elements and the sole obligatory position is that of the aspect-alignment suffix:

\medskip
\begin{center}
(\Neg{}) + \Stem{} + (\Antip{} / \Appl{}) + (\Caus{}) + \Asp{}.\textsc{align} + (\textsc{mood}) + (\textsc{evid})
\end{center}
\medskip

The \textsc{asp.align} slot fuses aspect and alignment into a single morpheme and is the one element no finite verb may lack. The mood slot may be analysed as containing a null morpheme in the indicative. The negative prefix \ird{zá-} stands before the entire stem complex. Evidential suffixes are attested in the literature but remain insufficiently described; they are accordingly omitted from the present chapter. Person is not morphologically expressed on the Iridian verb: there is no pronominal clitic series and no agreement layer. Overt person reference is conveyed by full pronoun DPs, demonstratives in the third person, and zero anaphora where discourse recoverability licenses argument omission.


\subsection{The Mirror Principle and functional projections}\label{sec:vm-overview}

\noindent Baker's (1985) \term{Mirror Principle}
predicts that morphological derivations directly reflect syntactic
derivations: each morpheme corresponds to a syntactic head, and the
order of morphemes on the verbal complex mirrors the order in which
those heads are merged. Reading outward from the root, the Iridian
template projects the following hierarchy of functional
projections:

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Morphological template positions and corresponding
  syntactic projections (innermost to outermost).}\label{tab:vm-projections}
  \medskip
  \begin{tabularx}{\textwidth}{l l L l}
    \toprule
    {\sc position} & {\sc morpheme} & {\sc syntactic projection} &
    {\sc see} \\
    \midrule
    Root         & \Stem{}           & $\surd$P / VP (lexical predicate)
                 & §\,\ref{sec:vm-root} \\
    Voice        & \textsc{trs} / \Antip{} / \Appl{} & VoiceP / ApplP
                 & §\,\ref{sec:vm-antipassive}, §\,\ref{sec:vm-n} \\
    Causative    & \Caus{}           & CausP
                 & §\,\ref{sec:vm-caus} \\
    Asp.Align    & \Asp{}.\textsc{align} & AspP (fused Asp + Case)
                 & §\,\ref{sec:vm-asp} \\
    Neg (prefix) & \Neg{}            & NegP (see §\,\ref{sec:vm-neg})
                 & §\,\ref{sec:vm-neg} \\
    Mood         & \textsc{mood}     & MoodP / IP
                 & §\,\ref{sec:vm-moodp} \\
    Evidential   & \textsc{evid}     & EvidP$_\text{inner}$
                 & §\,\ref{sec:vm-evid} \\
    \bottomrule
  \end{tabularx}
\end{table}

\noindent The prefix position of \Neg{} (§\,\ref{sec:vm-neg})
presents a principled complication: the Mirror Principle predicts
that NegP, being scopally above AspP, should correspond to a suffix
that appears \emph{outside} the Asp.Align morpheme. The Iridian
prefix instead appears before the root. This is a well-known
typological type --- prevalent in verb-final languages --- and its
theoretical resolution is discussed in §\,\ref{sec:vm-tension-neg}.



\subsubsection{The root and VP}
\label{sec:vm-root}

The lexical stem is the exponent of the Root merged with the
functional category V (or, in Distributed Morphology terms, the
root $\surd$ inserted under a categorizing head $v_\text{intr}$ for
intransitives, or $v^*$ for transitives). The Iridian verb allows no
citation form; the stem is bound and always requires at least the
Asp.Align suffix. This confirms that the root must be licensed by a
selecting functional head: a bare root without an Asp projection is
unlicensed in the language.

The VP formed by the root + its arguments constitutes the innermost
layer of the verbal clause. In a transitive clause the internal
argument (patient) is merged as the complement of V:

\begin{center}
\begin{forest}
  syntax tree
  [VP
    [{DP\\{\footnotesize patient}}]
    [V
      [{\textit{root}}]
    ]
  ]
\end{forest}
\end{center}

\noindent Because Iridian is rigorously head-final
(§\,\ref{sec:th-head-finality}), this complement surfaces to the
left of the verbal head at every embedded level. The preverbal
position of the object is therefore not derived by movement but is
the base-generated configuration of the head-final VP.



\section{The verb stem and thematic classes}

Iridian verbs admit no citation form. Lexicographers record the bare stem with a following hyphen to indicate its character as a bound form requiring at least one suffix; the verb rendered in English as \textit{to eat} is thus entered as \ird{piašt-}, where \ird{t} constitutes the \term{thematic consonant}.

The thematic consonant, being the final consonant of the stem, is phonologically active: it undergoes regular alternations when certain suffixes are adjoined. Two mutation systems must be distinguished. Some suffixes trigger \term{sibilant replacement}, a closed morphologized stem mutation whose chief exponent is the antipassive \ird{-aš-}. Others trigger ordinary \term{palatalization}, the productive softening caused by front-vocalic suffixes such as \ird{-ěv}, \ird{-ěc}, and the conjunctive linker \ird{-e}. The five thematic classes in Table~\ref{tab:verbclasses} govern sibilant replacement only.

\begin{table}[ht]
	\footnotesize\sffamily
	\caption{Thematic consonant classes.}\label{tab:verbclasses}
	\medskip
	\begin{tabularx}{\textwidth}{l L L L}
		\toprule
		{\sc class} & {\sc thematic consonants} & {\sc change before \Antip{}} & {\sc example} \\
		\midrule
		I    & -t, -k, -c, -č  & → -č                    & \ird{piašt-} → \ird{piašč-} \\
		II   & -s, -š          & → -š                    & \ird{tras-} → \ird{traš-} \\
		II-A & -l, -p          & add -š                  & \ird{dal-} → \ird{dalš-} \\
		III  & -d, -h, -j, -z, -ž & → -ž               & \ird{vad-} → \ird{váž-} \\
		III-A & all remaining  & add -ž                  & \ird{ščen-} → \ird{ščenž-} \\
		\bottomrule
	\end{tabularx}
\end{table}

Loan stems with surface \orth{g} are assigned analogically to the voiced velar/fricative series and do not require a native phoneme /g/. Palatalizing suffixes do \emph{not} follow Table~\ref{tab:verbclasses}; they invoke the broader front-vocalic mutation system described in Section~\ref{sec:front-palatalization} of Chapter~\ref{ch:phon}.

Outside the closed antipassive classes, verbs use the same general morphophonemic engine as nouns: suffix class selection, weak-\ird{e} deletion or retention, grade formation where licensed, soft-stem selection, and finally cluster and voicing cleanup. The difference is one of density rather than mechanism. Nouns expose that engine most heavily in their case paradigms, while verbs deploy it more selectively in the irrealis, the simultaneous converb, the conjunctive linker, and certain nominalizing formations.

\section{Aspect and alignment}
\label{sec:aspect}

Iridian grammaticalizes \term{aspect} rather than \term{tense} on the finite
verb. There is no dedicated past-, present-, or future-tense suffix; temporal
location is inferred from discourse context, temporal adverbials, and the
lexical semantics of the predicate. Aspect and alignment are fused into a
single obligatory suffix, the \term{aspect-alignment morpheme} (Asp.Align),
which every finite verb must bear: selecting an aspectual category
simultaneously determines how the event is construed and which participant
occupies the absolutive position.

The five Asp.Align categories fall into two coherent higher-order domains.
The \term{perfective domain} comprises the \term{perfective} and the
\term{retrospective}: both present an eventuality as bounded and externally
viewed---construed as a closed whole from an outside or posterior vantage
point---and both yield \term{ergative-absolutive alignment}, in which the
patient is the absolutive pivot. The \term{imperfective domain} comprises
the \term{continuous}, the \term{progressive}, and the \term{contemplative}:
none of these presents the eventuality as a bounded whole viewed from outside;
instead, each either locates the reference point within the event's internal
course or presents the event as projected relative to reference time, not yet
realized but anticipated, intended, or impending. All three yield
\term{nominative alignment}, in which the agent is the absolutive pivot.
These two domains differ not merely in which aspects they contain but in the
fundamental orientation they impose on the clause: the perfective domain
foregrounds culmination and the affected patient, while the imperfective domain
foregrounds the ongoing or prospective engagement of the agent.

The subsections that follow first lay out the shared temporal-semantic
machinery (including Reichenbachian framing and the formal Asp.Align analysis),
then treat the perfective domain (§\,\ref{sec:aspect-pf}) and the imperfective
domain (§\,\ref{sec:aspect-ipf}) in turn.

Three temporal notions must be kept distinct. Tense locates a situation in time
relative to a reference point, most commonly the moment of speaking. Grammatical
aspect does not locate the situation in this way, but construes its internal
temporal contour: whether it is viewed as bounded, ongoing, prior from a later
vantage point, or not yet realized. \term{Lexical aspect} or \term{Aktionsart},
by contrast, belongs to the predicate itself. A predicate may denote a state,
activity, accomplishment, or achievement independently of the inflection it
bears. Iridian aspectual morphology therefore does not determine the inherent
temporal type of the eventuality; rather, it imposes a viewpoint upon an
eventuality whose lexical contour is supplied by the verb phrase.

A useful descriptive frame is Reichenbach's distinction among speech time
(\textit{S}), reference time (\textit{R}), and event time (\textit{E}):
\textit{S} is the moment of utterance, \textit{E} the time at which the
situation itself takes place, and \textit{R} the temporal vantage point from
which that situation is presented. The usefulness of \textit{R} is that speakers
do not always describe an event directly from the moment of speaking, but may
instead locate it from some later or earlier point already established in the
discourse; English \textit{had left}, for example, places the leaving
(\textit{E}) before a reference point (\textit{R}) that is itself prior to the
speech moment (\textit{S}). In a tense-prominent system, finite morphology often
fixes the relation among all three. In Iridian, by contrast, the finite suffix
primarily construes the relation between \textit{E} and \textit{R}, while the
relation of \textit{R} to \textit{S} is often left to context, temporal
adverbials, and discourse sequencing. This is why the same aspectual category
may receive past-, present-, or future-oriented readings in different
environments without ceasing to be aspectual rather than temporal.

The five categories may be sketched in these terms. The \term{perfective}
presents an eventuality as bounded and viewed as a whole, with \textit{R}
aligned to the event in its entirety. The \term{retrospective} places \textit{E}
before \textit{R}, yielding a later vantage point from which present relevance,
result, prior experience, or other discourse effects of anteriority may be
construed. The \term{continuous} and \term{progressive} both place \textit{R}
within \textit{E}, but they differ in the kind of internal view imposed: the
continuous presents the situation as simply obtaining at reference time, without
foregrounding change, onset, or unfolding contour, whereas the progressive more
narrowly profiles a dynamic event as actively unfolding at the reference point.
The \term{contemplative} places \textit{E} after \textit{R}, presenting the
event as projected, intended, expected, or impending rather than as obtaining or
complete.

These five categories divide naturally into two families. The
\term{perfective domain}---the perfective and retrospective---consists of those
categories that locate the event from an external or posterior vantage and
present it as bounded or anteriorly packaged: in Reichenbachian terms, the
event time is enclosed within the reference span, or precedes it. The
\term{imperfective domain}---the continuous, progressive, and
contemplative---consists of those categories that do not present the event as a
bounded whole, but instead either locate reference time within the ongoing event
or situate reference time prior to an unrealized projection of the event. The
Reichenbachian figure (Figure~\ref{fig:iridian-aspect-reichenbach}) makes this
grouping visually explicit: the upper two rows belong to the perfective domain,
in which the event is viewed as bounded or anterior from outside; the lower
three rows belong to the imperfective domain, in which R lies within E or
precedes an unrealized E.

Because viewpoint aspect interacts with lexical aspect, the same inflection does
not yield identical interpretive effects with every predicate class. With states
and activities, the continuous naturally yields persistent or ongoing readings;
with accomplishments it typically selects an internal phase rather than a
completed endpoint; with achievements it may invite iterative, preparatory, or
inceptive coercions. Conversely, the perfective and retrospective readily
package telic predicates as bounded wholes or prior results, but with atelic
predicates they instead delimit a stretch, occasion, or state viewed from an
external vantage. The exact semantic and distributional consequences of these
interactions are discussed in the domain subsections below.

Typologically, this pattern is an instance of aspect-conditioned split
ergativity. Iridian does not maintain a single alignment pattern across all
finite clauses; rather, the aspectual category selected on the verb determines
whether the clause is built on an ergative-absolutive or a nominative-accusative
frame. Aspect in Iridian therefore has consequences at two levels at once: it
construes the temporal profile of the eventuality and it determines the
morphosyntactic alignment through which the clause is organized.

The mapping between aspect and alignment is not arbitrary. In bounded,
externally-viewed events (the perfective and retrospective), the clause is
fundamentally ``about'' what happened to a participant---the locus of
culmination---and the patient accordingly occupies the absolutive. In
unbounded, internally-viewed events (the continuous, progressive, and
contemplative), the clause is fundamentally ``about'' an agent's ongoing
activity or state, and the agent accordingly occupies the absolutive. This
alignment between aspectual viewpoint and case pivot is the semantic core of
Iridian's split-ergative system; the formal analysis is given in
§\,\ref{sec:vm-asp-case}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  x=1.05cm,y=1cm,
  evt/.style={line width=1.1pt},
  refpt/.style={circle,fill,inner sep=1.3pt},
  speech/.style={diamond,draw,inner sep=1.2pt},
  rowlab/.style={anchor=east,font=\small},
  every node/.style={font=\small}
]

% All rows hold S at x = 4.5 so the deictic baseline is visually constant.

% perfective (y = 0)
\node[rowlab] at (0,0) {perfective};
\draw[->] (0.5,0) -- (8.5,0);
\draw[evt] (3.0,0) -- (6.0,0);
\draw (3.0,0.15) -- (3.0,-0.15);
\draw (6.0,0.15) -- (6.0,-0.15);
\node[refpt] at (3.0,0) {};
\node[refpt] at (6.0,0) {};
% R bracket above the timeline, E label between bracket and line
\node at (4.5,0.30) {$E$};
\draw[decorate,decoration={brace,amplitude=3pt}]
  (3.0,0.55) -- (6.0,0.55)
  node[midway,above=4pt] {$R$};
% S below, matching spacing of other rows
\draw[dashed] (4.5,-0.08) -- (4.5,-0.42);
\node[speech] (s1) at (4.5,-0.55) {};
\node[below=2pt of s1] {$S$};

% retrospective (y = -2.0)
\node[rowlab] at (0,-2.0) {retrospective};
\draw[->] (0.5,-2.0) -- (8.5,-2.0);
\draw[evt] (1.8,-2.0) -- (3.4,-2.0);
\draw (1.8,-1.85) -- (1.8,-2.15);
\draw (3.4,-1.85) -- (3.4,-2.15);
\node at (2.6,-1.65) {$E$};
\node[refpt,label=above:{$R$}] at (4.5,-2.0) {};
\node[speech] (s2) at (4.5,-2.55) {};
\node[below=2pt of s2] {$S$};
\draw[dashed] (4.5,-2.08) -- (4.5,-2.42);

% continuous (y = -4.0)
\node[rowlab] at (0,-4.0) {continuous};
\draw[->] (0.5,-4.0) -- (8.5,-4.0);
\draw[dotted] (1.6,-4.0) -- (2.2,-4.0);
\draw[evt] (2.2,-4.0) -- (6.8,-4.0);
\draw[dotted] (6.8,-4.0) -- (7.4,-4.0);
\node at (3.0,-3.65) {$E$};
\node[refpt,label=above:{$R$}] at (4.5,-4.0) {};
\node[speech] (s3) at (4.5,-4.55) {};
\node[below=2pt of s3] {$S$};
\draw[dashed] (4.5,-4.08) -- (4.5,-4.42);

% progressive (y = -6.2)
\node[rowlab] at (0,-6.2) {progressive};
\draw[->] (0.5,-6.2) -- (8.5,-6.2);
\draw[evt] (2.4,-6.2) -- (7.0,-6.2);
\draw (2.4,-6.05) -- (2.4,-6.35);
\draw (7.0,-6.05) -- (7.0,-6.35);
\node at (6.5,-5.85) {$E$};
\draw[->] (2.8,-5.55) -- (4.2,-5.55)
  node[midway,above,font=\scriptsize] {dynamic phase};
\node[refpt,label=above:{$R$}] at (4.5,-6.2) {};
\node[speech] (s4) at (4.5,-6.75) {};
\node[below=2pt of s4] {$S$};
\draw[dashed] (4.5,-6.28) -- (4.5,-6.62);

% contemplative (y = -8.2)
\node[rowlab] at (0,-8.2) {contemplative};
\draw[->] (0.5,-8.2) -- (8.5,-8.2);
\node[refpt,label=above:{$R$}] at (4.5,-8.2) {};
\draw[evt] (5.8,-8.2) -- (7.5,-8.2);
\draw (5.8,-8.05) -- (5.8,-8.35);
\draw (7.5,-8.05) -- (7.5,-8.35);
\node at (6.65,-7.85) {$E$};
\node[speech] (s5) at (4.5,-8.75) {};
\node[below=2pt of s5] {$S$};
\draw[dashed] (4.5,-8.28) -- (4.5,-8.62);

\end{tikzpicture}
\caption{Reichenbachian schematic for Iridian aspect. $E$ = event time, $R$ =
reference time, $S$ = speech time. In the perfective, $R$ spans the bounded
event (marked by dots at the endpoints and a bracket); elsewhere, $R$ is a
point-like vantage. The upper two rows (perfective, retrospective) belong to the
\term{perfective domain}: in both, the event is construed as bounded or
anterior, viewed from outside or from a posterior vantage. The lower three rows
(continuous, progressive, contemplative) belong to the \term{imperfective
domain}: in all three, $R$ either lies within $E$ or precedes an unrealized $E$,
and no external terminus is foregrounded. Dashed connectors mark the unmarked
deictic baseline in which $R$ coincides with $S$. $S$~is held at a constant
horizontal position across all rows; in discourse, adverbials or sequencing may
shift $R$ independently of $S$.}
\label{fig:iridian-aspect-reichenbach}
\end{figure}

\subsection{Formal analysis: the Asp.Align morpheme}\label{sec:vm-asp}

The aspect-alignment suffix is the sole obligatory morpheme in every finite
Iridian verbal form. Its double burden---encoding both the aspect of the event
and the alignment pattern of case assignment---makes it the central theoretical
challenge of the Iridian verbal system. The five indicative aspect-alignment
suffixes are given in Table~\ref{tab:aspect}.

\begin{table}[ht]
	\footnotesize\sffamily
	\caption{Aspect-alignment suffixes, grouped by domain. The perfective domain
	yields ergative-absolutive alignment (patient as absolutive pivot); the
	imperfective domain yields nominative alignment (agent as absolutive
	pivot).}\label{tab:aspect}
	\medskip
	\begin{tabularx}{0.95\textwidth}{l L l L}
		\toprule
		{\sc domain} & {\sc aspect} & {\sc suffix} & {\sc notes} \\
		\midrule
		\multirow{2}{*}{\textit{Perfective}}
		  & Perfective (bare, intransitive only)    & \ird{-ek} / \ird{-ik}
		    & \ird{-ik} when stem vowel is \ird{ě} or \ird{e} (dissimilation;
		      see §\ref{sec:aspect-pf}). Licit only for inherently intransitive
		      predicates or ambitransitives in their intransitive reading.
		      Transitive perfective-domain clauses require overt Voice: \ird{-nik}
		      (= \ird{-n-} + \ird{-ik/-ek}) for the ergative transitive,
		      \ird{-aš-ek} for the antipassive. \\
		  & Retrospective (bare, intransitive only) & \ird{-ac}
		    & Bare form licit only for inherently intransitive predicates;
		      transitive retrospective clauses require overt Voice: \ird{-n-ac}
		      for the ergative transitive, \ird{-aš-ac} for the antipassive. \\
		\midrule
		\multirow{3}{*}{\textit{Imperfective}}
		  & Continuous   & \ird{-a}   & Unmarked member; situation holds at $R$,
		                                no culmination foregrounded. \\
		  & Progressive  & \ird{-ime} & Dynamic, marked member; event actively
		                                unfolding at $R$. \\
		  & Contemplative & \ird{-ach} & Projective member; event anticipated or
		                                projected at $R$, not yet realized. \\
		\bottomrule
	\end{tabularx}
\end{table}

\subsubsection{Split-ergative alignment}
\label{sec:vm-asp-case}

Asp.Align does two things at once. It encodes the viewpoint aspect of the
event---how the speaker construes the event's temporal contour---and it
determines which participant is the clause's \term{absolutive pivot}: the
argument that bears the direct case and serves as the informational anchor of
the assertion. These two functions are inseparable; the alignment of the clause
is a consequence of its aspectual category, and both are realized in the same
obligatory suffix.

The alignment defaults are:

\begin{enumerate}[wide]
\item ERG-aspect domain (perfective, retrospective). Asp.Align establishes
  patient-oriented absolutive as the privileged orientation of the domain: the
  patient is the absolutive pivot and the agent receives the transitive
  (ergative) case as the dependent case once a patient has been licensed in the
  absolutive. Crucially, Asp.Align determines this \emph{domain-level
  orientation}; it does not by itself license a transitive patient as a core
  argument. For a transitive predicate, an overt Voice head is additionally
  required: \ird{-n-} licenses the patient as a core argument and realizes the
  ordinary ergative transitive, while \ird{-aš-} removes the patient from core
  argument structure and realizes the antipassive. A bare perfective or
  retrospective without an overt Voice head is therefore well-formed only where
  the predicate is inherently intransitive, or where a lexical ambitransitive is
  used in its intransitive reading (§\,\ref{sec:ambitransitive}).
\item NOM-aspect domain (continuous, progressive, contemplative). The agent
  receives the direct (absolutive) case and the patient the indirect (oblique)
  case. The absolutive is therefore the \emph{agent}.
\end{enumerate}

The correlation between bounded aspect and patient-absolutive, and between
unbounded aspect and agent-absolutive, is not an arbitrary accident. It is
semantically motivated and cross-linguistically robust (Dixon 1994; Coon 2013).

In a bounded, externally-viewed event---the perfective and retrospective---the
event is construed as a completed whole, and the primary assertion of the clause
concerns what happened \emph{to} some participant. In transitive events the
locus of culmination is prototypically the patient: the bread was eaten, the
game was caught, the letter was brought. Placing the patient in the absolutive
encodes this culmination-focus: the patient is what the bounded event is
``about,'' and the case system reflects that primacy. The agent, while causally
indispensable, is the secondary specification---the dependent case falls on it
once the patient has taken the absolutive.

In an unbounded, internally-viewed event---the continuous, progressive, and
contemplative---there is no culmination foregrounded, and what is asserted is an
ongoing activity, state, or projected action. The most discourse-prominent
participant is the one engaged in or holding that activity: the agent. Placing
the agent in the absolutive encodes this activity-focus; the patient, if
present, is backgrounded as the indirect object of an activity rather than the
culmination-point of an event.

This structural asymmetry between the two aspect-domains reinforces the semantic
one. Non-perfective categories are cross-linguistically \emph{structurally
heavier}: progressive constructions typically involve copular or periphrastic
material (English \textit{be + V-ing}; Slavic imperfective morphology;
grammaticalized locative predicates), and the continuous often encodes states
that in other languages appear as nominal or stative predicates. Perfective, by
contrast, is the default, structurally simplest aspect---the unmarked member in
morphological paradigms, the first to emerge in acquisition and
grammaticalization---and it places the fewest additional demands on clause
structure. The richer, internally organized imperfective clause therefore
naturally yields nominative alignment, while the simpler, culmination-focused
perfective clause yields ergative alignment. Iridian's split is typologically
well-motivated, a surface reflex of an affinity between aspectual viewpoint and
case assignment that recurs across unrelated languages.

These patterns are directly visible in the case frames of indicative clauses.
In the NOM-aspect domain, the main argument of the verb bears the direct case
regardless of transitivity, while the patient of a transitive verb appears in
the indirect. This is a nominative-accusative frame, with the direct case
serving as the absolutive and the indirect as the oblique:

\pex
\a
\begingl
\gla Marek zmacime.//
\glb Marek.\Dir{} run-\Prog{}//
\glft \trsl{Marek is running.} \lingcontext{intransitive}//
\endgl
\a
\begingl
\gla Marek travě piastime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.} \lingcontext{transitive; P = \Ind{}}//
\endgl
\a
\begingl
\gla Vytam shoda.//
\glb shirt.\Dir{} wear-\Cont{}//
\glft \trsl{(I) am wearing a shirt.} \lingcontext{stative continuous}//
\endgl
\a
\begingl
\gla Vytam shodime.//
\glb shirt.\Dir{} wear-\Prog{}//
\glft \trsl{(I) am putting on a shirt.} \lingcontext{dynamic progressive}//
\endgl
\xe

In the ERG-aspect domain, by contrast, the patient of a transitive verb is
placed in the direct case, and the agent takes the transitive case. This is an
ergative-absolutive frame; what Iridian grammar calls the \emph{transitive case}
is, in typological terms, the ergative, and the \emph{direct case} serves as
the absolutive. In this domain, a transitive predicate requires the Voice head
\ird{-n-} to license the patient as a core argument; when the Voice head
\ird{-aš-} is present instead, the agent takes the absolutive and the patient is
removed from core argument structure (§\,\ref{sec:vm-antipassive}):

\pex
\a
\begingl
\gla Travat Marku piaštnik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} eat-\textsc{trs}-\Pf{}//
\glft \trsl{Marek ate the bread.} \lingcontext{ergative; P = \Dir{}}//
\endgl
\a
\begingl
\gla Marek piašček.//
\glb Marek.\Dir{} eat-\Antip{}-\Pf{}//
\glft \trsl{Marek ate.} \lingcontext{antipassive; A = \Dir{}}//
\endgl
\xe

The choice between these constructions is governed primarily by the discourse
status of the patient. The ergative construction foregrounds the patient: it is
the construction of a specific, discourse-trackable participant who is either
newly introduced or reactivated, and the event is asserted as bringing about a
change of state in that participant. It is accordingly characteristic of
narrative foreground, where the consequences of events for particular persons or
things constitute the substance of what is related. Patient definiteness
correlates strongly with the ergative, and the definite marker \ird{-t} most
commonly appears on the patient noun in this construction, though it is not
formally required.

The antipassive construction, by contrast, foregrounds the agent's activity
without reference to any outcome in a patient. It is the construction of choice
when the patient is non-referential, semantically incorporated, or of no
discursive moment: in habitual and generic statements, in activity-focus
contexts wherein the event itself and not its effect upon some participant is
what is asserted, and wherever the clause contains no patient at all. The
guiding principle is that the patient, where it exists, must either be present
through the ergative construction or wholly absent; the antipassive admits of no
intermediate state in which a patient is demoted to an oblique position.

Taken together, the choice between ergative and antipassive constructions in the
perfective and retrospective is a choice about where to locate the event's
absolutive pivot: on the patient (ergative, culmination-focus) or on the agent
(antipassive, activity-focus). This discourse-structural opposition is the
surface reflex of the same aspectual-semantic logic that separates the
ERG-domain aspects from the NOM-domain aspects at the level of the whole aspect
inventory: bounded events culminate in patients; unbounded activities centre on
agents.

Iridian's ergative case is \term{aspect-directed}: the Asp.Align suffix
specifies which argument is the absolutive, so the alignment varies with the
aspect rather than being fixed by structural position or thematic role alone.
Because the absolutive varies in this way---patient in the ERG domain, agent in
the NOM domain---the Asp.Align morpheme is the natural locus of alignment, not a
passive reflex of deeper case rules.

Within the domain Asp.Align establishes, the transitive (ergative) case behaves
as a \term{dependent case}: it falls on the agent when and only when a patient
has already received the absolutive. These two observations are complementary
rather than competing. Aspect-directed selection determines \emph{which
argument} is the absolutive; dependent case determines \emph{what the agent
receives} once the absolutive is fixed on the patient. The architecture is
aspect-driven at the inter-domain level and dependent-case at the intra-domain
level.

Selecting an aspect is simultaneously selecting which argument is the clause's
discourse topic. In the ERG domain, the patient-absolutive is characteristically
the discourse-prominent argument: the affected participant is the
culmination-point the event asserts something about. In the NOM domain, the
agent-absolutive is characteristically the discourse-prominent argument: the
active participant is whose activity or state the clause tracks. This is what it
means to say that aspect selection in Iridian is in part a pragmatic decision:
speakers choosing an aspect are choosing a case frame that encodes which
participant the clause is ``about,'' and the two decisions are one and the same
act.

The domain-level orientation established by Asp.Align is not itself sufficient to
realize a transitive clause. An overt Voice head is additionally required. The
morpheme \ird{-n-} is such a head: it licenses the patient as a core argument
and specifies patient absolutive (\textsc{[pat-abs]}). In the ERG-aspect domain
this is required for any transitive predicate---without \ird{-n-}, the patient
cannot be realized as a core argument, and the bare perfective or retrospective
is licit only for inherently intransitive predicates. In the NOM-aspect domain,
\ird{-n-} overrides the default agent-absolutive orientation, making the patient
the absolutive pivot and yielding the passive construction
(§\,\ref{sec:vm-n}, §\,\ref{sec:vm-passive}). The antipassive \ird{-aš-}
likewise constitutes an overt Voice head, but is subject to a structural input
condition: it may apply only where the patient already stands as the absolutive,
and it therefore surfaces exclusively in ERG-aspect clauses
(§\,\ref{sec:vm-antipassive}).

When an overt Voice head is present, its absolutive specification is satisfied
first, and the remaining non-absolutive argument receives the case predicted by
the aspect domain. Asp.Align determines the domain-level orientation; Voice
determines how the transitive base is realized within that domain.

The fusion of aspect and alignment into a single suffix is not an accident. It
reflects the language's deep integration of temporal viewpoint, case assignment,
and discourse-topic orientation---three properties that many other languages
encode separately, but that Iridian realizes in the one morpheme that no finite
verb may lack.

\subsection{The perfective domain}\label{sec:aspect-pf}

The perfective domain comprises those Asp.Align forms which construe an
eventuality as bounded and externally viewed. Unlike the imperfective domain,
which presents an event from within its internal course, the perfective domain
presents it as a temporally delimited whole. Iridian distinguishes two
categories within this domain: the \term{perfective} proper, which presents a
bounded event as a discrete occurrence, and the \term{retrospective}, which
presents a bounded prior event in relation to a present state, present
relevance, or present experience. Both categories interact regularly with
valency and alignment morphology, and both therefore belong to a single
structural and semantic domain rather than to unrelated aspectual classes.

Formally, the perfective suffix has two allomorphs. \ird{-ek} is the default
form, and \ird{-ik} appears when the final syllable of the stem contains
\ird{ě} or \ird{e} as its nucleus. This raising is dissimilatory in character:
direct suffixation of \ird{-ek} to such stems would produce a CěCe or CeCe
sequence across a single intervening consonant, which the language avoids. The
conditioning is thus vocalic and does not depend on consonant class. Where a
valency morpheme such as the antipassive \ird{-aš-}, the applicative \ird{-éb-},
or the causative \ird{-na-} intervenes between stem and suffix, that intervening
material breaks the disfavoured sequence and raising does not apply. The
ergative perfective \ird{-nik}, which follows the Voice head \ird{-n-}, is a
fixed form and is not derived by this alternation. The retrospective is marked
by \ird{-ac}; its derived forms are built compositionally, as in \ird{-n-ac} and
\ird{-aš-ac}. % Cross-reference: the \ird{-ek}/\ird{-ik} alternation may be
treated more fully in the phonology chapter.

The semantic opposition within the perfective domain is best stated as one of
\emph{event occurrence} versus \emph{post-event relevance}. The perfective
proper presents a bounded event as a single occurrence situated at a definite
point in narrative or discourse time, and is therefore the primary narrative
aspect. The retrospective likewise presupposes a bounded prior event, but does
not profile that event merely as an occurrence; rather, it presents it through
its present relevance, present consequences, or experiential force. In this
respect it corresponds in some uses to the English perfect. Its non-volitional
or accidental readings belong to the same general semantic profile: the event is
construed from the standpoint of its aftermath or present bearing, rather than
simply narrated as a bounded occurrence.

This contrast may be seen in the following pair:

\pex
\a
\begingl
\gla Pražě de stožac.//
\glb Prague.\Ind{} to go-\Ret{}//
\glft \trsl{(I) have been to Prague before.}//
\endgl
\a Compare:
\begingl
\gla Bych Pražě de stožek.//
\glb yesterday Prague.\Ind{} to go-\Pf{}//
\glft \trsl{Yesterday I went to Prague.} \lingcontext{specific past event; perfective}//
\endgl
\xe

Within the perfective domain, the ERG Asp.Align default assigns absolutive to
the patient. For a transitive predicate, an overt Voice head is required to
mediate between the stem and the Asp.Align suffix: the Voice head \ird{-n-}
licenses the patient as a core argument and places it in the direct case, while
the agent takes the transitive; the antipassive Voice head \ird{-aš-} instead
removes the patient from core argument structure, placing the agent in the
direct case as the clause's sole overt pivot. For an inherently intransitive
verb, no Voice head is required: the sole argument takes the direct case
directly, as the only available absolutive pivot.

This structural distinction is especially important in the retrospective. The
ergative retrospective (\ird{-n-ac}) locates present relevance in the outcome
state of the patient: something has been affected, and that effect continues to
matter in the discourse situation. The antipassive retrospective (\ird{-aš-ac})
instead locates present relevance in the experience or history of the agent: the
event matters as part of what the agent has done or undergone, not as a
persisting state of some affected patient. The contrast is parallel to that
between English \textit{the bread has been eaten}, which is outcome-relevant and
patient-oriented, and \textit{I have eaten}, which is experiential and
agent-oriented.

Negation further confirms the semantic distinction between the two members of
the perfective domain. When negated, the perfective normally carries the
implicature that the event was expected, intended, or contextually projected,
but failed to occur. The retrospective, by contrast, is the neutral form for
simple past negation without this implication. The opposition is illustrated
below:

\pex
\a
\begingl
\gla Marek zapiaščik.//
\glb Marek.\Dir{} \Neg{}-eat-\Antip{}-\Pf{}//
\glft \trsl{Marek didn't eat (but he should have).}//
\endgl
\a
\begingl
\gla Marek zapiaščac.//
\glb Marek.\Dir{} \Neg{}-eat-\Antip{}-\Ret{}//
\glft \trsl{Marek didn't eat (neutral past).}//
\endgl
\xe

\label{sec:ambitransitive}A small class of verbs shows particularly clearly how the perfective domain
interacts with argument structure. These are lexical ambitransitives: verbs
which possess both an intransitive reading, in which no patient is structurally
available, and a transitive reading, in which a patient is licensed. Neither
reading is derived from the other by ordinary valency operations; both are
lexical possibilities of the same stem. Two examples are \ird{lov-} \trsl{hunt,
forage [intr.]; catch, take [tr.]} and \ird{kěl-} \trsl{sing, perform [intr.];
sing a song, chant a text [tr.]}.

The diagnostic for membership in this class is whether the Voice head \ird{-n-}
is available in one reading but impossible in the other. In \ird{lov-}, the
intransitive perfective---formed without any Voice head, with the ERG Asp.Align
suffix attached directly to the stem---accesses only the intransitive reading:
the event is construed simply as hunting, with no patient structurally present
and no quarry included in the event frame as an affected participant. The
ergative and antipassive forms, by contrast, access the transitive reading. The
ergative perfective \ird{lovnik} is possible only when something is caught or
taken; the antipassive construes the event as directed toward prey, but
suppresses that prey from the clause's core argument structure.

\pex
\a
\begingl
\gla Marek bych lovek.//
\glb Marek.\Dir{} yesterday hunt-\Pf{}//
\glft \trsl{Marek went hunting yesterday.} \lingcontext{intransitive reading; no patient in event frame}//
\endgl
\a
\begingl
\gla Zvěrat Marku lovnik.//
\glb game.\Dir{}.\Def{} Marek.\Trs{} hunt-\textsc{trs}-\Pf{}//
\glft \trsl{Marek caught the game.} \lingcontext{transitive reading; ergative}//
\endgl
\a
\begingl
\gla Marek bych lovžek.//
\glb Marek.\Dir{} yesterday hunt-\Antip{}-\Pf{}//
\glft \trsl{Marek was hunting yesterday.} \lingcontext{transitive reading; antipassive; quarry conceptually present but not licensed}//
\endgl
\xe

The two non-ergative forms are therefore not interchangeable. \ird{Lovek}
belongs to the intransitive lexical reading: hunting is construed as an activity
with no patient present even in the event structure. \ird{Lovžek}, however,
belongs to the transitive lexical reading: the event is directed toward prey,
and that prey remains part of the conceptual frame even though the antipassive
removes it from core syntax. The distinction is thus not merely one of overt
argument expression, but of underlying event construal.

The same contrast appears even more clearly with \ird{kěl-}. Since the stem
vowel is \ird{ě}, the intransitive perfective takes the dissimilated allomorph
\ird{-ik}: \ird{kělek} would produce the avoided CěCe sequence across the
single intervening consonant \ird{-l-}. The antipassive \ird{kělšek} is
unaffected, since the antipassive morphology breaks that sequence, and the
ergative \ird{kělnik} is the fixed transitive form with inserted \ird{-n-}. The
three forms may be compared as follows:

\pex
\a \ird{kělik} \trsl{sang, performed} \lingcontext{intransitive; no song in the event frame}
\a \ird{kělnik} \trsl{sang it, sang a specific song} \lingcontext{transitive; ergative; patient licensed}
\a \ird{kělšek} \trsl{was singing} \lingcontext{transitive reading; antipassive; patient suppressed from core syntax}
\xe

The minimal pair \ird{kělik} / \ird{kělnik} is especially revealing. The two
forms differ only by the presence or absence of the inserted Voice head
\ird{-n-}, yet that difference is enough to distinguish an intransitive bounded
performance from a transitively bounded event directed toward a licensed
patient. Lexical ambitransitives of this kind therefore provide the clearest
evidence that the perfective domain must be analysed not merely in temporal
terms, but as the interaction of bounded viewpoint with the language's system of
alignment and valency.

\subsection{The imperfective domain: internal and projected viewpoint}
\label{sec:aspect-ipf}

The imperfective domain comprises those Asp.Align categories that do not
construe the eventuality as a bounded whole viewed from outside. Where the
perfective domain packages an event as a closed unit and foregrounds its
culmination or its anteriorly packaged result, the imperfective domain takes
no such external vantage: it either locates the reference point within the
ongoing event, without presenting its boundaries as salient, or presents the
event as projected relative to reference time---not yet realized, but
anticipated, intended, or impending. Because no culmination is foregrounded,
the clause is organized around the participant engaged in, controlling, or
holding the eventuality, and the alignment accordingly follows the nominative
pattern.

Three categories constitute the domain. The \term{continuous} is the unmarked
imperfective member; it presents a situation as simply obtaining at reference
time, without emphasis on change, onset, dynamic unfolding, or endpoint. The
\term{progressive} is the marked dynamic member; it profiles an event
specifically in its active unfolding phase. The \term{contemplative} is the
projective member; it presents the event as not yet realized at reference time
but projected from that point---intended, expected, impending, planned, or
otherwise prospectively construed. Together the three define an internal
opposition within the domain: static holding versus dynamic unfolding versus
prospective projection.

\paragraph{Alignment in the imperfective domain.}
The three imperfective categories share a common alignment profile
(the \textit{NOM-aspect domain} in Iridian's split-ergative system). In all
three, the main argument of the verb receives the direct case regardless of
transitivity, while the patient of a transitive verb appears in the indirect
case. This nominative pattern follows directly from the shared semantics of the
domain: because no culmination is foregrounded, no patient need occupy the
absolutive as the culmination-point of the event, and the agent instead takes
that structurally prominent position as the discourse anchor of the unbounded,
internally construed, or projected situation. The three imperfective categories
therefore form an alignment class not by historical accident but because they
share a viewpoint profile, and their nominative alignment is the morphosyntactic
reflex of that shared profile. The formal derivation is set out in
§\,\ref{sec:vm-asp}.

\paragraph{The continuous (\ird{-a}).}
The continuous is the unmarked member of the imperfective domain. It presents
a situation as obtaining at reference time without foregrounding change, onset,
active unfolding, or prospective projection. It is therefore the natural form
for stable states, contingent present conditions, broad durative situations, and
habitual or generic activity where what matters is that the situation holds
rather than how it is developing:

\pex
\a
\begingl
\gla Praže to mulša.//
\glb Prague.\Ind{} \textsc{loc} live-\Cont{}//
\glft \trsl{(I) live in Prague.} \lingcontext{stable durative state}//
\endgl
\a
\begingl
\gla Vytam shoda.//
\glb shirt.\Dir{} wear-\Cont{}//
\glft \trsl{(I) am wearing a shirt.} \lingcontext{contingent present condition}//
\endgl
\xe

Because the continuous does not foreground development or change, it is
compatible with predicates across the entire stative-dynamic spectrum, from
permanent properties to ongoing activities. Where the speaker wishes simply to
assert that a situation holds---without committing to whether it is in active
motion or stable---the continuous is the default choice. Its unmarked status
within the imperfective domain is reflected in the morphology: \ird{-a} is the
shortest and phonologically least specified of the three imperfective suffixes.

\paragraph{The progressive (\ird{-ime}).}
The progressive is the marked dynamic member of the imperfective domain.
Where the continuous merely asserts that a situation holds, the progressive
profiles an event specifically in its active unfolding phase---the period during
which it is dynamically occurring, not yet complete, and currently in motion.
It is accordingly most natural with activities and developing processes, where
the sense of an event in progress is pragmatically salient:

\pex
\a
\begingl
\gla Marek zmacime.//
\glb Marek.\Dir{} run-\Prog{}//
\glft \trsl{Marek is running.} \lingcontext{dynamic activity in progress}//
\endgl
\a
\begingl
\gla Marek travě piastime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.} \lingcontext{ongoing activity; patient in indirect}//
\endgl
\xe

The progressive is more marked than the continuous with stative predicates.
Where the continuous with a stative stem yields simple present-state
predication, the progressive imposes a developmental or ingressive
interpretation---the state is not merely current but is in the process of coming
to obtain. The minimal pair already shown in the alignment section illustrates
this:

\pex
\a
\begingl
\gla Vytam shoda.//
\glb shirt.\Dir{} wear-\Cont{}//
\glft \trsl{(I) am wearing a shirt.} \lingcontext{continuous: state holds at $R$}//
\endgl
\a
\begingl
\gla Vytam shodime.//
\glb shirt.\Dir{} wear-\Prog{}//
\glft \trsl{(I) am putting on a shirt.} \lingcontext{progressive: dynamic unfolding at $R$}//
\endgl
\xe

This contrast---between the continuous as the form of a holding situation and
the progressive as the form of an actively unfolding process---is the most
consequential internal opposition in the imperfective domain. The behaviour of
stative predicates provides especially sharp evidence for it and is examined in
detail in §\,\ref{sec:stative-stability}.

\paragraph{The contemplative (\ird{-ach}).}
The contemplative is the projective member of the imperfective domain. It
presents the eventuality as not yet realized at reference time but projected
from that point: the event is construed as intended, expected, impending,
planned, or otherwise prospectively oriented. The contemplative therefore does
not encode future tense---like all Iridian Asp.Align categories it is
indifferent to the relation between reference time and speech time---but rather
a prospective viewpoint in which reference time is interior to the projection
and the event lies ahead of it:

\pex
\a
\begingl
\gla Pražě de stożach.//
\glb Prague.\Ind{} to go-\textsc{contempl}{}//
\glft \trsl{(I) am about to go to Prague. / (I) intend to go to Prague.}//
\endgl
\a
\begingl
\gla Nužot mrocnach.//
\glb city.\Dir{}.\Def{} be.beautiful-\textsc{contempl}{}//
\glft \trsl{The city will be beautiful. / The city is expected to be beautiful.}//
\endgl
\xe

The projective semantics account for the range of English translations the
contemplative receives---\textit{about to}, \textit{intending to},
\textit{planning to}, or a simple future---depending on predicate type and
discourse context. With achievement predicates, which denote instantaneous
transitions, the contemplative is especially natural: instantaneous transitions
lend themselves readily to imminence and anticipation construals, where the
event is imminent or impending rather than in active progress.

The contemplative belongs to the imperfective domain because it shares the
defining property of that domain: it does not construe the event as a bounded
whole viewed from outside. Reference time precedes the unrealized eventuality,
which is prospective rather than complete. Culmination is not asserted, so the
agent---as the participant oriented toward or controlling the prospective
action---is the absolutive pivot, following the nominative alignment pattern of
the domain.

\paragraph{Interaction with lexical aspect.}
The three imperfective categories interact systematically with predicate class,
showing that they are distinct viewpoint operators rather than translation
variants of a single value. With \term{states}, the continuous gives ordinary
present-state predication; the progressive is marked and typically forces a
developmental or ingressive reading (\trsl{getting better}, \trsl{becoming
\ldots}); the contemplative gives an anticipated or expected-state reading
(\trsl{will be \ldots}, \trsl{is expected to be \ldots}). With
\term{activities}, the continuous gives broad ongoing or habitual
interpretation; the progressive gives a vivid, scene-internal reading in which
the activity is actively in course; the contemplative gives an intended or
pending-activity reading. With \term{accomplishments}, the continuous selects
the internal phase without projecting completion; the progressive highlights
active advancement toward culmination; the contemplative presents the event as
pending or aimed at, with the endpoint anticipated but not asserted. With
\term{achievements}, which denote instantaneous transitions, the continuous and
progressive typically require coercion into preparatory, iterative, incipient,
or extended interpretations, while the contemplative is often especially natural
because instantaneous transitions readily support imminence readings.

\paragraph{Discourse-pragmatic use.}
Because the imperfective domain places the agent in the absolutive, imperfective
clauses characteristically track the ongoing activity, condition, or orientation
of a discourse-prominent agent. The three categories play distinct roles within
this common frame. The continuous is the natural form for description and
framing---durable predications, stable conditions, and ongoing situations that
provide the setting or context for dynamic foreground events. It is the unmarked
imperfective of persistence and background. The progressive is suited to
scene-internal narration, vivid observation, and dynamic immediacy---contexts
where an event is presented as actively unfolding before the narrator's or
speaker's attention. It is characteristically a foregrounded imperfective: not
the background frame but the observed process. The contemplative serves
planning, expectation, prefiguration, and other forward-looking discourse
functions---expressing what an agent intends, what is expected to follow, or
what is imminent in the developing narrative.

\paragraph{Phasal particles and the imperfective domain.}
Imperfective categories are especially compatible with phasal modification
because they present the event internally or prospectively rather than as a
closed whole, leaving the event's phasal boundaries available for further
specification. The particle \ird{s} in its \trsl{still} reading combines
naturally with the continuous to assert the continuation of a situation:

\ex
\begingl
\gla S vytam shoda.//
\glb \textsc{still} shirt.\Dir{} wear-\Cont{}//
\glft \trsl{(I) am still wearing a shirt.}//
\endgl
\xe

The contemplative, presenting an event as prospective, is correspondingly
natural with \ird{s} in its \trsl{not yet} reading, describing a situation
that remains unrealized but is projected to obtain. The full phasal particle
system is described in Chapter~\ref{ch:function-words}; what matters here is
that the imperfective domain's open, non-bounded temporal profile is what
licenses phasal modification of this kind.

\paragraph{Evidence from stative predicates.}
Stative predicates furnish especially clear evidence for the contrast between
continuous and progressive, because with statives the interpretive difference
between the two is categorical. With a stable property predicate, the continuous
yields simple present-state predication: \ird{Nužot mrocna} \trsl{the city is
beautiful}. The progressive with the same predicate forces a developmental
reading: \ird{Nužot mrocnime} \trsl{the city is becoming beautiful}. The two
are not interchangeable. With temporary bodily or experiential conditions the
contrast is equally clear: the continuous describes the present condition as
simply obtaining (\trsl{is tired}), while the progressive more readily yields an
incipient or changing-state reading (\trsl{is getting tired}, \trsl{is starting
to feel sick}). This is not a stative-specific phenomenon; it instantiates the
general imperfective opposition described above. But the clarity with which
stative predicates exhibit it makes §\,\ref{sec:stative-stability} important
corroborating evidence for the distinction, and the reader is directed there for
fuller illustration and stative-specific consequences.


\subsection{Asp.Align under non-indicative Mood: impoverishment and reduced opposition}
\label{sec:vm-mood-asp}

No finite Iridian verb ever lacks an Asp.Align exponent. What changes under a
non-indicative Mood is not the presence of the Asp.Align head but the
\emph{richness} of its featural specification.

In the indicative, Asp.Align realizes the full five-way viewpoint inventory
(progressive, continuous, contemplative, perfective, retrospective) with the
associated split-ergative alignment. In the non-indicative moods---imperative,
prohibitive, hortative, and irrealis---a post-syntactic impoverishment rule
eliminates the fine-grained viewpoint features of the Asp node, retaining only
the coarser \term{alignment-domain} contrast that drives case assignment. The
result is a two-way \term{reduced Asp.Align} opposition shown in
Table~\ref{tab:reduced-asp}.

\begin{table}
  \footnotesize\sffamily
  \caption{Reduced Asp.Align exponents in non-indicative clauses.}
  \label{tab:reduced-asp}
  \medskip
  \begin{tabularx}{0.9\textwidth}{l l L}
    \toprule
    \textsc{domain} & \textsc{exponent} & \textsc{semantic range} \\
    \midrule
    Imperfective (IPFV) & $\emptyset$ & unbounded, ongoing, habitual, generic;
                                        processual commands and optatives \\
    Perfective (PFV)    & \ird{-ov-}  & bounded, single whole event, completive;
                                        ingressively directive \\
    \bottomrule
  \end{tabularx}
\end{table}

Crucially, the alignment consequences of these two domains are
identical to those operative in the full indicative system. The IPFV-domain
reduced Asp.Align triggers nominative-style alignment: the agent is the
absolutive (direct-case) argument and the patient receives the indirect. The
PFV-domain reduced Asp.Align (\ird{-ov-}) triggers ergative alignment: the
patient is the absolutive (direct-case) argument and the agent receives the
transitive case. No separate case-assigning principles are required for the
directive or irrealis domains; alignment is the output of the same head that
governs all finite clauses.

\paragraph{Morphological composition.}
The surface template for a non-indicative finite verb is therefore:

\begin{center}
\Stem{} + (\textsc{voice}) + (\Caus{}) + \textsc{reduced Asp.Align} + \textsc{mood}
\end{center}

\noindent where the reduced Asp.Align is silent ($\emptyset$) in the
IPFV domain and surfaces as \ird{-ov-} in the PFV domain. Representative
paradigms for the stem \ird{piašt-} `eat':

\begin{center}
\footnotesize\sffamily
\begin{tabular}{l l l l}
\toprule
\textsc{domain} & \textsc{imperative} & \textsc{prohibitive} & \textsc{hortative} \\
\midrule
IPFV ($\emptyset$) & \ird{piaštím}   & \ird{piašté}    & \ird{piaštká}   \\
PFV (\ird{-ov-})   & \ird{piaštovím} & \ird{piaštové}  & \ird{piaštovká} \\
\bottomrule
\end{tabular}
\end{center}

The PFV exponent \ird{-ov-} is invariant and does not
trigger the thematic consonant alternations conditioned by the mood
suffixes themselves; because \ird{-ov-} intervenes between the stem
and the palatalizing irrealis suffix \ird{-ěv}, stem palatalization
does not apply in the PFV irrealis (\ird{piaštověv}, not
\ird{*piaščověv}).

The IPFV/PFV opposition in directive clauses parallels, in function,
the well-known imperfective/perfective imperative distinction found in
West Slavic and other aspect languages: the IPFV imperative encodes an
ongoing, habitual, or processual directive (the event is construed as
unbounded), while the PFV imperative encodes a bounded, completive
directive (the event is construed as a single whole). In the irrealis,
the IPFV domain covers generic non-actual scenarios and standing
hypotheticals; the PFV domain covers single bounded counterfactuals.

In Distributed Morphology terms, the impoverishment rule may be stated as
follows. When MoodP bears a non-indicative value, a post-syntactic
impoverishment rule deletes the fine-grained viewpoint features
[$\pm$\textsc{bounded}, $\pm$\textsc{anterior}, \textsc{prospective}] of the Asp
node, leaving only the coarser feature [$\pm$\textsc{pfv-domain}]. Late
Vocabulary Insertion then maps this reduced bundle to $\emptyset$ (IPFV) or
\ird{-ov-} (PFV). The rule is:

\medskip
\noindent\begin{tabular}{@{}l@{\quad}l}
$[\textsc{Mood}: \text{non-indicative}]$ & $\Rightarrow$ suppress fine
aspectual contrasts within Asp.Align; \\
& \hspace*{1.5em}retain only $[\pm\textsc{pfv-domain}]$.
\end{tabular}
\medskip

Syntactically, Mood remains the higher head (MoodP $>$ AspP), consistent with
standard scope expectations and with Cinque's functional hierarchy.
Morphologically, however, Mood becomes \emph{dependent} on Asp.Align: Mood never
surfaces on a finite verb unless Asp.Align has been realized. The interaction is
therefore one of impoverishment rather than competition.

The pattern observable in directive clauses---patient or theme in the direct
case, overt speech-act participant subject in the transitive when it would
otherwise compete for the direct slot---follows from the interaction of reduced
Asp.Align with null-subject orientation, and requires no additional
case-assigning engine. In the default directive situation, the agent is null
(the understood second-person subject of an imperative or prohibitive is
unexpressed). Under IPFV-domain alignment, the agent is nominally the
absolutive, but with the agent unrealized, the patient absorbs the direct case
as the clause's sole overt core argument. Under PFV-domain alignment, the
patient is the absolutive by definition and the null agent is simply unrealized
in the transitive. In both domains, the patient or theme characteristically
occupies the direct case---not by a separate directive case principle, but by
the confluence of reduced Asp.Align and null-subject orientation. The
competition-driven transitive marking of overt speech-act participant subjects
(§\,\ref{sec:mood-directive-case}) is likewise a surface reflex of the
absolutive orientation already encoded by Asp.Align: an overt agent would
ordinarily be the absolutive under IPFV alignment, but when a competing direct
object is present, that structural pressure forces the subject into the
transitive to preserve the patient as the clause's single direct nominal.


\section{Valency operations}

Three derivational operations may be applied to the verb stem before the
aspect-alignment suffix: the antipassive, the applicative, and the causative.
Where more than one is present, the order is fixed: antipassive or applicative
precedes the causative.

\subsection{The antipassive (\ird{-aš-})}
\label{sec:vm-antipassive-desc}

The antipassive is an overt Voice head that applies only where the
patient is structurally the absolutive participant. It displaces that
patient from absolutive status, promotes the agent to absolutive, and
removes the patient from core argument structure. Because the
structural target it requires --- a patient standing as the
absolutive --- is available only in ERG-aspect clauses, the
antipassive is exclusively an ergative-domain construction. It is
therefore the obligatory non-ergative construction for any verb
capable of bearing the Voice head \ird{-n-} in
the perfective and retrospective aspects: such a verb must appear
with either \ird{-n-} (ergative transitive) or \ird{-aš-}
(antipassive). The bare perfective serves no antipassive function and
is reserved exclusively for inherently intransitive verbs, as
described in Section~\ref{sec:ambitransitive}.

The antipassive suffix is \ird{-aš-}, though its phonological realisation depends upon the thematic consonant class of the stem, following the alternations set out in Table~\ref{tab:verbclasses}. The stem-final consonant is replaced or augmented as shown; the underlying \ird{-aš-} string surfaces as the modified stem consonant followed by the relevant sibilant. Thus \ird{piašt-} yields \ird{piašč-} as its antipassive stem:

\pex
\a
\begingl
\gla Marek piašček.//
\glb Marek.\Dir{} eat-\Antip{}-\Pf{}//
\glft \trsl{Marek ate.}//
\endgl
\a
\begingl
\gla Marek houlam se piašček.//
\glb Marek.\Dir{} spoon.\Ind{} \textsc{inst} eat-\Antip{}-\Pf{}//
\glft \trsl{Marek ate with a spoon.} \lingcontext{instrument adjunct in \Ind{}}//
\endgl
\xe

The patient, having been removed from core argument structure by the antipassive, may not be expressed in any case form; it is accordingly inadmissible as an indirect-case argument, as may be seen from the following:

\ex
\begingl
\gla *Marek travě piašček.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Antip{}-\Pf{}//
\glft \trsl{(intended: Marek ate bread.)}//
\endgl
\xe

\noindent Where a patient is to be expressed in any form, the ergative construction is required. Instrument and adjunct arguments, which bear no patient relation to the predicate, remain fully admissible in the indirect case, as in the spoon example above. The absence of an expressible patient is not merely a discourse choice to background a participant that is otherwise retained in syntax: under the antipassive, the patient is removed from the clause's core argument structure and therefore cannot surface in any case form. The construction does not merely withhold the patient from prominence; it eliminates the patient from the overt argument structure of the clause. For an inherently transitive verb the patient is typically still understood as part of the conceptual event, but what is asserted is the agent's activity; the patient has no realization as a syntactic argument.

Equally, for a transitive verb, the bare perfective without either \ird{-n-} or \ird{-aš-} is ungrammatical:

\ex
\begingl
\gla *Marek piaštek.//
\glb Marek.\Dir{} eat-\Pf{}//
\glft \trsl{(intended: Marek ate.)} \lingcontext{bare perfective illicit for transitive verb}//
\endgl
\xe

\noindent\textit{Note:} The antipassive and the applicative are in complementary distribution; a verb cannot take both simultaneously.


\subsubsection{Minimalist analysis: the antipassive as VoiceP}
\label{sec:vm-antipassive}

The antipassive \ird{-aš-} is an overt Voice head (Kratzer 1996;
Alexiadou et~al. 2006) subject to a structural input condition: it
may be merged only in a clause whose input to VoiceP contains a
patient that is eligible for absolutive realization. It then removes
that patient from core argument structure and promotes the agent to
the absolutive position vacated by that removal.

\paragraph{Input condition.}
The antipassive requires, as the structural context for its
application, a clause in which the patient stands in the absolutive
position made available by the clause's alignment architecture. In
the ERG-aspect domain, Asp.Align's default assigns absolutive to the
patient (§\,\ref{sec:vm-asp-case}); the input condition is therefore
satisfied, and the antipassive is licensed. In the NOM-aspect domain,
the absolutive is assigned by default to the agent; no patient stands
in the absolutive position, and the structural target required by the
antipassive is absent. The antipassive is consequently unlicensed in
NOM-aspect clauses. This is a hard grammatical restriction, not a
preference or economy effect.

\paragraph{Argument-structure removal.}
Under the antipassive, the patient is not demoted to an oblique slot
within the clause. It is removed from the clausal core entirely.
Oblique expression of the patient is therefore not merely
dispreferred but impossible: the construction does not retain a
suppressed patient that has been reassigned to a peripheral position.
The patient ceases to be part of the overt argument structure of the
clause. This is why an indirect-case patient is ungrammatical with
the antipassive (see §\,\ref{sec:vm-antipassive-desc} for examples),
while instrument and adjunct arguments in the indirect case remain
fully admissible.

The template order --- antipassive between root and Asp.Align ---
confirms that VoiceP immediately dominates VP:

\begin{center}
\begin{forest}
for tree={align=center}
[AspP
  [{DP\\{\footnotesize agent}\\{\footnotesize\textsc{[abs]}}}]
  [{Asp$'$}
    [{Asp\\{\footnotesize\textsc{[erg]}}}]
    [VoiceP
      [{DP\\{\footnotesize agent}}]
      [{Voice$'$}
        [{Voice\\\ird{-aš-}\\{\footnotesize\textsc{[agt-abs]}}\\{\scriptsize requires p-abs}}]
        [VP
          [{\textit{root}}]
        ]
      ]
    ]
  ]
]
\end{forest}
\end{center}

\noindent The label \textsc{[p-abs required]} on the antipassive Voice
head represents the input condition: the head selects for a VP whose
internal argument would stand as the absolutive participant in the
relevant alignment environment. In ERG aspects this condition is met;
in NOM aspects it is not. The complementary distribution of the
antipassive and the applicative (\ird{-aš-} and \ird{-éb-} cannot
co-occur on the same stem) is a direct consequence of both occupying
the same VoiceP projection: only one Voice head may be merged at this
tier.

In ERG-aspect transitives, the patient requires overt Voice licensing
to be realized as a core argument (§\,\ref{sec:vm-n}). A verb bearing
neither \ird{-aš-} nor \ird{-éb-} nor \ird{-n-} in an ergative
aspect therefore fails for a different reason: the patient lacks the
overt patient-licensing head required for core realization. That
exclusion is distinct from the structural input condition that blocks
the antipassive in NOM aspects; the two empty cells in the alignment
system are each excluded by a separate principle (§\,\ref{sec:vm-asp-case}).


\subsection{The applicative (\ird{-éb-})}

The applicative promotes a peripheral argument --- most commonly a beneficiary or recipient, but also a goal, location, or instrument --- to core argument status, placing it in the absolutive. The suffix \ird{-éb-} derives from an older benefactive morpheme and still defaults to a beneficiary or recipient reading when such an interpretation is available. Where the lexical semantics of the verb or the wider clause exclude a benefactive reading, the applicative is interpreted instead as promoting the most discourse-salient non-core participant, typically a goal, location, or instrument. Only one applicative argument may be promoted in a single clause.

The contrast is most readily observed in nominative-alignment contexts, where the applied participant becomes the direct-case object that would otherwise have remained oblique:

\pex
\a
\begingl
\gla Marek bylam piastime.//
\glb Marek.\Dir{} child.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating for the child.} \lingcontext{beneficiary remains oblique}//
\endgl
\a
\begingl
\gla Marek bylam piaštébime.//
\glb Marek.\Dir{} child.\Dir{} eat-\Appl{}-\Prog{}//
\glft \trsl{Marek is feeding the child.} \lingcontext{beneficiary promoted to core status}//
\endgl
\xe

\noindent With verbs of motion and placement the same suffix promotes a goal or location rather than a beneficiary:

\pex
\a
\begingl
\gla Marek nužam stachime.//
\glb Marek.\Dir{} town.\Ind{} go-\Prog{}//
\glft \trsl{Marek is going toward the town.}//
\endgl
\a
\begingl
\gla Marek hrad stéběme.//
\glb Marek.\Dir{} town.\Dir{} go-\Appl{}-\Prog{}//
\glft \trsl{Marek is entering the town / making the town his destination.}//
\endgl
\xe

\noindent Instrumental promotion is likewise possible, though it is pragmatically more marked, being preferred where the instrument itself is the discourse-relevant participant:

\pex
\a
\begingl
\gla Marek houlam se piašček.//
\glb Marek.\Dir{} spoon.\Ind{} \textsc{inst} eat-\Antip{}-\Pf{}//
\glft \trsl{Marek ate with a spoon.} \lingcontext{instrument remains adjunct}//
\endgl
\a
\begingl
\gla Marek houlam piaštébime.//
\glb Marek.\Dir{} spoon.\Dir{} eat-\Appl{}-\Prog{}//
\glft \trsl{Marek is using the spoon for eating / is eating with the spoon as the relevant implement.}//
\endgl
\xe

\noindent In the imperative and hortative the same promotion yields benefactive readings by default. Directive clauses require the beneficiary to be overtly realized; the applicative is ungrammatical without one (see §\,\ref{sec:mood-directive-valency}):

\pex
\a
\begingl
\gla Trava bylam piaštébím.//
\glb bread.\Dir{} child.\Ind{} eat-\Appl{}-\Imp{}//
\glft \trsl{Feed the child bread.}//
\endgl
\a
\begingl
\gla Trava bylam piaštébká.//
\glb bread.\Dir{} child.\Ind{} eat-\Appl{}-\Hort{}//
\glft \trsl{Let's feed the child bread.}//
\endgl
\xe

\noindent The antipassive and the applicative are in complementary distribution. This is not an accidental gap: the antipassive removes the patient from core argument structure, whereas the applicative creates a new core object out of a peripheral participant. The two operations compete for the same VoiceP slot and cannot be combined on a single finite stem.


\subsubsection{Minimalist analysis: the applicative as ApplP}
\label{sec:vm-appl}

The applicative (\ird{-éb-}) promotes a peripheral (non-core)
participant to absolutive status. Following Pylkkänen
(2008), this can be analyzed as a
\term{high applicative} head (ApplP) that introduces the promoted
argument into the event structure between VoiceP and CausP:

\begin{center}
\begin{forest}
for tree={align=center}
[ApplP
  [{DP\\{\footnotesize applied argument}}]
  [{Appl$'$}
    [{Appl\\\ird{-éb-}}]
    [VoiceP
      [{\ldots}]
    ]
  ]
]
\end{forest}
\end{center}

\noindent The beneficiary, goal, or instrument promoted by the
applicative becomes the absolutive argument of the resulting vP,
effectively taking the slot that would otherwise be occupied by the
patient. The interaction with discourse is mediated at the AspP
level: which of the available absolutive candidates (patient or
applied argument) becomes the topic absolutive is determined by the
Asp.Align morpheme selected (§\,\ref{sec:vm-asp}).

%% STUB: The question of whether Iridian uses a high or low
%% applicative (Pylkkänen's distinction) should be tested
%% diagnostically. High applicatives introduce an argument that
%% relates to the event; low applicatives introduce an argument
%% related directly to the patient. In Iridian, beneficiary
%% promotion appears to be high (no entailment of patient
%% affectedness), while instrument promotion may be low.
%% This distinction bears on whether there are two distinct ApplP
%% projections, or a single one with type-sensitive semantics.


\subsection{The causative (\ird{-na-})}

Cross-linguistically, causation may be expressed lexically, analytically, or morphologically. Some languages employ distinct lexical verbs comparable to English \textit{kill}, while others use periphrastic constructions equivalent to \textit{make} or \textit{cause} plus a second predicate. Iridian belongs to the third type: causation is expressed by a productive morphological operation, realized synchronically as the suffix \ird{-na-}.

The causative \ird{-na-} adds an external causer argument to the event. In nominative-alignment contexts the causer stands in the direct case, the original agent or sole participant is demoted to the indirect, and any patient licensed by the base verb remains indirect. In ergative-alignment contexts the causer takes the transitive case. With transitive bases, the original patient remains the direct argument and the causee falls into the indirect; with intransitive bases, the causee itself occupies the direct as the sole affected participant. The causative thus increases valency by one while preserving the aspect-driven alignment pattern of the clause as a whole. Before a following vowel-initial aspect or mood suffix, \ird{-na-} inserts an epenthetic glide and surfaces as \ird{-naj-}; before consonant-initial material, including the Voice head \ird{-n-}, it remains \ird{-na-}.

\pex
\a
\begingl
\gla Marek bylam zmacnajime.//
\glb Marek.\Dir{} child.\Ind{} run-\Caus{}-\Prog{}//
\glft \trsl{Marek is making the child run.} \lingcontext{causative of intransitive base; nominative alignment}//
\endgl
\a
\begingl
\gla Travat Marku bylam piašt-na-nik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} child.\Ind{} eat-\Caus{}-\textsc{trs}-\Pf{}//
\glft \trsl{Marek made the child eat the bread.} \lingcontext{causative of transitive base; ergative alignment}//
\endgl
\xe

\noindent The semantic character of the causation is expressed primarily through aspect. In the perfective the causative presents the causer as exercising effective control over the caused event and therefore tends toward direct, interventionist, or coercive readings. In the retrospective the same morphology presents a looser relation between causer and event, so that the clause more readily receives indirect, delegated, accidental, or permissive interpretations. For the discourse consequences of this contrast, see Chapter~\ref{ch:discourse}.

\pex
\a
\begingl
\gla Marek bylam piaštnajime.//
\glb Marek.\Dir{} child.\Ind{} eat-\Caus{}-\Prog{}//
\glft \trsl{Marek is having the child eat / is feeding the child.}//
\endgl
\a
\begingl
\gla Travat Marku bylam piašt-na-n-ac.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} child.\Ind{} eat-\Caus{}-\textsc{trs}-\Ret{}//
\glft \trsl{Marek has let the child eat the bread / has ended up getting the child to eat it.}//
\endgl
\xe

\noindent The causative is fully compatible with the applicative, provided the applicative is formed first in accordance with the morphological template:

\pex
\a
\begingl
\gla Marek bylam piaštébnajime.//
\glb Marek.\Dir{} child.\Ind{} eat-\Appl{}-\Caus{}-\Prog{}//
\glft \trsl{Marek is making someone feed the child.}//
\endgl
\a
\begingl
\gla Trava bylam piaštnajím.//
\glb bread.\Dir{} child.\Ind{} eat-\Caus{}-\Imp{}//
\glft \trsl{Make the child eat the bread.} \lingcontext{directive; causee in \Ind{} required}//
\endgl
\a
\begingl
\gla Trava bylam piašt-na-ká.//
\glb bread.\Dir{} child.\Ind{} eat-\Caus{}-\Hort{}//
\glft \trsl{Let's make the child eat the bread.}//
\endgl
\xe

\noindent Synchronically, \ird{-na-} is distinct from the Voice head \ird{-n-}. The former is a valency-changing derivation occupying the causative slot of the verbal template (CausP); the latter is the Voice head (VoiceP) that licenses the internal argument and specifies patient absolutive, inserted immediately before the aspect suffix. Earlier notes used \ird{-ne-} for the causative, but the present grammar regularises \ird{-na-} in order to keep the causative head formally distinct from \ird{-n-}.


\subsubsection{Minimalist analysis: the causative as CausP}
\label{sec:vm-caus}

The causative (\ird{-na-}) introduces an external causer above the
base event structure, corresponding to a \term{CausP} projection
that dominates VoiceP/ApplP. The causer argument is merged in
Spec,CausP:

\begin{center}
\begin{forest}
for tree={align=center}
[CausP
  [{DP\\{\footnotesize causer}}]
  [{Caus$'$}
    [{Caus\\\ird{-na-}}]
    [ApplP
      [{\ldots}]
    ]
  ]
]
\end{forest}
\end{center}

\noindent The morphological order (applicative before causative
before Asp.Align) is predicted: CausP dominates VoiceP/ApplP, and
the Mirror Principle assigns them to the innermost positions relative
to AspP. The interaction with aspect is semantically significant
(§\,\ref{sec:aspect-caus}): perfective causation presents direct
causer control (CausP features trigger tight Agree with the event
head), while retrospective causation presents indirect or
permissive causation (looser Agree relation, admitting delegated or
accidental readings).

\subsubsection{The Voice head \ird{-n-}}
\label{sec:vm-n}

One doctrinal point should be made plain at the outset. \ird{-n-} is
not inherently a ``passive suffix,'' nor does it change its meaning
between aspectual domains. It is a single Voice head with constant
structural content: it licenses the internal argument (patient) as a
core participant in the clause's argument structure and specifies that
the patient is to stand as the absolutive (\textsc{[pat-abs]}). In
the ERG-aspect domain, where Asp.Align already establishes a
patient-oriented absolutive orientation, this produces the ordinary
ergative transitive: the patient is licensed as the absolutive pivot
and the agent receives the dependent transitive case. In the
NOM-aspect domain, where Asp.Align otherwise privileges the agent,
the same patient-absolutive Voice head overrides that default and
yields the construction described here as the passive. The passive
reading is therefore an effect of the interaction of \ird{-n-} with
the nominative domain, not the lexical meaning of \ird{-n-} itself.
This preserves the unity of the morphology and removes any appearance
that \ird{-n-} mysteriously changes meaning from one aspectual domain
to another.

The morpheme \ird{-n-} thus occupies the same VoiceP slot as the
antipassive \ird{-aš-} and the applicative \ird{-éb-}. Its two
defining properties are not environment-specific; they are always
present. What varies across aspectual domains is their structural
consequence within the Asp.Align system.

\paragraph{ERG-aspect behaviour.}
In the perfective and retrospective aspects the default absolutive
orientation of Asp.Align already favours the patient (§\,\ref{sec:vm-asp-case}).
The \textsc{[pat-abs]} specification contributed by \ird{-n-} is
therefore harmonious with the aspectual default. The indispensable
contribution of \ird{-n-} in this environment is accordingly not to
override the default but to license the patient as a core argument:
without an overt Voice head in VoiceP, a transitive predicate cannot
realize its patient as a properly licensed participant in the clause's
core argument structure. The bare perfective or retrospective is
consequently illicit for any transitive reading, as described in
§\,\ref{sec:ambitransitive}. With \ird{-n-} present, the patient is
licensed as the absolutive and the agent takes the transitive case;
the result is the ordinary ergative transitive construction.

\begin{center}
\begin{forest}
for tree={align=center}
[AspP
  [{DP\\{\footnotesize patient}\\{\footnotesize\textsc{[abs]}}}]
  [{Asp$'$}
    [{Asp\\{\footnotesize\textsc{[erg]}}}]
    [VoiceP
      [{DP\\{\footnotesize agent}\\{\footnotesize\textsc{[trs]}}}]
      [{Voice$'$}
        [{Voice\\\ird{-n-}\\{\footnotesize\textsc{[pat-abs]}}}]
        [VP
          [{DP\\{\footnotesize patient}}]
          [{\textit{root}}]
        ]
      ]
    ]
  ]
]
\end{forest}
\end{center}

\paragraph{NOM-aspect behaviour.}
In the continuous, progressive, and contemplative aspects the same
overt Voice head overrides the default agent-absolutive orientation
of Asp.Align. Because \ird{-n-} specifies \textsc{[pat-abs]}, the
patient rather than the agent becomes the direct-case pivot, and the
agent is demoted to the oblique domain. The result is a
\term{passive} construction. Asp.Align remains in the nominative
domain; it is the Voice head that imposes a patient-absolutive
realization upon the clause. The full analysis and examples are in
§\,\ref{sec:vm-passive}.

\ird{-n-} is thus not a pair of context-sensitive morphemes whose
function changes by aspect. It is a single Voice head whose constant
properties --- patient licensing and patient absolutive --- produce
different surface constructions depending on whether the aspectual
domain reinforces or is overridden by its absolutive specification.
The phonological resemblance to the causative \ird{-na-} is
addressed in §\,\ref{sec:vm-tension-n}.


\subsection{The passive}\label{sec:vm-passive}

The passive is the construction that arises when the Voice head
\ird{-n-} is merged in a NOM-aspect clause. Since
\ird{-n-} specifies patient absolutive (\textsc{[pat-abs]}) and
the NOM-aspect domain would by default assign absolutive to the agent,
the Voice head overrides the aspectual default: the patient surfaces
in the direct case as the absolutive pivot, and the agent is demoted
to the indirect (oblique) case. Asp.Align remains in the
nominative domain throughout; only the absolutive orientation of
the clause is altered by the overt Voice head.

The passive is thus the marked construction by which a
nominative-domain clause is reorganized around the patient, allowing
the patient rather than the agent to function as the discourse pivot.
Its counterpart in the ergative domain is the ordinary ergative
transitive, which arises from the same Voice head under ERG-aspect
alignment. The two constructions are formally identical at the Voice
level; their different surface alignment is a consequence of the
aspectual domain they are embedded in.

\subsubsection*{Morphological form}

The passive consists of a NOM-aspect suffix on a transitive stem
bearing \ird{-n-}. The NOM-aspect suffixes available for passive
formation are the continuous \ird{-a}, the progressive \ird{-ime},
and the contemplative \ird{-ach}; the patient-absolutive
specification of \ird{-n-} is realized in each case.

\subsubsection*{Case frame}

The patient surfaces in the direct case (\textsc{abs}/\textsc{dir});
the agent, displaced from the absolutive, surfaces in the indirect
case (\textsc{obl}/\textsc{ind}). The agent may be omitted when it
is recoverable or irrelevant.
\begin{center}
\begin{forest}
for tree={align=center}
[AspP
  [{DP\\{\footnotesize patient}\\{\footnotesize\textsc{[abs]}}}]
  [{Asp$'$}
    [{Asp\\{\footnotesize\textsc{[nom]}}}]
    [VoiceP
      [{DP\\{\footnotesize agent}\\{\footnotesize\textsc{[obl]}}}]
      [{Voice$'$}
        [{Voice\\\ird{-n-}\\{\footnotesize\textsc{[pat-abs]}}}]
        [VP
          [{DP\\{\footnotesize patient}}]
          [{\textit{root}}]
        ]
      ]
    ]
  ]
]
\end{forest}
\end{center}

\noindent Asp.Align remains nominative-domain (\textsc{Asp[nom]}),
but the overt Voice head requires patient absolutive; the agent
therefore surfaces in the oblique domain.

\subsubsection*{Examples}

With agent expressed in the oblique:

\pex
\a
\begingl
\gla Travat Marku piastnime.//
\glb bread.\Dir{}.\Def{} Marek.\Ind{} eat-\textsc{pass}-\Prog{}//
\glft \trsl{The bread is being eaten by Marek.} \lingcontext{passive; P = \Dir{}, A = \Ind{}}//
\endgl
\a
\begingl
\gla Vytam Marku shodnime.//
\glb shirt.\Dir{}.\Def{} Marek.\Ind{} wear-\textsc{pass}-\Prog{}//
\glft \trsl{The shirt is being put on by Marek.}//
\endgl
\xe

With agent omitted:

\pex
\a
\begingl
\gla Travat piastnime.//
\glb bread.\Dir{}.\Def{} eat-\textsc{pass}-\Prog{}//
\glft \trsl{The bread is being eaten.} \lingcontext{passive; agent omitted}//
\endgl
\a
\begingl
\gla Doma piasčna.//
\glb house.\Dir{} build-\textsc{pass}-\Cont{}//
\glft \trsl{The house is being built / is under construction.}//
\endgl
\xe

\noindent The discourse value of the passive is to constitute the
patient as the clause's absolutive topic, making it the argument
that subsequent clauses may pick up pronominally or by topic drop.
The agent, when expressed, appears in the indirect case as a
non-core oblique participant.


\subsection{Pragmatic use of valency operations}

The choice among ergative, antipassive, applicative, and causative is not purely
mechanical. Each construction offers the speaker a distinct way of packaging the
event for discourse. The antipassive is used when the patient is generic,
non-referential, semantically incorporated into the activity, recoverable from
context, or intentionally backgrounded. Its effect is to keep the agent in
absolutive position and to present the clause as being about what the agent did
rather than about what happened to some patient.

The applicative serves the opposite discourse function: it pulls a formerly
peripheral participant into the centre of the clause. A beneficiary,
destination, location, or instrument is promoted when that participant is the
one the discourse is tracking, when it will remain topical in subsequent
clauses, or when the speaker wishes to construe the event from its perspective
rather than from that of an unexpressed or backgrounded patient.

The causative likewise has a pragmatic life beyond its valency-changing effect.
By choosing a causative rather than a simple predicate, the speaker frames the
event in terms of inducement, delegated agency, or control. In direct contexts
this may foreground authority or coercion; in weaker retrospective contexts it
may soften the causer's responsibility, presenting the caused event as something
merely permitted, arranged, or indirectly brought about rather than personally
carried out.

The applicative and causative are compatible with the imperative and hortative
and occupy their regular positions in the morphological template before the mood
suffix.


\section{Negation}\label{sec:negation}

The negative prefix \ird{zá-} --- reduced to \ird{za-} in unstressed environments --- stands before the entire verbal complex. In the indicative it accordingly precedes the stem itself. In the imperative, hortative, and irrealis, negation is expressed by distinct mood-specific negative suffixes rather than by \ird{zá-}; see Section~\ref{sec:mood} and §\,\ref{sec:vm-moodp}.

This section describes verbal negation --- the use of \ird{zá-} with finite indicative predicates --- exclusively. Negation in equative clauses (zero-copula constructions) and existential clauses (the \ird{jec-}/\ird{zad-} alternation) is treated in §\,\ref{sec:clausal-negation} of Chapter~\ref{ch:simple}.

One framing point is worth flagging before the formal analysis. \ird{Zá-} is the most morphologically prominent of the three NegP exponents in Iridian, but it is also the structurally anomalous one: it surfaces as a prefix rather than in the head-final position that NegP occupies in the clause spine. In equative clauses, by contrast, Neg\textsuperscript{0} is realized as the free morpheme \ird{staly}, which surfaces exactly where the clause spine predicts --- clause-finally, after its two nominal arguments. \ird{Staly} thus provides the clearest direct evidence for the head-final status of NegP; the prefix position of \ird{zá-} is a PF-level cliticization effect, not the canonical position of the negation head. The full argument is in §\,\ref{sec:vm-neg-exponents}.

The absence of \ird{zá-} in the prohibitive (\ird{-é}), the negative hortative (\ird{-ku}), and the negative irrealis (\ird{-ir}) is not arbitrary. Each of these suffixes carries [+\textsc{neg}] as a lexically inherent feature of the MoodP Vocabulary Item; when such a head is selected, NegP is simply absent from the derivation and \ird{zá-} has no Neg\textsuperscript{0} node to spell out. There is no separate blocking principle at work: the absence of NegP is the full explanation. See §\,\ref{sec:vm-moodp} for the MoodP paradigm and analysis.


\subsection{Formal analysis: NegP and the prefix paradox}
\label{sec:vm-neg}
\label{sec:vm-neg-exponents}

The negative prefix \ird{zá-} stands before the entire verbal stem.
Within the clause spine (Table~\ref{tab:th-clause-spine}), NegP
occupies projection 4, above AspP but below IP. If the Mirror
Principle held perfectly, \ird{zá-} should appear \emph{after} the
Asp.Align suffix, since NegP dominates AspP and therefore corresponds
to a more peripheral morpheme. Instead it appears as a prefix, inside
the verbal complex to the left of the root.

Three analyses are possible. Under V-to-Neg head movement, the verbal
complex undergoes further movement to the Neg head in narrow syntax,
making \ird{zá-} the Neg head with the raised complex to its right;
but in a head-final language the Neg head should follow its complement,
predicting \ird{*piaštek zá-}, so the account requires either
surface-right adjunction or a PF reordering rule. A scope-based prefix
convention (Bybee 1985) treats the prefix position as a cross-linguistic
tendency for high-scope operators to surface as prefixes in verb-final
languages (cf.\ Turkish, Mongolian, Basque \textit{ez}), making
the Mirror Principle simply inapplicable to negation. The preferred
account is DM morphological incorporation at PF (Embick \& Noyer 2007):
NegP is a syntactic head above AspP, but \ird{zá-} is specified as a
prefix and at PF phonologically cliticizes to the left edge of the
verbal complex, leaving the syntactic hierarchy intact. The
procliticization itself is direct evidence: \ird{zá-} does not
form an independent prosodic word.

That the DM account is correct emerges most clearly from a unified
treatment of Neg\textsuperscript{0} across all clause types. Treating
\ird{zá-} as the sole reflex of NegP is incomplete: Neg\textsuperscript{0}
has three PF exponents conditioned by predicate type. With a finite
verbal predicate, Neg\textsuperscript{0} is realized as the procliticized
prefix \ird{zá-} attached to the left edge of the verbal complex; because
that complex is itself clause-final by head-finality, the prefix surfaces
immediately before the root, yielding propositional negation of the event
description. With a zero-copula equative or class-membership predicate,
there is no verbal host for \ird{zá-} to cliticize onto, and
Neg\textsuperscript{0} is instead spelled out as the free morpheme
\ird{staly}, which surfaces in the canonical head-final position following
its two nominal arguments --- the position expected of any clause-final
head, requiring no additional movement rule. When the predicative
complement is a plain DP the result is \term{predicational negation}
(\ird{Marek doktor staly}, \trsl{Marek is not a doctor}); when it is a
participant nominal (§\,\ref{sec:nmlz-predicative}) the result is
\term{identificational negation} (§\,\ref{sec:vm-ident-neg} below).
With the existential verb \ird{jec-}, Neg\textsuperscript{0} triggers a
DM readjustment: the existential root is realized as the suppletive
allomorph \ird{zad-} and Neg\textsuperscript{0} itself has no independent
PF exponent, negation being absorbed into the root via Vocabulary
Insertion --- the standard behavior of suppletive negative existentials
cross-linguistically (cf.\ Turkish \textit{yok}/\textit{var};
Japanese \textit{nai}/\textit{aru}).

The complete distribution, including mood-internal negation, is
summarized in Table~\ref{tab:neg-exponents} of Chapter~\ref{ch:simple}.
The three environments are a conditioned allomorphy of a single
Neg\textsuperscript{0} head, not distinct lexical negation strategies.
NegP occupies the same structural position (above AspP, below MoodP)
in all clause types; what varies is the PF exponent of
Neg\textsuperscript{0} and the semantic scope of negation, which is
determined by the semantics of the predication NegP takes in its scope.

\begin{figure}[ht]
  \footnotesize\sffamily
  \caption{Comparative exponence of Neg\textsuperscript{0} across clause types. In all three structures NegP occupies the same position above AspP or the zero-copula predication; what varies is the PF realization of Neg\textsuperscript{0}. Only the verbal clause shows the prefix-paradox effect, with \ird{zá-} cliticized to the left edge of the verbal complex rather than surfacing in canonical head-final position.}\label{fig:neg0-comparative}
  \medskip
  \centering
  \scriptsize
  \begin{minipage}[t]{0.30\textwidth}
    \centering
    \textsc{Verbal clause}\par
    \medskip
    \begin{forest}
      syntax tree, for tree={l sep=8pt,s sep=6pt,inner sep=0.5pt}
      [NegP
        [{DP\\{\footnotesize topic}}]
        [{Neg$'$}
          [{Neg$^{0}$}\\\ird{zá-}\\{\small(PF prefix)}]
          [AspP
            [{Asp$'$}
              [{Asp\\{\footnotesize\textit{asp.align}}}]
              [VP
                [{V\\{\footnotesize\textit{root}}}]
              ]
            ]
          ]
        ]
      ]
    \end{forest}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.30\textwidth}
    \centering
    \textsc{Equative clause}\par
    \medskip
    \begin{forest}
      syntax tree, for tree={l sep=8pt,s sep=6pt,inner sep=0.5pt}
      [NegP
        [{DP\\{\footnotesize topic}}]
        [{Neg$'$}
          [{Neg$^{0}$\\\ird{staly}}]
          [EquP
            [{DP\\{\footnotesize pred.}}]
            [{Eq$'$}
              [{Eq$^{0}$\\{$\varnothing$}}]
            ]
          ]
        ]
      ]
    \end{forest}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.30\textwidth}
    \centering
    \textsc{Existential clause}\par
    \medskip
    \begin{forest}
      syntax tree, for tree={l sep=8pt,s sep=6pt,inner sep=0.5pt}
      [NegP
        [{DP\\{\footnotesize pivot}}]
        [{Neg$'$}
          [{Neg$^{0}$\\{\small($\varnothing$: absorbed)}}]
          [AspP
            [{Asp$'$}
              [{Asp\\{\footnotesize\textit{asp}}}]
              [ExistP
                [{Exist$'$}
                  [{Exist$^{0}$\\\ird{zad-}\\{\small(suppletive)}}]
                ]
              ]
            ]
          ]
        ]
      ]
    \end{forest}
  \end{minipage}
\end{figure}

An important consequence follows for the prefix paradox. The paradox
arises specifically for verbal predicates, where \ird{zá-} must attach
to a phonological host; Figure~\ref{fig:neg0-comparative} makes clear
that this is the outlier among the three exponents, not the structural
default. In equative clauses, where no such host exists, \ird{staly}
surfaces in the expected head-final position --- after its complement
(the nominal predication) --- and therefore gives the clearest direct
evidence for the head-final status of NegP. In existential clauses,
\ird{zad-} shows the same Neg\textsuperscript{0} head being discharged
through suppletive Vocabulary Insertion at Exist\textsuperscript{0}
rather than by an independent overt exponent. The prefixhood of
\ird{zá-} is therefore a PF-level cliticization effect, not the
canonical behavior of Neg\textsuperscript{0}.

For equative clauses specifically, the word order reflects the
head-final NegP structure transparently:

\begin{center}
\begin{forest}
  syntax tree, for tree={l sep=8pt,s sep=6pt,inner sep=0.5pt}
  [NegP
    [{DP\\{\footnotesize topic}}]
    [{Neg$'$}
      [{DP\\{\footnotesize pred.}}]
      [{Neg$^{0}$\\\ird{staly}}]
    ]
  ]
\end{forest}
\end{center}

\noindent The two nominal constituents before \ird{staly} may permute
in accordance with the topic-comment architecture (§\,\ref{sec:topic-comment}),
while \ird{staly} is invariably clause-final.

The incompatibility of \ird{zad-} with \ird{zá-} also follows without additional
stipulation. If Neg\textsuperscript{0} triggers the suppletive Vocabulary
Insertion of \ird{zad-} at the root node, then Neg\textsuperscript{0} has
already been discharged structurally; stacking a second Neg\textsuperscript{0}
on top would create an uninterpretable double-negation configuration. The
prohibition is a consequence of the analysis rather than a separate blocking
rule.

The relationship between Neg\textsuperscript{0} and the inherently negative mood
forms (prohibitive \ird{-é}, negative hortative \ird{-ku}, negative irrealis
\ird{-ir}) parallels the \ird{zad-} case in its effect on NegP projection: when
MoodP carries [+\textsc{neg}] as a VI-internal feature, NegP is absent from the
derivation and \ird{zá-} accordingly does not surface. The full analysis of
MoodP [+\textsc{neg}] --- including the semantic distinction between
directive-scoped and propositional negation, and the restricted palatalizing
behavior of \ird{-ir} --- is in §\,\ref{sec:vm-moodp}.

The DM morphological-incorporation account makes the following empirical
commitments. (a)~\ird{Zá-} does not constitute an independent prosodic word: it
is a bound clitic on the verbal complex and cannot receive independent stress or
be separated from the stem by an intervening focus particle. (b)~Nothing may
intervene between \ird{zá-} and the stem; evidence of such separation would
require the cliticization analysis to be revised. (c)~In ellipsis and fragment
contexts, \ird{zá-} cannot appear stranded without the full verbal complex,
unlike a free-standing negative head that could survive ellipsis of its
complement. Two further commitments bear on the three-exponent analysis:
(d)~\ird{staly} is fully invariant, bearing no aspect, person, mood, or
iterative marking; paradigmatic variation in future data would favour a
defective-verbal-root analysis over the Neg\textsuperscript{0} account. (e)~The
three exponents (\ird{zá-}, \ird{staly}, \ird{zad-}) are in strict complementary
distribution, conditioned by the presence or absence of a finite verbal host;
the equative/stative boundary is treated in §\,\ref{sec:vm-tension-stative}.
Cross-linguistic typology is consistent with this picture: prefix negation is
well attested in SOV languages (Dryer 2013), suggesting the prefix position of
\ird{zá-} reflects a systematic PF convention rather than a Mirror Principle
violation.

\subsubsection{Identificational negation and the scope of \ird{staly}}
\label{sec:vm-ident-neg}

The availability of participant nominals as equative predicates
(§\,\ref{sec:nmlz-predicative} of Chapter~\ref{ch:verbs}) introduces a
semantically significant variation within the \ird{staly} environment. The
contrast can be stated precisely in terms of what NegP takes in its scope.

In a plain equative (\ird{Marek doktor}), the predication is
\term{predicational}: it ascribes a property or class to the topic. The negated
version (\ird{Marek doktor staly}) denies property-membership. NegP takes a
property-ascription in its scope and yields propositional negation of that
ascription.

In an identificational equative (\ird{Lobek Jancí piaštnikou}, \trsl{The apple
is what Janek ate}), the predication is \term{specificational} or
\term{identificational}: it asserts that the topic DP is the unique entity
satisfying the nominalized description. The semantic content of the predicate
(the participant nominal) is inherently exhaustive --- it picks out exactly one
referent. NegP over a specificational predication yields \term{identificational
negation}:

\begin{center}
$\lnot$[\textsc{topic} = \textsc{unique-satisfier}(\textit{nom.\ description})]
\end{center}

The contrastive implication --- that some other entity is the actual satisfier
--- follows from the exhaustivity of the original predication, not from any
special semantic feature of \ird{staly}. This is an automatic consequence of
negating an exhaustive identification, parallel to the contrastive implication
of cleft negation in English (\textit{It was not the apple that Janek ate}).

Crucially, the semantic difference between propositional \ird{zá-} and
identificational \ird{staly}+nominal is not ambiguity in the NegP head but a
difference in what NegP scopes over. \ird{Zá-} scopes over a finite event
description; \ird{staly} over a (zero-copula) equative predication, which may be
either property-ascribing or identificational depending on the type of its
predicate DP.

\begin{tabularx}{\linewidth}{l L L}
\toprule
& \textsc{propositional neg.\ (\ird{zá-})} & \textsc{identificational neg.\ (\ird{staly}+nom.)} \\
\midrule
Event presupposed?       & no                        & yes \\
Contrastive implication? & no                        & yes \\
NegP scope               & event-VP                  & equative predication \\
PF of Neg\textsuperscript{0} & prefix to verbal complex & free morpheme, clause-final \\
\bottomrule
\end{tabularx}

\medskip

he distinction between these rows is what the user should consult when deciding
which construction to employ: if the question is whether an event occurred,
\ird{zá-} is used; if the question is which entity was involved in a presupposed
event, \ird{staly} with a participant nominal is used. Both are structurally
expressions of a single NegP.

Note also that the nominal-domain negative operators (\ird{zide} \trsl{no one},
\ird{niho} \trsl{nothing}, etc.; see §\,\ref{sec:nm-pron-operators} of
Chapter~\ref{ch:nouns}) are not part of the clausal NegP system described here.
Their interaction with the NegP projection --- in particular, whether they
project their own NegP, co-occur with \ird{zá-}, or block NegP projection --- is
an open descriptive question discussed at the relevant section of
Chapter~\ref{ch:nouns}.

\section{Mood}\label{sec:mood}

The mood system comprises the indicative, the imperative, the prohibitive, the
hortative, and the irrealis. Mood is expressed by a suffix occupying the
position immediately outside the aspect-alignment morpheme; in the indicative
this suffix is null. Where a non-indicative mood suffix is present, the
Asp.Align head remains in the clause but its full five-way viewpoint inventory
is reduced by post-syntactic impoverishment to a two-way
imperfective/perfective-domain contrast, as described in
§\,\ref{sec:vm-mood-asp}. Every finite verb, indicative or not, therefore
carries an Asp.Align exponent. The imperative, prohibitive, and hortative
together form the directive domain, a unified subsystem whose case properties
follow from the interaction of reduced Asp.Align with null-subject orientation;
it is treated as a group beginning in §\,\ref{sec:mood-directive}.

\subsection{Indicative}

The indicative is the unmarked mood. It alone hosts the full five-way aspect
system and the associated split-ergative alignment described in
Section~\ref{sec:aspect}. Non-indicative moods carry a reduced two-way
aspect-domain contrast in Asp.Align (§\,\ref{sec:vm-mood-asp}), which
preserves the ergative/nominative alignment split but collapses the five-way
viewpoint inventory. The antipassive-transitive distinction, which depends on
the full structural availability of the ERG-domain Asp.Align in its indicative
form, is operative only in the indicative.

\subsection{The directive domain}\label{sec:mood-directive}

The imperative, the prohibitive, and the hortative form a unified directive
domain: a compact finite subsystem used for commands, negative commands,
suggestions, proposals, soft requests, permissive uses, and wishes. Where
the indicative provides the full five-way aspect system and the split-ergative
alignment described in Section~\ref{sec:aspect}, directive moods carry a
reduced two-way aspect-domain contrast in Asp.Align: the imperfective domain
($\emptyset$ exponent, nominative-style alignment) and the perfective domain
(\ird{-ov-} exponent, ergative alignment). The directive mood suffix follows
Asp.Align in the verbal template, so the order is always \textsc{stem}
+ (\textsc{voice}) + \textsc{reduced Asp.Align} + \textsc{mood}. See
§\,\ref{sec:vm-mood-asp} for the formal account.

The observable case geometry of the directive domain follows from the
interaction of reduced Asp.Align with the null-subject convention characteristic
of direct commands. The understood subject of imperative and prohibitive clauses
is second person and is normally left unexpressed, being recoverable from the
speech situation. With the agent unrealized, the patient or theme is the clause's
most prominent overt argument and accordingly occupies the direct (absolutive)
case under both alignment types: under PFV-domain (ergative) alignment the
patient is the absolutive by definition; under IPFV-domain (nominative)
alignment the agent is nominally the absolutive, but the null agent leaves the
direct slot available for the theme. In either case the patient or theme
characteristically appears in the direct. Where an overt speech-act participant
subject would compete with an overt direct object for that direct slot, it is
expressed in the transitive case instead, preserving the affected participant as
the clause's sole direct nominal. Beneficiaries licensed by the applicative and
causees licensed by the causative are obligatorily overt and appear in the
indirect. Antipassive morphology is excluded from all directive moods.

\subsection{Imperative and prohibitive}\label{sec:mood-imp}

The imperative (\ird{-ím}) is the ordinary mood for direct commands,
instructions, warnings, and immediate discourse-management directives addressed
to the hearer. The prohibitive (\ird{-é}) is its dedicated negative counterpart,
used for negative commands and preventative directives. The two are treated
together because they share the same null-subject pattern and the same case
geometry.

The subject of both the imperative and the prohibitive is second person and is
syntactically understood but ordinarily unexpressed; no overt pronoun is
required in fluent discourse. The theme or patient of the clause occupies the direct case as the
clause's single core direct nominal:

\pex
\a
\begingl
\gla Trava piaštím.//
\glb bread.\Dir{} eat-\Imp{}//
\glft \trsl{Eat the bread.}//
\endgl
\a
\begingl
\gla Tak hrabním.//
\glb \Dem{}.\Prox{}.\Dir{} hear-\Imp{}//
\glft \trsl{Hear this.}//
\endgl
\xe

The prohibitive carries the negative force within the suffix itself and does not
co-occur with the ordinary negative prefix \ird{zá-}. This is a structural
consequence of the morphology: the prohibitive suffix bears the inherent
negative value of the MoodP Vocabulary Item, rendering NegP absent from the
derivation; see §\,\ref{sec:vm-moodp} for the formal account. A prohibition
accordingly takes the following form:

\pex
\a
\begingl
\gla Trava piašté.//
\glb bread.\Dir{} eat-\Proh{}//
\glft \trsl{Do not eat the bread.}//
\endgl
\xe

Overt second-person subjects are permitted in both the imperative and the
prohibitive, but they are pragmatically marked, expressing foregrounding,
contrast, insistence, or rebuke toward the addressee. The second-person singular
pronoun is \ird{já} in the direct and \ird{jí} in the transitive; the
second-person plural is \ird{tová} in the direct and \ird{tvy} in the
transitive. In a transitive imperative or prohibitive clause where an overt
direct object is already present, an overt speech-act participant subject
appears in the transitive case rather than the direct, so that the affected
participant may remain the clause's sole direct nominal:

\pex
\a
\begingl
\gla Jí trava piaštím.//
\glb 2\textsc{sg}.\Trs{} bread.\Dir{} eat-\Imp{}//
\glft \trsl{You eat the bread.} \lingcontext{marked; overt 2\textsc{sg} subject in \Trs{}}//
\endgl
\a
\begingl
\gla Tvy trava piašté.//
\glb 2\textsc{pl}.\Trs{} bread.\Dir{} eat-\Proh{}//
\glft \trsl{You lot do not eat the bread.} \lingcontext{marked; overt 2\textsc{pl} subject in \Trs{}}//
\endgl
\xe

\noindent The pressure for the transitive case arises specifically from
competition: an overt subject would otherwise displace the direct object from
the clause's single direct slot. In intransitive directive clauses lacking a
competing direct object, no such competition arises, and the rule does not force
the transitive. The rule is therefore competition-sensitive rather than
absolute.

\subsection{Hortative}\label{sec:mood-hort}

The hortative covers a broader range of directive uses than the imperative:
inclusive proposals of the type \textit{let us}, softened direct commands,
permissive uses of the type \textit{let the child come}, and optative
expressions of wish or hope. It serves as the default polite register in
ordinary speech, typically preferred over the imperative in everyday
conversation. The hortative takes two forms: the positive \ird{-ká} and the
negative \ird{-ku}. Like the imperative and prohibitive, it excludes the
ordinary aspect suffixes and operates within the single-direct principle of the
directive domain.

The hortative defaults to a null subject whose interpretation is supplied by
clause type and context. In proposal uses the null subject is most readily read
as first person plural, inclusive of the addressee; in softened directive uses
the same null subject may receive a second-person reading. Null-subject
hortatives are the unmarked form:

\pex
\a
\begingl
\gla Piaštká.//
\glb eat-\Hort{}//
\glft \trsl{Let's eat.} \lingcontext{null subject; inclusive proposal}//
\endgl
\a
\begingl
\gla Tak hrabnká.//
\glb \Dem{}.\Prox{}.\Dir{} hear-\Hort{}//
\glft \trsl{Please hear this.} \lingcontext{null subject; softened directive with direct object}//
\endgl
\xe

The negative hortative \ird{-ku} is used for proposals to refrain, cautious
collective restraint, and mild negative requests. Like the prohibitive, it is an
inherently negative Vocabulary Item and accordingly excludes \ird{zá-}:

\pex
\a
\begingl
\gla Tak piaštku.//
\glb \Dem{}.\Prox{}.\Dir{} eat-\Neg{}\Hort{}//
\glft \trsl{Let's not eat this.} \lingcontext{negative hortative; collective restraint}//
\endgl
\xe

Overt subjects in the hortative fall into two structurally distinct classes. The
first class comprises overt speech-act participant subjects --- first person
plural, second person singular, second person plural --- appearing in proposal
or soft-command uses. These behave exactly like marked subjects in the
imperative and prohibitive: where they would otherwise compete with an overt
direct object for the clause's single direct slot, they appear in the transitive
case:

\pex
\a
\begingl
\gla Mně trava piaštká.//
\glb 1\textsc{pl}.\Trs{} bread.\Dir{} eat-\Hort{}//
\glft \trsl{Let us eat the bread.} \lingcontext{overt 1\textsc{pl} subject in \Trs{}}//
\endgl
\a
\begingl
\gla Jí tak hrabnká.//
\glb 2\textsc{sg}.\Trs{} \Dem{}.\Prox{}.\Dir{} hear-\Hort{}//
\glft \trsl{Please, you hear this.} \lingcontext{marked; overt 2\textsc{sg} subject in \Trs{}}//
\endgl
\xe

The second class comprises overt lexical subjects in optative or permissive
hortatives, where the clause expresses a wish or permission directed toward a
third participant rather than a command to the addressee. Such clauses are
typically intransitive. Their lexical subject occupies the direct case as the
clause's single core direct nominal --- the same position that in transitive
directive clauses would be held by the theme or patient. Because there is no
competing direct object, no competition for the direct slot arises and the
transitive case is not required:

\pex
\a
\begingl
\gla Byl ščenká.//
\glb child.\Dir{} arrive-\Hort{}//
\glft \trsl{May the child come.} \lingcontext{intransitive optative; lexical subject in \Dir{}}//
\endgl
\a
\begingl
\gla Je ščenká.//
\glb \Dem{}.\Dist{}.\Dir{} arrive-\Hort{}//
\glft \trsl{May that come to pass.} \lingcontext{intransitive optative; distal demonstrative in \Dir{}}//
\endgl
\xe

\noindent The contrast between these two subject classes captures the essential
geometry of the directive domain. An overt speech-act participant subject
appears in the transitive when a direct object is present because it would
otherwise compete with that object for the single direct slot. A lexical subject
of an intransitive optative or permissive hortative faces no such competition
and therefore remains in the direct. The reason for the difference is not
grammatical person but structural: in the first class a competing direct nominal
is present; in the second it is not.

\subsection{Case and subject expression in directive clauses}\label{sec:mood-directive-case}

The case properties of the directive domain follow from the general principles
of reduced Asp.Align alignment (§\,\ref{sec:vm-mood-asp}) operating under
the null-subject convention characteristic of direct commands. The analysis is
unified: no directive-specific case rule is required.

Directive clauses characteristically show only one overt core nominal in the
direct case. This is not a stipulated constraint but the convergent outcome of
alignment plus null-subject orientation. The understood second-person subject
of imperative and prohibitive clauses, and the understood subject of hortative
clauses, are syntactically present but normally left unexpressed. With the agent
unrealized, the patient or theme---which is the absolutive by PFV-domain (ergative)
alignment, and by absolutive absorption under null-agent IPFV-domain (nominative)
alignment---surfaces as the clause's single direct nominal. In intransitive uses,
the clause's sole argument occupies the direct in both alignment types.

Overt speech-act participant subjects are pragmatically marked throughout the
directive domain. Where their expression would otherwise compete with an overt
direct object, they appear in the transitive case rather than the direct,
preserving the affected participant as the clause's sole direct nominal. This
transitive marking is a surface reflex of the same absolutive competition that
governs the overt-subject/overt-patient interaction in any Asp.Align-governed
clause. The following minimal pair makes the contrast plain:

\pex
\a
\begingl
\gla Trava piaštím.//
\glb bread.\Dir{} eat-\Imp{}//
\glft \trsl{Eat the bread.} \lingcontext{null subject; theme in \Dir{}}//
\endgl
\a
\begingl
\gla Jí trava piaštím.//
\glb 2\textsc{sg}.\Trs{} bread.\Dir{} eat-\Imp{}//
\glft \trsl{You eat the bread.} \lingcontext{overt subject in \Trs{}; theme in \Dir{}}//
\endgl
\xe

\noindent By contrast, an overt lexical subject of an intransitive optative or
permissive hortative may remain in the direct when no competing direct nominal
is present. The competition-sensitive nature of the transitive-subject pattern
means it is contingent on the presence of a competing absolutive nominal, not
a categorical restriction on subject case across all directive clauses.

\subsection{Valency-changing morphology in directive clauses}\label{sec:mood-directive-valency}

Applicative and causative morphology are compatible with the directive moods.
Both occupy their regular positions in the morphological template immediately
before the directive mood suffix, which is accordingly the outermost suffix in
the verbal complex within the mood domain. The directive suffix thus remains
outer to all valency morphology.

The applicative (\ird{-éb-}) is licensed in directive clauses only if the
beneficiary is overtly realized. The beneficiary appears in the indirect case,
typically with the postposition \ird{my}, while the theme retains the direct.
This fits the directive case geometry without disturbance: the direct holds the
theme, the indirect holds the beneficiary, and an overt addressee subject, if
present, occupies the transitive:

\pex
\a
\begingl
\gla Trava bylam piaštébím.//
\glb bread.\Dir{} child.\Ind{} eat-\Appl{}-\Imp{}//
\glft \trsl{Feed the child bread.} \lingcontext{applicative; beneficiary in \Ind{}}//
\endgl
\a
\begingl
\gla Jí trava bylam piaštébím.//
\glb 2\textsc{sg}.\Trs{} bread.\Dir{} child.\Ind{} eat-\Appl{}-\Imp{}//
\glft \trsl{You feed the child bread.} \lingcontext{overt subject in \Trs{}}//
\endgl
\xe

\noindent Without an overt beneficiary the applicative is ungrammatical in
directive clauses:

\pex
\rec{trava piaštébím} \trsl{intended: `Feed someone bread.'} \lingcontext{applicative without beneficiary}
\xe

The causative (\ird{-na-}) likewise requires an overtly realized causee in
directive clauses. The causee appears in the indirect case, keeping the direct
available for the theme. The directive case pattern for causatives therefore
parallels that for applicatives exactly: theme in the direct, causee in the
indirect, overt addressee subject in the transitive if present. Before the
vowel-initial mood suffixes (\ird{-ím}, \ird{-é}), the causative takes its
epenthetic form \ird{-naj-}; before the consonant-initial hortative \ird{-ká} it
remains \ird{-na-}:

\pex
\a
\begingl
\gla Trava bylam piaštnajím.//
\glb bread.\Dir{} child.\Ind{} eat-\Caus{}-\Imp{}//
\glft \trsl{Make the child eat the bread.} \lingcontext{causative; causee in \Ind{}}//
\endgl
\a
\begingl
\gla Jí trava bylam piaštnajím.//
\glb 2\textsc{sg}.\Trs{} bread.\Dir{} child.\Ind{} eat-\Caus{}-\Imp{}//
\glft \trsl{You make the child eat the bread.} \lingcontext{overt subject in \Trs{}}//
\endgl
\xe

\noindent Without an overt causee the causative is ungrammatical in directive clauses:

\pex
\rec{trava piaštnajím} \trsl{intended: `Make someone eat the bread.'} \lingcontext{causative without causee}
\xe

The requirement for overt beneficiaries and causees is categorical. The
directive subsystem resolves case pressure not by multiplying direct nominals
but by reserving the direct for the theme and assigning added non-theme
participants to the indirect. An applicative without a beneficiary or a
causative without a causee would leave an indirect role structurally implied but
unresolved. Since the language tolerates multiple indirect nominals, a directive
clause bearing both applicative and causative morphology is structurally
possible, even if stylistically heavy; the principle throughout is that added
participants enter the indirect while the theme retains the direct.

Antipassive morphology is excluded from all directive moods. The antipassive
in the indicative performs a specific discourse function: it demotes the patient
from absolutive status in order to foreground the agent's activity in
ERG-domain clauses where a patient would otherwise occupy the direct. This
motivation is absent in directive clauses, where null-subject orientation has
already removed the agent-versus-patient absolutive competition that the
antipassive resolves in the indicative. Moreover, the reduced Asp.Align of the
directive domain does not provide the full structural conditions under which the
antipassive is licensed as an overt Voice head (§\,\ref{sec:vm-antipassive-desc}).
The five-way antipassive-transitive distinction, which depends on the full
indicative Asp.Align, is therefore operative only in the indicative; what the
directive domain preserves is the coarser ERG/NOM alignment contrast of
reduced Asp.Align.

The directive subsystem as a whole is characterized by null-subject orientation,
a reduced two-way aspect-domain contrast in place of the full five-way
indicative inventory, the characteristic appearance of the patient or theme in
the direct case (derived from reduced Asp.Align plus null-subject orientation),
marked overt speech-act participant subjects in the transitive when they would
otherwise compete for the direct slot, and the structured integration of
applicative and causative morphology in which beneficiaries and causees enter
the indirect. This pattern is not the output of a separate case system: it is
the convergent surface effect of the same Asp.Align-driven alignment engine
operating under null-subject conditions. The direct nominal remains central, and
competition within the clause core is prevented, because Asp.Align places the
absolutive on the patient (PFV domain) or on the agent (IPFV domain), and the
null agent resolves the latter into the same patient-direct surface pattern.


\subsection{Irrealis}

The irrealis (\ird{-ěv}) is used for counterfactual, hypothetical, or non-actual
situations. It belongs to the palatalizing, not the sibilant-replacing, domain:

\ex
\ird{piašt-} + \ird{-ěv} → \ird{piaščev}
\xe

In stems with the appropriate grade alternations, the front vowels /ɛ/ and /eː/
further raise to /i/ and /iː/ before this palatalizing suffix:

\ex
\ird{zděk-} `to bring' → \ird{zdíčev} \lingcontext{-e- raised to -í-; -k- palatalized to -č-}
\xe

The irrealis is compatible with all three converb moods and serves as the
main-clause mood in concessive constructions; both uses are treated in
Chapter~\ref{ch:complex}.

The negative counterpart of the irrealis is \ird{-ir}, a competing negative
Vocabulary Item at MoodP rather than a suppletive form in the strict sense. Like
the prohibitive \ird{-é} (negative imperative) and the negative hortative
\ird{-ku}, it is inherently negative and therefore does not co-occur with the
negative prefix \ird{zá-}. Phonologically, \ird{-ir} does not trigger the full
palatalizing pattern of \ird{-ěv}: stems in \ird{-k} show the restricted
alternation \ird{k $\rightarrow$ c}, but there is no broader \ird{-ěv}-type
palatalization and no associated vowel raising. Thus the same stem that yields
\ird{piaščev} in the affirmative irrealis yields \ird{piaštir} in the negative
irrealis, while \ird{zděk-} forms \ird{zděcir} rather than \ird{zdíčev}. The
formal analysis is in §\,\ref{sec:vm-moodp}.

\pex
\a \ird{piaštir} \trsl{would/might not eat; irrealis non-eating}
\a \ird{zděcir} \trsl{would not bring} \lingcontext{restricted \ird{k $\rightarrow$ c}; no vowel raising; cf.\ affirmative \ird{zdíčev}}
\xe


\subsection{Formal analysis: the IP domain}\label{sec:vm-ip}

\subsubsection{MoodP}
\label{sec:vm-moodp}

The mood suffixes occupy the projection immediately outside AspP. In the
indicative, MoodP is present but headed by a null morpheme. In the
non-indicative moods (imperative, hortative, irrealis), a lexical mood morpheme
is inserted at MoodP and simultaneously triggers a post-syntactic impoverishment
rule on the Asp node: the fine-grained viewpoint features of Asp are deleted,
leaving only the coarser $[\pm\textsc{pfv-domain}]$ feature that drives
reduced Asp.Align (see §\,\ref{sec:vm-mood-asp}). AspP is therefore always
projected in finite clauses; what varies is the richness of features available
for Vocabulary Insertion at that node.

The ordering of Mood above Asp is consistent with the cross-linguistic tendency
for tense and mood to dominate aspect (Cinque's (1999) functional hierarchy
places Mood$_\text{irrealis}$ above Asp$_\text{perfective/imperfective}$). The
mood paradigm consists of competing Vocabulary Items inserted at the MoodP head:

\begin{center}
\begin{tabular}{l l l}
\toprule
\textsc{mood} & \textsc{affirmative vi} & \textsc{negative vi} \\
\midrule
Indicative   & $\emptyset$      & ---\quad (NegP projected; \ird{zá-}) \\
Imperative   & \ird{-ím}        & \ird{-é} \hspace{3pt}(prohibitive) \\
Hortative    & \ird{-ká}        & \ird{-ku} \\
Irrealis     & \ird{-ěv}        & \ird{-ir} \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\noindent Three of the eight cells above carry a lexically specified
[+\textsc{neg}] feature as part of their Vocabulary Item: the prohibitive
\ird{-é}, the negative hortative \ird{-ku}, and the negative irrealis \ird{-ir}.
This is a property of the individual VI, not a productive morphological
operation; there is no synchronic rule that derives the negative cell from the
affirmative. The affirmative/negative pairing is an artifact of their
paradigmatic organization at MoodP, not of compositional negation.

When a MoodP head bearing [+\textsc{neg}] is selected, NegP is absent
from the derivation entirely: the clause is assembled with the negative MoodP
merging directly above AspP, and no Neg\textsuperscript{0} head enters the
numeration. \ird{Zá-}, as the PF exponent of Neg\textsuperscript{0}, therefore
has no node at which to be inserted and does not surface. No independent
blocking principle is required; the absence of \ird{zá-} is a simple consequence
of NegP not being projected.

An important asymmetry follows. The indicative is the only mood that projects
NegP and uses \ird{zá-}: it is the only cell in the MoodP paradigm whose
negative value is expressed by a separate functional projection rather than by
an inherently negative VI. The three non-indicative moods each have their own
negative VI, rendering NegP superfluous in those environments.

The semantic character of MoodP [+\textsc{neg}] differs from that of NegP
[+\textsc{neg}]. NegP negation is propositional: it operates over a
truth-evaluable event description and inverts its truth conditions. MoodP
[+\textsc{neg}] is \term{illocutionary/modal-scoped}: it negates the force or
modality encoded by the MoodP head, not a proposition. Within this class, two
sub-types are distinguished. The prohibitive \ird{-é} and the negative hortative
\ird{-ku} are \term{directive-scoped}: the negation qualifies the illocutionary
force of a command or suggestion. A prohibitive such as \ird{piaštéb-é}
\trsl{Don't feed anyone!} does not have truth conditions in the ordinary sense;
its negation qualifies the directive illocution, not a predicate. The negative
irrealis \ird{-ir}, by contrast, is \term{modal/hypothetical-scoped}: the
negation denies the relevance or applicability of a hypothetical
event-description --- that is, it operates over the non-actual world introduced
by the irrealis, not over a directive force. \ird{Piaštir} \trsl{would not eat;
irrealis non-eating} asserts that the hypothetical eating does not obtain,
rather than prohibiting it. The two sub-types of MoodP [+\textsc{neg}] are
formally parallel in their blocking of NegP projection; they are semantically
distinct in scope (directive illocution vs.\ hypothetical modality) and
therefore appear in separate rows of the negation summary table
(Table~\ref{tab:neg-exponents}).

The negative irrealis \ird{-ir} shows only a restricted front-vocalic effect. In
stems ending in \ird{-k}, it triggers the alternation \ird{k $\rightarrow$ c},
but it does not otherwise trigger the broader verbal palatalization pattern
associated with \ird{-ěv}: there is no \ird{č}-output and no accompanying vowel
raising. This partial behavior sets it apart both from fully palatalizing
\ird{-ěv} and from wholly non-palatalizing \ird{-ic}. Accordingly, \ird{zděc-ir}
is the correct negative irrealis of \ird{zděk-} \trsl{bring}, not \ird{*zděkir}
and not \ird{*zdíčir}; stems such as \ird{piašt-} still surface simply as
\ird{piaštir}.


\subsubsection{Inner evidential (EvidP$_\text{inner}$)}
\label{sec:vm-evid}

An evidential suffix occupies the outermost position in the finite inflectional
spine, immediately inside the complement linker \ird{-e}. This
\term{inner evidential} (EvidP$_\text{inner}$) is distinct from the clausal
evidential and speech-act operators in the left periphery
(§\,\ref{sec:th-particle-framework}), which are Zone~I particles occupying
EvidP$_\text{outer}$ and SpeechActP in the clause spine.

The morphological evidence for the inner evidential is the position of the
indirect evidential suffix (tentatively \ird{-ejte}) in the verbal template: it
precedes the complement linker \ird{-e} and follows MoodP, giving the order:

\begin{center}
\textsc{stem} + \textsc{asp.align} + (\textsc{mood}) +
(\textsc{ind.evid}) + \ird{-e} + \ird{po}
\end{center}

\noindent The corresponding syntactic projection is therefore above MoodP and
constitutes the outermost layer of the finite inflectional spine. The boundary
with the clausal EvidP$_\text{outer}$ (a left-peripheral projection) must be
kept clear: the inner evidential marks the \emph{source of knowledge} as it
bears on the proposition expressed by the verbal complex; the outer EvidP
particles (Zone~I) operate at the clause-level and take scope over the full
proposition including its discourse-relational frame.

%% STUB: The inner evidential suffix is insufficiently described
%% (as noted in the main verb chapter). A full treatment must:
%% (a) establish the form of the suffix with certainty;
%% (b) determine whether it occupies a projection above or below
%%     MoodP (the current template places it above, i.e. outside);
%% (c) investigate its interaction with the outer EvidP particles
%%     (can they co-occur? What is their combined interpretation?);
%% (d) test whether the indirect evidential is restricted to
%%     matrix clauses or also appears freely in embedded clauses.



\section{Non-finite forms: converb morphology}\label{sec:converb-morphology}

% ============================================================
% STATUS: MORPHOLOGY-ONLY SECTION
% This section documents what the converb suffixes look like and how
% they attach. The syntax and semantics of converb constructions are
% treated in Chapter~\ref{ch:complex}.
% ============================================================

Converb suffixes replace the aspect-alignment slot entirely. A verb carrying a converb suffix accordingly bears no aspect, no mood, and no alignment. The morphological structure of a converb is:

\medskip
\begin{center}
(\textsc{lex-prefix}) + \Stem{} + (\Appl{}) + (\Caus{}) + \textsc{converb-suffix}
\end{center}
\medskip

The converb suffixes and the stem mutations they occasion are set out in the following table:

\begin{table}[ht]
	\footnotesize\sffamily
	\caption{Converb suffixes and their phonological effects.}\label{tab:converb-morph}
	\medskip
	\begin{tabularx}{\textwidth}{L L L L}
		\toprule
		{\sc converb} & {\sc suffix(es)} & {\sc stem mutation} & {\sc notes} \\
		\midrule
		Sequential    & \ird{-u}, \ird{-ení} & none & \ird{-ní} after applicative/causative \\
		Simultaneous/manner & \ird{-ěc} & palatalization & front-vocalic trigger; may coincide with older sibilantized outputs in some stems \\
		Temporal (6)  & \ird{-ezak}, \ird{-eškady}, \ird{-eeby}, \ird{-ešhoume}, \ird{-eebymy}, \ird{-esny} & none & all add \ird{-e-} linker \\
		Concessive    & \ird{-ouc} & none & \\
		Conditional   & \ird{-ic}, \ird{-emy}, \ird{-ezmy}, \ird{-ena} & none for \ird{-ic}; see note & \ird{-ic} does not trigger palatalization \\
		Purposive     & \ird{-it} & irregular; mainly Class I & \\
		Contrastive   & \ird{-ema} & none & \\
		\bottomrule
	\end{tabularx}
\end{table}

The simultaneous/manner converb \ird{-ěc} and the irrealis mood \ird{-ěv} belong to the palatalizing domain, while the antipassive \ird{-aš-} is the clearest instance of true sibilant replacement. The conditional \ird{-ic} does not palatalize despite its initial front vowel; the historical reasons for this exception are discussed in Section~\ref{sec:cond-nopala} of Chapter~\ref{ch:complex}. Stronger irregularity is concentrated instead in a few high-frequency linker and quotative forms, where analogical leveling has obscured the older outputs.


\subsection{Formal analysis: the syntactic status of converbs}
\label{sec:vm-converbs}

\subsubsection{Converbs as aspect-deficient vP/VoiceP projections}
\label{sec:vm-converb-aspless}

Converbs replace the Asp.Align slot entirely. A verb bearing a
converb suffix therefore projects a vP/VoiceP but no AspP. The
converb suffix itself can be analyzed as the Vocabulary Item
inserted at a \term{defective AspP} node: a position in the
functional hierarchy that would normally host an Asp.Align morpheme
but instead hosts a converb-class morpheme that lacks aspect and
alignment features.

Alternatively, converbs project no AspP at all: the converb suffix
is the head of a Voice or vP projection that merges directly with
the higher adverbial or complementizer position without passing
through AspP. The absence of the ergative case marker \ird{-n-}
inside converb clauses (§\,\ref{sec:converb-chains}) is predicted
by either account: if AspP is absent, the Asp head that would have
triggered ergative-absolutive alignment is not present to assign
the transitive case to the agent.

\subsubsection{Case alignment within converb clauses}
\label{sec:vm-converb-case}

The nominative-accusative alignment that is invariant within
converb clauses (absolutive agent, indirect patient) is consistent
with the absence of an Asp head: without the perfective/
retrospective Asp head to assign ergative case to the agent, the
agent defaults to absolutive (nominative), and the patient receives
oblique/indirect case. This is in effect the same pattern as the
nominative-aspect main clause, with the difference that in the main
clause it is the Asp head (continuous/progressive/contemplative)
that assigns this pattern, whereas in the converb clause it is the
absence of any Asp head.

\subsubsection{Converb clauses as adjuncts and embedded topics}
\label{sec:vm-converb-adjunct}

In the clause-combining hierarchy, converb clauses function as
adverbial adjuncts to the matrix clause. They are merged as
adjuncts to the matrix IP or vP, not as arguments in Spec or
Complement positions. This explains their preverbal (and pre-matrix)
position: as adjuncts to IP, they precede all material within IP
and surface before the matrix arguments in the linear string,
consistent with the head-final adjunction pattern.

Switch-reference in converb chains (§\,\ref{sec:switch-reference}) is a
discourse-pragmatic phenomenon: different-subject readings are licensed by
the discursive contrast of two distinct referents in topic position, not by
any morphological clitic on the converb. When both clauses bear full nominal
subjects, those DPs suffice for disambiguation; when a subject is absent,
discourse recoverability determines the reading.


\subsection{The nominalized verb forms}\label{sec:nominalized}

% ============================================================
% STATUS: STUB — to be developed
% The 2020 archive describes the following nominalized forms:
%   - Generic nominal in -ou (from inflected stem)
%   - Participant nominals (agent nominal -kou, patient nominal)
%   - Event nominals
% These interact with definiteness, argument structure, and focus
% (the "nominalization for indefiniteness" construction described in Ch. 4).
% ============================================================

Iridian possesses several forms that nominalize the verb. One is the lexical or resultant nominal in \ird{-ou}, often semi-productive and lexeme-specific, which yields nouns such as \ird{piaštou} \trsl{food}. Another is the inherited gerundal strategy, which remains the productive event-nominal formation. Participant nominals likewise employ \ird{-ou}, but they are built from stems whose morphology already fixes which participant occupies the absolutive pivot.

Antipassive stems yield agent-oriented participant nominals, unmarked transitive stems patient-oriented ones, and applicative stems nominals referring to the promoted participant.

All five indicative aspects are used productively in these participant nominals: perfective, retrospective, continuous, progressive, and contemplative. The same voice-sensitive stem that determines the nominal's orientation also carries the aspectual morphology, so the nominal may refer to a completed participant, an experientially or result-relevant past participant, an ongoing or habitual participant, a participant in an event now in progress, or a participant associated with an intended or contemplated event. In this formation \ird{-ek + -ou} contracts to \ird{-ikou}, retrospective \ird{-ac} simply yields \ird{-acou}, \ird{-a + -ou} to \ird{-ova}, and \ird{-ime + -ou} to \ird{-imou}; the contemplative stem takes \ird{-ou} without further special contraction. Thus \ird{piaščikou} \trsl{the one who ate}, \ird{piaščacou} \trsl{the one who has eaten}, \ird{piaščova} \trsl{the one who eats / is eating}, \ird{piaščimou} \trsl{the one who is in the act of eating}, \ird{piaščachou} \trsl{the one who is about to eat}, \ird{piaštnikou} \trsl{that which was eaten}, \ird{piaštnacou} \trsl{that which has been eaten}, and \ird{piaštébikou} \trsl{the one who was fed}. Their syntax, definiteness properties, and argument structure are treated in Section~\ref{sec:event-nominals} of Chapter~\ref{ch:nouns}.

\subsubsection{Predicative use: participant nominals in identificational equatives}
\label{sec:nmlz-predicative}

Participant nominals may serve as the predicative complement of a zero-copula equative clause, with the topic DP identified as the entity satisfying the nominalized description. This construction has the schematic form:

\begin{center}
[\textit{topic-DP}] \ \ [\textit{(internal-arg.\Ind{})} \ \ \textit{participant-nominal}]
\end{center}

\noindent Compare the base transitive clause with its patient-nominal equative counterpart:

\pex
\a
\begingl
\gla Lobek Janku piaštnik.//
\glb apple.\Dir{} Janek.\Trs{} eat-\textsc{trs}-\Pf{}//
\glft \trsl{Janek ate the apple.} \lingcontext{standard ergative clause}//
\endgl
\a
\begingl
\gla Lobek Jancí piaštnikou.//
\glb apple.\Dir{} Janek.\Ind{} eat-\textsc{trs}-\Pf{}-\textsc{nmlz}//
\glft \trsl{The apple is what Janek ate.} \lingcontext{patient nominal as equative predicate}//
\endgl
\xe

\noindent And with an agent-nominal equative:

\ex
\begingl
\gla Janek lubcí piaščikou.//
\glb Janek.\Dir{} apple.\Ind{} eat-\Antip{}-\Pf{}-\textsc{nmlz}//
\glft \trsl{Janek is the one who ate the apple.} \lingcontext{agent nominal as equative predicate}//
\endgl
\xe

Two features of the case marking are notable. First, the agent of the base
transitive clause, which bears the transitive case in the finite verb
construction (\ird{Janku}, \Trs{}), shifts to the indirect case
(\ird{Jancí}, \Ind{}) when it appears as an internal argument of the
patient nominal. Second, the patient argument removed from core argument structure by the antipassive
in a finite clause may reappear in the indirect case as an internal argument
of the agent nominal (\ird{lubcí} in the second example above). In both
cases all internal arguments of the nominalized event appear in the indirect
case, since the absolutive/direct slot of the equative clause is occupied
by the external topic DP and no verbal case-assigning machinery operates
within the nominalized predicate.

The semantic effect of predicating a participant nominal is
\term{identificational}: the topic DP is being identified as the unique
referent satisfying the description encoded in the nominal. This contrasts
with simple property-ascription equatives (\ird{Marek doktor}, \trsl{Marek
is a doctor}) and with the plain verbal assertion (\ird{Lobek Janku
piaštnik}, \trsl{Janek ate the apple}). Negation of identificational
equatives with \ird{staly} and its semantic contrast with the verbal
\ird{zá-} are treated in §\,\ref{sec:ident-equative} and
§\,\ref{sec:clausal-negation} of Chapter~\ref{ch:simple}.


\section{The conjunctive / quotative linker}\label{sec:conjunctive-linker}

Iridian has a verbal linking form in \ird{-e}, used in clause linking and quotative constructions. Synchronically this \ird{-e} is a front-vocalic palatalizing suffix added to the fully inflected verb. It therefore belongs with the palatalization system described in Chapter~\ref{ch:phon}, not with sibilant replacement.

\pex
\a \ird{piašček} $\rightarrow$ \ird{piaščice}
\a \ird{piaščac} $\rightarrow$ \ird{piaščace}
\xe

The exact output depends on the segment immediately preceding the linker. Velars typically yield \orth{c/ce}; other endings show the regular soft reflexes of the final consonant or syllable in standard spelling, without introducing separate haček letters for soft \orth{t}, \orth{d}, or \orth{l}. Some older conjunctive formations are analogically levelled and remain morphologized, but the synchronic generalization is clear: \ird{-e} selects a palatalized form of the preceding material. The formal syntactic analysis of the linker is developed in the following subsection.


\subsection{Formal analysis: the conjunctive linker \ird{-e}}
\label{sec:vm-tension-linker}

The linker \ird{-e} appears at the right edge of embedded finite
clauses before \ird{po}/\ird{poby} and also in the conjunctive
clause-chaining function described above.
Its dual function raises the question of whether it is a single
morpheme with a unified syntactic status or two homophonous
morphemes. A full resolution requires integrating the phonological
behavior of \ird{-e} with its morphosyntactic distribution and its
position in the verbal template.

\subsubsection{Phonological characterization}

\paragraph{Trigger class and palatalization.}
Synchronically, \ird{-e} is a bound suffix that belongs to the
productive front-vocalic palatalization system described in
§\,\ref{sec:front-palatalization}. Its trigger class is the same as
that of \ird{-ě}, \ird{-í}, and the irrealis suffix \ird{-ěv}: all
are front-vocalic and all soften the segment to which they attach.
The conditional converb \ird{-ic}, though it begins with a front
vowel, is the systematic exception: it does not trigger
palatalization, because it was grammaticalized during a period when
its initial vowel lacked palatalizing force
(§\,\ref{sec:cond-nopala}). This exception is historically
motivated and synchronically opaque; it does not disturb the
generalization that \ird{-e} is a palatalizing trigger in the
productive system.

\paragraph{Segment-specific outputs.}
The palatalizing effect of \ird{-e} targets the final segment of the
inflected form to which it attaches, not the thematic consonant of
the bare stem. The outcomes depend on the place and manner of the
final consonant:

\begin{itemize}
  \item \textit{Velars} (/k/, /g/): the velar yields the palatalized
    affricate series. The most common output is \orth{c} for /k/,
    as in \ird{piašček} $\rightarrow$ \ird{piaščice} (the final
    velar of the perfective suffix \ird{-ek/-ik} palatalizes to
    \orth{c} before \ird{-e}).

  \item \textit{Coronals} (/t/, /d/, /n/, /l/, /r/): the regular
    soft reflexes apply. Standard spelling does not introduce haček
    diacritics for softened \orth{t}, \orth{d}, or \orth{l} in this
    context; the softness is indicated by the following vowel graph
    or is left orthographically unmarked for the consonant itself.
    Thus \ird{piaščac} (the retrospective, whose suffix now ends in
    \ird{-c}) yields \ird{piaščace} under \ird{-e}: the final
    affricate simply surfaces with the linker vowel as \ird{-ce}.

  \item \textit{Labials} (/p/, /b/, /m/, /v/) and liquids (/r/):
    secondary palatalization (a palatal off-glide) is added, giving
    the same soft-series output as before nominal \ird{-í}
    (§\,\ref{sec:palatalization}). These outcomes are less
    phonologically salient than velar or coronal softening and are
    typically not marked distinctively in orthography.
\end{itemize}

\noindent The key generalization is that \ird{-e} selects the soft
allomorph of whichever ending it attaches to, following the same
mechanism as other palatalizing suffixes. The output is therefore
predictable from the palatalization system already operative in
nominal and verbal morphology, without requiring a separate
phonological rule.

\paragraph{Interaction with vowel raising.}
When the final syllable of the inflected form contains a short
front vowel (\ird{e} or \ird{ě}), that vowel is subject to raising
to \ird{i}/\ird{í} in the syllable immediately preceding the
palatalizing trigger (§\,\ref{sec:vowel-raising}). In the derivation
\ird{piašček} + \ird{-e}: the thematic \ird{-e-} of the perfective
suffix raises to \ird{-i-} before the palatalized output, yielding
\ird{piaščice} rather than *\ird{piaščece}. This raising is the same
rule that operates in \ird{zděk-} + \ird{-ěv} $\rightarrow$
\ird{zdíčev} (the \ird{-e-} of the root raises before the
front-vocalic verbal suffix). The interaction of raising with
palatalization is thus not specific to \ird{-e}: it is a general
consequence of attaching any front-vocalic suffix to a stem whose
final syllable contains a short front vowel.

\paragraph{Analogically levelled forms.}
Some high-frequency conjunctive forms are morphologized: they
preserve older palatalizations or otherwise deviate from the
synchronic rule. These are most concentrated in the finite forms
of verbs with irregular thematic consonants or with stems that
underwent earlier lenition or vowel contraction. The synchronic
system tolerates these as exceptions without compromising the
productivity of the rule. Their existence does, however,
complicate any derivational analysis that requires strict
compositionality: the linker form of a high-frequency verb may
need to be stored as a listed item alongside its productive
derivation.

\paragraph{Prosodic status.}
The linker \ird{-e} is a bound morpheme: it does not constitute an
independent prosodic word and does not shift the stress of the
verbal complex. Stress in Iridian is fixed to the lexical stem
and is not influenced by suffixation in the outer inflectional
layers. The prosodic invariance of \ird{-e} is consistent with its
status as a suffix rather than a clitic or particle. If it were a
prosodic word or particle, the stress pattern of \ird{piaščice}
might be expected to differ from that of \ird{piaščik}; there is no
evidence that it does.

\subsubsection{Morphosyntactic distribution}

\paragraph{Template position.}
Within the verbal complex, \ird{-e} attaches outside all finite
inflectional morphology. The full template is:

\begin{center}
\Stem{} + \Asp{}.\textsc{align} + (\textsc{mood}) +
(\textsc{evid}) + \ird{-e}
\end{center}

\noindent The linker is therefore exterior to EvidP$_\text{inner}$, the
outermost projection in the finite inflectional spine
(§\,\ref{sec:vm-evid}). This position is significant: if \ird{-e} were an
agreement or mood morpheme, it would be expected inside the inflectional spine.
Its consistent extra-clausal position supports an analysis in which it marks the
edge of the finite complex as it interfaces with a higher projection --- a
CP-boundary marker, not an inflectional head.

\paragraph{Attested environments.}
The linker appears in exactly three environments:

\begin{enumerate}
  \item \textit{The finite \ird{po}-complement}
    (§\,\ref{sec:vm-po-comp}): the linker marks the right edge of
    the embedded finite TP before the complement C head \ird{po} or
    \ird{poby}. It is obligatory here; an embedded finite clause
    without the linker is ungrammatical before \ird{po}/\ird{poby}.
    The two complementizers attach not to a bare proposition versus a
    bare \textit{whether}-clause, but to assertive versus non-assertive
    finite complements respectively; the non-assertive type includes
    embedded wh-content clauses as well as embedded polar issues (see
    §\,\ref{sec:poby} of Chapter~\ref{ch:complex}).

  \item \textit{The conjunctive chain} (§\,\ref{sec:conjunctive-linker}):
    the linker marks non-final finite clauses in a clause chain
    where the chained clauses are interpreted as coordinated or
    sequentially related propositions. This use resembles the
    clause-linking \textit{-ko} of Korean or the converbal
    linker of other verb-final languages, but unlike those, the
    Iridian form attaches to a fully finite, fully inflected verb.

  \item \textit{The matrixless \ird{po}-construction}
    (§\,\ref{sec:complementation}): even when the matrix predicate
    is absent and \ird{po} functions as a discourse-level
    attribution marker, the linker \ird{-e} is retained on the
    finite verb. This shows that \ird{-e} is not licensed by
    the presence of a lexical matrix predicate but by the presence
    of the C-domain element \ird{po} itself.
\end{enumerate}

\paragraph{Non-environments.}
The linker is absent in two structurally revealing contexts. First,
independent finite clauses lack \ird{-e}: when no C head embeds the
TP, the linker is not licensed. Second, converb forms --- the
non-finite clause-linking morphology of simultaneous, sequential,
conditional, and other converbs --- do not take \ird{-e}. This
follows naturally: converbs are already non-finite (they lack
Asp.Align and Mood morphology), so they do not participate in the
domain of the finite-clause boundary that \ird{-e} marks.

\subsubsection{Competing analyses}

Three main analyses are available; they diverge principally on
whether \ird{-e} has a unified syntactic identity across its two
main environments (complement and conjunctive chain).

\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
  \item[\upshape Unified CP-edge analysis.] Under this analysis,
    \ird{-e} is a single morpheme throughout: the phonological
    exponent of a \textsc{[+fin, +sub]} feature complex that is
    valued on the embedded TP when it enters the scope of a C
    head. In the complement construction, C is overt (\ird{po} or
    \ird{poby}). In the conjunctive chain, C is a null clause-juncture
    head (following the general principle that coordination in
    head-final languages may involve a null C$^0$ that takes a
    finite TP as its specifier or complement). The linker is the
    spell-out of the embedded T's contact with this C head, and its
    front-vocalic palatalizing nature is the PF signature of
    \textsc{[+fin, +sub]} in Iridian. The absence of \ird{-e} in
    independent clauses follows without stipulation: no C head is
    present to value the feature on T.

    The central challenge for the unified analysis is the
    distributional asymmetry: in the complement construction,
    \ird{-e} co-occurs with an overt C head (\ird{po}/\ird{poby});
    in the conjunctive construction, no overt C appears. The unified
    analysis must postulate a null conjunctive C$^0$ whose presence
    is evidenced only by the linker itself, which risks circularity
    unless independent evidence for this head can be adduced.

  \item[\upshape PF boundary-marker analysis.] Under this analysis,
    \ird{-e} is not a syntactic morpheme but a phonological
    convention: a boundary signal inserted at any right edge of a
    finite clause when that clause is followed by further clausal
    material. Its front-vocalic palatalizing character is then a
    phonological property of the boundary diacritic rather than a
    reflex of any syntactic feature. This account handles the
    conjunctive environment straightforwardly (every non-final
    finite clause is followed by another clause) and is empirically
    efficient.

    The drawback is that a PF boundary marker need not itself be a
    palatalizing morpheme: phonological boundary tones and
    juncture markers in other languages do not normally trigger
    segmental mutation. The palatalization caused by \ird{-e} is
    identical in type to the palatalization triggered by inflectional
    suffixes (\ird{-ě}, \ird{-í}, \ird{-ěv}), suggesting that
    \ird{-e} is a genuine suffix, not a boundary diacritic. The PF
    account therefore purchases distributional coverage at the cost
    of explaining a morphologically specific segmental process by
    appeal to an otherwise unmotivated phonological rule.

  \item[\upshape Two-homophone analysis.] Under this analysis,
    \ird{-e}$_1$ (the complement linker) and \ird{-e}$_2$ (the
    conjunctive linker) are distinct morphemes that happen to be
    phonologically identical and share the palatalizing trigger
    class. \ird{-e}$_1$ is the exponent of the TP-CP boundary
    in complement clauses; \ird{-e}$_2$ is a clause-chaining
    morpheme lexically specified for the conjunctive domain. Their
    coincidence in form is not synchronically active but is a
    historical convergence (or retention from a common source).

    This analysis avoids the null-C problem but is the most
    stipulative: it requires two separate lexical entries sharing
    phonological form and the same segmental-mutation property,
    with their distribution jointly covering the environments
    of both complement embedding and clause chaining. The burden
    of proof falls on demonstrating that the two environments
    are genuinely syntactically disjoint in a way that the
    unified analysis cannot accommodate.
\end{description}

\subsubsection{Preferred account and outstanding questions}

The unified CP-edge analysis is preferred here on parsimony grounds.
Its core claim is that \ird{-e} spells out the feature
\textsc{[+fin, +sub]} --- roughly, `finite clause under
subordination' --- on the embedded TP's outermost layer, and that
this feature is valued whenever a C head (overt or null) selects the
finite TP. The palatalizing nature of \ird{-e} is then the PF
realization of this morphosyntactic feature, fully parallel to the
palatalizing effect of other front-vocalic suffixes in the
productive system.

The position of \ird{-e} outside EvidP$_\text{inner}$ in the template
supports this analysis directly: the linker is the last element
added to the verbal complex before the clause exits the finite
inflectional domain, which is exactly where a TP-edge or C-interface
marker would be expected.

Two empirical questions remain open:

\begin{enumerate}
  \item \textit{The null conjunctive C$^0$.} The postulated null C
    head in the conjunctive construction should in principle be
    detectable through its interaction with the argument structure of
    the chained clauses: if the conjunctive C$^0$ takes the finite
    TP as its specifier and itself projects a CP, then the chained
    clause is a CP adjunct. This predicts that binding relations and
    scope interactions across conjunctive-chain clauses should pattern
    with embedding rather than flat coordination. Corpus investigation
    of binding and scope in conjunctive chains would provide a
    diagnostic.

  \item \textit{Interaction with the evidential.} The indirect
    evidential \ird{-ejte} appears inside \ird{-e} in the template
    (\ird{-ejte-e}), confirming that the evidential is a finite
    inflectional morpheme whereas the linker is not. But it is not
    yet established whether the palatalizing context of \ird{-e}
    applies to the final segment of \ird{-ejte} or directly to the
    verb stem. If \ird{-ejte} itself ends in a front vowel, the
    palatalizing context would overlap with it, potentially yielding
    an opaque surface form. Documenting the attested surface outputs
    of \ird{-ejte + -e} sequences across all five aspect forms would
    clarify whether the palatalization is invariant (expected under
    the unified analysis) or subject to additional sandhi conditions.
\end{enumerate}

\noindent The front-vocalic palatalizing nature of \ird{-e} is the
strongest evidence against treating it as a mere PF convention, and
the strongest argument for its status as a genuine morphosyntactic
suffix --- one that happens to appear at the clause's structural
edge, but whose phonological behavior is fully integrated with the
inner inflectional system of Iridian.


\section{Stative verbs}\label{sec:stative}

A class of stative verbs describes states rather than events. These verbs are
the language's primary means of expressing properties predicatively: where
English often uses an adjective, Iridian normally uses a stative verb. Stative
verbs are true verbs and accordingly require no copula to function as
predicates.

\subsection{Property roots and the verbalising suffix \ird{-n-}}

A small closed class of property roots each denotes an abstract property as a
nominal. The verbalising suffix \ird{-n-} is added to the root to yield a
stative verb stem, which then conjugates as a regular intransitive verb:

\begin{table}[ht]
	\footnotesize\sffamily
	\caption{Property roots and derived stative verbs.}\label{tab:statroots}
	\medskip
	\begin{tabularx}{\textwidth}{l L l L}
		\toprule
		{\sc root} & {\sc gloss} & {\sc stative stem} & {\sc gloss} \\
		\midrule
		\ird{hrad} & goodness  & \ird{hradn-}  & to be good \\
		\ird{mroc} & beauty    & \ird{mrocn-}  & to be beautiful \\
		\ird{trpa} & badness   & \ird{trpan-}  & to be bad \\
		\ird{svor} & bigness   & \ird{svorn-}  & to be big \\
		\bottomrule
	\end{tabularx}
\end{table}

The derived stative verb conjugates across the full five-way aspect system.
The continuous is closest to an equative property predication; the other
aspects carry characteristically dynamic or inchoative readings. The full
paradigm of \ird{mrocn-} \trsl{be beautiful} illustrates the system:

\pex
\a
\begingl
\gla Nužot mrocna.//
\glb city.\Dir{}.\Def{} be.beautiful-\Cont{}//
\glft \trsl{The city is beautiful.} \lingcontext{continuous: current state}//
\endgl
\a
\begingl
\gla Nužot mrocnime.//
\glb city.\Dir{}.\Def{} be.beautiful-\Prog{}//
\glft \trsl{The city is becoming beautiful.} \lingcontext{progressive: entering the state}//
\endgl
\a
\begingl
\gla Nužot mrocnik.//
\glb city.\Dir{}.\Def{} be.beautiful-\Pf{}//
\glft \trsl{The city became beautiful.} \lingcontext{perfective: ingressive}//
\endgl
\a
\begingl
\gla Nužot mrocnac.//
\glb city.\Dir{}.\Def{} be.beautiful-\Ret{}//
\glft \trsl{The city was beautiful.} \lingcontext{retrospective: past state}//
\endgl
\a
\begingl
\gla Nužot mrocnach.//
\glb city.\Dir{}.\Def{} be.beautiful-\textsc{contempl}//
\glft \trsl{The city will be beautiful.} \lingcontext{contemplative: expected state}//
\endgl
\xe

In bounded or developmental aspects --- the progressive, perfective, and
retrospective --- stative verbs are normally interpreted ingressively unless
context strongly resists it. The perfective thus yields \trsl{became
beautiful}, not simply \trsl{was beautiful}; that reading belongs to the
retrospective, which describes a past state conceived as a complete episode, a
state that no longer obtains, or one that is no longer at issue in the current
discourse. The continuous, by contrast, gives plain property predication
without any implication of change, and is the unmarked form for simple
description.

Stative verbs in the perfective take the raised allomorph \ird{-ik} rather
than \ird{-ek}; the verbalising suffix \ird{-n-} conditions this allomorphy in
the same manner as a stem-final \ird{e} or \ird{ě} (§\ref{sec:aspect-pf}).

\subsection{Attributive and adverbial derivatives}\label{sec:stative-attr}

Unlike regular verbs, stative verbs derived from property roots possess a
dedicated attributive form built by adding \ird{-á} directly to the verbalised
root. This form is tenseless, carries no aspectual or alignment marking, and
is used exclusively as a direct modifier within the noun phrase:

\pex
\a \ird{mrocná nuž} \trsl{a beautiful city}
\a \ird{hradná byl} \trsl{a good child}
\a \ird{svorná dum} \trsl{a big house}
\xe

The \ird{-á} form cannot be used predicatively. For predication, the full
stative verb form is required:

\pex
\a \ird{Nužot mrocna.} \trsl{The city is beautiful.} \lingcontext{predicative: stative verb required}
\a \ird{*Nužot mrocná.} \lingcontext{ungrammatical: attributive form in predicative position}
\xe

\noindent The \ird{-á} attributive functions as the adnominal counterpart of
the continuous stative, expressing the property directly without tense or
aspect. Its position within the noun phrase relative to determiners and other
modifiers is treated in Chapter~\ref{ch:nouns}.

Alongside the attributive in \ird{-á}, property-root-based stative stems
productively form a dedicated class of derived manner adverbs in \ird{-ě}
\label{sec:stative-adv}. This formation is derivational rather than
inflectional: it yields a lexical adverbial adjunct, not a further verbal
form, and it is distinct both from the adnominal attributive and from the
converb system discussed in Chapter~\ref{ch:complex}. Schematically, the
pattern is as follows:

\begin{center}
\term{property root $\rightarrow$ stative stem in \ird{-n-} $\rightarrow$ attributive in \ird{-á} / manner adverb in \ird{-ě}}
\end{center}

\noindent A representative paradigm is the following:

\pex
\a \ird{skor} \trsl{quickness}
\a \ird{skorn-} \trsl{be quick}
\a \ird{skorná} \trsl{quick, fast} \lingcontext{attributive}
\a \ird{skorně} \trsl{quickly}
\xe

The resulting system preserves a clear three-way distinction: the finite
stative predicate expresses property predication, the \ird{-á} form modifies a
noun within the DP, and the \ird{-ě} form modifies the event as a manner
adverb. Thus \ird{skorn-} means \trsl{be quick}, \ird{skorná} means
\trsl{quick} as a noun modifier, and \ird{skorně} means \trsl{quickly} as a
clausal adjunct. The syntactic position of these derived manner adverbs in the
preverbal field is discussed in Chapter~\ref{ch:simple}; compare also the
simultaneous / manner converb in \ird{-ěc} in Chapter~\ref{ch:complex}, which
remains a non-finite verbal form rather than a lexical adverb.

\paragraph{Degree prefixes.}
Scalar degree on the stative class is expressed by a closed set of prefixes
that attach to the stative stem before any derivational material, including
the verbalising \ird{-n-} and the attributive \ird{-á}. The five prefixes are
mutually exclusive and occupy a single slot:

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Degree prefixes on stative verbs.}\label{tab:degpfx}
  \medskip
  \begin{tabularx}{\textwidth}{l l L L}
    \toprule
    {\sc prefix} & {\sc value} & {\sc finite} & {\sc attributive} \\
    \midrule
    \ird{po-}  & slight; somewhat      & \ird{pomrocna}  & \ird{pomrocná}  \\
    \ird{do-}  & moderate; quite       & \ird{domrocna}  & \ird{domrocná}  \\
    \ird{pre-} & strong; very          & \ird{premrocna} & \ird{premrocná} \\
    \ird{od-}  & elevated; exceedingly & \ird{odmrocna}  & \ird{odmrocná}  \\
    \ird{nad-} & excessive; too        & \ird{nadmrocna} & \ird{nadmrocná} \\
    \bottomrule
  \end{tabularx}
\end{table}

\noindent The prefixes are fully productive across the stative class. Because
they precede the stem complex, they appear before \ird{-n-} in finite forms
and before the root in attributives.

\paragraph{The emphatic suffix \ird{-ěsk-}.}
The suffix \ird{-ěsk-} is distinct from the degree prefix system and does not
occupy the same grammatical slot. It is restricted to the \ird{-á} attributive
and carries a marked, heightened, or literary flavour rather than a simple
scalar value --- closer in function to an emphatic particle than to a degree
adverb. The contrast with a prefixed form is one of register and framing
rather than degree alone: \ird{premrocná} states a degree as a plain fact;
\ird{mrocněská} colours the attribution, implying that the beauty is
remarkable, conspicuous, or worth dwelling on. In ordinary speech
\ird{-ěsk-} attributives may sound affected; in verse, formal oratory, and
deliberate rhetorical prose they are appropriate and expected.

The suffix is not fully productive. It is most natural with roots that carry
inherent aesthetic or evaluative content:

\pex
\a \ird{mrocněská} \trsl{strikingly beautiful; beautiful indeed} \lingcontext{natural in literary register}
\a \ird{hradněská} \trsl{truly good; good in the full sense} \lingcontext{natural in literary register}
\xe

\noindent With dimensional or manner-denoting roots the formation is possible
but tends to sound stilted or unidiomatic outside verse:

\pex
\a \ird{svorněská} \trsl{(strikingly big)} \lingcontext{odd in prose; possible in elevated style}
\a \ird{skorněská} \trsl{(remarkably quick)} \lingcontext{marginal; feels forced outside verse}
\xe

\noindent \ird{trpěská} occupies a middle ground: the negative valence of
\ird{trpa} (\trsl{badness}) makes its \ird{-ěsk-} form feel ironic or
rhetorical rather than sincerely emphatic, and its use is consequently rare
and stylistically loaded.

\ird{-ěsk-} may co-occur with a degree prefix. The two are additive rather
than competing: the prefix supplies the scalar value and \ird{-ěsk-} supplies
the rhetorical register.

\pex
\a \ird{premrocněská} \trsl{very beautiful} \lingcontext{emphatic; degree stated and dwelt upon}
\a \ird{odmrocněská} \trsl{exceedingly beautiful} \lingcontext{elevated register; characteristic of panegyric}
\a \ird{nadmrocněská} \trsl{excessively, ruinously beautiful} \lingcontext{ironic or elegiac; beauty to a fault}
\xe

\noindent Such combinations are licensed but marked; they are characteristic
of formal verse and deliberate oratorical prose. In ordinary speech, a
prefixed form without \ird{-ěsk-} is the normal choice.

\subsection{Nominalisation of stative verbs}

Finite stative verbs are nominalised by the same \ird{-ou} mechanism that
applies to all verbs (§\ref{sec:nominalized}). The nominalised form yields a
substantive or classifying reading rather than a simple description: the
speaker is not merely ascribing a property but identifying a referent by that
property or setting it in contrast to others.

\pex
\a \ird{Nužot mrocna.} \trsl{The city is beautiful.} \lingcontext{simple property predication}
\a \ird{Nužot mrocnova.} \trsl{The city is the beautiful one.} \lingcontext{substantive: identifying or contrastive}
\xe

\pex
\a \ird{Bylot svornik.} \trsl{The child became big (i.e.\ grew up).} \lingcontext{ingressive perfective}
\a \ird{Bylot svornikou.} \trsl{The child is the one who grew up.} \lingcontext{nominalised perfective}
\xe

When a nominalised stative verb is used adnominally, it takes the standard
\ird{-ovní} form:

\ex
\begingl
\gla Svornikovní bylot Marek.//
\glb be.big-\Pf{}-\textsc{nmlz}-\textsc{attr} child.\Dir{}.\Def{} Marek.\Dir{}//
\glft \trsl{Marek is the child who grew up.}//
\endgl
\xe

The continuous is an exception: the continuous stative nominalised in
\ird{-ova} cannot be used adnominally with \ird{-ovní}. In adnominal position
the \ird{-á} attributive must be used instead:

\pex
\a \ird{mrocná nuž} \trsl{a beautiful city} \lingcontext{correct: \ird{-á} attributive}
\a \ird{*mrocnavní nuž} \lingcontext{ungrammatical: \ird{-ovní} built on continuous stem}
\xe

\subsection{State types and aspectual interpretation}\label{sec:stative-stability}

Not all states predicated by stative verbs are equally permanent. Some roots
denote enduring or characteristic properties---canonical evaluative qualities
such as \ird{hrad} \trsl{goodness} or \ird{mroc} \trsl{beauty}, and
dimensional properties such as \ird{svor} \trsl{bigness}---while others denote
contingent or temporary conditions: \ird{ptakn-} \trsl{be tired}, \ird{smrn-}
\trsl{be sick}, \ird{hvodn-} \trsl{be open}, \ird{žbirn-} \trsl{be wet}. This
distinction is not encoded grammatically, but it shapes the naturalness and
interpretation of aspectual forms.

The general contrast between continuous and progressive---laid out for all
predicate classes in §\,\ref{sec:aspect-ipf}---takes a particularly sharp form
with stative predicates, because here the interpretive difference between the
two is categorical rather than merely gradient. For a stable property, the
continuous yields plain predication and the progressive is distinctly marked,
carrying a developmental reading that implies a process of change
(\trsl{getting better}, \trsl{getting bigger}). For a temporary bodily or
experiential condition, the continuous describes the present condition as
simply obtaining (\trsl{is tired right now}), while the progressive more
readily receives an ingressive reading (\trsl{is starting to feel tired},
\trsl{is getting sick}). In both cases the continuous is the default
imperfective, giving simple present-state predication, while the progressive
imposes the dynamic, unfolding interpretation that is its characteristic
contribution across the predicate system.

Three semantic classes within the stative system may accordingly be
distinguished: \term{canonical quality statives}, which denote evaluative
properties of things (\trsl{be good}, \trsl{be beautiful}, \trsl{be bad});
\term{dimensional statives}, which denote extent or scale properties
(\trsl{be big}); and \term{experiential and bodily-condition statives}, which
denote states of persons (\trsl{be ill}, \trsl{be hungry}, \trsl{be ashamed}).
The last class most readily admits progressive and retrospective readings, since
temporary and personally felt states support aspectual dynamics more naturally
than stable properties do.

\subsection{Valency operations with stative verbs}\label{sec:stative-valency}

Stative verbs admit the causative \ird{-na-}, following the same alignment
conditions as other verbs (§\ref{sec:aspect}): nominative alignment in the
progressive and contemplative, ergative alignment in the perfective and
retrospective. Because stative stems end in \ird{-n-}, the causative suffix
surfaces as \ird{-an-} rather than \ird{-na-}, the order being reversed to
avoid a geminate cluster at the morpheme boundary:

\pex
\a
\begingl
\gla Marek problemam svornanime.//
\glb Marek.\Dir{} problem.\Ind{} be.big-\Caus{}-\Prog{}//
\glft \trsl{Marek is making the problem bigger.} \lingcontext{causative progressive; nominative alignment}//
\endgl
\a
\begingl
\gla Problem Marku svornanik.//
\glb problem.\Dir{} Marek.\Trs{} be.big-\Caus{}-\textsc{trs}-\Pf{}//
\glft \trsl{Marek made the problem bigger.} \lingcontext{causative perfective; ergative alignment}//
\endgl
\xe

The applicative \ird{-éb-} is admitted with stative verbs but is semantically
more restricted than with active verbs. It yields a circumstantial reading in
which a causally implicated participant --- typically the entity or situation
responsible for bringing about the state --- is introduced. This use is close
to an experiencer-causative in English (\trsl{gladden}, \trsl{please}). The
causing entity stands in the direct case; the affected participant appears in
the indirect:

\ex
\begingl
\gla Senica Marcí všihnébik.//
\glb news.\Dir{} Marek.\Ind{} be.happy-\Appl{}-\Pf{}//
\glft \trsl{The news gladdened Marek.} \lingcontext{\ird{všihn-}: be happy, be glad}//
\endgl
\xe

Stative verbs do not possess a transitive form and accordingly do not admit the
antipassive. The antipassive is licensed only where an ergative transitive
reading would otherwise be available; since stative verbs are inherently
intransitive, no such reading exists.

Stative verbs may further function as attributive modifiers of nouns and remain
the productive attributive strategy in Iridian. Alongside them stands the small
closed set of paired adnominal property-root forms just mentioned; this use is
described in Chapter~\ref{ch:nouns}.

Property-root-based stative stems also feed the productive derived manner
adverbs in \ird{-ě} (§\,\ref{sec:stative-adv}), yielding lexical adverbial
adjuncts such as \ird{skorně} \trsl{quickly}. These belong to the clause's
preverbal adjunct structure, not to the attributive system proper.

\subsection{Syntactic analysis of stative verbs}
\label{sec:vm-tension-stative}

Under the analysis adopted here, stative verbs are regular finite verbs rather
than a partially exceptional predicate class. A lexical property root combines
with the verbalising suffix \ird{-n-}, analysed syntactically as a stativising
verbal head, to yield an intransitive predicate denoting a state. That
predicate is then selected by the same Asp head that selects other finite
verbs in Iridian, so the full five-way Asp.Align system applies to stative
bases exactly as it does elsewhere in the verbal system.

This account captures the central semantic contrast of the paradigm without
requiring a separate copular projection for ordinary property predication. In
the continuous, Asp selects the stative predicate as a state holding at the
reference time, yielding plain predications such as \ird{Nužot mrocna}
\trsl{the city is beautiful}. In the progressive, perfective, retrospective,
and contemplative, the same Asp system contributes developmental, bounded, or
temporally displaced viewpoints over that stative base; the resulting readings
are therefore naturally inchoative or ingressive (\trsl{be becoming
beautiful}, \trsl{became beautiful}) rather than simple present-state
descriptions. Asp does not cease to apply to stative predicates; rather, its
semantic contribution is interpreted over a predicate whose lexical content is
stative.

A consequence of this analysis for negation deserves explicit statement,
since the surface form of stative predications can resemble zero-copula
equatives. Consider the contrast:

\begin{center}
\begin{tabular}{l l l}
\toprule
\textsc{predication type} & \textsc{affirmative} & \textsc{negative} \\
\midrule
Zero-copula equative & \ird{Marek doktor} & \ird{Marek doktor staly} \\
Stative verbal       & \ird{Nužot mrocna} & \ird{Nužot zámrocna} \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\noindent The two clause types are superficially similar: both surface as [DP
Pred V?] with no overt copula and no aspectual suffix. Yet their negation
strategies differ sharply. The zero-copula equative takes \ird{staly} (the
free-morpheme Neg\textsuperscript{0} exponent), because there is no verbal host
for \ird{zá-} to cliticize onto. The stative verbal predication takes \ird{zá-},
because \ird{mrocna} is a finite verb (the continuous form of \ird{mrocn-}
\trsl{be beautiful}), which provides the verbal host that cliticization requires.

The structural distinction is therefore the presence or absence of a finite
verbal head. In zero-copula equatives, the predicate is a bare DP (or
participant nominal); there is no V head in the clause and AspP is not
projected. In stative verbal predications, the property root combines with
the stativizing suffix \ird{-n-} to yield a proper verbal stem, which is
inflected for Asp.Align in the usual way. Neg\textsuperscript{0} therefore
takes a verbal complement and its exponent is \ird{zá-}. The two predicate
types are formally distinguished by the presence of \ird{-n-} (and the
resulting finite Asp slot), not by anything special about the NegP layer.

The boundary becomes potentially unclear only with underived property adjuncts
that lack \ird{-n-}, if any such items exist; the present analysis treats all
property predications as going through the \ird{-n-} stativizer. See
§\,\ref{sec:vm-tension-neg} for the related open question about
complementarity of the three Neg\textsuperscript{0} exponents.

The syntactic infrastructure of the stative domain therefore distinguishes
three outputs from the same lexical base. First, the finite stative verb heads
the clause's verbal predicate and requires no copula in ordinary property
predication. Second, the \ird{-á} form is a non-finite attributive derivative
inside the DP, corresponding to the adnominal strategy discussed in
Chapter~\ref{ch:nouns}; it is not a finite predicate and cannot bear aspect or
alignment. Third, the \ird{-ě} form is a lexical adverbial derivative that
functions as a clausal adjunct in the preverbal field, not as an inflectional
member of the verbal paradigm. The same lexical base thus feeds clause-level
predication, DP-internal attribution, and adverbial modification through three
distinct syntactic realisations.

Because underived stative predicates are intransitive, the alignment component
of Asp.Align is trivial in their simple finite use: the sole argument is the
direct-case absolutive/nominative argument in every aspect. The ordinary
alignment alternations re-enter the system only when valency-changing
morphology is added. Causativised statives behave like other derived
transitives, showing nominative alignment in the progressive and contemplative
and ergative alignment in the perfective and retrospective. The stative system
thus does not challenge the broader verbal theory; it instantiates it with a
lexically stative base whose aspectual interpretations follow from the
interaction of a regular Asp head with a regular intransitive predicate.


\section{Verbal complements and head finality: preverbal position}
\label{sec:vm-comp-headfinal}

This section develops the formal structural analysis of verbal complements within the head-final clause: how the head-final parameter determines the preverbal position of nominal arguments and embedded clauses, and what syntactic projection each complement type occupies.

\subsection{The base position of the object}
\label{sec:vm-obj-base}

In a head-final language, the complement of a head $H$ is to the
left of $H$ in the linear string. For the lexical predicate V, its
internal argument (patient DP) is the complement of V and therefore
surfaces to V's left:

\begin{center}
\term{[VP [DP patient] V]}
\end{center}

\noindent This is a \emph{base-generated} preverbal position, not
the result of A-bar movement or scrambling. In the Minimalist
framework, head-finality is parameterized as the linearization
convention that assigns $<$Complement, Head$>$ ordering (Kayne's
1994 antisymmetric alternative --- deriving all
SOV orders from movement --- is not adopted here; the parametric
head-final approach is more consistent with the thoroughgoing
head-finality of Iridian across all phrase types).

The patient DP, while base-generated as the complement of V, may
also undergo A-movement to a higher specifier position for topic
purposes. In the ergative aspects, the patient is the absolutive
(topic) argument and moves from [Spec-VP] or [Compl-VP] to
Spec,TopP:

\medskip
\begin{center}
[TopP [patient$_i$] [Top$'$ Top [AspP ... [VP $t_i$ V]]]]
\end{center}
\medskip

\noindent In this derivation, the patient originates as the
complement of V (preverbal by head-finality), then moves to
Spec,TopP (clause-initially) to receive [absolutive] case and
topic status. The surface position of the patient (clause-initial,
absolutive) is therefore derived by A-movement, not base-generation
in TopP.

In the nominative aspects, the patient remains in its base position
as the complement of V (preverbal in the SOV order), while the
agent moves to Spec,TopP:

\medskip
\begin{center}
[TopP [agent$_j$] [Top$'$ Top [AspP ... [vP [VP [patient] V] $t_j$]]]]
\end{center}
\medskip

\noindent The surface word order in both cases is SOV (or more
precisely, absolutive-oblique-V), with the absolutive topic
first and the verb final.

\subsection{The po-clause as a nominal CP complement}
\label{sec:vm-po-comp}

The finite propositional complement (\ird{po}-clause) is a CP that
occupies the complement position of the matrix predicate. The
structure of the complement CP is:

\begin{center}
\term{[CP [TP embedded-clause-\ird{e}] \ird{po}]}
\end{center}

\noindent \ird{Po} is the head of the complement CP (a
complementizer), and the embedded finite clause is the TP whose
head is marked by the linker \ird{-e}. In head-final structure,
the C head follows its complement TP:

\begin{center}
\begin{forest}
  syntax tree
  [CP
    [{TP\\\textit{embedded clause}\\\textit{(verb + -e)}}]
    [{C\\\ird{po}}]
  ]
\end{forest}
\end{center}

\noindent This CP is then the complement of the matrix verb:

\begin{center}
\begin{forest}
  syntax tree
  [VP
    [{CP\\{[\textit{emb-clause}-\ird{e}] [\ird{po}]}}]
    [V]
  ]
\end{forest}
\end{center}

\noindent Head-finality places the CP complement to the left of the
matrix V. The surface word order of a propositional complement
construction is therefore:

\begin{center}
\term{[embedded-TP + \ird{-e}] [\ird{po}] [matrix arguments] [matrix V]}
\end{center}

\noindent This is precisely the attested order (§\,\ref{sec:th-po}).
The analysis requires no movement: the complement CP is
base-generated in [Complement, matrix-VP], and head-finality places
it before the matrix V.

A further consequence: the matrix absolutive argument (the topic)
in a propositional complement clause is most often null (topic
drop). This follows from the heaviness of the CP complement: in
a head-final clause, the CP already occupies the preverbal position
that would be available to an overt absolutive DP, and the topic
is preferentially dropped when the complement clause is highly
informative and heavy.

\paragraph{Po as C$^0$: evidence.}

The analysis of \ird{po} as a C head (and not, for example, as an
adposition or particle) is supported by the following diagnostics:
\begin{enumerate}
\item The embedded clause is fully finite (retains aspect,
  alignment, mood, and person morphology); this is what is
  expected of a CP complement, not a reduced VP.
\item \ird{Po} appears after the embedded clause-final verb
  (the right edge of TP in a head-final CP), in exactly the
  position predicted for a head-final C.
\item \ird{Poby} (the non-assertive counterpart) parallels \ird{po}
  structurally and has the semantics of a complementizer: it
  types the embedded clause as a polar-interrogative content
  (\textit{whether}) rather than a declarative. Complementizer
  alternation is a CP-level phenomenon.
\item The linker \ird{-e} on the embedded verb is the edge-marking
  morpheme at the TP-CP boundary; this is consistent with its role
  as a phonological reflex of \mbox{[Spec,CP] → C} movement or of
  the TP entering the scope of C (§\,\ref{sec:vm-tension-linker}).
\end{enumerate}

\subsection{The purposive converb complement}
\label{sec:vm-purp-comp}

The purposive converb (\ird{-it}) serves as the complement of
desiderative, volitional, and modal predicates. Unlike the
\ird{po}-clause, it is non-finite: it lacks aspect, alignment, and
mood. Structurally it is an aspectless vP or VoiceP projection ---
a verbal phrase headed by the root + any applicable valency
morpheme, terminated by the purposive suffix \ird{-it} rather
than an Asp.Align morpheme.

The purposive converb complement is therefore not a CP but a
smaller projection, most plausibly a bare VoiceP or AspP$_\varnothing$
(an AspP whose head is phonologically null, licensed only when
embedded under a controlling predicate). This is the standard
\term{event complement} (Stowell 1982; Wurmbrand 2001): the predicate denotes an
event rather than a proposition, and it is controlled by the matrix
subject in a typical control configuration.

In head-final structure, the purposive converb complement precedes
the matrix predicate:

\begin{center}
\term{[matrix agent-\textsc{abs}] [purposive-\ird{-it}] [matrix V]}
\end{center}

\noindent The shared-subject (control) reading is the default, with
an implicit PRO subject in the converb clause coreferent with the
matrix absolutive (topic).

%% STUB: The formal structure of the purposive converb complement
%% (VoiceP vs. AspP_null vs. TP with reduced features) should be
%% examined through the following diagnostics:
%% (a) Can the purposive converb take its own topic arguments
%%     distinct from the matrix clause? (→ if yes, it has a
%%     full TP or CP projection)
%% (b) Can the subject of the purposive converb be overtly
%%     expressed, or must it be PRO? (→ tests control vs. raising)
%% (c) Does the purposive converb license aspect-sensitive
%%     alternations in the matrix clause? (→ aspect interaction)
%% These tests bear directly on how high a projection the
%% purposive converb clause actually projects.

\subsection{The preverbal zone: a structural hierarchy}
\label{sec:vm-preverbal-zone}

The preceding analysis allows a unified characterization of the
preverbal zone in the Iridian clause. Reading from left to right,
the preverbal constituents occupy the following structural positions:

\begin{center}
\footnotesize
\begin{tabular}{l l l}
  \toprule
  {\sc position} & {\sc constituent type} & {\sc structural slot} \\
  \midrule
  1 (leftmost) & Topic DP (absolutive)     & Spec,TopP \\
  2            & Zone I peripheral particles & Spec/Head of EvidP, SpeechActP \\
  3            & Clause-typing particle \ird{by} & Head of PolP \\
  4            & Wh-phrase (in questions)  & Spec,FocP \\
  5            & Non-topic arguments       & Spec,IP or IP-adjuncts \\
  6            & Comment-domain operators   & Spec/Head of AspP (zone II) \\
  7            & Phasal particles           & Spec/Head of PhasP (zone III) \\
  8            & CP complement (po-clause) & Compl,VP (innermost) \\
  8 (alt.)     & Purposive converb         & Compl,VP or Spec,vP \\
  9 (rightmost) & Finite verb               & Head of VP / vP \\
  \bottomrule
\end{tabular}
\end{center}

\noindent The verbal complement (whether NP, CP, or converb) is
structurally the innermost preverbal constituent, closest to the
verb. This reflects its base position as the complement of V in the
VP (positions 8 in the table). Higher functional projections
(AspP, IP, TopP) are realized by progressively more leftward
material in the string. The verb itself appears at the absolute
right edge, consistent with the head-final parameter operating
recursively at every projection.

\section{Tension points and areas requiring development}
\label{sec:vm-tensions}

The following issues represent the principal theoretical challenges
in the current analysis. Each will require targeted empirical
investigation, corpus diagnostics, or theoretical elaboration before
the verbal morphosyntax chapter can be considered fully settled.

\subsection{The Asp.Align portmanteau and the Mirror Principle}
\label{sec:vm-tension-asp}

The most fundamental tension in the morphological template is the
\term{portmanteau} character of the Asp.Align morpheme: a single
morpheme does double duty, encoding both the aspectual value of the
predicate and the case-alignment pattern of the clause. The Mirror
Principle predicts that each morpheme corresponds to a single
syntactic head. If Asp and Align are two distinct syntactic
features (housed in AspP and CaseP respectively), the Mirror
Principle predicts separate morphemes for each.

Three responses are available:
\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
\item[\upshape Portmanteau morpheme by fusion (DM).] In DM,
  morphological Fusion (Halle \& Marantz 1993)
  allows two syntactic heads to be realized by a single Vocabulary
  Item when they are adjacent in the morphological structure. AspP
  and CaseP are adjacent in the Iridian verbal complex; their
  adjacency triggers fusion, and the resulting portmanteau is the
  Asp.Align suffix. This is the most parsimonious DM account.
\item[\upshape Single head with bundled features.] AspP is a
  single syntactic head whose feature bundle includes both
  $[$\textsc{aspect}: pf/retro/cont/prog/contempl$]$ and
  $[$\textsc{case-pattern}: erg/nom$]$. This is not a violation
  of the Mirror Principle if the feature bundle defines a single
  head, not two. The objection is theoretical: are aspect and
  alignment independent features (motivated independently) or are
  they necessarily co-specified?
\item[\upshape Aspect drives alignment, not vice versa.] The
  alignment pattern is not an independent feature but a
  \emph{consequence} of the aspect value under the Agree mechanism:
  perfective/retrospective Asp features trigger one Agree pattern
  (ergative), nominative Asp features trigger another (nominative).
  There is only one head (AspP); the alignment is derived by the
  case assignment properties of that head. This is the approach
  assumed in §\,\ref{sec:vm-asp-case} and is consistent with
  aspect-driven ergativity in the broader typological literature
  (Coon 2010; Legate 2008).
\end{description}

\noindent The third approach is preferred here. The desideratum
is to formalize exactly which feature of AspP triggers which
case-assigning behaviour, without reducing the analysis to mere
labelling of the phenomenon.

\subsection{Mood and the reduced Asp.Align: open questions}
\label{sec:vm-tension-mood-asp}

The interaction of Mood and Asp.Align is analyzed in
§\,\ref{sec:vm-mood-asp} as \term{impoverishment}: a non-indicative
Mood triggers the deletion of fine-grained viewpoint features from
the Asp.Align node, reducing the full five-way inventory to a
two-way IPFV/PFV-domain contrast. Asp.Align is always present in
finite clauses; what varies is the richness of its featural content.
Several questions remain empirically open.

First, the distribution of PFV-domain (\ird{-ov-}) versus
IPFV-domain ($\emptyset$) directive and irrealis forms requires
fuller corpus documentation. The semantic contrast---bounded,
completive directive versus ongoing, processual directive---is
typologically well-motivated by analogy with the perfective/
imperfective imperative distinction in West Slavic, but the
frequency, pragmatic conditioning, and any lexical restrictions on
PFV-domain non-indicatives in Iridian are not yet systematically
described.

Second, the interaction between the reduced Asp.Align and
information structure in directive clauses is understudied. In the
indicative, aspect choice encodes the discourse-topic orientation of
the clause (§\,\ref{sec:vm-asp-case}). Whether a parallel
topic-sensitivity operates for the IPFV/PFV distinction in
directive clauses---and whether the null agent of typical imperative
clauses interacts with this---has not been investigated.

Third, the irrealis (\ird{-ěv}) in its PFV-domain form
(\ird{-ov-ěv}) may exhibit distributional restrictions with certain
predicate classes. The degree to which the bounded/unbounded
irrealis distinction tracks predicate telicity (in the way the
indicative perfective/imperfective distinction does) remains to be
established through elicitation and text data.

\subsection{The Voice head \ird{-n-} and the causative \ird{-na-}: shared form, distinct functions}
\label{sec:vm-tension-n}

The morpheme \ird{-n-} (the Voice head that licenses the patient as a
core argument and specifies patient absolutive, appearing before
aspect suffixes on transitive stems) and the causative
\ird{-na-} share a common phonological shape. The analysis adopted
in §\,\ref{sec:vm-n} treats them as formally distinct morphemes
occupying distinct functional positions: \ird{-n-} is the Vocabulary
Item of the VoiceP head, while \ird{-na-} is the head of the higher
CausP projection (§\,\ref{sec:vm-caus}). They are not allomorphs of
a single morpheme, and their distribution is mutually reinforcing
rather than in competition: \ird{-na-} may co-occur with \ird{-n-}
on the same stem (the causative takes a transitive complement), while
\ird{-n-} and \ird{-aš-} cannot (they compete for the same VoiceP
slot).

The outstanding question is whether the phonological resemblance
between \ird{-n-} and \ird{-na-} is synchronically active or merely
a residue of a historical relationship. Two positions are possible.
On the \term{historical coincidence} view, the two morphemes derive
from a common ancestor that bifurcated into CausP and VoiceP
exponents in the history of Iridian, and their phonological overlap
is no longer productive. On the \term{morphological family} view, the
shared \ird{n} is a synchronically transparent marker of
argument-introducing functional heads, realized in full (\ird{-na-})
when an additional argument (the causee) is introduced and in reduced
form (\ird{-n-}) when no new argument is added. The grammar takes no
definitive position on this historical question, but regularizes
\ird{-na-} for the causative and \ird{-n-} for the VoiceP head
throughout.

\subsection{The absolutive--topic mapping: Agree or movement?}
\label{sec:vm-tension-abs-top}

The grammar's analysis treats the absolutive argument as the clause
topic, occupying Spec,TopP. Two formal options:

\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
\item[\upshape A-movement to Spec,TopP.] The absolutive DP moves
  from its base position (complement or specifier of VP/vP) to
  Spec,TopP in the narrow syntax. This movement is driven by a
  [\textsc{top}] feature on the Top head, which probes for a
  matching DP. Case is assigned in Spec,TopP (absolutive as
  a structural case of the topic position). This parallels subject
  raising to Spec,TP in English.
\item[\upshape Long-distance Agree without movement.] The absolutive
  DP remains in its base position; the Top head Agrees with it
  long-distance and values its case as absolutive. The DP's
  surface-initial position is then a consequence of base-ordering
  (head-finality places the specifier of the highest projection
  first) rather than movement.
\end{description}

\noindent The movement account is more standard and predicts
reconstruction effects and binding asymmetries that could in
principle be tested. The long-distance Agree account is more
parsimonious but makes weaker predictions. Which analysis is
correct has consequences for how topic drop is modeled: movement
predicts that topic drop is pro-drop in Spec,TopP; long-distance
Agree predicts that topic drop is simply the absence of a lexical
DP in the base position of the absolutive argument.

\subsection{Ergative case: dependent, structural, or inherent?}
\label{sec:vm-tension-erg}

The transitive/ergative case on the agent in ergative-aligned
aspects can be assigned by several mechanisms:

\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
\item[\upshape Dependent case (Marantz 1991).] The
  ergative is assigned to the agent because another DP (the patient)
  has already received absolutive in the same domain. The ergative
  is thus not assigned by any head but is computed as a dependent
  case: marked case on the DP in the presence of another marked-for-
  absolutive DP.
\item[\upshape Structural case assigned by $v^*$.] The transitive
  light verb $v^*$ assigns ergative to the external argument (agent)
  in its specifier. This is the standard Kratzer/Legate analysis for
  ergative languages.
\item[\upshape Asp-driven ergative.] The Asp head, as argued here,
  is responsible for the case pattern. In the ergative aspects, the
  Asp head's case-assigning operation assigns absolutive to the
  patient and ergative to the agent simultaneously. This requires
  the Asp head to case-assign two DPs, which is unusual in standard
  Agree but not unprecedented (cf.\ Baker 2015 two-probe model).
\end{description}

\noindent The Asp-driven account is consistent with the analysis
developed in §\,\ref{sec:vm-asp-case} and is preferred here. Its
key prediction is that the ergative case is aspect-sensitive: only
in perfective and retrospective aspects does the agent receive
ergative, and this is because only those Asp heads have the
case-assigning feature bundle that triggers the ergative pattern.


\subsection{The conjunctive linker \ird{-e}: see §\,\ref{sec:conjunctive-linker}}

The formal analysis of the conjunctive linker \ird{-e} --- including its phonological characterization, morphosyntactic distribution, the competing unified CP-edge, PF boundary-marker, and two-homophone analyses, and the outstanding empirical questions --- is developed in full in the conjunctive linker section above (§\,\ref{sec:conjunctive-linker}).

\subsection{The NEG prefix vs.\ NegP position: see §\,\ref{sec:vm-neg}}
\label{sec:vm-tension-neg}

The prefix-paradox tension and its resolution via the DM
morphological-incorporation account are discussed in full in
§\,\ref{sec:vm-neg}; the empirical commitments of that account
are listed at the end of §\,\ref{sec:vm-neg-exponents}.


%% ================================================================
%%  SUMMARY TABLE
%% ================================================================

\section{Summary: the verbal complex as a syntactic structure}
\label{sec:vm-summary}

Table~\ref{tab:vm-full-structure} gives the full proposed
hierarchical structure of the Iridian finite verbal clause, from
the highest projection (SpeechActP) to the lowest (VP/Root), with
the morphological exponents of each relevant layer.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{The Iridian verbal clause: functional hierarchy and
  morphological exponents.}\label{tab:vm-full-structure}
  \medskip
  \begin{tabularx}{\textwidth}{l l l L}
    \toprule
    {\sc level} & {\sc projection} & {\sc morphological exponent}
    & {\sc notes} \\
    \midrule
    Highest  & \textsc{SpeechActP} & Zone I particles (peripheral)
    & Post-topic; speech act framing \\
    & \textsc{EvidP}$_\text{outer}$ & \ird{oče, hlavdy, ktady, či}
    & Clause-level evidential/epistemic \\
    & \textsc{TopP} & [absolutive DP]
    & Topic = absolutive; A-movement target \\
    & \textsc{PolP} & \ird{by}
    & Clause typing: polar interrogative \\
    & \textsc{FocP} & [wh-phrase]
    & Focus/wh movement target \\
    \midrule
    & \textsc{EvidP}$_\text{inner}$ & \ird{-ejte} (?)
    & Indirect evidential; inside verbal complex \\
    & \textsc{MoodP} & \ird{-ím, -é, -ká, -ku, -ěv, -ir}, $\emptyset$
    & Non-indicative moods; suppresses Asp when overt \\
    & \textsc{NegP} & \ird{zá-} (prefix)
    & Clausal negation; PF-cliticized to verbal complex \\
    & \textsc{AspP} & \ird{-ek/-ik, -ac, -a, -ime, -ach} (fused w/ align)
    & Sole obligatory head; aspect + case assignment \\
    \midrule
    & \textsc{CausP} & \ird{-na-}
    & Adds external causer; dominates Voice/ApplP \\
    & \textsc{VoiceP / ApplP} & \ird{-aš-} (antipassive) / \ird{-éb-} (applicative)
    & Valency operations; complementary distribution \\
    & \textsc{VP} & Root stem
    & Lexical predicate; complement of V = preverbal \\
    Lowest & \textsc{VP complement} & NP / CP (\ird{po}) / VoiceP$_{\text{-it}}$
    & Object, po-clause, or purposive converb; all preverbal \\
    \bottomrule
  \end{tabularx}
\end{table}

\noindent The preverbal position of all verbal complements --- NP
objects, CP complements (\ird{po}-clauses), and purposive converb
complements --- follows uniformly from the head-final parameter
applied at the VP level. No movement or extraposition is required:
the complement is base-generated to the left of the verb and
surfaces there in the linear string. Higher projections (AspP,
NegP, MoodP, and the inner evidential) are realized as suffixes on the verbal
complex, which surfaces at the right edge of the clause as required by
pervasive head-finality.

%% ----------------------------------------------------------------
%% END: verb-minim.tex
%% ----------------------------------------------------------------
