\chapter{Nouns and noun phrases}
\label{ch:nouns}

%% ================================================================
%%  PART I: THE EXTENDED NOMINAL PROJECTION IN MINIMALIST TERMS
%% ================================================================

\section{Overview}
\label{sec:nm-overview}

The nominal morphological template of Iridian, reading from the
lexical root outward to the rightmost morpheme, is:

\medskip
\begin{center}
\textsc{root} + (\textsc{case}) + (\textsc{anch})
\end{center}
\medskip

\noindent Prenominal material precedes the root in the surface
string:

\medskip
\begin{center}
(\ird{ne}) + (\ird{si}/\ird{druh}) + (\textsc{dem}) +
(\textsc{quant}/\textsc{num}) + (\textsc{poss}.\Ind{}) +
(\textsc{prop.root} / \textsc{stative} / \textsc{\ird{-ovní}}) +
\textsc{root} + (\textsc{case}) + (\textsc{anch})
\end{center}
\medskip

This template is interpreted under the
\term{DP Hypothesis} (Abney 1987): the nominal
expression is not simply an NP but a full functional projection with
an articulated D-domain. In this chapter I use
\term{anchoring} as the theoretical label for the category expressed
by \ird{-to}; \emph{definiteness} is retained only as the
traditional descriptive name. The suffix \ird{-to} is therefore
treated as the overt realization of an anchoring head, labelled
\textsc{Anch\textsuperscript{0}}, when the feature
[\textsc{+anch}] is present.

Only the case and anchoring layers are analyzed here as ordinary
suffixal heads. The prenominal items \ird{ne}, the demonstrative
system, and \ird{si}/\ird{druh} are treated instead as overt
realizations of distinct higher nominal projections whose exponents
surface at the left edge of their domain. This avoids the present
oscillation between calling them projections in the prose and drawing
them as unstructured edge material in the trees. The hierarchical
organization of functional projections, reading from innermost to
outermost, is given in Table~\ref{tab:nm-projections}.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Functional projections of the Iridian DP (innermost to
  outermost).}\label{tab:nm-projections}
  \medskip
  \begin{tabularx}{\textwidth}{l L L l}
    \toprule
    {\sc position} & {\sc morpheme/item} & {\sc syntactic projection} &
    {\sc see} \\
    \midrule
    Root          & lexical stem        & NP / $\surd$P
                  & §\,\ref{sec:nm-root} \\
    Property root & small closed class  & PropP (adjunct to NP)
                  & §\,\ref{sec:nm-prop} \\
    Possessor     & oblique noun phrase & PossP (prenominal possessor layer inside KP)
                  & §\,\ref{sec:nm-poss} \\
    Case          & \Trs{} / \Ind{} endings & KP (case layer)
                  & §\,\ref{sec:nm-case} \\
    Anchoring     & \ird{-to}        & AnchP (suffixal D-layer)
                  & §\,\ref{sec:nm-def} \\
    Demonstrative & proximal \ird{tak}/\ird{tam}/\ird{ten};
                       medial \ird{tuto}/\ird{toho}/\ird{taly};
                       distal \ird{je}/\ird{jaky}/\ird{jení}
                  & DemP (noun-internal realization of the
                       demonstrative system)
                  & §\,\ref{sec:nm-dem} \\
    Quantification & numerals, adnominal interrogatives,
                     universal and negative quantifiers
                   & quantificational modifier field in the upper
                     nominal domain; exact cartographic slot left
                     conservative
                   & §\,\ref{sec:nm-quant} \\
    Specificity   & \ird{si}, \ird{druh} & SpecificityP
                       (left-peripheral specifically indefinite layer)
                  & §\,\ref{sec:nm-spec} \\
    Number        & \ird{ne}            & NumP
                       (plural semantics with left-edge spellout)
                  & §\,\ref{sec:nm-num} \\
    \bottomrule
  \end{tabularx}
\end{table}

The Mirror Principle (Baker 1985) predicts
that the morphological order reflects the syntactic order. Reading
the template from root outward, \textsc{case} appears closer to the
root than \textsc{anch}; this means KP is dominated by AnchP. The
same suffixal order follows directly from head-finality. By contrast,
the higher projections realized prenominally do not need to be
treated as ordinary suffixal heads. The simpler analysis is that
Iridian has an articulated nominal left periphery whose overt
exponents realize NumP, DemP, and SpecificityP at the left edge of
their domain, while the suffixal spine remains strictly head-final.

The head-final DP thus mirrors the head-final clause: just as the
finite verb — the head of the clause's highest functional projection
— surfaces at the right edge of the clause, the anchoring suffix
— the overt head of the noun phrase's suffixal D-layer — surfaces at
the right edge of the noun phrase. A convenient master tree follows.

\begin{center}
\begin{forest}
  syntax tree
  [NumP
    [{\ird{ne}}]
    [SpecificityP
      [{\ird{si}/\ird{druh}}]
      [DemP
        [{\textit{demonstrative series}}]
        [QP
          [{\textit{numeral / quantifier}}]
          [AnchP
            [KP
              [PossP
                [{\textit{oblique possessor}}]
                [{NP\\\textit{root}}]
              ]
              [{K\textsuperscript{0}\\{\footnotesize case}}]
            ]
            [{Anch\textsuperscript{0}\\\ird{-to} / $\varnothing$}]
          ]
        ]
      ]
    ]
  ]
\end{forest}
\end{center}

\noindent The tree keeps the nominal spine uniform across the
subsections below. Overt exponents of higher nominal projections are
shown directly at the edge of the projection they realize, and silent
heads are not inserted merely to force a specifier analysis. Only the
case and anchoring heads are represented as overt heads because only
they are consistently realized as suffixal morphology. The QP layer is
drawn only as a conservative cover term for prenominal quantificational
material; the sections below do not assume a finer numeral-vs-quantifier
split unless the data require it.


\subsection{The root and NP}
\label{sec:nm-root}

The lexical noun occupies the head position of NP. Unlike the verbal
root, which is licensed by a selecting functional head (the Asp.Align
morpheme is obligatory on every finite verb; cf.\
§\,\ref{sec:vm-root}), the nominal root may surface without any overt
functional morphology: the direct case is the unmarked form, with zero
ending. The absence of \ird{-to} in the bare noun is therefore not
the absence of nominal structure. Rather, it is the absence of overt
anchoring morphology: the noun phrase is interpreted as unanchored,
and \textsc{Anch\textsuperscript{0}} receives no overt spellout.

I continue to use the DP Hypothesis as the general claim that Iridian
nominals project beyond NP, but the trees below label the relevant
D-layer as AnchP so that the analysis remains terminologically stable:
the theoretical feature is anchoring, while \emph{definiteness}
remains the traditional descriptive label for one common use of that
layer.


%% ================================================================
%%  PART II: LOW NOMINAL MORPHOLOGY
%% ================================================================

\section{The case system}
\label{sec:nm-case}

\subsection{Case, KP, and the alignment-sensitive argument system}
\label{sec:nm-case-system}

The lowest overt functional layer of the noun phrase is KP, the case
projection. Modern Iridian has three productive structural cases:
\term{direct}, \term{transitive}, and \term{indirect}; a fourth case,
the \term{vocative}, survives as a nonproductive form limited to
direct address in lexically specified items. The direct case is
morphologically unmarked and supplies the citation form. The
transitive and indirect are suffixal realizations of
\textsc{K\textsuperscript{0}}, while the vocative is a specialized
speech-act-adjacent realization of the same low nominal domain.

\begin{table}
  \footnotesize\sffamily
  \caption{Case values in the Iridian nominal domain.}
  \label{tab:nm-cases}
  \medskip
  \begin{tabularx}{\textwidth}{l L L}
    \toprule
    {\sc case} & {\sc primary function} & {\sc structural interpretation} \\
    \midrule
    Direct & citation form; absolutive argument; predicate nominal &
    absence of overt K exponent \\
    Transitive & agent of perfective/retrospective transitives &
    ergative realization of K \\
    Indirect & patient in nominative-alignment aspects; possessors;
    obliques; governed complements of postpositions &
    general non-core K realization \\
    Vocative & direct address in lexically specified names, kin
    terms, and a small residue of nouns &
    speech-act linked case in the low nominal domain \\
    \bottomrule
  \end{tabularx}
\end{table}

The split alignment facts belong to the clause, but the
nominal contribution is straightforward: KP makes available the case
values that the clausal architecture selects. In the perfective and
retrospective aspects the patient remains direct while the agent is
transitive; in the nominative-alignment aspects the agent remains
direct while the patient appears in the indirect.

\pex
\a
\begingl
\gla Marek zmacime.//
\glb Marek.\Dir{} run-\Prog{}//
\glft \trsl{Marek is running.}//
\endgl
\a
\begingl
\gla Travat Marku piaštnik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} eat-\Pf{}//
\glft \trsl{Marek ate the bread.}//
\endgl
\a
\begingl
\gla Marek travě piastime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.}//
\endgl
\xe

\noindent Four points about the architecture deserve explicit statement
before the declensional detail.

First, the three productive cases serve distinct syntactic functions. The
direct is the citation form and the unmarked absolutive. The transitive
marks the agent in perfective and retrospective aspects. The indirect is
the all-purpose non-core case: patient in nominative-alignment aspects,
possessor, oblique, and postpositional complement alike. A fourth case,
the vocative, survives nonproductively for direct address.

Secondly, declension is not a mere inventory of endings. It is the
interaction of case assignment with stem class and stem grade: each noun
carries a class specification that determines which K-exponents it
realizes and which stem operations those exponents trigger. The classes
and processes are described in §§\,\ref{sec:nm-stemform}--\ref{sec:nm-declension}.

Thirdly, anchoring is assembled outside case and may therefore attach to
any case form: direct, transitive, or indirect alike. Anchoring in
non-direct case forms is possible, though more marked pragmatically, as
noted in the discourse chapter (§\,\ref{sec:disc-def}). The structural
relation AnchP $>$ KP makes this prediction automatically and grounds the
full anchored paradigms described in §\,\ref{sec:nm-def-nondirect}.

Fourthly, the morphological template from root to right edge is:

\medskip
\begin{center}
\textit{root} $\rightarrow$
\textit{class-selected stem grade} $\rightarrow$
K\textsuperscript{0} exponent $\rightarrow$
Anch\textsuperscript{0} exponent
\end{center}
\medskip

\noindent That template is the stable backbone of the rest of this
section.

\subsection{Stem formation and morphophonemic processes}
\label{sec:nm-stemform}

Before the declensional paradigms can be read properly, the nominal
morphophonemic processes recruited by the declension classes must be set out.
These are not peripheral irregularities, but productive operations that help
constitute class membership itself. A nominal class is defined not only by its
endings, but also by the morphophonemic operations it permits or requires.

Weak-vowel deletion affects a lexical stem vowel, most commonly \ird{e} or a
phonologically reduced counterpart in the final stem syllable. This vowel
deletes productively before a vowel-initial suffix. The process is triggered by
the K-suffix and applies systematically in those consonant-stem classes that
admit it: \ird{perek} $\rightarrow$ \ird{prku}, \ird{lobek} $\rightarrow$
\ird{lubku}. The resulting form is the \term{reduced} or \term{zero-grade} stem.

Raised-grade formation follows weak-vowel deletion in certain classes. When
deletion produces an obstruent or liquid cluster, the output is regularised by
grade raising: \ird{o} $\rightarrow$ \ird{u} in back-raised subclasses, as in
\ird{lobek} $\rightarrow$ \ird{lubk-}, and \ird{e/ě/é} $\rightarrow$ \ird{i/í}
in front-raised subclasses. In such classes, the raised stem is the one selected
by the oblique endings.

Front-vocalic mutation is triggered by indirect endings with a front vowel and
affects the immediately preceding consonant. Velars undergo affrication, as in
\ird{k} $\rightarrow$ \ird{c} and \ird{g} $\rightarrow$ \ird{z}, while coronals
are fronted, as in \ird{t} $\rightarrow$ \ird{č} and \ird{d} $\rightarrow$
\ird{dž}. The class of \ird{perek} illustrates the interaction of these
processes: \ird{perek} $\rightarrow$ \ird{prce}, where deletion, raising, and
affrication all apply.

Liquid resolution arises in weak-vowel stems ending in a liquid. When deletion
removes the supporting vowel, the liquid may surface in a syllabic, reduced, or
metathesised form: \ird{uzel} $\rightarrow$ \ird{uzla}. This process determines
the shape of the reduced stem in a distinct subset of nouns.

Analogical levelling is found in certain class members that have generalised a
single reduced stem across several oblique forms, thereby suppressing
alternations that were formerly regular. Such levelled forms are lexically
specified and will be noted under the individual classes below.

Taken together, these processes provide the basis for identifying the stem
grades that appear in the paradigms that follow. Once they are understood, the
larger declensional patterns appear structurally motivated rather than
arbitrary.


\subsection{Declension class as the lexical specification of KP}
\label{sec:nm-declension}

KP is not realized by a single invariant suffix. Rather, the noun
lexically selects a declension class whose case exponents interact
with the morphophonemic processes described in
§\,\ref{sec:nm-stemform}. That lexical information is part of the
nominal root's stored feature bundle: the root enters the derivation
already specified for which K-exponents it realizes and which stem
grades those exponents select.

Six macroclasses are recognized. Five are broadly productive structural
classes; the sixth is a small closed relic class of inherited stems
with front-vocalic oblique exponents not generalisable to the main classes.
The macroclasses are identified by stem type and by the exponents they
use across the direct, transitive, and indirect cases.

\paragraph{Class I: Hard consonant stems.}
The core class for ordinary consonant-final nouns, including personal
names and a large body of lexical nouns. The direct is the bare citation
stem. The transitive adds a back-vocalic exponent, typically \ird{-a}
(and \ird{-u} in certain subclasses). The indirect adds a back or
front-vocalic exponent, typically \ird{-am} or a related ending, and
may trigger softening of the final consonant in subclasses with
front-coded obliques.

\pex
\a \ird{byl} $\rightarrow$ \ird{byla} / \ird{bylam}
\xe

\noindent This class is semantically open and is one of the chief
productive defaults for consonant-final nouns. Subclasses differ in
the precise vocalism of their oblique exponents, but share the
fundamental property of a bare-stem direct and suffixal case endings.

\paragraph{Class II: Weak-vowel consonant stems.}
The morphophonemically rich consonant-stem class. Stems carry an
unstable final \ird{e} or similar weak vowel that deletes productively
before vowel-initial oblique endings. Many subclasses then undergo
raising or front-vocalic mutation. This class should be understood as
the most morphologically active productive consonant-stem family.

\pex
\a \ird{lobek} $\rightarrow$ \ird{lubku} / \ird{lubcí}
\a \ird{perek} $\rightarrow$ \ird{prku} / \ird{prce}
\xe

\noindent Back-raised subclasses show \ird{o} $\rightarrow$ \ird{u};
front-raised subclasses show \ird{e/ě} $\rightarrow$ \ird{i}.

\paragraph{Class III: \ird{-a} stems.}
The chief vowel-stem class and the largest open class among ordinary
lexical nouns. The direct ends in \ird{-a}. The transitive replaces the
final vowel with a back or central exponent, typically \ird{-y}. The
indirect takes a front exponent, typically \ird{-ě}. This class is
highly productive and relatively transparent.

\pex
\a \ird{trava} $\rightarrow$ \ird{travy} / \ird{travě}
\xe

\paragraph{Class IV: Soft and sibilant stems.}
Nouns whose citation stem already ends in a historically soft, palatal,
or sibilant segment. Their oblique formation typically involves smaller
overt alternations because softness is already built in. These stems
are defined less by weak-vowel deletion and more by soft-grade
preservation, sibilant alternation, and front-vocalic compatibility.

\pex
\a \ird{red} $\rightarrow$ \ird{rezy} / \ird{reže}
\xe

\paragraph{Class V: Vowel-replacing stems.}
Nouns such as \ird{ráno} whose oblique formation reshapes the final
vowel rather than simply concatenating a suffix. These are historically
old vowel stems whose modern inflection shows replacement endings or
inherited vowel alternation. The class is semi-productive, with many
lexically fixed members.

\pex
\a \ird{ráno} $\rightarrow$ \ird{ránou} / \ird{ráne}
\xe

\paragraph{Class VI: Relic front-vocalic stems.}
A small, closed class of inherited stems whose oblique exponents — a
transitive in \ird{-im} and an indirect in \ird{-y} — are not
generalisable to the main productive classes. These items form a tight
microclass learned as a lexical residue.

\pex
\a \ird{pět} $\rightarrow$ \ird{pětim} / \ird{pěty}
\xe

\noindent The essential Minimalist point is unchanged: declension class
is not a rival to syntactic structure but the lexical instruction that
tells PF how a given noun realizes the K head merged in the syntax.
The syntax provides KP; the lexicon and PF provide the class-sensitive
spellout.

\subsection{Stem grades as allomorph selection under K}
\label{sec:nm-stem-grades}

Case endings do not simply attach to a single invariant stem. They
select among recurring stem grades: the basic stem, zero-grade stem,
front-raised stem, back-raised stem, soft stem, and lexically levelled
stem. These are contextually conditioned allomorphs of the same root,
triggered by the K head and by the morphophonemic processes described
in §\,\ref{sec:nm-stemform} above.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Case endings and their typical stem selection across the six
  macroclasses.}
  \label{tab:nm-stem-selection}
  \medskip
  \begin{tabularx}{0.95\textwidth}{L L L}
    \toprule
    {\sc case slot} & {\sc typical endings by macroclass} &
    {\sc stem selected} \\
    \midrule
    Direct & zero (all classes) & basic stem \\
    Transitive & I--II: \ird{-u}; III: \ird{-y}; IV: \ird{-y}/\ird{-ovy};
    V: vowel-replacing; VI: \ird{-im} &
    basic, zero-grade, raised, or vowel-replacing stem by class \\
    Indirect & I: \ird{-í}/\ird{-ení}; II: \ird{-í}/\ird{-ení}/\ird{-e};
    III: \ird{-ě}; IV: \ird{-ě}/\ird{-ově}; V: vowel-replacing;
    VI: \ird{-y} &
    soft or front-vocalic stem for front endings; zero-grade or raised
    stems where licensed; replacement in V and VI \\
    Vocative & \ird{-o}, \ird{-e}, zero &
    basic or soft stem; nonproductive \\
    \bottomrule
  \end{tabularx}
\end{table}

Representative outputs from the macroclass system confirm that the
nominal system is rule-governed rather than suppletive:

\pex
\a \ird{byl} $\rightarrow$ \ird{byla} / \ird{bylam} \lingcontext{Class I: hard consonant}
\a \ird{perek} $\rightarrow$ \ird{prku} / \ird{prce} \lingcontext{Class II: weak-vowel deletion plus softening}
\a \ird{trava} $\rightarrow$ \ird{travy} / \ird{travě} \lingcontext{Class III: \ird{-a} stem}
\a \ird{red} $\rightarrow$ \ird{rezy} / \ird{reže} \lingcontext{Class IV: soft/sibilant stem}
\a \ird{ráno} $\rightarrow$ \ird{ránou} / \ird{ráne} \lingcontext{Class V: vowel-replacing}
\a \ird{pět} $\rightarrow$ \ird{pětim} / \ird{pěty} \lingcontext{Class VI: relic front-vocalic stem}
\xe

\subsection{Vocative and proper-name behavior}
\label{sec:nm-vocative}

The vocative is the most restricted member of the case system. It is
nonproductive in the modern language. A marked vocative is viable only
for lexically specified items: inherited proper names, older
integrated loan proper names, kinship terms, and a restricted residue
of concrete and abstract nouns. The latter class is lexically
determined item by item and does not follow systematically from
membership in the inherited vocabulary. Outside those lexicalized
items, direct address uses the unmarked direct case. In other words,
in the absence of a marked vocative, the language defaults to the
direct case. The simplest analysis is that the vocative realizes the
low nominal domain in direct relation to the speech-act layer,
bypassing the ordinary anchoring opposition. This is why the vocative
rejects \ird{-to}: direct address identifies the referent
deictically by the act of address itself.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Kinship terms retaining lexicalized vocative forms.}
  \label{tab:nm-vocative}
  \medskip
  \begin{tabularx}{0.7\textwidth}{L l l}
    \toprule
    {\sc direct} & {\sc vocative} & {\sc meaning} \\
    \midrule
    \ird{papka}  & \ird{papko}  & father \\
    \ird{mamka}  & \ird{mamko}  & mother \\
    \ird{mlazka} & \ird{mlazko} & brother \\
    \ird{voska}  & \ird{vosko}  & sister \\
    \bottomrule
  \end{tabularx}
\end{table}

Proper names otherwise behave like ordinary DPs: they may bear the
ordinary case suffixes, may host the honorific use of \ird{ne}, and
may participate in possessive and demonstrative constructions. What is
special about names is not their exemption from the nominal spine, but
their lexically strong identifiability, which makes bare-name
reference especially natural and makes overt anchoring more marked than
with common nouns unless a discourse-contrastive reading is intended.


%% ================================================================
%%  PART III: THE ANCHORING SUFFIX AS THE NOMINAL D-LAYER
%% ================================================================

\section{Anchoring and definiteness}
\label{sec:nm-def}

\subsection{The suffix \ird{-to} and its allomorphy}
\label{sec:nm-def-allomorph}

The anchoring suffix \ird{-to} has two surface-distinct allomorphs whose
distribution is conditioned by the phonological shape of the immediately
preceding morpheme — the case ending, or the bare stem where no case ending is
present.

\begin{enumerate}[wide]
  \item The full allomorph \ird{-te/-to} is inserted when the immediately
  preceding segment is a \textbf{vowel}. The full form is preserved because no
  cluster-breaking is required.
  \item A reduced allomorph \ird{-e/-o} is inserted when the immediately preceding
  segment is a \textbf{consonant}. The epenthetic vowel breaks up a potential
  consonant cluster. No final \ird{-t} is present on the surface: the historic
  \ird{*-to/-te} loses its stop in the consonant-adjacent environment, leaving
  only the vowel.
\end{enumerate}

In both allomorphs, the choice between front (\ird{-e}, \ird{-te}) and back
(\ird{-o}, \ird{-to}) vocalism is governed by class-conditioned harmony rather
than purely by the surface vocalism of the immediately preceding case ending.
Some declension classes preserve front anchoring in the direct case but
generalise back-vocalic anchoring in the oblique cases; see
§\,\ref{sec:nm-def-nondirect} for the full paradigmatic account.

The traditional descriptive label for this suffix is \emph{definiteness}, but in
the present discussion the theoretical feature is anchoring. In Distributed
Morphology terms (Halle \& Marantz 1993; Embick \& Noyer 2007), the reduced and
full variants are two Vocabulary Items competing for insertion at the same
\textsc{Anch\textsuperscript{0}} node. The choice is made by phonological
context at PF: the reduced allomorph (\ird{-e/-o}) is the consonant-context
spellout, while the full allomorph (\ird{-te/-to}) is the vowel-context
spellout.

Representative paradigms are given in
Table~\ref{tab:nm-def-allomorphy}.

\begin{table}
    \footnotesize
    \caption{Allomorphy of the anchoring suffix across cases and
    declension classes.}\label{tab:nm-def-allomorphy}
    \medskip
    \begin{tabularx}{\textwidth}{L l l l l l l l}
      \toprule
      {\sc noun} & {\sc class} & {\sc dir} & {\sc trs} & {\sc ind} & {\sc dir.def} & {\sc trs.def} & {\sc ind.def} \\
      \midrule
      \irdp{byl}{child}  & I & \ird{byl}   & \ird{byla}  & \ird{bylam} & \ird{byle}  & \ird{bylato}  & \ird{bylamo} \\
      \irdp{lobek}{apple} & II  & \ird{lobek}  & \ird{lubku} & \ird{lubcí} & \ird{lubko}  & \ird{lubkuto}  & \ird{lubcíte} \\
      \irdp{perek}{body} & II & \ird{perek}  & \ird{prku}  & \ird{prce}  & \ird{prko}  & \ird{prkuto}  & \ird{prcete} \\
      \irdp{trava}{bread} & III & \ird{trava}  & \ird{travy}  & \ird{travě} & \ird{travato}  & \ird{travyte}  & \ird{travíte} \\
      \irdp{red}{blade} & IV & \ird{red}  & \ird{rezy}  & \ird{reže}  & \ird{rede}  & \ird{rezyte}  & \ird{režete} \\
      \irdp{ráno}{hand} & V & \ird{ráno}  & \ird{ránou}  & \ird{ráne}  & \ird{ránó}  & \ird{ránuto}  & \ird{ránete} \\
      \irdp{pět}{sun} & VI & \ird{pět}  & \ird{pětim}  & \ird{pěty}  & \ird{pěto}  & \ird{pětimo}  & \ird{pětyte} \\
      \bottomrule
    \end{tabularx}
\end{table}


\medskip
\begin{center}
\textsc{Anch\textsuperscript{0}}[\textsc{+anch}] $\rightarrow$
\ird{-e/-o} / $[_{\sigma}\ldots C]$ \underline{\quad},\quad
\ird{-te/-to} / $[_{\sigma}\ldots V]$ \underline{\quad}
\end{center}
\medskip

\noindent where \ird{e}/\ird{te} is front-coded and \ird{o}/\ird{to} is
back-coded. The conditioning environment is the phonological shape of the
immediately preceding morpheme; the vocalism is governed by the declension class
of the host, as elaborated in §\,\ref{sec:nm-def-nondirect}.


\subsection{Case stacking and the ordering AnchP $>$ KP}
\label{sec:nm-def-stacking}

The surface fact is straightforward: \ird{-to} attaches to the
\emph{already case-inflected} noun. The morphological order is
therefore:

\begin{center}
\textsc{root} $+$ \textsc{case-ending} $+$ \textsc{anch}
\end{center}

\noindent The interpretive consequence is that case determines the
noun's clausal role, while anchoring is added outside that case layer.
The simplest structural analysis is therefore KP dominated by AnchP:

\begin{center}
\begin{forest}
  syntax tree
  [AnchP
    [KP
      [NP
        [{\textit{root}}]
      ]
      [{K\textsuperscript{0}\\{\footnotesize case ending}}]
    ]
    [{Anch\textsuperscript{0}\\\ird{-to} / $\varnothing$}]
  ]
\end{forest}
\end{center}

\noindent This tree deliberately omits the higher left-peripheral
layers. Demonstratives, \ird{ne}, and \ird{si}/\ird{druh} are treated
below as distinct edge material outside AnchP rather than as
specifiers of AnchP itself. The right edge of KP is therefore the only
place where the overt anchoring head is expected to surface.

This analysis parallels the verbal domain exactly. The Asp.Align
morpheme — the verbal analogue of the overt nominal D-layer — is also
the sole obligatory element of the finite verb, surfaces at the right
edge of the verbal complex, and mediates between the lexical root and
the higher inflectional layers (§\,\ref{sec:vm-asp}). In both the
verbal and nominal domains, the rightmost morpheme is the head of the
highest overt suffixal projection, and the directionality of
morphological layering is predicted by head-finality.


\subsection{Anchoring in non-direct case forms}
\label{sec:nm-def-nondirect}

The morphological template predicts that anchoring, being outside KP,
may attach to any case form: direct, transitive, or indirect alike.
This prediction is borne out. Anchored transitive and indirect forms
exist, though they are more marked pragmatically than anchored direct
forms, in keeping with the observation in the discourse chapter
(§\,\ref{sec:disc-def}) that anchoring in non-direct cases signals a
higher degree of contrastive individuation or resumptive tracking
rather than simple anaphoric definiteness.

This asymmetry matters for the clause-level analysis as well. If the agent of a
perfective transitive were the language's neutral discourse center simply by
virtue of agenthood, one would expect anchored transitive agents to be the least
marked pattern. The facts point in the opposite direction: anchoring on the
direct-case absolutive is the neutral way to maintain a discourse referent,
whereas anchoring on transitive or indirect DPs signals extra contrastive or
resumptive work. The anchoring system therefore supports an absolutive-pivot
analysis of discourse prominence rather than a uniform subject-prominence model.

The allomorphy rule applies uniformly across cases. Since the
transitive and indirect endings are typically vowel-final, they
systematically trigger the full allomorph \ird{-te/-to}. The direct,
which for consonant-stem classes ends in a consonant, triggers the
reduced allomorph \ird{-e/-o}. The following paradigm illustrates a
full anchored series for \ird{byl} \trsl{child} (Class II):

\pex
\a \ird{byl} [\Dir{}] $\rightarrow$ \ird{byle}
\a \ird{byla} [\Trs{}] $\rightarrow$ \ird{bylato}
\a \ird{bylam} [\Ind{}] $\rightarrow$ \ird{bylamo}
\xe

\noindent Two features of this paradigm deserve explicit comment.
First, the reduced allomorph on the direct is front-coded (\ird{-e}),
while the full allomorphs on the transitive and indirect are
back-coded (\ird{-to}, \ird{-o}). This asymmetry does not reflect the
surface phonology of the immediately preceding vowel — \ird{byla}
ends in \ird{-a}, a low back vowel, yet the anchored form is
\ird{bylato} rather than \ird{*bylate}. It reflects instead the
\emph{declension class} of the case ending.

Secondly, the anchored indirect \ird{bylamo} shows the reduced
allomorph (\ird{-o}), not the full allomorph, because \ird{bylam}
ends in the consonant \ird{m}: the general consonant-context rule
applies at the immediate surface boundary, regardless of the vowel
that precedes the consonant.

The recommended analysis is therefore the following. The anchoring
suffix selects its reduced or full allomorph from the phonological
shape of the immediately preceding segment (consonant $\rightarrow$
reduced; vowel $\rightarrow$ full). Within the reduced allomorph, the
choice of \ird{-e} versus \ird{-o} is class-governed: the direct
bare stem of \ird{byl} belongs to the front-coded subclass of Class II
and thus attracts \ird{-e}. Within the full allomorph, the choice of
\ird{-te} versus \ird{-to} is also class-governed: the oblique endings
\ird{-a} (transitive) and \ird{-am} (indirect) belong to back-coded
sub-paradigms in this class and therefore attract \ird{-to/-o}.

The result is a principled, learnable paradigmatic asymmetry: front
reduced anchoring in the direct; back full or back reduced anchoring in
the obliques. The general rule is:

\medskip
\begin{center}
Anchoring vocalism = front in the direct for front-coded classes; \\
back in the obliques for back-coded oblique exponents.
\end{center}
\medskip

\noindent For a vowel-stem class (Class III), where the direct is
itself vowel-final, the full allomorph applies throughout the paradigm
but the vocalism pattern is similarly class-governed:
\ird{trava} $\rightarrow$ \ird{travato} (\ird{-to}, back-coded direct);
\ird{travě} $\rightarrow$ \ird{travěte} (\ird{-te}, front-coded indirect).


\subsection{A historical note on the anchoring suffix and the
  declension classes}
\label{sec:nm-def-history}

The modern case endings are not all descendants of one uniform ending
per case. Rather, each case had a small family of historical allomorphs
distributed by stem type, and the modern declensional classes are the
result of those old exponents passing through regular sound change and
later analogical levelling. The following six-stage account makes
the modern system coherent.

\paragraph{Proto-stage.} Proto-Iridian had a small set of nominal
endings: a bare direct with no overt suffix; a transitive marker of the
shape \ird{-a} or \ird{-u}, distributed by stem class; an indirect
marker with front vocalism, perhaps \ird{-i} or \ird{-e}; and an
anchoring enclitic of the shape \ird{*-to/-te}. Nominal stems fell into
inherited classes: hard consonant stems, weak-vowel stems, \ird{-a}
stems, vowel stems, and a residue of soft or sibilant stems. These are
the direct ancestors of the six modern macroclasses.

\paragraph{Oblique vowel differentiation.} The transitive and indirect
endings began to specialise phonologically. Back-vocalic obliques
remained associated with transitive marking, while front-vocalic
obliques favoured indirect functions and triggered palatalisation or
fronting of the preceding consonant. This explains why the indirect
in the modern language is the natural locus of softening and mutation,
as described in §\,\ref{sec:nm-stemform}.

\paragraph{Weak-vowel loss.} In consonant stems with a weak final
\ird{e}, that vowel was lost before vowel-initial endings wherever
the resulting cluster was admissible. The grammar already has this as
a living synchronic process (§\,\ref{sec:nm-stemform}); historically
it is the source of the modern reduced-stem alternants:
\ird{*perek-u} $>$ \ird{prku}, \ird{*perek-i} $>$ \ird{prce}.

\paragraph{Grade formation and raising.} Dense clusters arising from
weak-vowel deletion were regularised by raised-grade formation in
certain classes: \ird{o} $\rightarrow$ \ird{u}, \ird{e/ě}
$\rightarrow$ \ird{i/í}. This yielded the modern stem-grade
alternations that characterise Class II.

\paragraph{Class reanalysis.} Once the phonological outputs stabilised,
speakers reinterpreted them as characteristic stem grades of distinct
declension classes. Some classes levelled a reduced stem across several
forms; others preserved alternations only in one case. This class
reanalysis is the source of what modern speakers experience as the
``gut feel'' of noun class membership.

\paragraph{Reduction of the anchoring suffix.} The old anchoring
suffix \ird{*-to/-te} was retained in full after vowel-final case
forms, but reduced after consonant-final forms by loss of the stop,
yielding modern \ird{-o/-e}. In weak-vowel stems the reduced anchor
selects the same reduced stem used in oblique formation. The result
is the paradigmatic system described in §\,\ref{sec:nm-def-nondirect}:
\ird{Marek} $\rightarrow$ \ird{Marko};
\ird{byl} $\rightarrow$ \ird{byle};
\ird{byla} $\rightarrow$ \ird{bylato};
\ird{bylam} $\rightarrow$ \ird{bylamo}.

\paragraph{Relic preservation in Class VI stems.} A small set of
inherited stems preserved older oblique exponents longer than ordinary
nouns, yielding the modern Class VI forms. The paradigm
\ird{pět} / \ird{pětim} / \ird{pěty} illustrates this: the
front-vocalic oblique exponents \ird{-im} (transitive) and \ird{-y}
(indirect) are not generalisable to the main declensional classes yet
remain internally coherent as a microclass. Because \ird{pět} is
consonant-final in the direct, its anchored direct is \ird{pěte}
(reduced front allomorph). Its oblique anchored forms take the full
allomorph: \ird{pětimto}, \ird{pětyto} (or \ird{pětyte} if \ird{-y}
is treated as front-coded in this class).


\subsection{The \ird{-ou\,+\,-to} contraction}
\label{sec:nm-def-contraction}

The sequence arising when the anchoring suffix follows a noun ending
in \ird{-ou} (Class~18, including all deverbal nominals in the direct
case; §\,\ref{sec:event-nominals}) undergoes a distinct PF operation:

\begin{center}
/\textsc{root}\ird{-ou}/ +
\textsc{Anch\textsuperscript{0}}[\textsc{+anch}]
$\longrightarrow$ \ird{-ovat}
\end{center}

\noindent Thus \ird{piaščikou} combined with the anchoring head surfaces as
\ird{piaščikovat}, not \ird{*piaščikoute} or \ird{*piaščikouto}. In
the other cases of the same noun, where the vowel-final endings are
\ird{-ovy} and \ird{-ově}, the regular vowel-context rule applies
without contraction: \ird{piaščikovyte} and \ird{piaščikověte}.

The simplest analysis is a PF morphological fusion: when the
anchoring node is adjacent to the \ird{-ou} ending, the two are
spelled out together as \ird{-ovat}. An alternative truncation
analysis is conceivable, but the stem-extension pattern in the rest of
the paradigm makes the fusion account the better fit at present.

The contraction has a grammatical consequence. Because the direct-case
form of deverbal nominals ends in \ird{-ou}, the anchored form
\ird{-ovat} is visually and phonologically distinctive from both the
plain form and the other case forms. This distinctiveness is exploited
in one disambiguation strategy for complex gerunds
(§\,\ref{sec:event-nominals}): when a gerund contains two
indirect-case nominal arguments, marking the theme nominal with
\ird{-to} identifies it unambiguously as the theme.


\subsection{Anchored reference and the [\textsc{+anch}] feature}
\label{sec:nm-def-semantics}

The descriptive generalization is that nouns with \ird{-to}
normally pick out a referent that the speaker presents as already
recoverable or singled out in the discourse. This includes the
familiar uses traditionally grouped under \emph{definiteness}, but the
range is slightly broader: the suffix also appears where the relevant
effect is contrastive individuation or resumptive tracking rather than
simple article-like definiteness.

The interpretive effect is therefore best described as
\term{anchoring}. A noun phrase marked by \ird{-to} is construed as
an anchored discourse token rather than as a merely lexical type or an
unintroduced individual. The most common contexts are the following:

\begin{enumerate}[wide]
\item Anaphoric:The anchored noun resumes a referent
  introduced in prior discourse.
\item Situationally unique: The anchored noun refers to an
  entity that is salient in the speech situation, though Iridian is
  more selective here than languages with obligatory articles.
\item Contrastive or resumptive: The suffix appears where
  the noun is being re-identified, resumed, or singled out as the
  discourse-relevant token.
\item Generic/kind: The bare noun is the ordinary vehicle
  for generic and kind-denoting reference. A noun with
  [\textsc{+anch}] generally resists a generic reading.
\end{enumerate}

The theoretical inference, stated cautiously, is that the Iridian
D-layer is best modeled as \textsc{Anch\textsuperscript{0}} carrying a
feature [\textsc{+anch}]. That feature instructs the hearer to recover
an individuated referent from the discourse model. In that sense,
\emph{definiteness} is the traditional descriptive name, while
\emph{anchoring} is the theoretical term used throughout the rules and
trees below.


\subsection{Distribution restrictions}
\label{sec:nm-def-restrictions}

Several contexts exclude the anchoring suffix:

\paragraph{The vocative.}
The vocative case is incompatible with \ird{-to}. It is also
nonproductive: only lexically specified vocatives survive, especially
with inherited and older integrated loan proper names, kinship terms,
and a small residue of concrete and abstract nouns. Elsewhere direct
address uses the direct case. The simplest analysis is that direct
address already identifies its referent at the speech-act level and
therefore does not enter the ordinary anchoring opposition.

\paragraph{Generic and kind-denoting NPs.}
As noted in §\,\ref{sec:nm-def-semantics}, [\textsc{+anch}] is
incompatible with the generic reading. Generics are expressed by the
bare noun because they denote a class or kind rather than a discourse
anchored token.

\paragraph{Approximate quantification with \ird{ne} and numerals.}
When the plural clitic \ird{ne} combines with a numeral or quantifier
to yield an approximate reading (\ird{ne hrona byl} \trsl{about three
children}), the anchoring suffix is excluded. The distribution
suggests that an approximately quantified expression is not being
packaged as an anchored nominal individual, but as an imprecisely
bounded plurality.


\subsection{Double-marking: demonstrative + \ird{-to}}
\label{sec:nm-def-double}

The grammar (§\,\ref{sec:definiteness}) notes that a demonstrative may
co-occur with the anchoring suffix:
\ird{tak byle} \trsl{this very child; this particular child here}.
This construction is marked rather than impossible.

The surface fact is therefore one of \emph{layering}, not strict
incompatibility. The demonstrative contributes deixis, while
\ird{-to} contributes anchoring. A plain demonstrative phrase is
structurally unmarked. A demonstrative plus anchoring suffix is
typically more emphatic, resumptive, contrastive, or strongly
individuating than its plain counterpart.

The theoretical analysis adopted here is that demonstratives occupy a
DemP above AnchP (§\,\ref{sec:nm-dem}). Co-occurrence is grammatical
because the two layers are distinct; it is marked because the noun
phrase is being both deictically located and additionally singled out
as the anchored discourse token.

\pex
\a
\begingl
\gla tak byl//
\glb \Dem{}.\Prox{} child.\Dir{}//
\glft \trsl{this child}//
\endgl
\a
\begingl
\gla tak byle//
\glb \Dem{}.\Prox{} child.\Dir{}.\Def{}//
\glft \trsl{this very child; this particular child here}//
\endgl
\xe


%% ================================================================
%%  PART IV: THE UPPER NOMINAL LAYERS
%% ================================================================

\section{The upper nominal functional sequence: demonstratives,
  quantification, number, and specific indefiniteness}
\label{sec:nm-upper}

\subsection{The demonstrative system and noun-internal DemP}
\label{sec:nm-dem}

Iridian demonstratives form a three-way deictic system with proximal,
medial, and distal values. The same paradigm has three distinct
morphosyntactic realizations: an adnominal demonstrative, a
pronominal demonstrative, and an indirect-case possessive
demonstrative. The noun-internal analysis developed here concerns the
adnominal realization, namely DemP, but the wider demonstrative
system must be acknowledged because the same case-inflecting paradigm
is also used pronominally and possessively; see
§\,\ref{sec:pronouns} for the fuller descriptive discussion and
§\,\ref{sec:np-structure} for the overt noun-phrase order.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Demonstrative paradigm relevant to noun-internal and
  pronominal syntax.}\label{tab:nm-dem-paradigm}
  \medskip
  \begin{tabularx}{0.7\textwidth}{L l l l}
    \toprule
                 & {\sc direct} & {\sc transitive} & {\sc indirect} \\
    \midrule
    Proximal     & \ird{tak}  & \ird{tam}  & \ird{ten} \\
    Medial       & \ird{tuto} & \ird{toho} & \ird{taly} \\
    Distal       & \ird{je}   & \ird{jaky} & \ird{jení} \\
    \bottomrule
  \end{tabularx}
\end{table}

\noindent Within the noun phrase, the adnominal demonstrative
realizes DemP, the outermost noun-internal demonstrative projection
above AnchP. The hierarchy required by the overt ordering facts is
therefore DemP $>$ AnchP $>$ KP $>$ PossP $>$ NP, matching the
descriptive order \textsc{demonstrative} $>$
\textsc{poss}.\Ind{} $>$ \textsc{head} given in
§\,\ref{sec:np-structure}. Demonstratives stand outside the
possessive grouping, whereas possessors remain inside the KP/NP
domain. The co-occurrence facts in §\,\ref{sec:pronouns}, including
\ird{tuto že tom}, show directly that demonstratives and possessors
cannot be analyzed as occupants of one and the same structural slot.

\begin{center}
\begin{forest}
  syntax tree
  [DemP
    [{\textit{adnominal demonstrative}}]
    [AnchP
      [KP
        [PossP
          [{\textit{oblique possessor}.\Ind{} / $\varnothing$}]
          [{NP\\\textit{head noun}}]
        ]
        [{K\textsuperscript{0}\\{\footnotesize case}}]
      ]
      [{Anch\textsuperscript{0}\\\ird{-to} / $\varnothing$}]
    ]
  ]
\end{forest}
\end{center}

The core semantics of the demonstrative layer is deictic: proximal
indexes a referent near the speaker, medial a referent near the
addressee, and distal a referent away from both. The
topic-proximal, addressee-linked medial, and obviative or
backgrounded distal uses described in §\,\ref{sec:pronouns} are
pragmatic extensions of those lexical deictic values, not
replacements for them. The formal analysis therefore keeps lexical
deixis and discourse tendency distinct.

Because DemP and AnchP are distinct layers, the relation between a
demonstrative and \ird{-to} is one of marked layering rather than
true exclusion. A plain demonstrative phrase is structurally
unmarked. A demonstrative plus anchoring suffix yields a marked
emphatic, resumptive, contrastive, or strongly individuating
interpretation.

\paragraph{Indirect-case demonstratives as possessors.}
The same demonstrative paradigm also feeds possessive syntax. The
indirect forms \ird{ten}, \ird{taly}, and \ird{jení} may occupy the
ordinary possessor slot before the noun, as in \ird{ten dum}. In
such forms the demonstrative is not the outer deictic determiner of
the noun phrase; it is an indirect-case possessor inside PossP. The
grammar therefore allows two distinct noun-internal positions fed by
the same demonstrative system: adnominal DemP above AnchP, and
demonstrative possessors inside PossP.

\begin{center}
\begin{forest}
  syntax tree
  [DemP
    [{\textit{adnominal demonstrative} / $\varnothing$}]
    [AnchP
      [KP
        [PossP
          [{\ird{ten}/\ird{taly}/\ird{jení}}]
          [{NP\\\textit{head noun}}]
        ]
        [{K\textsuperscript{0}\\{\footnotesize case}}]
      ]
      [{Anch\textsuperscript{0}\\\ird{-to} / $\varnothing$}]
    ]
  ]
\end{forest}
\end{center}

\subsection{Pronominal demonstratives and third-person reference}
\label{sec:nm-dem-pron}

Iridian lacks independent third-person personal pronouns. The full
demonstrative paradigm therefore serves as the ordinary third-person
referential series (§\,\ref{sec:pronouns}). This pronominal use is
not identical to noun-internal DemP, because no lexical head noun is
present inside the expression, but it is derived from the same
underlying demonstrative system and must be kept distinct from the
noun-internal ordering facts summarized in §\,\ref{sec:np-structure}.

The descriptive discourse tendencies can be compressed as follows.
Proximal forms commonly track the current topic or most accessible
referent, medial forms a secondary or addressee-linked referent, and
distal forms a backgrounded, contrastive, or obviative referent. In
formal terms, these effects are best treated as interpretive
extensions available to pronominal demonstratives, not as separate
lexical paradigms. The fuller descriptive account remains in
§\,\ref{sec:pronouns}; the present chapter cites it only insofar as
it bears on the referential force of the demonstrative system.

\ex
\begingl
\gla Tak vrbnik, tuto spasnik, je těsnik.//
\glb \Prox{}.\Dir{} stand.up-\Pf{}, \Med{}.\Dir{} sit.down-\Pf{}, \Dist{}.\Dir{} leave-\Pf{}//
\glft \trsl{He stood up, he sat down, that one left.}//
\endgl
\xe

\subsection{The reduced anaphoric series}
\label{sec:nm-dem-reduced}

Alongside the full paradigm, Iridian has a reduced anaphoric series
\ird{i}/\ird{ji}/\ird{ni}, corresponding to distal
\ird{je}/\ird{jaky}/\ird{jení}. These forms are restricted to
pronominal use, only when the referent is already highly accessible
in the discourse, and they are excluded as ordinary adnominal
demonstratives of the type described in §\,\ref{sec:np-structure}.

Semantically and pragmatically, the reduced series signals weak
anaphoric resumption rather than full deictic pointing. Formally, it
belongs to the demonstrative system, but it is not the spellout of
noun-internal DemP. A natural DM treatment is to analyze
\ird{i}/\ird{ji}/\ird{ni} as specialized pronominal Vocabulary Items
competing with the full distal series under a discourse-anaphoric
high-accessibility condition; see again §\,\ref{sec:pronouns}.

\pex
\a \ird{je} \trsl{that one; fully deictic distal}
\a \ird{i} \trsl{he/she; weak anaphoric resumption}
\xe


\subsection{The plural marker \ird{ne} and the number layer}
\label{sec:nm-num}

The surface distribution of \ird{ne} shows that it appears at the left
edge of the constituent within its scope:
\ird{ne byl} \trsl{children},
\ird{ne byle} \trsl{the children},
\ird{ně Marcí tóm} \trsl{the students' book}, and
\ird{Marcí ne tóm} \trsl{the student's books}.

This distribution is best understood by keeping apart semantic scope
and morphophonological placement. Semantically, \ird{ne} realizes the
number layer. Morphophonologically, it is a left-edge plural clitic:
it attaches to the leftmost element inside the constituent whose
plurality it marks. Its position at the left edge of the noun phrase
therefore does not by itself show that it is an ordinary overt head of
a head-final projection.

The same point explains its compatibility with anchoring. Plurality
and anchoring are independent:
\ird{ne byl} and \ird{ne byle} differ in anchoring, not in number.
The anchored form \ird{byle} carries the reduced front allomorph
characteristic of the Class II direct stem.
The simplest analysis is therefore a NumP semantics with left-edge
spellout of the plural marker.

\pex
\a
\begingl
\gla ne byl//
\glb \Pl{} child.\Dir{}//
\glft \trsl{children}//
\endgl
\a
\begingl
\gla ne byle//
\glb \Pl{} child.\Dir{}.\Def{}//
\glft \trsl{the children}//
\endgl
\a
\begingl
\gla ne si byl//
\glb \Pl{} \Spec{} child.\Dir{}//
\glft \trsl{some particular children}//
\endgl
\xe

When \ird{ne} combines with a numeral or other quantifier, the usual
interpretation is approximate rather than exactly plural
(\ird{ne hrona byl} \trsl{about three children}). It is safer to treat
that approximation as an interpretive effect of the construction than
to posit a separate feature at this stage. In those approximate uses,
anchoring is excluded, and specifically indefinite marking with
\ird{si}/\ird{druh} is also ordinarily avoided.

\paragraph{Mass, count, and lexical restrictions.}
The plural clitic is most natural with count nouns. With mass nouns it
is ordinarily excluded unless a covert measure noun is understood, as
in the familiar type \ird{ne nebo} \trsl{glasses of water}. A small
set of pluralia-tantum nouns, especially names of paired body parts and
paired objects, also resist ordinary pluralization because the bare
form already denotes the pair.

\paragraph{Existential and honorific uses.}
In existential clauses the plural clitic appears before the existential
predicate rather than on the noun, yielding \ird{Ne ješ bžem}
\trsl{There are bees}. The same phonological item also has a distinct
honorific use with personal names and kin terms
(\ird{ne Stám Kovárž}, \ird{ne mlazka}), which does not instantiate
NumP at all and is therefore best treated as a lexicalized discourse
clitic homophonous with the plural marker. In addition, \ird{ne} may
occur in \emph{wh}-questions when the speaker expects a plural answer,
as in \ird{ne jede bych ščenžkou?} \trsl{Who are those who arrived
yesterday?}

\subsection{Quantifiers, numerals, and the prenominal quantificational
  field}
\label{sec:nm-quant}

Iridian places numerals and quantifiers prenominally. They belong to
the upper nominal field, but the presently available evidence does not
force a fine-grained cartographic split among cardinals, adnominal
interrogatives, and universal or negative quantifiers. I therefore
treat them conservatively as occupants of a single quantificational
modifier field between the demonstrative/specificity domain and the
lower NP-internal material.

Numerals interact with number in a characteristic way. Because nouns
are not obligatorily marked for singular or plural, a bare numeral plus
noun already yields an exact cardinal interpretation. Adding \ird{ne}
to the same phrase shifts the reading from exact cardinality to
approximation. This supports the separation of QP from NumP: the
numeral itself provides quantification, while \ird{ne} contributes a
separate number-related discourse effect.

\pex
\a \ird{hrona byl} \trsl{three children}
\a \ird{ne hrona byl} \trsl{about three children}
\a \ird{ona pive štava} \trsl{one mug of beer}
\xe

Adnominal interrogatives pattern with the same prenominal field.
\ird{Jak} selects from a contextually restricted set, whereas the
other interrogatives simply introduce the relevant open variable. In
determiner use they modify the noun phrase directly rather than
occupying the clause-level \emph{wh}-position themselves.

\pex
\a \ird{jede ruščevní svirkošc} \trsl{what Russian author}
\a \ird{jak ruščevní svirkošc} \trsl{which Russian author}
\xe

Universal and negative quantifiers can likewise be adnominal:
\ird{nět študent} \trsl{all students}, \ird{niže jec}
\trsl{all dogs}, \ird{nách študent} \trsl{each student},
\ird{žé tóm} \trsl{not one book}. Under the current grammar's
negative-existential analysis, forms such as \ird{niho} are retained as
nominal quantifiers only when they modify a noun phrase directly; they
are not analyzed as the sentential negative existential head.


\subsection{The specifically indefinite particles \ird{si} and
  \ird{druh}}
\label{sec:nm-spec}

The surface distribution is clear: \ird{si} and \ird{druh} are
preposed nominal particles that introduce a referent as particular,
but not yet anchored in the common ground. They therefore mark
\term{specific indefiniteness}.

The interpretive contrast with \ird{-to} is equally clear.
\ird{-to} presents a noun phrase as anchored; \ird{si}/\ird{druh}
present it as specific but unanchored. This resembles the familiar
contrast between specificity and definiteness discussed in the
literature, including Enç (enc:1991), but the present
analysis borrows only that descriptive distinction, not Enç's full
feature geometry.

The theoretical analysis adopted here is a SpecificityP within the
nominal left periphery. I avoid the label \emph{SpecP}, since
\emph{Spec} is already reserved for \emph{specifier}. Both particles
occupy the same structural position. Their incompatibility with
\ird{-to} is best treated as a selectional restriction:
SpecificityP selects an unanchored complement. Put informally,
\ird{si}/\ird{druh} require a noun phrase that has not already been
anchored.

The difference between the two particles is lexical-pragmatic rather
than syntactic. \ird{Si} is the lighter and less marked form, common
in ordinary discourse. \ird{Druh} is weightier, more selective, and
often more formal or emphatic. No additional syntactic machinery is
required to capture that contrast.

\pex
\a
\begingl
\gla si byl//
\glb \Spec{} child.\Dir{}//
\glft \trsl{a certain child; some particular child}//
\endgl
\a
\begingl
\gla druh byl//
\glb \Spec{} child.\Dir{}//
\glft \trsl{one given child; a certain particular child}//
\endgl
\xe

Their left-edge position follows directly from the analysis adopted
here. They are not ordinary head-final heads that would need a
separate PF-placement rule. They are left-peripheral particles in
SpecificityP, and they therefore appear where they are described as
appearing in the prose and where they are shown in the trees.


%% ================================================================
%%  PART V: THE PRONOMINAL DOMAIN
%% ================================================================

\section{Pronouns, anaphors, and nominal operators}
\label{sec:nm-pron}

\subsection{Personal pronouns as full DPs}
\label{sec:nm-pron-personal}

First- and second-person pronouns are best treated as fully projected
DPs whose nominal content is supplied by person features rather than by
a lexical noun root. They inflect for the same three productive cases
as lexical nouns; their indirect forms serve as oblique possessors in
the analytic possessive construction described in §\,\ref{sec:nm-poss}.
Iridian lacks a separate third-person personal-pronoun paradigm; that
function is taken over by the demonstrative system described in
§§\,\ref{sec:nm-dem}--\ref{sec:nm-dem-reduced}.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Personal pronouns.}
  \label{tab:nm-personal-pronouns}
  \medskip
  \begin{tabularx}{0.65\textwidth}{L l l l}
    \toprule
                  & {\sc direct} & {\sc transitive} & {\sc indirect} \\
    \midrule
    1\textsc{sg}  & \ird{dá}   & \ird{dam}  & \ird{že}   \\
    2\textsc{sg}  & \ird{já}   & \ird{jí}   & \ird{jena} \\
    1\textsc{pl}  & \ird{mé}   & \ird{mně}  & \ird{mi}   \\
    2\textsc{pl}  & \ird{tová} & \ird{tvy}  & \ird{tě}   \\
    \bottomrule
  \end{tabularx}
\end{table}

The indirect case forms serve as the possessors in analytic possessive
expressions: possession is structurally encoded through the indirect
domain, and the possessor occupies PossP as a prenominal indirect-case
DP (§\,\ref{sec:nm-poss}).

\pex
\a \ird{že dum} \trsl{my house}
\a \ird{mi dum} \trsl{our house}
\xe

\subsection{The reflexive as local nominal anaphor}
\label{sec:nm-pron-refl}

The reflexive is not a discourse-tracking pronoun but a true anaphor.
Its interpretation depends on a local antecedent in the clausal domain,
and it competes with demonstrative third-person forms precisely by
forcing coreference where demonstratives would permit disjoint
reference.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Reflexive anaphor forms.}
  \label{tab:nm-reflexive}
  \medskip
  \begin{tabularx}{0.5\textwidth}{L l}
    \toprule
    {\sc case} & {\sc form} \\
    \midrule
    Direct     & \ird{svem} \\
    Transitive & \ird{svě}  \\
    Indirect   & \ird{svy}  \\
    \bottomrule
  \end{tabularx}
\end{table}

\pex
\a
\begingl
\gla Svem Marku vednik.//
\glb \Refl{}.\Dir{} Marek.\Trs{} see-\Pf{}//
\glft \trsl{Marek saw himself.}//
\endgl
\a
\begingl
\gla I Marku vednik.//
\glb 3.\Dist{}.\Dir{} Marek.\Trs{} see-\Pf{}//
\glft \trsl{Marek saw him.}//
\endgl
\xe

The older reflexive-as-\trsl{own} use remains compatible with the
current system if analyzed as emphatic spellout in the local anaphoric
domain rather than as a separate possessive adjective. In other words,
the intensifying reading is an interpretive extension of local
coreference, not a distinct lexical category.

\subsection{Interrogatives, indefinites, and negative nominal operators}
\label{sec:nm-pron-operators}

Interrogatives and their derived indefinite or negative counterparts
behave as nominal operators. They can stand pronominally or modify a
noun phrase adnominally. In the latter use they pattern with the
quantificational field discussed in §\,\ref{sec:nm-quant}; in the
former they project as DPs without an overt lexical noun.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Core interrogative nominal operators.}
  \label{tab:nm-interrogatives}
  \medskip
  \begin{tabularx}{0.8\textwidth}{L L L L}
    \toprule
    \ird{jede} & who & \ird{jak} & which \\
    \ird{ježe} & what & \ird{zajehu} & why \\
    \ird{jehát} & whom & \ird{jiká} & how many \\
    \ird{jehu} & how & \ird{jišká} & how much \\
    \ird{jemí} & when & \ird{jeně} & to where \\
    \ird{jena} & where & \ird{jení} & from where \\
    \bottomrule
  \end{tabularx}
\end{table}

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Selected universal and negative nominal operators.}
  \label{tab:nm-indefinites}
  \medskip
  \begin{tabularx}{0.8\textwidth}{L L L L}
    \toprule
    {\sc negative form} & {\sc meaning} & {\sc universal form} &
    {\sc meaning} \\
    \midrule
    \ird{zide} & no one & \ird{nět} & everyone \\
    \ird{niho} & nothing & \ird{niže} & everything \\
    \ird{žena} & nowhere & \ird{nina} & everywhere \\
    \ird{žemě} & never & \ird{nimě} & always \\
    \ird{žé} & not one & \ird{nách} & each \\
    \bottomrule
  \end{tabularx}
\end{table}

In pronominal use, these items are full operators over the discourse
model. In adnominal use, they narrow or negate the denotation of the
following noun phrase. The older archive's existential uses of forms
like \ird{niho} are not imported as part of the nominal system here,
because sentential negative existence is now handled elsewhere by the
suppletive existential predicate.

An important structural clarification concerns the division of labour between
these nominal operators and the clause-level interrogative system.
Wh-hood --- the property of introducing an open variable --- is a lexical or
DP-internal feature of these forms: they are interrogative operators that may
occupy argument or adjunct positions inside the clause. Questionhood proper,
by contrast, is clausal and is located in \textsc{PolP}: the clause is typed
as a question by the Pol\textsuperscript{0} head (overt \ird{by} or null), not
by the presence of a wh-operator in Spec,FocP (see §\,\ref{sec:by-pol-head}
and §\,\ref{sec:th-wh-interrogation} in Chapter~\ref{ch:simple}). A
wh-expression may remain in its base argument or adjunct position (in situ) or
move to Spec,FocP for marked focus; neither option alone determines whether the
clause is interrogative. The same interrogative nominals listed in
Table~\ref{tab:nm-interrogatives} appear both in matrix content questions and
in embedded non-assertive \ird{poby}-complements (§\,\ref{sec:poby} of
Chapter~\ref{ch:complex}), where they function as wh-operators inside
an issue-denoting embedded clause rather than as direct-question operators.

The relationship between these nominal negative operators and the
clausal NegP (§\,\ref{sec:negation} and §\,\ref{sec:clausal-negation}
of Chapter~\ref{ch:simple}) is not yet fully described and should be
treated as an open area. Several questions bear on the interaction.
First, does a clause containing \ird{zide} or \ird{niho} project NegP
independently --- yielding \ird{zá-} on the finite verb in addition to
the nominal operator --- or do these items themselves satisfy the
[+\textsc{neg}] requirement of the clause, rendering \ird{zá-} redundant
or ungrammatical? Second, do the negative nominals license negative
polarity items that the indicative \ird{zá-} also licenses, suggesting
they participate in the same NegP projection? Third, in existential
clauses, the suppletive root \ird{zad-} already expresses the
non-existence of a referent; it is unclear whether \ird{niho} can
co-occur with \ird{zad-} (double-marking of the same semantic
content) or whether one blocks the other. Pending systematic elicitation
on these points, the interaction of nominal negative operators with
clausal NegP should not be assumed to follow the pattern of either
standard negative concord or strict complementary distribution.

\pex
\a \ird{jede bych ščenžkou?} \trsl{who arrived yesterday?}
\a \ird{jak na ne ruščevní svirkošta} \trsl{which of these Russian authors}
\a \ird{nět študent} \trsl{all students}
\a \ird{zide byl} \trsl{no child}
\xe


%% ================================================================
%%  PART VI: THE POSSESSOR AND THE INDIRECT CASE IN DP
%% ================================================================

\section{Possessors, the indirect case, and the PossP layer}
\label{sec:nm-poss}

\subsection{The oblique possessor in the DP}
\label{sec:nm-poss-struct}

Iridian expresses possession by means of an oblique (indirect-case)
noun phrase in prenominal position (§\,\ref{sec:indirect-uses}).
The oblique possessor \ird{Marcí} in \ird{Marcí dum} \trsl{Marek's
house} occupies a structural position immediately above the head
NP and inside the KP domain, below AnchP and below noun-internal
DemP. In terms of the DP hierarchy, this is the
prenominal dependent position of PossP.

The indirect case of the possessor is not a special possessive case
but the language's general non-core case, deployed here as it is
throughout the grammar for dependents that are not the main
absolutive argument. There is no separate structural Case assignment
operation for possessors: the possessor's indirect case arises from
the same K head that assigns case to other indirect-case NPs in the
clause. The structural uniformity of the indirect case across nominal
and clausal environments is a design feature of the grammar, not an
accident.

\begin{center}
\begin{forest}
  syntax tree
  [AnchP
    [KP
      [PossP
        [{\textit{oblique possessor}.\Ind{}}]
        [{NP\\\textit{head noun}}]
      ]
      [{K\textsuperscript{0}\\{\footnotesize case ending}}]
    ]
    [{Anch\textsuperscript{0}\\\ird{-to} / $\varnothing$}]
  ]
\end{forest}
\end{center}

\noindent The possessor appears before the head noun in the surface
string, consistent with head-final base-generation of specifiers.
The entire possessive phrase [PossP oblique-N head-N] forms a
constituent that is then inflected as a whole for case by K and
optionally for anchoring by \textsc{Anch\textsuperscript{0}}.

\subsection{Distribution of \ird{ne} across the possessive DP}
\label{sec:nm-poss-ne}

The plural clitic \ird{ne} has two attachment sites within a
possessive DP, yielding different scope interpretations:

\pex
\a \ird{ně Marcí tóm} --- \trsl{the students' book} (students
plural; book possibly singular)
\a \ird{Marcí ne tóm} --- \trsl{the student's books} (book plural;
student possibly singular)
\xe

\noindent In the first case, \ird{ne} is proclitic to the NumP
headed by the possessive phrase: plurality scopes over the whole
possessive expression. In the second, \ird{ne} is proclitic to
the head noun directly, within the PossP complement: plurality
scopes over the head alone. The structural analysis predicts these
readings by the standard scope-by-hierarchy principle: the NumP
that \ird{ne} heads takes scope over whatever it dominates.



%% ================================================================
%%  PART VII: NOUN PHRASE STRUCTURE IN MINIMALIST TERMS
%% ================================================================

\section{The head-final DP: word order and phrase structure}
\label{sec:nm-np-structure}

\subsection{The functional sequence and linear order}
\label{sec:nm-funcseq}

The canonical linear order of the Iridian noun phrase —

\begin{center}
\ird{ne} $>$ \ird{si}/\ird{druh} $>$ \textsc{dem} $>$ \textsc{quant/num} $>$
\textsc{poss}.\Ind{} $>$ \textsc{prop.root / stative / -ovní} $>$
\textsc{head} + \textsc{case} + \ird{-to}
\end{center}

— is derived from the functional hierarchy in
Table~\ref{tab:nm-projections} by the head-final principle. Each
overt exponent of a higher nominal projection appears before the noun
phrase domain with which it is associated, while suffixal morphology
continues to follow the head-final principle. The rightward drift of
functional morphology (case suffix, then anchoring suffix)
corresponds to the innerward drift of those same heads in the
syntactic projection: the case head K\textsuperscript{0} is lower than
\textsc{Anch\textsuperscript{0}}, but its suffix appears closer to the
root because the root is the complement that K\textsuperscript{0}
takes, and in a head-final system the complement precedes the head.

Reading the noun phrase from left to right is therefore equivalent
to reading the functional projection from outermost (NumP) to
innermost (NP), mirroring the left-to-right = high-to-low mapping
in the clause (§\,\ref{sec:th-head-finality}).

This order should be understood conservatively. The robust general
facts are that number, specificity, demonstratives, and
quantificational material belong to the prenominal left field;
possessors stand closer to the noun than demonstratives do; and
property roots, stative modifiers, and \ird{-ovní} phrases stand
immediately before the head. The current data are not yet rich enough
to force a finer ordering among every member of the prenominal field,
so the chapter does not claim more cartographic precision than the
evidence warrants.

\pex
\a \ird{tak študencí tóm} \trsl{this student's book}
\a \ird{jak na ne ruščevní svirkošta} \trsl{which of these Russian authors}
\xe

\subsection{Stative verbs and property roots as NP modifiers}
\label{sec:nm-prop}

Iridian's primary attributive modification strategy deploys stative
verbs in the dedicated attributive form in \ird{-á} directly before the
head noun. The small set of lexical property roots that feed those
stative verbs are analyzed as heads of a PropP projection that
immediately precedes the NP and is itself immediately preceded by the
PossP layer.

The head-final principle correctly predicts that all prenominal
modifiers — whether stative-verb-based or property-root-based —
appear before the head noun, never after it. Postnominal modification
is not a productive construction in standard Iridian; the prenominal
restriction on modifiers is a direct consequence of the head-final
parameter.

\pex
\a \ird{mrocná nuž} \trsl{a beautiful city}
\a \ird{hradná byl} \trsl{a good child}
\a \ird{svorná dum} \trsl{a big house}
\xe

\noindent This \ird{-á} form is restricted to adnominal attributive use within
the DP. Clause-level manner modification is expressed instead by the derived
adverbs in \ird{-ě} discussed in §\,\ref{sec:stative-adv} of
Chapter~\ref{ch:verbs}; the attributive itself is not extended to ordinary
adverbial use.

\subsection{The \ird{-ovní} attributive and its embedded DP}
\label{sec:nm-ovni}

The attributive suffix \ird{-ovní} derives from the same verbal stem
as the participant nominal \ird{-ou} and functions as a prenominal
modifier (§\,\ref{sec:event-nominals}). The \ird{-ovní} form heads a
nominalized predication that is embedded as a modifier within the
host DP.

The key structural property of the \ird{-ovní} construction is that
it forms a self-contained syntactic domain: non-privileged arguments
of the embedded verbal predication appear in the indirect case
within the \ird{-ovní} phrase, regardless of the case requirements
of the host DP. This is the nominal correlate of the boundary
condition that governs finite embedded clauses: just as a finite
\ird{po}-clause is syntactically independent of its matrix, the
\ird{-ovní} nominalized phrase is syntactically independent of the
host DP in which it is embedded.

The head noun of the host DP receives its case from the host clause;
the \ird{-ovní} modifier preceding it is not inflected for the case
of the host but maintains its own internal case pattern. In nested
\ird{-ovní} constructions, the inner head noun shifts to the
indirect case of the outer embedding domain, as described in
§\,\ref{sec:event-nominals}. This shifting is the nominal counterpart
of the case valuation that occurs at clausal boundaries in the verbal
domain.

The present chapter adopts the reduced-relative analysis explicitly.
The \ird{-ovní} form heads a nominalized predication inside a RelP
adjoined low in the nominal structure, most naturally to NP or to the
NP-side edge of PossP. A null operator in Spec,RelP is interpreted as
the privileged participant identified by the head noun. This ties the
construction directly to the broader relativization system discussed in
§\,\ref{sec:th-relativization}, while preserving the head-final fact
that the modifying material precedes the noun it restricts.

\pex
\a \ird{vravět obrohnikovní pres} \trsl{the man who read the letter}
\a \ird{[nazam zasdnikovní vravět] obrohnikovní pres}
   \trsl{the man who read the letter that the woman wrote}
\xe

\subsection{The \ird{ty}-supplement}
\label{sec:nm-appos}

Iridian possesses an appositional particle \ird{ty} that introduces a
right-peripheral supplement: a phrase that is coreferent with a preceding
nominal and presents it as a clarification, identification, or elaboration.
The semantics are those of identity---\textit{that is}, \textit{namely}---and
the \ird{ty}-phrase is always postnominal, set off from its antecedent by a
prosodic break.

The \ird{ty}-phrase is not a restrictive modifier. It does not narrow the
denotation of the head noun the way an adjective or relative clause does: a
phrase like \ird{Karel ty mlazka} does not pick out a subset of Karels who
are brothers, it simply adds the information that the referent already
identified as Karel is the speaker's brother. The \ird{ty}-phrase is therefore
best analysed as a \term{supplement}---a nominally-anchored, non-restrictive
phrase that is coreferent with its antecedent but structurally peripheral to
it, not integrated into the NP's extended projection. This distinguishes
\ird{ty}-phrases categorically from adjectives and relative modifiers, which
are adjuncts within the NP and compositionally restrict the head. Because
\ird{ty}-phrases are supplements rather than adjuncts, they are not governed
by the head-final principle: modifiers fall within the NP and must precede
their head; supplements are peripheral and their position---always to the
right of the antecedent---is determined by information structure and discourse
organisation.

The antecedent nominal is introduced first; \ird{ty} then opens the supplement
phrase. When the supplement consists of a bare nominal, no further element is
required: the nominal is itself the head and the identity relation is immediate.

\pex
\a \ird{Karel ty mlazka}
   \trsl{Karel---that is, my brother}
\xe

When the supplement consists of non-nominal material---an adjective,
a \ird{-ovní}-relative, a postpositional phrase---a pronominal head is required
to host it. Modifiers in Iridian must precede a nominal head; without one, the
modifier cannot form a well-formed phrase. This head is supplied by the
anaphoric pronoun \ird{i}, which appears at the right edge of the supplement
and corefers with the antecedent nominal.

\pex
\a \ird{nuž ty mrocná i}
   \trsl{the city---that is, the beautiful one}
\a \ird{pres ty že vednikovní i}
   \trsl{the man---that is, the one I saw}
\xe

\noindent The internal structure of the \ird{ty}-phrase respects head-finality
throughout: \ird{mrocná} and \ird{že vednikovní} precede their head \ird{i} in
the ordinary left-to-right modifier-before-head order. Head-finality is not
suspended in the supplement; it applies within it. The distribution of \ird{i}
follows directly: it is required whenever the \ird{ty}-phrase contains a
non-nominal modifier that needs a nominal head to attach to, and absent when
the \ird{ty}-phrase already contains a nominal head.

Ordinary attributive modification---adjectives, \ird{-ovní}-relatives, and
similar prenominal material---remains the unmarked strategy for adnominal
description and carries no special discourse implication. The \ird{ty}-supplement
is marked for deliberate elaborative intent: it presents the clarification as a
distinct informational act, separate from the introduction of the referent.
In spoken language this implicature is weak, since production pressures
naturally favour establishing the referent before specifying it; in formal and
written language the implicature is stronger, and a \ird{ty}-phrase reads as
emphatic, contrastive, or rhetorically foregrounded. The construction is not
excluded from formal writing but is marked there, and is most natural for heavy
adnominal material---long nominalised relative clauses, for instance---that
would be unwieldy in prenominal position.

The full discourse-interactional properties of the \ird{ty}-supplement,
including its use as a repair and afterthought device in spoken interaction, are
treated in Chapter~\ref{ch:spoken} (\S\,\ref{sec:spoken-ty}).


%% ================================================================
%%  PART VIII: NOMINAL DEPENDENCE BEYOND POSSESSORS
%% ================================================================

\section{Indirect-case nominal dependence and postpositional phrases}
\label{sec:nm-ext}

\subsection{Material, partitive, source, and other NP-external
  dependencies}
\label{sec:nm-ext-indirect}

The indirect case is the general dependency form of the nominal
system. Possessors are simply its most tightly integrated DP-internal
use. The same K-form appears with material nouns, partitives, source
expressions, and other semantically dependent nominals that are not the
privileged head of the phrase. In this wider domain the indirect does
not mark a special inventory of separate morphological cases; rather, it
is the single oblique realization of K whose interpretation is narrowed
by the lexical relation between nouns or by the head that selects the
dependent nominal.

\pex
\a \ird{kuní prosc} \trsl{silver spoon}
\a \ird{ona pive štava} \trsl{one mug of beer}
\a \ird{Študencí na hroná} \trsl{three of the students}
\a \ird{Marcí z houba} \trsl{a gift from Marek}
\xe

Material, partitive, and source dependents therefore stand on a
continuum with possessors. The possessor is the most tightly integrated
prenominal use of indirect KP inside the DP. Material and partitive
dependents are looser nominal relations. Postpositional phrases are the
next step outward: the same indirect nominal domain serves as the
complement of a relational head that closes the phrase from the right.

\subsection{Postpositions as heads selecting an indirect nominal
  complement}
\label{sec:nm-postpositions}

Iridian adpositions are head-final and are best analyzed as
\term{postpositions}. The constant generalization is that the nominal
complement bears indirect case. In the smallest representation, the
postposition combines with an indirect KP. In fuller noun phrases, that
KP may itself be embedded inside AnchP or higher nominal layers, but the
selectional requirement remains the same: the nominal core is oblique,
and the postposition closes the complement phrase at its right edge.

This analysis replaces the older archive treatment that described the
same forms as prepositions or distributed them across several narrower
case labels. Synchronically the simpler statement is the better one: the
nominal complement is inflected in the indirect, and the following
postposition specifies the spatial, participant, source, privative, or
substitutive relation.

\pex
\a \ird{dumě na} \trsl{in the house; at home}
\a \ird{hradě de} \trsl{into the town; to the town}
\a \ird{Jancí še} \trsl{with Janek}
\a \ird{Marcí z} \trsl{from Marek}
\xe

\subsection{Internal structure of the postpositional phrase}
\label{sec:nm-postp-structure}

The structural asymmetry is straightforward. The nominal complement is
assembled first: the root combines with PossP, receives indirect case in
K, and may then be expanded by anchoring or other higher nominal
material. Only after that nominal phrase is complete does the
postposition merge as the head of the larger postpositional phrase.
Head-finality then yields the surface order \textit{complement} +
\textit{postposition}.

\begin{center}
\begin{forest}
  syntax tree
  [PostP
    [AnchP
      [KP
        [PossP
          [{\textit{oblique possessor} / $\varnothing$}]
          [{NP\\\textit{head noun}}]
        ]
        [{K\textsuperscript{0}\\\Ind{}}]
      ]
      [{Anch\textsuperscript{0}\\\ird{-to} / $\varnothing$}]
    ]
    [{Post\textsuperscript{0}\\\ird{na} / \ird{de} / \ird{še} / \dots}]
  ]
\end{forest}
\end{center}

\noindent The tree is intentionally conservative. What matters is not
whether every postpositional complement is always projected up to AnchP,
but that the complement is at least an indirect KP and that any higher
nominal material remains inside the complement phrase rather than
intervening between the complement and the postposition.

\subsection{Attachment to the right edge of the host phrase}
\label{sec:nm-postp-attachment}

The postposition attaches to the right edge of the governed phrase as a
whole, not to the noun stem in isolation. A bare noun, an anchored noun,
a possessed noun phrase, and a noun phrase with prenominal modifiers all
behave alike in this respect: the nominal material comes first, and the
postposition closes the entire phrase.

\pex
\a \ird{dumě na} \trsl{in the house; at home}
\a \ird{bylamo my} \trsl{for the child}
\a \ird{Marcí dumě na} \trsl{at Marek's house}
\a \ird{svorná mostě nad} \trsl{over the big bridge}
\xe

\noindent The same logic extends to larger complements. Demonstratives,
possessors, property-root modifiers, stative modifiers, and
\ird{-ovní} phrases all belong to the nominal complement if they modify
the noun phrase itself. The postposition does not split them apart or
attach to the lexical noun before they are assembled. This is the chief
diagnostic for analyzing the forms as postpositions over phrasal
complements rather than as suffixes on nouns or as free adverbs loosely
associated with them.

\subsection{Semantic classes of postpositions}
\label{sec:nm-postp-classes}

The common postpositions form a compact but semantically varied system.
Their lexical meanings differ, but their syntax is uniform.

\begin{table}[ht]
  \footnotesize\sffamily
  \caption{Core postpositions and their main semantic classes.}
  \label{tab:nm-postpositions}
  \medskip
  \begin{tabularx}{\textwidth}{l l L}
    \toprule
    {\sc class} & {\sc members} & {\sc core relation} \\
    \midrule
    Locative & \ird{na} & location, interiority, surface contact \\
    Goal & \ird{de} & direction, endpoint \\
    Proximity & \ird{u} & nearness \\
    Accompaniment & \ird{še} & accompaniment, comitative association \\
    Benefactive & \ird{my} & benefit, intended recipient, purpose-oriented relation \\
    Source / separation & \ird{z} & source, origin, movement away, extraction \\
    Privative & \ird{nam} & absence \\
    Replacement & \ird{pale} & substitution, `instead' \\
    Exception & \ird{ahte} & exclusion, `except' \\
    \bottomrule
  \end{tabularx}
\end{table}

\noindent The core spatial and participant relations are illustrated
below:

\pex
\a \ird{dumě na} \trsl{in the house}
\a \ird{hradě de} \trsl{to the town}
\a \ird{hradě u} \trsl{near the town}
\a \ird{bylam my} \trsl{for the child}
\a \ird{Jancí še} \trsl{with Janek}
\a \ird{Marcí z} \trsl{from Marek}
\a \ird{záhárí nam} \trsl{without sugar}
\a \ird{Jancí pale} \trsl{instead of Janek}
\a \ird{Jancí ahte} \trsl{except for Janek}
\xe

\subsection{Specialized spatial and path postpositions}
\label{sec:nm-postp-extended}

Alongside the broad locative and goal markers, Iridian also admits a
set of more specialized spatial and path postpositions. These behave
like the core set in selecting the same indirect nominal complement and
in appearing at the right edge of the phrase.

\pex
\a \ird{stolě pod} \trsl{under the table}
\a \ird{mostě nad} \trsl{over the bridge}
\a \ird{domě měz} \trsl{between the houses; among the houses}
\a \ird{hradě do} \trsl{up to the town}
\a \ird{lesě mně} \trsl{through the forest; via the forest}
\xe

\noindent These extended forms matter analytically because they show
that the postpositional construction is not a tiny lexical quirk tied
only to \ird{na}. The same head-final syntax generalizes across broad
location, bounded path, vertical relation, interstitial relation, and
traversal.

\subsection{Postpositional phrases in clause structure}
\label{sec:nm-postp-clausal}

The postpositional phrase is not confined to nominal modification. The
same structure appears as a clausal adjunct, as the location complement
of existential predication, and as a dependent nominal expression inside
larger noun phrases. The PP discussion therefore belongs in the nominal
chapter because of its internal structure, but it interfaces directly
with clause structure as well.

\pex
\a
\begingl
\gla Marek dumě na jeca.//
\glb Marek.\Dir{} house.\Ind{} in \Exst{}-\Cont{}//
\glft \trsl{Marek is at home.}//
\endgl
\a
\begingl
\gla Kroumaště na po zma jeca pivo.//
\glb refrigerator.\Ind{} in still few \Exst{}-\Cont{} beer.\Dir{}//
\glft \trsl{There is still some beer left in the refrigerator.}//
\endgl
\a
\begingl
\gla Bylam my Jancí še stóžách.//
\glb child-\Ind{} for Janek-\Ind{} with go-\Ctp{}//
\glft \trsl{(I am) going for the child with Janek.}//
\endgl
\a
\begingl
\gla Marcí z houba.//
\glb Marek-\Ind{} from gift.\Dir{}//
\glft \trsl{a gift from Marek}//
\endgl
\xe

\noindent The existential examples connect directly to
§\,\ref{sec:jec} in Chapter~\ref{ch:simple}: the entity whose
existence is asserted stands in the direct case, while the locative PP
contributes the place relation. More broadly, postpositional phrases
behave like ordinary head-final dependent phrases in the clause. They
precede the final predicate and preserve their internal noun-first,
head-last organization.

\subsection{Analytical status: postpositions, clitics, and orthographic
  practice}
\label{sec:nm-postp-status}

Phonologically, the common postpositions are weak. They lean prosodically
on the preceding complement and show the reduced independence typical of
clitic-like material. For that reason it is reasonable to describe them
as \term{postposed clitic adpositions}. Orthographically, however, this
grammar writes them as separate words for ease of reading and because
their scope is phrasal rather than stem-bound.

The important analytical point is that phonological weakness does not
force an affixal analysis. The host is not a noun stem but a fully
formed nominal phrase with indirect case at its core. Conversely, the
head-final position and obligatory adjacency to that nominal phrase
argue against treating the forms as free clause-level adverbs. The
construction is therefore most transparently analyzed as a
postpositional phrase whose head is phonologically weak but
syntactically regular.

Older descriptions sometimes treated some of these relations as
prepositions or distributed them across several special cases. The
present grammar instead unifies them under a single analysis: one
indirect nominal domain, followed by a head-final relational marker
whose lexical meaning specifies the precise relation.


%% ================================================================
%%  PART IX: DERIVED NOMINALS AND NOMINALIZED PREDICATION
%% ================================================================

\section{Derived nominals, gerunds, and participant reference}
\label{sec:nm-derived}

\subsection{Event nominals and the \ird{-ou} class}
\label{sec:nm-derived-event}

The \ird{-ou} class occupies a central place in the nominal system
because it is where verbal structure most transparently enters the DP.
Event nominals and participant nominals are fully nominal in their
external syntax: they inflect for case, host anchoring, admit plural
marking, and enter the ordinary possessive and demonstrative syntax.
Internally, however, they preserve argument structure inherited from
the verbal stem.

\pex
\a \ird{piaščikou} \trsl{the one who ate}
\a \ird{piaščikovat} \trsl{the one who ate, anchored / identified}
\a \ird{piaščikovět} \trsl{of the one who ate / to the one who ate}
\xe

\subsection{Simple and complex gerunds}
\label{sec:nm-derived-gerund}

Gerunds in \ird{-ou} may be simple, denoting an event without overt
participants, or complex, hosting the participants of that event inside
the noun phrase. In complex gerunds both the actor and the theme are
typically expressed in the indirect case. This follows from the general
Iridian strategy of encoding non-privileged nominal dependents through
the indirect domain.

\pex
\a \ird{Jancí podohletou} \trsl{Janek's remembering / the remembering of Janek}
\a \ird{Marcí travěte piaštou} \trsl{Marek's eating of the bread}
\xe

The resulting ambiguity is structural rather than accidental: two
indirect KPs inside one nominalized domain are not morphologically
distinguished. Speakers disambiguate either by ordering the actor
before the theme or by anchoring the theme noun, especially when the
theme itself is discourse-recoverable.

\subsection{Participant nominals and privileged-participant
  reference}
\label{sec:nm-derived-participant}

Participant nominals identify whichever argument is privileged by the
voice/aspect stem to which \ird{-ou} attaches. Antipassive-based forms
pick out the actor, plain transitive stems the patient, and
applicative stems the promoted non-patient participant. The nominal
system therefore inherits the voice architecture of the verbal domain
without ceasing to be nominal externally.

\pex
\a \ird{piaščikou} \trsl{the one who ate}
\a \ird{piaštnikou} \trsl{that which was eaten}
\a \ird{piaštébikou} \trsl{the one who was fed}
\xe


%% ================================================================
%%  PART X: ANCHORING AND THE CLAUSE-LEVEL INTERFACE
%% ================================================================

\section{The discourse interface: anchoring, topicality, and
  the absolutive}
\label{sec:nm-topic-interface}

\subsection{Anchoring and the topic-absolutive connection}
\label{sec:nm-topic-abs}

The grammar notes (§\,\ref{sec:definiteness}) that the anchoring
suffix is \emph{especially frequent} on direct-case (absolutive)
patient nouns in transitive clauses of the perfective and
retrospective aspects. This frequency effect is not a syntactic
requirement — the suffix is never imposed by aspect or by
argument structure — but a pragmatic tendency with a formal
correlate.

The formal correlate is the link between the absolutive argument
and the topic position TopP (§\,\ref{sec:th-topic-comment-theory}).
In the perfective/retrospective, the patient occupies the direct
case (absolutive) and surfaces in Spec,TopP as the clause topic.
A topical argument is by definition discourse-salient and
discourse-recoverable — precisely the condition that licenses the
\textsc{Anch\textsuperscript{0}}[\textsc{+anch}] feature. The
co-occurrence of anchoring and absolutive-topicality is therefore not a
construction-specific rule but the interaction of two independent
feature systems: [\textsc{+top}] in the clause domain and
[\textsc{+anch}] in the nominal domain.

This interaction can be formalized as a soft interface constraint
(an optimization preference, not a hard Agree condition):

\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
\item[\upshape Topicality-Anchoring Harmony:] A DP bearing the feature
  [\textsc{+top}] in Spec,TopP is preferentially realized with
  \textsc{Anch\textsuperscript{0}}[\textsc{+anch}] in its nominal functional
  hierarchy. The constraint is violable (topic-drop, bare nouns as topics, and
  pragmatically foregrounded indefinites are all attested), but it captures the
  statistical correlation observed.
\end{description}

\noindent The constraint also predicts the opposite tendency: DPs in non-topic
positions (agents in the ergative case, oblique adjuncts) are less frequently
anchored, because they are not in Spec,TopP and therefore carry no
[\textsc{+top}] feature to harmonize with [\textsc{+anch}].

\subsection{Anchoring in non-direct cases: marked uses}
\label{sec:nm-topic-oblique}

The grammar observes (§\,\ref{sec:definiteness}) that the anchoring suffix may
appear on nouns in the transitive and indirect cases, but that such use is
\emph{generally less neutral and more plainly expressive of topical, resumptive,
contrastive, or otherwise salient reference}.

This asymmetry follows from the topic-absolutive architecture. Transitive-case
(ergative) nouns are agents in the perfective aspects; they are not the
absolutive argument and are not in Spec,TopP. A [\textsc{+anch}] agent is
therefore marked: its anchoring is not discourse-default but serves an
additional function — contrastive individualization, resumption, or topical
promotion — that requires the anchoring suffix to perform pragmatic work beyond
its baseline anaphoric use. Indirect-case nouns similarly bear the suffix only
when the speaker wishes to signal that the oblique participant is specifically
tracked in the discourse.

One useful way to formalize the marked anchored agent is to say that
[\textsc{+anch}] combines there not with ordinary topic continuity but with
contrastive prominence. Because the ergative aspect already assigns topichood to
the absolutive patient, an anchored agent is most naturally interpreted as
focally individuated rather than merely anaphoric. This predicts the familiar
contrastive or resumptive flavor of anchored transitive-case DPs without forcing
a second topic position inside the clause.

\subsection{The bare noun as an AnchP with null anchoring}
\label{sec:nm-bare-noun}

Within the present analysis, the bare noun in Iridian is not an NP that has
failed to project upward. Rather, it is an AnchP whose anchoring head bears no
[\textsc{+anch}] feature and therefore receives no overt Vocabulary Item at PF.
This treatment parallels the analysis of topic-drop in the clausal domain: just
as a null topic in Spec,TopP is a syntactically present but phonologically empty
argument (§\,\ref{sec:th-topic}), the null anchoring head is a syntactically
present but phonologically empty head.

The theoretical advantage of this treatment is that it derives the optionality
of anchoring morphology without stipulation. There is no constraint that forces
\textsc{Anch\textsuperscript{0}} to receive an overt Vocabulary Item; the DM
system simply inserts \ird{-to} when the feature [\textsc{+anch}] is present and
inserts nothing when it is absent. The parallelism between optional anchoring
and optional topic-marking reflects a broader property of the grammar:
discourse-level features are never obligatorily expressed in Iridian, but they
are consistently and transparently encoded when they are present.


%% ================================================================
%%  PART XI: THEORETICAL TENSIONS AND OPEN QUESTIONS
%% ================================================================

\section{Theoretical tensions and open questions}
\label{sec:nm-tensions}

\subsection{The projection of AnchP: DP or a separate anchoring layer?}
\label{sec:nm-tension-dp}

The standard DP Hypothesis (Abney 1987) places D\textsuperscript{0} as the
single highest head of the nominal extended projection, unifying articles,
demonstratives, and similar functional elements under one projection. An
alternative tradition (Longobardi 1994; Alexiadou et~al. 2007) distinguishes a
dedicated definiteness layer from a higher DP layer (which may encode
referential anchoring, logophoricity, or other discourse properties).

For Iridian, the evidence from morphological ordering supports a single high
functional head for the anchoring suffix, but the interaction with
demonstratives and specificity particles suggests that at least three distinct
functional layers are involved in the upper nominal domain: AnchP (\ird{-to}),
DemP (demonstratives), and SpecificityP (\ird{si}/\ird{druh}). Whether these
should be unified under an articulated DP or organized as separate projections
is left open pending more detailed cartographic work on the upper nominal
domain.

\subsection{The anchoring suffix and Case theory}
\label{sec:nm-tension-case}

A potential tension arises within Case theory. In the standard
Minimalist account (Chomsky 2001), case is
assigned to DPs by an Agree relation with a functional head (v*,
T, or an applicative head). If the nominal extended projection is
capped by AnchP, then the outermost head of the DP is D\textsuperscript{0},
not K\textsuperscript{0}. Case should in principle be visible on
the DP as a whole (from the perspective of the external clause
structure), but the morphological encoding of case is on the K
layer inside the DP, below D\textsuperscript{0}.

This can be resolved by one of two strategies: (a) the D head
is transparent to case, meaning the Agree probe from the clausal
head targets K\textsuperscript{0} or the NP root directly through
AnchP; or (b) case features percolate from K\textsuperscript{0} to
D\textsuperscript{0} by feature inheritance within the DP, and D
presents the valued case feature to the external probe. Strategy (b)
has the advantage of preserving the standard Agree architecture;
strategy (a) requires relaxing the usual opacity of DP boundaries.

\subsection{The anchoring suffix and phase theory}
\label{sec:nm-tension-phase}

Within phase theory (Chomsky 2000, 2001), DP may constitute a phase alongside CP and vP. If the Iridian DP
is a phase, then the internal structure of the DP (including the
K and NP layers) becomes inaccessible to clausal operations once
AnchP (= DP-phase edge) is assembled. This has consequences for
case assignment (as noted above) and for the possessor construction:
if the possessive indirect-case NP is inside the DP phase, it
cannot be directly targeted by a clausal probe. The standard
resolution is to treat possessors as phase-edge elements (moved
or base-generated at the edge of the DP phase), which is consistent
with their prenominal, edge-adjacent position in the surface string.

\subsection{Converb-based relative clauses and their nominal
  embedding}
\label{sec:nm-tension-rel}

The \ird{-ovní} attributive construction is Iridian's primary
relativization strategy (§\,\ref{sec:th-relativization}). A full
syntactic analysis of this construction — and in particular its
interaction with the anchoring suffix on the head noun —
requires a formal account of how a nominalized verbal phrase
(the \ird{-ovní} clause) integrates into the DP-internal structure.
The head noun following an \ird{-ovní} modifier receives case from
the host clause and optionally bears the anchoring suffix,
suggesting that it remains within the AnchP/KP layer of the host DP
and is not incorporated into the nominalized domain. This is
consistent with the analysis of the \ird{-ovní} construction as
a prenominal adjunct, but the precise adjunction site (to NP,
to PossP, or to a dedicated RelP layer) requires further
specification.

The main text has adopted the reduced-relative option, and the open
question is therefore narrower than before: not whether \ird{-ovní}
instantiates RelP at all, but where that RelP adjoins inside the lower
nominal spine and how it interacts with possessive material. That is a
real structural question, but it no longer blocks the broader analysis
of \ird{-ovní} as Iridian's default prenominal relativizing
construction.
