\chapter{Function Words and Particles}\label{ch:function-words}

\section{Overview}

Function words in Iridian form several small, largely closed classes whose
importance is out of proportion to their size. Unlike nouns and verbs, they
contribute little lexical content of their own; instead they organise the
clause, mark relations between constituents, and encode the speaker's stance
toward what is being said. Among them, the preverbal particle series is
especially central, since it interacts directly with predicate structure,
aspect, interrogation, and discourse framing.

\section{Adverbial particles}\label{sec:adv-particles}

\subsection{In general}\label{sec:adv-particles-general}

Adverbial particles are a small series of words that are phonologically
proclitic and syntactically tied to the left edge of the predicate domain. In
their unmarked position they stand immediately before the predicate they scope
over: before a finite verb in verbal clauses, before the existential verb in
existential clauses, and before the predicate nominal in zero-copula
predications. Their proclitic phonology is therefore best understood as a
surface reflex of a fixed structural position, not as the sole reason for
treating them as a class.

The traditional Iridian term for these items is \ird{zemo} `grain, particle'.
The label \term{adverbial particle} is retained here because many of them behave
broadly like clause adverbs, but the class is wider than ordinary manner or
temporal adverbs. Particles may contribute phasal contour, addition,
affirmation, reportative force, evidential stance, mirativity, and other
clause-level meanings. What unifies them is not a single semantic type but a
shared positional behaviour and a fixed hierarchy of scope.

One element that appears in a similar linear region but belongs to a different
syntactic class is the interrogative particle \ird{by}. It is treated in
§\,\ref{sec:by-pol-head} immediately after the three adverbial particle classes,
where its status as the overt exponent of \synhead{Pol\textsuperscript{0}} —
a dedicated clause-typing head above the comment domain — is developed. It
is separated from the adverbial particle classes here because grouping it with
them would misrepresent its structural position.

\begin{table}[ht]
	\footnotesize
	\caption{Adverbial particle classes.}\label{tab:adv-particle-classes}
	\medskip
	\begin{tabularx}{\textwidth}{l L L l}
		\toprule
		{\sc class} & {\sc members} & {\sc core function and scope} & {\sc position} \\
		\midrule
		Class 1 &
		\ird{v}, \ird{s} &
		event-internal or predicational phasal contour; closest scope over the event or state itself &
		closest to predicate \\
		Class 2 &
		\ird{och}, \ird{moc}, \ird{či}, \ird{kamo}/\ird{kam}, \ird{dlí}, \ird{jse} &
		propositional and illocutionary meanings such as addition, affirmation, hearsay stance, exclusivity, temporary priority, and contrast &
		middle \\
		Class 3 &
		\ird{oče}, \ird{nadě}, \ird{hlavdy}, \ird{ktady}, \ird{mlada} &
		speaker stance over the whole utterance, including mirative, consequential, inferential, speculative, and hypothetical meanings &
		farthest from predicate \\
		\bottomrule
	\end{tabularx}
\end{table}

The basic ordering of the series is thus as follows:

\pex
\a \ird{Class 3 > Class 2 > Class 1 > predicate}
\xe

\noindent This ordering is best explained semantically: lower particles take
narrower scope and higher particles take scope over progressively larger units
of meaning. The familiar phonological tendency for shorter particles to stand
closer to the predicate and longer ones farther away is real, but it is better
treated as secondary evidence of grammaticalisation depth than as the primary
source of the ordering itself.

The unmarked placement of all three classes is within this preverbal series.
The highest particles may, in marked discourse use, be placed farther left as
sentence-framing elements; this possibility is most plausible for the Class~3
particles and is taken up when those particles are discussed individually. In
the ordinary clause, however, the default pattern is strictly preverbal.

It is important to distinguish these particles from ordinary lexical adverbs.
Derived manner adverbs in \ird{-ě}, such as \ird{skorně} \trsl{quickly}, are
not members of the particle series and do not occupy one of its fixed
functional slots. They are lexical adjuncts in the preverbal field, whereas the
particles discussed here are a closed series of scope-ordered functional items.

\pex
\a
\begingl
\gla Janek v piašček.//
\glb Janek.\Dir{} already eat-\Antip{}-\Pf{}//
\glft \trsl{Janek has already eaten.}//
\endgl
\a
\begingl
\gla Tak tóm Janku oče by v koupnik?//
\glb this book Janek.\Trs{} mirative question already buy-\Trs{}-\Pf{}//
\glft \trsl{Has Janek, surprisingly, already bought this book?}//
\endgl
\a
\begingl
\gla Marek v doktor.//
\glb Marek.\Dir{} already doctor.\Dir{}//
\glft \trsl{Marek is already a doctor.}//
\endgl
\xe

\section{Lexical adverbs outside the particle series}
\label{sec:lex-adv}

\subsection{Manner and frequency adverbs}

Not all adverbial meanings in Iridian are expressed by preverbal particles.
Alongside the particle system stands a smaller open class of lexical adverbs.
Two types are especially relevant. First, manner adverbs may be derived from
property-root-based stative stems by the suffix \ird{-ě}
(§\,\ref{sec:stative-adv}), yielding forms such as \ird{skorně}
\trsl{quickly}. Second, frequency adverbs form a lexical subclass not derived
from stative verbs. They modify the frequency or regularity of the event but
are not part of the rigid particle hierarchy.

The core examples of lexical frequency adverbs are:

\pex
\a \ird{ohly} \trsl{sometimes}
\a \ird{roca} \trsl{often}
\a \ird{člíně} \trsl{infrequently}
\a \ird{sabo} \trsl{never}
\xe

\noindent These items behave as preverbal lexical modifiers in the clause, but
they remain distinct from adverbial particles in both status and syntax. They
should therefore not be analysed as members of the Class~1--3 particle series,
and the set should not be taken as closed.

\subsection{Class 1: \ird{v} and \ird{s}}\label{sec:class1-particles}

The lowest slot in the particle sequence is occupied by either \ird{v} or
\ird{s}, never by both at once. These are best analysed as phasal particles
anchored to a reference time, not as direct exponents of the aspect-alignment
suffixes described in Section~\ref{sec:aspect}. Broadly speaking, \ird{v}
signals that the relevant state or event has already been attained, or is taken
to be close to attainment, by the reference point. \ird{s}, by contrast,
presents the relevant interval as still open, still continuing, or not yet
exhausted at that same point.

Because these meanings interact with viewpoint aspect, \ird{v} tends to occur
most naturally with bounded, attained, or resultative predications, while
\ird{s} prefers ongoing, persistent, or still-pending predications. The
correlation is strong, but it is not absolute. Class~1 particles do not replace
aspect morphology; rather, they add a phasal interpretation to the aspectual
viewpoint supplied elsewhere in the predicate.

\noindent\textit{Morphophonological note.} Before a following word beginning
with \ird{v}, \ird{v} surfaces as \ird{ve}; before a following word beginning
with a sibilant, \ird{s} surfaces as \ird{se}.

\subsubsection{Core meaning relative to a reference time}

With present-time reference, the contrast between \ird{v} and \ird{s} is often
most transparent with stative or nominal predicates. \ird{V} marks the state as
already reached; \ird{s} presents it as continuing.

\pex
\a
\begingl
\gla Marek doktor.//
\glb Marek.\Dir{} doctor.\Dir{}//
\glft \trsl{Marek is a doctor.}//
\endgl
\a
\begingl
\gla Marek v doktor.//
\glb Marek.\Dir{} already doctor.\Dir{}//
\glft \trsl{Marek is already a doctor.}//
\endgl
\a
\begingl
\gla Marek s doktor.//
\glb Marek.\Dir{} still doctor.\Dir{}//
\glft \trsl{Marek is still a doctor.}//
\endgl
\xe

The same placement rule holds with existential predicates. Here \ird{s}
indicates persistence, while \ird{v} tends to mark a newly attained or already
established situation.

\pex
\a
\begingl
\gla Marek dumě na jeca.//
\glb Marek.\Dir{} house.\Ind{} in \Exst{}-\Cont{}//
\glft \trsl{Marek is at home.}//
\endgl
\a
\begingl
\gla Marek dumě na s jeca.//
\glb Marek.\Dir{} house.\Ind{} in still \Exst{}-\Cont{}//
\glft \trsl{Marek is still at home.}//
\endgl
\xe

\subsubsection{With explicit temporal adjuncts}

When the clause contains an explicit time expression, the reference point
introduced by that expression becomes central to the interpretation of
\ird{v} and \ird{s}. With future-oriented predicates, \ird{v} presents the
event as already close from the speaker's current vantage point, or as one that
will have been reached by the named time; \ird{s} presents the interval leading
up to that time as still open and not yet exhausted.

\pex
\a
\begingl
\gla Janek sobotě ščenách.//
\glb Janek Saturday.\Ind{} arrive-\Ctpl{}//
\glft \trsl{Janek will arrive on Saturday.}//
\endgl
\a
\begingl
\gla Janek sobotě v ščenách.//
\glb Janek Saturday.\Ind{} already arrive-\Ctpl{}//
\glft \trsl{Janek will already have arrived by Saturday / Saturday already feels close for Janek's arrival.}//
\endgl
\a
\begingl
\gla Janek sobotě se ščenách.//
\glb Janek Saturday.\Ind{} still arrive-\Ctpl{}//
\glft \trsl{Janek will arrive on Saturday, and there is still time before then.}//
\endgl
\xe

With past-oriented predicates, \ird{s} may likewise measure the interval
between the event time and the present, yielding the sense that a considerable
stretch of time has elapsed since that event. In the same environment,
\ird{v} usually retains its ordinary already-attained reading instead of
contributing the same interval effect.

\pex
\a
\begingl
\gla Janek sobotě ščenik.//
\glb Janek Saturday.\Ind{} arrive-\Pf{}//
\glft \trsl{Janek arrived on Saturday.}//
\endgl
\a
\begingl
\gla Janek sobotě se ščenik.//
\glb Janek Saturday.\Ind{} still arrive-\Pf{}//
\glft \trsl{Janek arrived on Saturday, and it has been quite some time since then.}//
\endgl
\xe

\subsubsection{Interaction with aspect morphology}

The relation between Class~1 particles and aspect is systematic, but it is not
a matter of simple one-to-one matching. \ird{V} is especially natural with
predicates that present a bounded result, a completed event, or an attained
state: perfectives, retrospectives, zero-copula predications, and some
contemplatives. \ird{S} is especially natural with predicates that present an
ongoing state, an open interval, or a still-relevant course of action:
continuous and progressive predicates, persistent existential clauses, and some
contemplatives.

This means that \ird{v} overlaps partly with the retrospective, but the two are
not equivalent. The retrospective contributes prior-event perspective and
continuing relevance; \ird{v} adds the scalar meaning that the relevant result
has already been reached by the reference point. A perfective clause with
\ird{v} therefore often resembles an English \trsl{already}-perfect, while a
retrospective clause with \ird{v} places that already-attained result within a
broader frame of prior relevance.

\pex
\a
\begingl
\gla Janek ščenik.//
\glb Janek.\Dir{} arrive-\Pf{}//
\glft \trsl{Janek arrived.}//
\endgl
\a
\begingl
\gla Janek v ščenik.//
\glb Janek.\Dir{} already arrive-\Pf{}//
\glft \trsl{Janek has already arrived.}//
\endgl
\a
\begingl
\gla Pražě de stožac.//
\glb Prague.\Ind{} to go-\Ret{}//
\glft \trsl{(I) have been to Prague before.}//
\endgl
\xe

\noindent In a framed retrospective clause, \ird{v} shows that the relevant
prior event had already been completed by the framing event:

\pex
\begingl
\gla Marek ščenu Janek v piaščac.//
\glb Marek.\Dir{} arrive-\Seq{} Janek.\Dir{} already eat-\Antip{}-\Ret{}//
\glft \trsl{When Marek arrived, Janek had already eaten.}//
\endgl
\xe

With contemplatives, the same contrast appears in a different form: \ird{v}
often yields an imminent reading (\trsl{already about to}), whereas \ird{s}
preserves the sense that the projected event or intention is still in force
(\trsl{still going to}, \trsl{still intending to}).

\subsubsection{Negated predicates}

Under negation, the core meanings of \ird{v} and \ird{s} remain visible as a
scope interaction. \ird{V} with a negative predicate usually marks cessation or
loss of a previously attained state, yielding translations such as
\trsl{no longer} or \trsl{no more}. \ird{S} with a negative predicate marks
non-attainment, delay, or an interval that remains open, yielding
\trsl{not yet}. Stronger emphatic readings are possible in context, but these
follow from the basic contrast rather than replacing it.

\pex
\a
\begingl
\gla Pivo v zada.//
\glb beer.\Dir{} already \NegExst{}-\Cont{}//
\glft \trsl{There is no more beer.}//
\endgl
\a
\begingl
\gla Pivo s zada.//
\glb beer.\Dir{} still \NegExst{}-\Cont{}//
\glft \trsl{There is no beer yet.}//
\endgl
\a
\begingl
\gla Marek doktor v staly.//
\glb Marek.\Dir{} doctor.\Dir{} already \NegCop{}//
\glft \trsl{Marek is no longer a doctor.}//
\endgl
\a
\begingl
\gla Marek doktor s staly.//
\glb Marek.\Dir{} doctor.\Dir{} still \NegCop{}//
\glft \trsl{Marek is not yet a doctor.}//
\endgl
\xe

\subsubsection{Extended uses of \ird{s}}

Besides its core continuative meaning, \ird{s} has several extended uses that
are best understood as developments from the idea of an interval that remains
open or of an additional step on a scale. With perfective predicates,
especially in narratives of inconvenience, effort, or unexpected inclusion,
\ird{s} may be translated as \trsl{even}. This use often carries annoyance,
commiseration, or rhetorical emphasis; the affirmative particle \ird{moc} may
reinforce the interpretation, though the details belong properly to the Class~2
discussion.

\pex
\a
\begingl
\gla Magazinam de bych s stojnik!//
\glb store.\Ind{} into yesterday still go-\Pf{}//
\glft \trsl{I even went to the store yesterday!}//
\endgl
\a
\begingl
\gla Marek Štěpánovímně sprivam moc se stojnik.//
\glb Marek.\Dir{} Štěpánov neighborhood.\Ind{} toward affirmative still go-\Pf{}//
\glft \trsl{Marek even went to Štěpánov.}//
\endgl
\xe

In questions, the same extension yields \trsl{else} or \trsl{in addition}:

\pex
\begingl
\gla Mám Marku s piaštnik?//
\glb what Marek.\Trs{} still eat-\Trs{}-\Pf{}//
\glft \trsl{What else did Marek eat?}//
\endgl
\xe

These meanings should not be treated as the core semantics of \ird{s}. They are
contextually licensed extensions of the general continuative/additive profile of
the particle.

\subsubsection{Yes-no questions}

The overt interrogative particle \ird{by} is what makes a clause a question
(§\,\ref{sec:by-pol-head}); \ird{v} and \ird{s} do not themselves form
questions. Rather, they specify whether the speaker asks about an already
attained result, an ongoing situation, or an interval that remains open. This
distinction becomes especially clear when the question concerns completion or
timing relative to a reference point.

\begin{table}[ht]
\footnotesize
\caption{Class 1 particles in yes-no questions.}\label{tab:v-s-questions}
\medskip
\begin{tabularx}{\textwidth}{l L L}
\toprule
{\sc aspect} & {\sc with \ird{v}} & {\sc with \ird{s}} \\
\midrule
Perfective &
\ird{Janek by v piašček?} \newline
\trsl{Has Janek already eaten?} &
\ird{Janek by s piašček?} \newline
\trsl{Did Janek still manage to eat?} \\
Progressive &
\ird{Janek by v piaštime?} \newline
\trsl{Is Janek already eating?} &
\ird{Janek by s piaštime?} \newline
\trsl{Is Janek still eating?} \\
Contemplative &
\ird{Janek by v piaštách?} \newline
\trsl{Is Janek already about to eat?} &
\ird{Janek by s piaštách?} \newline
\trsl{Is Janek still planning to eat?} \\
\bottomrule
\end{tabularx}
\end{table}

\noindent The particle \ird{by} that forms polar questions is analysed as
\synhead{Pol\textsuperscript{0}} in §\,\ref{sec:by-pol-head}; the full syntax of polar
and content questions is treated in §\,\ref{sec:th-polar-questions} and
§\,\ref{sec:th-wh-interrogation} of Chapter~\ref{ch:simple}. What matters
here is the independent contribution of the Class~1 particle to the
temporal-phasal profile of the question.

\subsection{Class 2: propositional particles}\label{sec:class2-particles}

The middle slot of the preverbal particle series is occupied by particles whose
scope is broader than the event-internal phasal meanings of Class~1 but
narrower than the sentence-wide stance meanings of Class~3. Class~2 particles
contribute propositional and illocutionary meanings such as addition,
affirmation, hearsay stance, exclusivity, temporary priority, and contrast.
Their ordinary position is between any Class~3 particle and any Class~1
particle. The interrogative clause-typing particle \ird{by} formerly appeared
in this class in some descriptive traditions, but it is set apart from the
Class~2 inventory in the present analysis and given a dedicated structural
account in §\,\ref{sec:by-pol-head}; the formal distinction is explained
there.

\paragraph{Och} \ird{Och} is primarily an additive particle. It presents the
proposition as holding for an additional discourse referent, or as an
additional fact alongside one already under discussion. In positive clauses it
is usually translated as \trsl{also} or \trsl{too}; under negation it often
corresponds to \trsl{either} or \trsl{neither.}

\pex
\a
\begingl
\gla Janek och zdravime.//
\glb Janek.\Dir{} also sleep-\Prog{}//
\glft \trsl{Janek is also sleeping.}//
\endgl
\a
\begingl
\gla Matuš och záščenik.//
\glb Matuš.\Dir{} also \Neg{}-arrive-\Pf{}//
\glft \trsl{Matuš has not arrived either.}//
\endgl
\xe

With bounded predicates, especially perfectives and retrospectives,
\ird{och} may develop a scalar phasal reading such as \trsl{finally} or
\trsl{at last}. This is best treated as a contextual extension rather than as a
separate lexical meaning: the particle still adds the event to an already
active discourse sequence, but the relevant addition is now a long-awaited
attainment. The reading is especially natural with \ird{v}. With \ird{s} and a
negative predicate, a related extension yields \trsl{still not} or
\trsl{not yet}, often with an overtone of prolonged waiting.

\pex
\a
\begingl
\glpreamble \lingcontext{After a long wait, the speaker sees Janek at the
door.}//
\gla Janek och v ščenik.//
\glb Janek.\Dir{} also already arrive-\Pf{}//
\glft \trsl{Janek has finally arrived.}//
\endgl
\a
\begingl
\glpreamble \lingcontext{The speaker has been waiting for Janek for some
time and is becoming impatient.}//
\gla Janek och se záščenik.//
\glb Janek.\Dir{} also still \Neg{}-arrive-\Pf{}//
\glft \trsl{Janek still has not arrived.}//
\endgl
\xe

\paragraph{Moc} \ird{Moc} is an affirmative or confirmatory particle. It marks
the speaker's ratification of the proposition as true, often where that truth
has been doubted, denied, questioned, or otherwise made salient in the
preceding discourse. Its stable contribution is therefore confirmatory rather
than emotional. Depending on context, the same particle may sound reassuring,
insistent, triumphant, annoyed, or reproachful.

\pex
\begingl
\glpreamble \lingcontext{A tells B that Janek has arrived. B doubts this, then
sees Janek in person and reports back.}//
\gla Janek moc v ščenik.//
\glb Janek.\Dir{} affirm already arrive-\Pf{}//
\glft \trsl{Janek really has arrived.}//
\endgl
\xe

In imperatives and hortatives, \ird{moc} keeps this ratifying force, but now
the speaker presents the urged course of action as one that should by now be
carried out. The result is often urgency, coaxing, or emphatic insistence.
With \ird{v}, the reading commonly approaches \trsl{come on, already}.

\pex
\begingl
\gla Moc v piaštím!//
\glb affirm already eat-\Imp{}//
\glft \trsl{Go on, eat already!}//
\endgl
\xe

\paragraph{Či} \ird{Či} is a hearsay or attributed-information particle. It
marks the proposition as one the speaker is passing on from another source
rather than presenting as newly verified by direct personal knowledge. The
source may be a prior report, common talk, something just heard, or information
already circulating in the discourse. Its effect is often lighter than a fully
formal reportative: the speaker indicates that the proposition is being relayed,
but need not thereby dramatize distance or strong scepticism.

\pex
\a
\begingl
\glpreamble \lingcontext{The speaker has just heard that Janek has arrived and
passes the information on.}//
\gla Janek či v ščenik.//
\glb Janek.\Dir{} reportedly already arrive-\Pf{}//
\glft \trsl{They say Janek has already arrived.}//
\endgl
\a
\begingl
\glpreamble \lingcontext{The speaker is repeating what they have been told about
Matuš.}//
\gla Matuš či záščenik.//
\glb Matuš.\Dir{} reportedly \Neg{}-arrive-\Pf{}//
\glft \trsl{They say Matuš has not arrived.}//
\endgl
\xe

\noindent What \ird{či} adds is not simply uncertainty. A speaker may use it
whether or not they personally find the proposition plausible; the essential
point is that the information is being framed as relayed or attributed rather
than as a fresh firsthand assertion.

\noindent This function should be kept distinct from complement constructions
built with an overt speech or cognition predicate. Those constructions
syntactically subordinate one proposition under another, as in \trsl{X said
that P} or \trsl{X thinks that P}. \ird{Či}, by contrast, does not embed the
proposition at all: it marks the speaker's stance toward the main clause
itself, presenting that clause as hearsay or relayed information. Accordingly,
\ird{či} is not redundant with quotative framing. The two constructions do
genuinely different jobs and may co-occur when both meanings are wanted:
syntactic attribution to a speech frame on the one hand, and speaker-level
hearsay stance toward the clause on the other.

\paragraph{Kamo and kam} \ird{Kamo} and \ird{kam} are exclusive particles
meaning \trsl{only}. The two are semantically equivalent; \ird{kam} is lighter
and more colloquial, while \ird{kamo} is more formal or careful in style. Their
scope depends on the focused constituent in context, so English translations
such as \trsl{only}, \trsl{just}, or \trsl{alone} reflect information
structure as much as lexical meaning.

\pex
\a
\begingl
\gla Janek Roubžu na by mulša?//
\glb Janek.\Dir{} Roubže.\Ind{} in question live-\Cont{}//
\glft \trsl{Does Janek live in Roubže?}//
\endgl
\a
\begingl
\gla Janek Roubžu na kamo by mulša?//
\glb Janek.\Dir{} Roubže.\Ind{} in only question live-\Cont{}//
\glft \trsl{Is Janek the only one living in Roubže?}//
\endgl
\xe

In the second example, the most natural focus is on the identity of the
inhabitant, so the exclusive reading is \trsl{only Janek} rather than
\trsl{only in Roubže}. A fuller treatment should make the interaction between
particle placement and focus structure more explicit.

\paragraph{Dlí} \ird{Dlí} is a particle of temporary priority or interim
course. It presents the predication as the action to be carried out first, for
the time being, or in advance of some later step that remains contextually
available. In English it is often rendered by \trsl{first}, \trsl{for now},
\trsl{for the meantime}, or \trsl{just for the moment}. The particle therefore
does not merely locate an event earlier in time. Rather, it frames the event as
an immediate or provisional move within a larger sequence of actions or
decisions.

\pex
\a
\begingl
\glpreamble \lingcontext{The speaker explains the immediate plan before deciding
what to do next.}//
\gla Dá dlí zdrava.//
\glb I.\Dir{} for.now sleep-\Cont{}//
\glft \trsl{I'll sleep for now.}//
\endgl
\a
\begingl
\glpreamble \lingcontext{The speaker urges the addressee to do the eating step
before anything else.}//
\gla Dlí piaštím!//
\glb first eat-\Imp{}//
\glft \trsl{Eat first.}//
\endgl
\xe

The broader interaction of \ird{dlí} with imperatives, future-oriented
contexts, and clause chaining remains to be worked out in detail. The working
analysis, however, is that \ird{dlí} belongs to the propositional layer because
it manages the place of the whole predication within an intended sequence or
temporary plan, rather than merely describing the internal phase of the event
itself.

\paragraph{Jse} \ird{Jse} is a discourse-reorienting contrast particle. It
often introduces an alternative or contrast, but its use need not be confined
to strong semantic opposition. It may also mark topic shift, alternation,
turn-taking, or a looser \trsl{as for X} framing in which attention is
redirected to another participant or proposition.

\pex
\begingl
\gla Janek zdrava. Marek jse znohtime.//
\glb Janek.\Dir{} sleep-\Cont{} Marek.\Dir{} contrast study-\Cont{}//
\glft \trsl{Janek is sleeping. Marek, for his part, is studying.}//
\endgl
\xe

This broader reorienting value also underlies corrective and defensive uses.
Here the speaker contrasts the current proposition not necessarily with an
explicitly stated opposite, but with some assumption, expectation, or implied
line of reasoning in the discourse.

\pex
\begingl
\glpreamble \lingcontext{The speaker's doctor tells them that they should start
eating more vegetables. They respond defensively.}//
\gla Dá graže jse piašta.//
\glb I.\Dir{} vegetable.\Ind{} contrast eat-\Cont{}//
\glft \trsl{But I do eat vegetables.}//
\endgl
\xe

Because \ird{jse} reorients discourse rather than encoding contradiction by
itself, identical predicates are not automatically excluded. Such examples are
usually odd if no shift in participant set or discourse management is intended,
but they become natural when the second clause is interpreted as parallel,
alternating, or explicitly additive. Combination with \ird{och} is especially
natural in this use.

\pex
\begingl
\gla Marek zdravime. Janek och jse zdravime.//
\glb Marek.\Dir{} sleep-\Prog{} Janek.\Dir{} also contrast sleep-\Prog{}//
\glft \trsl{Marek is sleeping. Janek, meanwhile, is sleeping too.}//
\endgl
\xe

When \ird{jse} combines with the Class~1 particles \ird{v} and \ird{s}, the
result is not merely a mechanical accumulation of meanings. The combination
with \ird{v} is the more lexicalised of the two. Since \ird{v} presents the
relevant state or event as already attained, while \ird{jse} sets that
attainment against a prior course of events or a previously active discourse
frame, \ird{jse v} often yields a renewed-occurrence reading best translated as
\trsl{again}, \trsl{once again}, or \trsl{again already}. The core idea is not
simple repetition in the abstract, but the return of a now-salient event or
state. For this reason, \ird{jse v} often sounds affectively coloured in
context, suggesting weary recognition, annoyance, resignation, or
counterexpectation, though such colouring is pragmatic rather than encoded.

\pex
\a
\begingl
\glpreamble \lingcontext{Janek was expected to stay for a while, but he has
left once more.}//
\gla Janek jse v varžek.//
\glb Janek.\Dir{} contrast already leave-\Pf{}//
\glft \trsl{Janek has left again.}//
\endgl
\a
\begingl
\glpreamble \lingcontext{The speaker notices a familiar and unwelcome pattern
returning.}//
\gla Janek jse v záščenik.//
\glb Janek.\Dir{} contrast already \Neg{}-arrive-\Pf{}//
\glft \trsl{Janek has once again not arrived.}//
\endgl
\xe

The combination with \ird{s} is less lexicalised and remains more transparently
compositional. Here \ird{s} contributes continuance, non-completion, or
ongoing validity, while \ird{jse} reorients that continuing situation against
some expectation, worry, competing line of action, or shared background
assumption. As a result, \ird{jse s} often yields translations such as
\trsl{still, though}, \trsl{still, after all}, \trsl{still, as you know}, or
\trsl{not yet, anyway}. This cluster is especially natural in concessive,
mitigating, reassuring, or regretive environments.

\pex
\a
\begingl
\glpreamble \lingcontext{Someone worries that there is no time left. The
speaker reassures them.}//
\gla Čas jse s jeca.//
\glb time.\Dir{} contrast still exist-\Cont{}//
\glft \trsl{There is still time, though.}//
\endgl
\a
\begingl
\glpreamble \lingcontext{The speaker softens a negative update.}//
\gla Tak jse se staly.//
\glb this.\Dir{} contrast still \NegCop{}//
\glft \trsl{That is not the case yet, though.}//
\endgl
\xe

The same compositional logic extends to rhetorical interrogative environments.
When \ird{jse} combines with the interrogative particle \ird{by}, the result is often
best understood not as a neutral request for information but as interrogative
form plus discourse stance. The interrogative force is contributed by
\ird{by}; \ird{jse} then reorients the proposition against what should already
be evident, expected, or inferable in the current discourse. As a result,
\ird{jse by} often sounds retortive, incredulous, reproachful, or
challenge-like: roughly \trsl{come now, P?}, \trsl{P, after all?}, or
\trsl{was it not obvious that P?}, depending on context. The special flavour is
therefore pragmatic and compositional, not the fixed lexical meaning of an
idiom.

This same analysis also clarifies the larger clusters. \ird{Jse by v} combines
interrogative framing with an already-attained or now-salient result, and so
is especially natural where the speaker throws that result back into the
discourse as the obvious explanation for something else; this configuration
readily yields exasperated or \trsl{what else did you expect?} overtones.
\ird{Jse by s}, by contrast, combines rhetorical interrogation with an open or
still-valid interval, and so more often suggests \trsl{is it not still the
case, after all?}, \trsl{is there not still time?}, or \trsl{has this not yet
been settled, then?} The contrast mirrors the broader difference between
\ird{jse v} and \ird{jse s}: the former foregrounds a returned or already
attained state of affairs, while the latter foregrounds persistence,
non-completion, or continuing relevance under a discourse-reorienting frame.

With imperatives and hortatives, \ird{jse} often signals that the urged course
of action should already be evident from the shared situation. Depending on the
context, this may sound gentle, coaxing, impatient, teasing, or mildly
reproachful.

\pex
\begingl
\glpreamble \lingcontext{In a library, one speaker has been making noise for
some time. The other finally intervenes.}//
\gla Jse okouzé!//
\glb contrast be:noisy-\Proh{}//
\glft \trsl{Please do not be so noisy.}//
\endgl
\xe

\subsection{Class 3: stance particles}\label{sec:class3-particles}

The highest slot in the preverbal particle series is occupied by particles
whose scope extends over the clause as an utterance. Unlike Class~1 particles,
which shape the internal phasal contour of the predicate, and unlike Class~2
particles, which contribute propositional or illocutionary meanings, Class~3
particles encode the speaker's stance toward the whole proposition. The
present section is adapted from the older lexnotes draft, but the inventory has
been updated to the current system: \ird{oče}, \ird{nadě}, \ird{hlavdy},
\ird{ktady}, and \ird{mlada}. Earlier archive notes discussed
\ird{izdy} and \ird{děne}; the corresponding semantic space is now covered
instead by \ird{hlavdy} and \ird{ktady}.

\paragraph{Oče} \ird{Oče} is the mirative particle. It marks the proposition
as surprising, suddenly discovered, contrary to expectation, or otherwise
arriving at a moment when the speaker is not mentally prepared for it. The
relevant surprise may come from new information, abrupt realisation,
counterexpectation, or simple recollection of something momentarily forgotten.

\pex
\a
\begingl
\glpreamble \lingcontext{The speaker is surprised to learn that Janek has
left; they had expected him still to be there.}//
\gla Janek oče ve varžek.//
\glb Janek.\Dir{} mirative already leave-\Pf{}//
\glft \trsl{Janek has already left, unexpectedly.}//
\endgl
\a
\begingl
\glpreamble \lingcontext{The speaker suddenly remembers that it is raining.}//
\gla Oče pozebževí.//
\glb mirative rain-\Cont{}//
\glft \trsl{Oh, it's raining.}//
\endgl
\xe

\paragraph{Nadě} \ird{Nadě} is the consequential particle. It presents the
proposition as the upshot of facts already in view, whether those facts have
just been stated overtly or are recoverable from the shared situation. English
translations often use \trsl{so}, \trsl{therefore}, or \trsl{that's why}, but
the particle is best analysed as a stance marker rather than as a simple
coordinator: it signals that the speaker is packaging the clause as a
conclusion that follows from what is already given.

\pex
\begingl
\glpreamble \lingcontext{The office is dark and Janek's coat is gone. The
speaker draws the obvious conclusion.}//
\gla Janek nadě ve varžek.//
\glb Janek.\Dir{} therefore already leave-\Pf{}//
\glft \trsl{So Janek has already left.}//
\endgl
\xe

\paragraph{Hlavdy} \ird{Hlavdy} marks an inference grounded in some available
evidence while leaving room for uncertainty. It is more committal than open
speculation: the speaker presents the proposition as the likeliest conclusion
from what is known, but still as an inference rather than as directly verified
fact. Older notes contrasted \ird{hlavdy} with a weaker inferential particle
\ird{izdy}; under the current inventory, \ird{hlavdy} is the dedicated
inferential member of Class~3.

\pex
\a
\begingl
\glpreamble \lingcontext{The speaker sees Janek's car outside the house and
infers that he has arrived, though they have not yet seen him in person.}//
\gla Janek hlavdy v ščenik.//
\glb Janek.\Dir{} infer already arrive-\Pf{}//
\glft \trsl{Janek must have arrived already.}//
\endgl
\a
\begingl
\glpreamble \lingcontext{The speaker has already seen Janek in person, so the
statement is no longer merely inferential.}//
\gla \ljudge{\#}Janek hlavdy v ščenik.//
\glb Janek.\Dir{} infer already arrive-\Pf{}//
\glft \trsl{Janek must have arrived already.}//
\endgl
\xe

\paragraph{Ktady} \ird{Ktady} is the speculative particle. It presents the
proposition as conjectural, exploratory, or tentatively entertained rather
than as a conclusion anchored in clear evidence. It is especially natural in
questions or other non-assertive environments where the speaker invites the
addressee to consider a possibility without strongly committing to it. In this
respect it occupies part of the domain covered in older notes by
\ird{děne}, though the interaction of \ird{ktady} with the renewed
question-particle system still requires fuller exemplification.

\pex
\begingl
\glpreamble \lingcontext{The speakers are trying to guess Marek's profession
from a brief first impression.}//
\gla Marek ktady doktor?//
\glb Marek.\Dir{} speculative doctor.\Dir{}//
\glft \trsl{Could Marek be a doctor, do you think?}//
\endgl
\xe

\paragraph{Mlada} \ird{Mlada} has essentially the same functions as Tagalog
\ird{sana}. It is a hypothetical or desiderative particle which frames the
clause as wished for, hoped for, counterfactual, or entertained against the
actual course of events. With conditional forms it yields ordinary
counterfactual protases and concessive hypotheticals; in isolation it may
function as an optative \trsl{if only} or a wish directed toward an uncertain
future outcome. Its meaning is therefore not merely temporal or modal in the
narrow sense, but explicitly speaker-oriented: the particle presents the whole
proposition as belonging to a non-actual, imagined, desired, or hoped-for
alternative.

\pex
\begingl
\glpreamble \lingcontext{The speaker hopes that Janek will still come.}//
\gla Janek mlada s ščenik.//
\glb Janek.\Dir{} desiderative still arrive-\Pf{}//
\glft \trsl{I hope Janek still comes. / If only Janek would still come.}//
\endgl
\xe

\section{\ird{By}: the canonical exponent of \synhead{Pol\textsuperscript{0}}}
\label{sec:by-pol-head}

The particle \ird{by} is the canonical overt exponent of
\synhead{Pol\textsuperscript{0}}, the head of \synhead{PolP} in the clause
spine (§\,\ref{sec:th-clause-structure}). \synhead{PolP} is a dedicated
clause-typing projection situated above \synhead{FocP} and below \synhead{TopP};
it is therefore structurally distinct from, and higher than, the
comment-domain tier occupied by the Class~2 adverbial particles. This is a
synchronic structural fact, not merely a terminological preference: \ird{by}
types the clause as interrogative by contributing an operator at
\synhead{Pol\textsuperscript{0}}, whereas Class~2 particles such as \ird{och},
\ird{moc}, and \ird{či} contribute additive, affirmative, and reportative
meanings within the \synhead{AspP}-level comment domain. The two categories have
different structural addresses, different scope domains, and different
interactions with higher projections.

\medskip
\noindent\textit{Note on earlier descriptive accounts.}
Older descriptive traditions grouped \ird{by} with the middle-preverbal
particles because it surfaces in the same broad left-of-predicate region.
Under the present analysis, that linear grouping is retained only as a
surface generalisation, not as a claim about syntactic constituency or
structural level. The distinction between a broad \term{surface zone} and a
specific \term{syntactic projection} is critical here: particles may share a
linear neighbourhood while occupying structurally distinct functional heads.
\ird{By} and, say, \ird{och} appear in adjacent surface regions, but they are
the exponents of projections at different levels of the clause spine. The
former is the present analysis; the latter was a valid observational shorthand
that must not be reinterpreted as a structural claim.

The analysis of \ird{by} as \synhead{Pol\textsuperscript{0}} entails the following
structural claims. First, \ird{by} is the overt exponent of \synhead{Pol\textsuperscript{0}}
in matrix interrogatives generally: it types the whole clause as interrogative
and scopes over the full \synhead{IP}, including any comment-domain material. Second,
\synhead{Pol\textsuperscript{0}} may be phonologically null. In polar questions, overt
\ird{by} is the ordinary and unmarked explicit strategy; in content questions,
both overt \ird{by} and null \synhead{Pol\textsuperscript{0}} are standardly available
(see §\,\ref{sec:th-wh-interrogation}). The choice between them in content
questions reflects a difference in discourse profile --- how explicitly and
directly the clause is presented as a question in its interactional context ---
rather than a grammatical asymmetry. Third, \synhead{Pol\textsuperscript{0}} is not a
focus-licensing head; Spec,\synhead{FocP} remains a distinct position hosting a fronted
wh-phrase or other focused constituent on independent grounds, as discussed in
§\,\ref{sec:th-focus}. The movement of a wh-phrase to the left-peripheral
focus position is independent of whether \synhead{Pol\textsuperscript{0}} is overt or
null.

The contribution of the Class~1 phasal particles \ird{v} and \ird{s} to a
polar question is therefore independent of \ird{by} itself. Where \ird{v}
and \ird{s} specify the phasal profile of the predicate — whether an event
is already attained or an interval still open — \ird{by} contributes the
clause-level interrogative force. Because \synhead{PolP} is above \synhead{FocP}, which is above
\synhead{AspP} and \synhead{PhasP}, the higher particles (\synhead{EvidP}, \synhead{SpeechActP}) naturally scope
over the interrogative force introduced by \ird{by}: a mirative question
(\ird{oče~by}) presents the polar question as counterexpectational; a
reportative question (\ird{či~by}) presents it as relayed from another
source. These scope facts are exactly what the \synhead{PolP} position predicts and
are discussed further in §\,\ref{sec:th-polar-questions}.

\pex
\a
\begingl
\gla Marek by zdravime?//
\glb Marek.\Dir{} \textsc{q} sleep-\Prog{}//
\glft \trsl{Is Marek sleeping?}//
\endgl
\a
\begingl
\gla Janek by v piašček?//
\glb Janek.\Dir{} \textsc{q} already eat-\Antip{}-\Pf{}//
\glft \trsl{Has Janek already eaten?}//
\endgl
\a
\begingl
\gla Tak tóm oče by Janku v koupnik?//
\glb this book.\Dir{} \textsc{mir} \textsc{q} Janek.\Trs{} already buy-\Trs{}-\Pf{}//
\glft \trsl{Has Janek, surprisingly, already bought this book?}//
\endgl
\xe

\noindent In ordinary conversation, intonation and wh-marking may license a
null-Pol surface in many contexts, but overt \ird{by} remains the clearest
and most explicit matrix interrogative marker across both polar and content
questions. Its use is especially expected in careful, explicit, or formally
written style, in interactionally foregrounded questions, and whenever the
speaker wants the interrogative framing to be unambiguous to the addressee.
The full structural analysis of polar and content questions --- including the
four-combination paradigm of \ird{by}/null Pol crossed with
fronted/in-situ wh --- is in §\,\ref{sec:th-polar-questions} and
§\,\ref{sec:th-wh-interrogation} of Chapter~\ref{ch:simple}.

\section{Connectives and conjunctions}

The remaining function-word classes are treated in later sections of this
chapter.

\section{Discourse-connective particles}

Many discourse-connective particles arise historically from older predicative or
clitic material. Their interaction with the sentence-level particle system will
be described below.

\section{Interjections}

Interjections stand outside ordinary clause structure and are not part of the
preverbal particle sequence described in Section~\ref{sec:adv-particles}.
