\chapter{Complex Sentences and Clause Chaining}\label{ch:complex}

\section{Overview: clause-combining strategies}

Iridian employs several strategies for combining clauses. The dominant strategy is \textbf{converb chaining}, in which all non-final clauses are expressed as non-finite converb forms linked to a single main (finite) clause. Additional strategies include coordination with conjunctions, and finite complement clauses: the \ird{po}-clause for finite propositional content and the \ird{poby}-clause for polar non-assertive content, including indirect questions of the \textit{whether/if} type.

Across these strategies, subordination in Iridian follows the same structural principle that governs the language more generally: strong head-finality. In clause combining, this means that a dependent clause normally precedes the higher clause in which it is embedded, while the matrix verbal head remains final. The unmarked pattern is therefore not \textit{main clause + subordinate clause}, as in many European languages, but rather:

\begin{center}
\textbf{[SUBORDINATE CLAUSE] [MATRIX-CLAUSE MATERIAL] [MATRIX VERB]}
\end{center}

\begin{figure}[ht]
	\centering
	\begin{forest}
		for tree={
			align=center,
			parent anchor=south,
			child anchor=north,
			l sep=12pt,
			s sep=16pt,
		}
		[\textsc{s}
			[\textsc{Subordinate}\\\textsc{clause}]
			[\textsc{Matrix predicate}
				[\textsc{Matrix-clause}\\\textsc{material}]
				[\textsc{Matrix verb}]
			]
		]
	\end{forest}
	\caption{Head-final organization of subordinate constructions.}
\end{figure}

The same head-final logic holds inside the subordinate clause itself. Its arguments and adjuncts precede its own verbal head, so the subordinate clause is internally verb-final, while the larger construction as a whole culminates in the final verb of the matrix clause. Iridian subordination is thus characterized primarily by ordering: dependent clausal material precedes the higher clause, and the matrix predicate closes the sentence. This generalization holds across the major subordinate clause types, even though those types differ morphologically in how the dependency is marked.

In practical terms, adverbial subordination is expressed chiefly through converbs, while finite subordination is handled by the \ird{po}/\ird{poby} complement system. The centrality of the converb to Iridian clause-combining means that what would be expressed as multiple finite clauses in a Slavic language typically appears in Iridian as a chain of converbs followed by a single finite verb. This has consequences for information structure (only the final, finite verb carries aspect and alignment), for switch-reference tracking, and for discourse organization (see Chapter~\ref{ch:discourse}).

\section{Converb chains: structure and alignment}\label{sec:converb-chains}

\subsection{The finite-verb-final rule}

In a converb chain, all verbs except the last are non-finite (converbs). Only the final verb is finite and carries aspect, mood, and alignment. This rule is absolute in standard written Iridian; in colloquial and dialectal speech, finite verbs may occasionally appear in non-final positions in certain coordinated structures.

\subsection{Accusative alignment within converb clauses}

Case marking within converb clauses always follows nominative-accusative alignment: the absolutive argument is the agent or sole argument, and the patient appears in the indirect (oblique) case. The ergative case is never found inside a converb clause, because converbs carry no aspect and the ergative alignment is triggered only by the perfective and retrospective aspects.

\subsection{Switch-reference and topic continuity}\label{sec:switch-reference}

By default, the subject of a converb clause is coreferent with the subject of the following clause (same-subject, SS reading). When the subjects differ (different-subject, DS), a pronominal clitic is attached to the converb to mark the switch. If both clauses have full nominal subjects, the full NPs disambiguate and no clitic is required.

\section{Sequential constructions}\label{sec:sequential}

\subsection{Forms: \ird{-u} and \ird{-ení}}

The sequential converb has two forms with subtly different temporal force. The \ird{-u} form implies a temporally bounded, completed sub-event preceding the main clause event. The \ird{-ení} form is temporally underspecified and is compatible with both sequential and simultaneous readings.

\pex
\a
\begingl
\gla Marek zdravení Janek znovime.//
\glb Marek sleep-\Seq{} Janek study-\Prog{}//
\glft \trsl{While Marek sleeps, Janek is studying.} \lingcontext{simultaneous reading with \ird{-ení}}//
\endgl
\xe

\subsection{The copula converb and the discourse particle \ird{nu}}\label{sec:seq-copula}

The marked copular verb \ird{neh-} (see §\,\ref{sec:neh} of Chapter~\ref{ch:simple}) has the sequential converb form \ird{nesu}, a suppletive form established independently of the regular aspect paradigm. In colloquial speech, \ird{nesu} is contracted to \ird{nu}. The form \ird{nu} has been grammaticalized in some dialects as a discourse particle inserted between clauses in the manner of English `and then', without strong sequential implication and without retaining the inchoative/copular force of the underlying verb.

\subsection{Interaction with applicative and causative}

\pex
\a \ird{piašt-u} / \ird{piašt-ení} \trsl{having eaten / and [then] ate\ldots}
\a \ird{piašt-éb-u} / \ird{piašt-éb-ní} \trsl{having fed someone}
\a \ird{piašt-nau} / \ird{piašt-na-ní} \trsl{having made someone eat}
\xe

\section{Temporal constructions}\label{sec:temporal}

\subsection{The simultaneous / manner converb (\ird{-ěc})}

The simultaneous converb (\ird{-ěc}) describes an action that is ongoing or co-occurring with the main clause event, and also functions as a manner converb. The suffix belongs to the palatalizing domain of front-vocalic morphology (see Section~\ref{sec:front-palatalization}), not to the special sibilant-replacement system of the antipassive. When the converb and main clause have different subjects, the sequential \ird{-ení} is preferred for purely simultaneous meaning.

\subsection{The six temporal converbs}

\begin{table}[ht]
	\footnotesize\sffamily
	\caption{Temporal converb suffixes.}\label{tab:tempconverbs}
	\medskip
	\begin{tabularx}{0.9\textwidth}{L l L}
		\toprule
		{\sc suffix} & {\sc meaning} & {\sc function} \\
		\midrule
		\ird{-ezak}    & until              & main clause holds until converb-clause event \\
		\ird{-eškady}  & when, around when  & loose temporal co-occurrence \\
		\ird{-eeby}    & since              & from the time of the converb-clause event \\
		\ird{-ešhoume} & as soon as         & immediately upon the converb-clause event \\
		\ird{-eebymy}  & after              & after the converb-clause event \\
		\ird{-esny}    & before             & before the converb-clause event \\
		\bottomrule
	\end{tabularx}
\end{table}

\section{Causal constructions}\label{sec:causal}

% ============================================================
% STATUS: STUB — causal converb suffix not yet documented.
% The 2026 notes list converb types 1, 2, 4, 5 and skip 3 (causal).
% This section is a placeholder until the suffix is confirmed.
% ============================================================

A causal converb introduces the reason or cause for the main clause event (`because X, Y'). The suffix is not yet confirmed in the current revision and will be documented here once established.

\section{Concessive constructions}\label{sec:concessive}

The concessive converb (\ird{-ouc}) introduces a conceded or contrasted clause (`although X, Y'). It licenses the irrealis mood in the main clause when the conceded content is contrary to expectations.

\section{Conditional constructions}\label{sec:conditional}

\subsection{The four conditional converbs}

\begin{table}[ht]
	\footnotesize\sffamily
	\caption{Conditional converb suffixes.}\label{tab:conditional}
	\medskip
	\begin{tabularx}{0.8\textwidth}{L l L}
		\toprule
		{\sc suffix} & {\sc meaning} & {\sc notes} \\
		\midrule
		\ird{-ic}   & if (possibility)       & does not trigger palatalization \\
		\ird{-emy}  & if not (possibility)   & \\
		\ird{-ezmy} & if/when (expectation)  & \\
		\ird{-ena}  & if/when not (expected) & \\
		\bottomrule
	\end{tabularx}
\end{table}

\subsection{No palatalization with \ird{-ic}}\label{sec:cond-nopala}

Despite beginning with the front vowel \ird{-i-}, the conditional converb \ird{-ic} does not trigger palatalization of the thematic consonant. This is a historical accident: the suffix was grammaticalized during a period when this \ird{-i-} lacked palatalizing force. The conditional is therefore distinguishable from the irrealis (\ird{-ěv}), the simultaneous converb (\ird{-ěc}), and the conjunctive linker \ird{-e}, all of which belong to the productive palatalizing domain.

\section{Purposive constructions}\label{sec:purposive}

The purposive converb (\ird{-it}) expresses purpose (`in order to X') and serves as the complement of modal and desiderative verbs. Palatalization is irregular, applying mainly to Class~I stems.

\subsection{The intentional future reanalysis}

In colloquial speech, the purposive converb has been reanalyzed as a future/intentional marker with first-person clitics when no governing verb is present:

\ex
\begingl
\gla Dá trave piašcitem.//
\glb \textsc{1sg} bread.\Ind{} eat-\Purp{}-\textsc{1sg}//
\glft \trsl{I'm gonna eat bread.} \lingcontext{colloquial intentional future}//
\endgl
\xe

\section{Contrastive constructions}\label{sec:contrastive}

The contrastive converb (\ird{-ema}) introduces a contrasting clause (`X, but/whereas Y'), covering simple contrast, adversative contrast, and parallel contrast.

\section{Coordination}\label{sec:coordination}

% ============================================================
% STATUS: STUB — systematic treatment needed.
% Key conjunctions: a (additive), še (additive / complementizer),
% má / ozná (adversative), disjunctives.
% ============================================================

Coordination is expressed by \ird{a} (additive), \ird{še} (additive, also complementizer), \ird{má} / \ird{ozná} (adversative), and a set of disjunctives.

\section{Clausal complementation}\label{sec:complementation}

A complement clause is a subordinate clause which stands to a higher
predicate as one of its arguments. In Iridian such clauses are used for the
content of saying, thought, belief, report, memory, inference, perception
understood as reportable content, evaluation, and the like. Earlier
descriptions often treated much of this domain under the heading
\textit{quotative}. That term is not adopted here as the principal category,
since the construction extends well beyond quotation in the narrow sense.

Thus, where English says \textit{Marek said that Janek left}, or \textit{It
seems odd to me that he has not called}, Iridian treats the bracketed clause
as the complement of the higher predicate. Two chief constructions must be
distinguished. A full proposition, stated as such, is expressed by the finite
\ird{po}-construction. An event viewed as intended, attempted, controlled,
incipient, or otherwise projected is expressed instead by the purposive
converb, which serves here as a controlled or prospective event complement.

\begin{figure}[ht]
  \centering
  \begin{forest}
    for tree={
      align=center,
      parent anchor=south,
      child anchor=north,
      l sep=10pt,
      s sep=14pt,
    }
    [\textsc{sentence}
      [\textsc{clause}
        [\textsc{clause}
          [\textsc{core}
            [\textsc{arg}\\\textit{Lobek}]
            [\textsc{arg}\\\textit{Marku}]
            [\textsc{nuc}
              [\textit{piaštnic-e}]
            ]
          ]
          [\textsc{comp}\\\ird{po}]
        ]
        [\textsc{core}
          [\textsc{arg}\\\textit{Janku}]
          [\textsc{nuc}
            [\textit{dalnik}]
          ]
        ]
      ]
    ]
  \end{forest}
  \caption{RRG clausal juncture (subordinate nexus): \ird{po}-complement construction.}
\end{figure}

\subsection{The finite \ird{po}-construction}

The ordinary means of embedding a full proposition is the finite
\ird{po}-clause. Its form is:

\begin{center}
\textbf{[finite embedded clause + linker \ird{-e}] \ird{po}}
\end{center}

The embedded clause remains fully finite. It preserves the same verbal
categories as an independent clause, including aspect, the alignment
distinctions associated with aspect, voice, and, where required by the sense,
mood. The construction is therefore not a reduced verbal form, but a true
finite clause made subordinate by the linker \ird{-e} and the complement
marker \ird{po}.

An independent clause such as the following:

\ex
\begingl
\gla Lobek Marku piaštnik.//
\glb apple.\Dir{} Marek.\Trs{} eat-\Pf{}//
\glft \trsl{Marek ate the apple.}//
\endgl
\xe

may be embedded without loss of finiteness:

\ex
\begingl
\gla Lobek Marku piaštnice po Janku dalnik.//
\glb apple.\Dir{} Marek.\Trs{} eat-\Pf{}-\Lnk{} \Compl{} Janek.\Trs{} say-\Pf{}//
\glft \trsl{Janek said that Marek ate the apple.}//
\endgl
\xe

The same construction is used after predicates of saying, reporting,
thinking, knowing, believing, remembering, inferring, perceiving a
proposition, and evaluating a proposition. In all such cases the complement
is construed as a full proposition, not merely as a projected event.

The \ird{po}-clause may also appear without an overt matrix predicate. Such
forms were formerly described as \textit{bare quotatives}; synchronically it
is better to treat them as matrixless uses of the same finite complement
construction. In such cases the proposition is presented not as a simple
speaker-assertion, but as attributed content. According to context, this may
yield a reportive, hearsay, echoic, reflective, distancing, ironic, or
reservation-marking effect.

\ex
\begingl
\gla Lobek Marku piaštnice po.//
\glb apple.\Dir{} Marek.\Trs{} eat-\Pf{}-\Lnk{} \Compl{}//
\glft \trsl{Marek ate the apple, I hear.}//
\endgl
\xe

The precise omitted frame need not be supplied in every instance. One may
understand such clauses loosely as equivalent to \textit{it is said},
\textit{I hear}, \textit{one gathers}, \textit{so they say}, or the like;
but grammatically the essential fact is that the proposition is cast as
attributed rather than directly asserted.

\subsection{Prospective and controlled event complements}

The finite \ird{po}-construction must be kept distinct from the complement
used for intended, attempted, controlled, or projected events. This second
construction belongs to the domain of the purposive converb. It is employed
with predicates such as \textit{want}, \textit{try}, \textit{intend},
\textit{go in order to}, \textit{begin}, and similar verbs whose complement
denotes not a proposition as such, but an event regarded as prospective,
volitional, incipient, or purposive.

In this use the complement is non-finite. It does not present an independent
state of affairs as true, reported, believed, or remembered, but rather an
event toward which the matrix predicate is directed. The distinction is
therefore one between propositional complementation and event
complementation. The former is the province of finite \ird{po}-clauses; the
latter, of the purposive converb.

\subsection{The non-assertive complement \ird{poby}}\label{sec:poby}

Alongside \ird{po} there stands a parallel construction in \ird{poby}, used
when the embedded content is not presented as declarative content, but as an
open question, unresolved alternative, or indirect polar interrogative of
the \textit{whether} type. Its form is:

\begin{center}
\textbf{[finite embedded clause + linker \ird{-e}] \ird{poby}}
\end{center}

Here again the embedded clause remains fully finite. The opposition between
\ird{po} and \ird{poby} is thus broadly comparable to that between English
\textit{that} and \textit{whether}. One may say, in effect, that \ird{po}
embeds a proposition as content, while \ird{poby} embeds it as an unsettled
issue.

\ex
\begingl
\gla Lobek Marku piaštnice poby Janku houstnik.//
\glb apple.\Dir{} Marek.\Trs{} eat-\Pf{}-\Lnk{} \Compl{} Janek.\Trs{} ask-\Pf{}//
\glft \trsl{Janek asked whether Marek ate the apple.}//
\endgl
\xe

This construction is accordingly found after verbs such as \textit{ask},
\textit{wonder}, \textit{inquire}, \textit{investigate}, \textit{check},
\textit{doubt}, \textit{decide}, and predicates of \textit{knowing whether}.
It is not the ordinary complement of verbs such as \textit{say},
\textit{believe}, \textit{remember}, or \textit{know that}, which require
the declarative \ird{po}-construction.

\subsection{Indirect evidentiality and attributed content}

Finite clauses in Iridian are, in themselves, evidentially unmarked. The
ordinary declarative does not grammatically specify whether the speaker knows
the matter directly or indirectly, though in practice it is often compatible
with direct knowledge. Where the source of knowledge is explicitly indirect,
Iridian employs a special indirect evidential. This marks that the
proposition rests on report, inference, visible result, or post hoc
discovery rather than on firsthand evidence. In suitable contexts a mirative
colour may arise from this meaning, especially where the information is
sudden or contrary to expectation; but the grammatical value remains that of
indirect evidence.

This category must not be confused with the matrixless \ird{po}-construction.
The indirect evidential concerns the source of knowledge. A matrixless
\ird{po}-clause concerns the mode of presentation in discourse. The two
frequently approach one another in use, since hearsay lies near the border of
both; but they are not the same thing.

Morphologically the indirect evidential belongs to the finite verbal complex.
It stands before the complement linker \ird{-e}; person clitics, when
present, also remain inside the embedded finite complex and likewise precede
the linker. The order is therefore as follows:

\begin{center}
\textsc{stem} + \textsc{asp.align} + (\textsc{mood}) + (\textsc{ind.evid}) + (\textsc{person}) + \ird{-e} + \ird{po}/\ird{poby}
\end{center}

If the indirect evidential has the form \ird{-ejte}, then an embedded finite
clause will show:

\begin{center}
\textbf{[finite clause-\ird{ejte}-\ird{-e}] \ird{po}}
\end{center}

Compare the simple finite clause:

\ex
\begingl
\gla Lobek Marku piaštnicejte.//
\glb apple.\Dir{} Marek.\Trs{} eat-\Pf{}-\textsc{ind.evid}//
\glft \trsl{Marek ate the apple, apparently.}//
\endgl
\xe

When the speaker chiefly wishes to indicate that the information is
second-hand, inferred, or otherwise non-firsthand, the indirect evidential is
the ordinary means. When the speaker chiefly wishes to cast the proposition
as attributed discourse, with reportive, distancing, sceptical, ironic, or
echoic effect, the matrixless \ird{po}-construction is the ordinary means.
The two may co-occur, but such co-occurrence is marked, and yields a layered
interpretation: a proposition already marked as indirectly known is then
further presented as attributed content. This is suitable for rumour,
mediated narration, gossip, and similar discourse types, but should not be
treated as the unmarked way of expressing hearsay.

The two categories are therefore to be kept distinct. The indirect
evidential is a grammatical marker of indirect source of knowledge. The
matrixless \ird{po}-clause is a construction of attributed content, whose
pragmatic effects often include distance, reserve, scepticism, irony, or
reduced commitment, but whose core meaning is not evidential in itself.

\section{Fixed expressions with converbs}\label{sec:fixed-converbs}

% ============================================================
% Thanks, apologies, congratulations, greetings that are historically
% grammaticalized converb forms — to be documented.
% ============================================================
