\chapter{The Simple Sentence}\label{ch:simple}

% ============================================================
% STATUS: SCAFFOLD — largely new content; some material available in
% archive/2020/chapters/07-clause-structure.tex.
% Key theoretical framing: the syntax chapters adopt a Minimalist,
% head-final clause spine with a Cartographic left periphery and a
% Distributed Morphology view of spellout. Topic-comment structure is
% important, but it is implemented within that architecture rather than
% replacing it.
% ============================================================

\section{Theoretical background and analytical conventions}
\label{sec:th-background}

The analyses in Chapters~\ref{ch:simple} and~\ref{ch:complex} adopt a
Minimalist architecture in which clause structure is built by Merge,
interpreted through hierarchies of lexical and functional projections, and
linearized according to the strongly head-final character of Iridian. The
goal is not to derive each clause type from construction-specific templates,
but to show how the recurrent properties of the language follow from a
stable structural spine: a predicate domain, a small set of inflectional and
operator projections, and an articulated left periphery.

Within that architecture, the clause edge is treated cartographically. The
high field of the clause is not a single undifferentiated CP, but a sequence
of projections for topic, focus, clause typing, evidential/epistemic
operators, and speech-act marking. This is the machinery that the later
sections of Chapters~\ref{ch:simple} and~\ref{ch:complex} actually use to
analyse topic placement, wh-licensing, polar questions, complementation, and
the ordering of high particles. Information structure is therefore not an
alternative organizing principle to syntax; it is encoded in designated
syntactic positions, especially in \synhead{TopP} and \synhead{FocP}, within a
broader head-final clause spine.

The morphology discussed in these chapters is interpreted in the same way as
elsewhere in the grammar: broadly within a Distributed Morphology model.
Syntactic structure is built first, and the exponence of heads and feature
bundles is determined at spellout. This matters directly for Iridian because
many contrasts that look at first glance like separate `constructions' are
better treated as alternative realisations of the same structural heads:
Neg\textsuperscript{0} surfaces as \ird{zá-}, \ird{staly}, or zero-plus-root
suppletion depending on the host available at PF; converb suffixes realise
feature bundles on non-finite heads; and case and agreement morphology spell
out configurations established in the syntax rather than replacing them.

The central descriptive claim of the two syntax chapters is therefore more
specific than a general appeal to `topic-comment structure'. Iridian clause
structure is organised by a head-final functional sequence in which the
absolutive pivot is strongly privileged for topical interpretation, but topic,
focus, polarity, evidentiality, and clause combining are all represented by
independent structural positions. The split-ergative alignment, the ordering
of particles, the placement of subordinate clauses, the distribution of
wh-phrases, and the complement system are analysed as consequences of that
shared architecture. This section sets out the main assumptions and
terminological conventions used throughout both chapters and relates them to
the particle analysis developed in Chapter~\ref{ch:function-words}.

\subsection{Clause structure and the functional sequence}
\label{sec:th-clause-structure}

The Iridian clause is built up from the predicate outward through a
structured sequence of functional projections. Table~\ref{tab:th-clause-spine}
gives the full spine from innermost to outermost; the projections directly
implicated in the particle system are discussed in detail in
§\,\ref{sec:th-particle-framework}.

\begin{table}[ht]
	\footnotesize
	\caption{The Iridian clause spine (innermost to outermost).}\label{tab:th-clause-spine}
	\medskip
	\begin{tabularx}{\textwidth}{l l L}
		\toprule
		{\sc level} & {\sc projection} & {\sc content} \\
		\midrule
		1 (lowest) & \synhead{PhasP} & Phasal contour: attained state (\ird{v}) vs.\ ongoing interval (\ird{s}) \\
		2 & \synhead{PredP} & Lexical predicate; root + aspect-alignment morphology \\
		3 & \synhead{AspP} & Comment-domain operators: additive, exclusive, affirmative,
		                     interim-sequential \\
		4 & \synhead{NegP} & Clausal negation: \ird{zá-} (prefix, verbal predicates);
		                     \ird{staly} (free morpheme, equative/zero-copula predicates);
		                     suppletive root \ird{zad-} (existential predicates, Neg\textsuperscript{0}
		                     absorbed into root) \\
		5 & \synhead{IP} & Inflection: finiteness, mood \\
		6 & \synhead{FocP} & Focus position; optionally hosts one fronted wh-phrase or other focused constituent \\
		7 & \synhead{PolP} & Interrogative clause typing: \ird{by} (overt) / $\varnothing$ (null) \\
		8 & \synhead{TopP} & Topic: typically the absolutive argument \\
		9 & \synhead{EvidP} & Evidential and epistemic operators:
		                      \ird{oče}, \ird{hlavdy}, \ird{ktady}, \ird{či} \\
		10 (highest) & \synhead{SpeechActP} & Illocutionary and discourse-relational
		                      framing: \ird{nadě}, \ird{mlada}, \ird{jse} \\
		\bottomrule
	\end{tabularx}
\end{table}

\noindent Because Iridian is rigorously head-final, specifiers and
complements both precede their heads. The predicate therefore surfaces last;
projections that scope over it surface progressively earlier in the string.
Reading the clause from left to right is thus equivalent to reading the
functional sequence from the highest (\synhead{SpeechActP}) to the lowest (\synhead{PhasP})
projection. This principle, elaborated in §\,\ref{sec:th-head-finality},
provides the primary explanatory tool for understanding both particle ordering
and clause-combining word order.

One projection in the spine requires special comment. \synhead{NegP} (projection 4)
has a single structural position in all Iridian clause types, but its
phonological realisation varies by predicate type. In verbal clauses,
Neg\textsuperscript{0} is spelled out as the procliticized prefix \ird{zá-},
which attaches to the left edge of the clause-final verbal complex. In
zero-copula equative clauses, where there is no verbal host available,
Neg\textsuperscript{0} is spelled out as the free morpheme \ird{staly},
which surfaces in the canonical clause-final position expected of a
head-final \synhead{NegP} head. In existential clauses, Neg\textsuperscript{0}
triggers suppletive root replacement: the existential root \ird{jec-} is
replaced by \ird{zad-}, and Neg\textsuperscript{0} itself has no
independent PF exponent. These are three phonological reflexes of one
syntactic head, not three independent negation strategies; the structural
position of \synhead{NegP} is the same in all three cases.

Within the zero-copula domain, a further distinction in semantic scope
arises from the type of the predicative complement. When the predicate is
a plain \synhead{DP}, \ird{staly} produces predicational negation. When the predicate
is a participant nominal (a nominalised verb form carrying its own internal
argument structure; see §\,\ref{sec:nmlz-predicative} of
Chapter~\ref{ch:verbs}), the zero-copula equative becomes identificational
and \ird{staly} produces identificational negation --- the denial that the
topic \synhead{DP} is the unique satisfier of the nominalised description, with the
contrastive implication that the correct referent exists. This scope
difference does not require any modification to the clause spine or to the
analysis of \synhead{NegP}; it is a consequence of the semantics of the predication
that \synhead{NegP} scopes over. The full analysis is developed in
§\,\ref{sec:vm-neg-exponents} and §\,\ref{sec:vm-ident-neg} of the verbal
morphosyntax companion and
§\,\ref{sec:clausal-negation} of Chapter~\ref{ch:simple}.

The clause fields are schematized in Figure~\ref{fig:th-clause-fields};
Figure~\ref{fig:th-left-periphery} shows the detailed hierarchical structure
of the left periphery.

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}[font=\footnotesize]
		\node[syn band, minimum width=18mm] (top)  {Topic\\\synhead{TopP}};
		\node[syn band, minimum width=42mm, right=0pt of top]  (op)
			{Speech-act \& evidential field\\\synhead{SpeechActP} / \synhead{EvidP}};
		\node[syn band, minimum width=16mm, right=0pt of op]   (pol)
			{Clause typing\\\synhead{PolP}};
		\node[syn band, minimum width=34mm, right=0pt of pol]  (tp)
			{Comment domain\\\synhead{IP} / \synhead{FocP}};
		\node[syn band, minimum width=20mm, right=0pt of tp]   (asp)
			{Phasal\\\synhead{PhasP}};
		\node[syn band, minimum width=14mm, right=0pt of asp]  (pred)
			{Pred.\\\synhead{PredP}};
	\end{tikzpicture}
	\caption{Clause fields of the Iridian clause.}\label{fig:th-clause-fields}
\end{figure}

\begin{figure}[ht]
	\footnotesize
	\caption{Hierarchical structure of the Iridian left periphery (\synhead{SpeechActP}
	to \synhead{IP}).}\label{fig:th-left-periphery}
	\medskip
	\centering
	\begin{forest}
		syntax tree
		[\synhead{SpeechActP}
			[{SpeechAct\\\ird{nadě} / \ird{mlada} / \ird{jse}}]
			[\synhead{EvidP}
				[{Evid\\\ird{oče} / \ird{hlavdy} / \ird{ktady} / \ird{či}}]
				[\synhead{TopP}
					[{\synhead{DP}\\{\footnotesize topic}}]
					[{Top$'$}
						[Top]
						[\synhead{PolP}
							[{Pol\\\ird{by} / $\varnothing$}]
							[{Pol$'$}
								[Pol]
								[\synhead{FocP}
									[{XP\\{\footnotesize\textit{wh / focus}}}]
									[{Foc$'$}
										[Foc]
										[\synhead{IP} [{\ldots}]]
									]
								]
							]
						]
					]
				]
			]
		]
	\end{forest}
\end{figure}

\subsection{Topic-comment structure as the primary information-structural
dimension}
\label{sec:th-topic-comment-theory}

The Iridian clause is organised around a \term{topic-comment bipartition}.
The \term{topic} --- the constituent the clause is semantically `about' ---
is realised in the absolutive (direct) case and occupies the specifier of
\synhead{TopP}, the highest argument-hosting projection in the left periphery. The
\term{comment} --- the predication made about the topic --- corresponds
broadly to the \synhead{IP}-internal material and the predicate.

Three notions must be kept distinct in this system: \term{subject},
\term{absolutive pivot}, and \term{discourse topic}. Iridian does not require a
language-wide subject relation as the primitive around which clause structure is
organised. The stable cross-clausal privilege belongs instead to the
absolutive/direct argument: it is the argument that most readily serves as the
clausal pivot for topichood, topic drop, and discourse continuation. In the
nominative-domain aspects this pivot is usually the agent or sole argument,
which creates a superficial resemblance to familiar subject-prominent systems.
In the ergative-domain aspects it is instead the patient. The term
\emph{subject} is therefore only a descriptive convenience for those local
configurations in which the absolutive pivot happens also to be the agent; it is
not the organizing primitive of Iridian syntax.

This architecture has consequences for the alignment system. The split-ergative
pattern of Iridian (perfective/retrospective aspects trigger ergative-absolutive
case marking; the nominative aspects trigger nominative-absolutive marking) is
not a syntactically arbitrary quirk: both patterns converge on placing the
discourse-central participant in the absolutive (direct) case, thereby
licensing it as the canonical clause topic. In the perfective/retrospective, the
patient is the typical pivot-topic and therefore receives the absolutive. In
the nominative aspects, the agent or sole participant is the typical
pivot-topic and receives the absolutive accordingly. The aspect system is
therefore partly an information-structural resource, allowing the speaker to
select which argument will serve as the default discourse center
(see §\,\ref{sec:topic-comment} for detailed illustration).

The ergative-domain agent is not thereby removed from the core syntax. It
remains the initiator, controller, and often the most volitional participant of
the event, but its ordinary place is inside the comment domain as a non-pivot
core argument. When specially anchored or fronted, it receives a marked
contrastive, resumptive, or topic-shift interpretation rather than the neutral
clause-topic interpretation associated with the absolutive.

The practical effects of this architecture for word order are described in
§\,\ref{sec:word-order}: the topic-initial, verb-final order is the unmarked
pattern, and any departures from it are driven by information-structural
rather than purely grammatical requirements.

\subsection{Head-finality}
\label{sec:th-head-finality}

Iridian is a rigorously head-final language. Both phrases and clauses are
internally verb-final or head-final: arguments precede their predicates,
modifiers precede their heads, adpositions follow their nominal complements
(as postpositions, cf.\ §\,\ref{sec:adpositions}), and subordinate clauses
precede the matrix predicate. This property, which is statable as a general
principle of phrase structure and applies at every level of embedding, has
several major consequences that recur throughout the syntactic analysis.

For the clause in isolation, head-finality means that the finite verb is
always clause-final in standard Iridian: it is the head of the highest
verbal projection and accordingly surfaces after all its specifiers and
complements. For clause-combining, it means that subordinate clauses
precede the matrix clause whose head they are complements of (see
§\,\ref{sec:th-subordination} below). For the particle system, it means
that particles in higher projections surface earlier in the linear string:
an operator that takes scope over the full \synhead{IP} will surface before the topic
and before \synhead{IP}-internal material, because the specifier of a high projection
must precede the head of a lower one.

Head-finality also has consequences for relativization and
complementation. Relative clauses precede the head noun they modify;
finite complement clauses precede the matrix predicate that embeds them;
and purpose-clause converbs precede the verb they are oriented toward.
These properties are not stipulated separately but follow from the single
head-final phrase structure combined with the clause-spine in
Table~\ref{tab:th-clause-spine}.

\subsection{The particle framework and its integration into the syntax
chapters}
\label{sec:th-particle-framework}

The adverbial particle system, analysed in detail in
§\,\ref{sec:adv-particles} of Chapter~\ref{ch:function-words}, establishes
the main empirical evidence for the left-peripheral structure summarised in
Figures~\ref{fig:th-clause-fields} and \ref{fig:th-left-periphery}. The
particles serve here as syntactic diagnostics: their obligatory ordering
relative to one another, their scope properties, and their ability or
inability to license fragment answers are all direct reflex of the
projections they occupy.

Four structural positions are distinguished in the particle system. The first
three correspond to the three adverbial particle classes of
Chapter~\ref{ch:function-words}; the fourth is \ird{by}, which occupies a
distinct structural level and is treated separately in
§\,\ref{sec:by-pol-head} and §\,\ref{sec:th-polar-questions}.

\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
\item[\upshape Clause-typing operator (\synhead{PolP}).] The interrogative
	particle \ird{by} is the overt exponent of
	\synhead{Pol\textsuperscript{0}}, the head of \synhead{PolP}
	(§\,\ref{sec:by-pol-head}). \synhead{PolP} occupies level~7 in the clause
	spine, above \synhead{FocP} and below \synhead{TopP}. It is structurally
	and functionally distinct from the adverbial particle classes: it is a
	clause-typing projection, not a comment-domain or stance operator. Because
	it is below \synhead{TopP}, \ird{by} surfaces post-topic in the linear
	string (topic $>$ \ird{by} $>$ focus field $>$ predicate). Because
	\synhead{EvidP} and \synhead{SpeechActP} occupy higher positions in the
	spine, their particles scope over the interrogative force introduced by
	\ird{by} and surface to \ird{by}'s left.
\item[\upshape Zone~\textsc{i} (upper peripheral: evidential and speech-act).]
	Speech-act and discourse-relational particles (\ird{nadě}, \ird{mlada},
	\ird{jse}) occupy \synhead{SpeechActP}. Evidential and epistemic particles
	(\ird{oče}, \ird{hlavdy}, \ird{ktady}, \ird{či}) occupy \synhead{EvidP}.
	Both sets are structurally above \synhead{TopP} (levels~9--10 in the
	spine) and scope over the full proposition, including the topic and any
	interrogative operator in \synhead{PolP}.
\item[\upshape Zone~\textsc{ii} (comment-domain).] Operators whose scope is
	restricted to the comment domain --- \ird{och} (additive), \ird{moc}
	(affirmative), \ird{kamo}/\ird{kam} (exclusive), \ird{dlí} (interim
	sequential) --- occupy the \synhead{AspP} tier below \synhead{IP}. They
	are focus-sensitive where applicable and scope over the predicate domain,
	not the full proposition.
\item[\upshape Zone~\textsc{iii} (phasal).] The phasal particles \ird{v}
	(\trsl{already}) and \ird{s} (\trsl{still} / \trsl{not yet}) occupy
	\synhead{PhasP} immediately above \synhead{PredP}. Their scope is strictly
	predicational.
\end{description}

\noindent The surface ordering from leftmost to rightmost ---
Zone~\textsc{i} particles $>$ topic $>$ \ird{by} $>$ Zone~\textsc{ii} $>$
Zone~\textsc{iii} $>$ predicate --- is a direct consequence of the clause
spine combined with head-finality. The important analytical point is that
\ird{by}'s surface proximity to the Zone~\textsc{ii} particles does not
indicate shared constituency: \ird{by} is in \synhead{PolP} between
\synhead{TopP} and \synhead{FocP}, while Zone~\textsc{ii} items are in
\synhead{AspP} below \synhead{IP}. Cross-references from the simple and
complex sentence chapters to §\,\ref{sec:adv-particles} and
§\,\ref{sec:by-pol-head} are intended to be followed for formal motivation.


\section{The left periphery: topics, operators, and clause typing}
\label{sec:th-left-periphery}

\subsection{Topic and topic drop}
\label{sec:th-topic}

The specifier of \synhead{TopP} hosts the clause topic. In canonical declaratives the
topic is realised by the absolutive pivot of the predicate; it surfaces
clause-initially and is invariably in the direct case. Because Iridian permits
pro-drop, the topic position may be phonologically empty when the referent is
recoverable from the discourse. Topic drop is the default in contexts of topical
continuity;
overt topicalization of a pronominal --- especially with the peripheral
particles --- carries contrastive or focus-related force.

When the topic is dropped, Zone~\textsc{i} particles appear at the left
edge of the string without any overt nominal to their left. This is a
phonological consequence of null-topic, not a separate fronting operation:
the particles remain in their regular post-topic positions (§\,\ref{sec:adv-particles}).

\pex
\a
\begingl
\gla Janek zdravime.//
\glb Janek.\Dir{} sleep-\Prog{}//
\glft \trsl{Janek is sleeping.} \lingcontext{canonical topic-initial}//
\endgl
\a
\begingl
\gla Zdravime.//
\glb sleep-\Prog{}//
\glft \trsl{[He] is sleeping.} \lingcontext{topic drop; referent established}//
\endgl
\a
\begingl
\gla Oče zdravime.//
\glb \textsc{mir} sleep-\Prog{}//
\glft \trsl{Oh, [he] is sleeping!} \lingcontext{topic drop with peripheral mirative; \ird{oče} in post-topic position}//
\endgl
\xe

\subsection{Focus and focus movement}
\label{sec:th-focus}

The \synhead{FocP} position immediately below \synhead{PolP} is the target for
focus movement. In the unmarked declarative clause, focus falls in situ on the
rightmost comment-domain constituent immediately before the verb (the
`comment-final' position), consistent with the general word-order tendency
toward late-focus. A focused constituent or wh-phrase may optionally move to
\synhead{SpecFocP}, surfacing after any peripheral particles. This movement is
not syntactically obligatory; it is discourse-driven, signalling marked focus,
contrast, topicalization, or discourse prominence. The absence of overt
movement is therefore the neutral or unmarked option, especially in content
questions, where a wh-phrase may remain in situ (see §\,\ref{sec:th-wh-interrogation}).

It follows from this architecture that \synhead{SpecFocP} is not an exclusively
interrogative position: it is the target of focus movement in general. A
focalized non-wh constituent moves there in much the same way as a wh-phrase.
The functional difference --- between a focused declarative and a content
question --- comes from the clause-level interrogative operator (\ird{by} or
a phonologically null counterpart in \synhead{PolP}), not from the focus
position itself.

\subsection{The clause-typing operator \ird{by} and matrix interrogatives}
\label{sec:th-polar-questions}

Matrix interrogatives project \synhead{PolP}. Pol\textsuperscript{0} may be
spelled out overtly as \ird{by} or left phonologically null. In polar
questions, overt \ird{by} is the ordinary and unmarked explicit strategy. In
content questions, both overt \ird{by} and null Pol\textsuperscript{0} are
standardly available: a wh-question may appear with or without \ird{by}, and
the choice reflects the speaker's discourse position rather than a grammatical
asymmetry between question types. The full discussion of the overt/null
alternation in content questions is in §\,\ref{sec:th-wh-interrogation};
the present subsection focuses on the clause-typing function of \ird{by} itself
and on the properties shared across all matrix interrogatives
(§\,\ref{sec:by-pol-head}).

\pex
\a
\begingl
\gla Marek by zdravime?//
\glb Marek.\Dir{} \textsc{q} sleep-\Prog{}//
\glft \trsl{Is Marek sleeping?}//
\endgl
\a
\begingl
\gla Janek by v ščenik?//
\glb Janek.\Dir{} \textsc{q} already arrive-\Pf{}//
\glft \trsl{Has Janek already arrived?}//
\endgl
\xe

The interaction of \ird{by} with Zone~\textsc{i} particles (evidential and
speech-act operators) follows directly from their relative positions in the
functional hierarchy. Because those particles occupy projections above
\synhead{PolP}, they scope over the interrogative force introduced by \ird{by}:
a mirative question (\ird{oče~by}) presents the polar question as
counterexpectational; a reportative question (\ird{či~by}) presents it as
relayed from another source.

\pex
\a
\begingl
\gla Tak tóm oče by Janku v koupnik?//
\glb this book.\Dir{} \textsc{mir} \textsc{q} Janek.\Trs{} already buy-\Trs{}-\Pf{}//
\glft \trsl{Has Janek, surprisingly, already bought this book?}//
\endgl
\xe

A constituent may be fronted to Spec,\synhead{FocP} or set as a topic in Spec,\synhead{TopP}
inside a polar question, while \ird{by} in Pol\textsuperscript{0} continues to
type the clause as interrogative. The fronting of a non-wh constituent inside a
polar question is not a special interrogative mechanism but ordinary
information structure operating inside an interrogative clause. A direct-case
topic may remain in Spec,\synhead{TopP}; a narrow-focus non-wh constituent may move to
Spec,\synhead{FocP}; either operation is independent of the interrogative clause typing
supplied by \ird{by}. This point is important for understanding argument and
adjunct fronting in questions and prevents conflating topic/focus movement with
the mechanism that licenses interrogative force: interrogative force belongs to
\synhead{PolP}, while focus and topic placement belong to \synhead{FocP} and
\synhead{TopP} respectively.


\section{Content questions and \textit{wh}-interrogation}
\label{sec:th-wh-interrogation}

\subsection{Licensing of content questions: \synhead{PolP} and wh-operators}
\label{sec:th-single-wh}

Content questions are licensed by the combination of an interrogative
operator in \synhead{PolP} and a wh-expression inside the clause. They are
not licensed by movement alone. \textsc{Pol}\textsuperscript{0} may be
phonologically null in neutral content questions, but it is structurally
present as the clause-typing head; the wh-operator within its scope domain
is the interrogative variable that the question is resolved against.

The distributional rule for wh-phrases is as follows. In neutral content
questions, a single wh-expression may remain in situ in its base-generated
argument or adjunct position. One wh-expression may optionally move to
Spec,\synhead{FocP} when assigned marked focus; this movement signals that the
wh-phrase is singled out against an already salient frame, or contributes
contrast, insistence, or other discourse prominence. No clause permits more
than one wh-phrase to front overtly to Spec,\synhead{FocP}.

\pex
\a
\begingl
\gla Marek jena mulša?//
\glb Marek.\Dir{} where live-\Cont{}//
\glft \trsl{Where does Marek live?} \lingcontext{neutral; wh in situ}//
\endgl
\a
\begingl
\gla Jena Marek mulša?//
\glb where Marek.\Dir{} live-\Cont{}//
\glft \trsl{Where does Marek live?} \lingcontext{marked; wh to \synhead{SpecFocP} for discourse prominence}//
\endgl
\a
\begingl
\gla Tak tóm jeda koupnik?//
\glb this book.\Dir{} who buy-\Pf{}//
\glft \trsl{Who bought this book?} \lingcontext{neutral; wh in situ}//
\endgl
\a
\begingl
\gla Jeda tak tóm koupnik?//
\glb who this book.\Dir{} buy-\Pf{}//
\glft \trsl{Who bought this book?} \lingcontext{marked; wh to SpecFocP}//
\endgl
\xe

A fronted wh-phrase does not itself license interrogative force; its
movement signals discourse prominence, contrast, or singled-out status of the
wh-variable. The interrogative clause type is established by the \synhead{PolP} layer
independently of whether any fronting has occurred. A content question is
therefore structurally defined by the conjunction of an interrogative
operator in \synhead{PolP} and a wh-expression within \synhead{IP}, not by the
position of the wh-phrase at the surface.

Matrix content questions permit both overt and null Pol\textsuperscript{0},
and these are \emph{both} standard options, not a default with an emphatic
variant. The two differ in discourse profile. A content question with null
Pol\textsuperscript{0} is the structurally lighter form, appropriately neutral
in ordinary information-seeking contexts. A content question with overt
\ird{by} is more explicitly typed as a direct question: it is more
interactionally foregrounded, conversationally engaged, or oriented toward
a specific addressee in the current exchange. The difference is a pragmatic
tendency, not a semantic or grammatical opposition. Both forms are equally
grammatical in standard Iridian.

The same independence holds between \ird{by} and the fronting of the wh-phrase.
\ird{By} may appear in a content question regardless of whether the wh-phrase
has moved to the left-peripheral focus position or remains in situ. This gives
four available combinations: null Pol with in-situ wh, null Pol with fronted
wh, overt \ird{by} with in-situ wh, and overt \ird{by} with fronted wh. All
four are well-formed; they differ only in the independent contributions of
clause-typing explicitness (\ird{by}/null Pol) and discourse prominence
(fronting/in situ).

The following paradigm illustrates all four for the content question
\trsl{where does Marek live}:

\pex\label{ex:wh-four-combinations}
\a
\begingl
\gla Marek jena mulša?//
\glb Marek.\Dir{} where live-\Cont{}//
\glft \trsl{Where does Marek live?} \lingcontext{null Pol; wh in situ: neutral}//
\endgl
\a
\begingl
\gla Marek jena by mulša?//
\glb Marek.\Dir{} where \textsc{q} live-\Cont{}//
\glft \trsl{Where does Marek live?} \lingcontext{overt \ird{by}; wh in situ: explicit, interactionally foregrounded}//
\endgl
\a
\begingl
\gla Jena Marek mulša?//
\glb where Marek.\Dir{} live-\Cont{}//
\glft \trsl{Where does Marek live?} \lingcontext{null Pol; wh fronted: wh-phrase discourse-prominent}//
\endgl
\a
\begingl
\gla Jena by Marek mulša?//
\glb where \textsc{q} Marek.\Dir{} live-\Cont{}//
\glft \trsl{Where does Marek live?} \lingcontext{overt \ird{by}; wh fronted: both clause and wh-phrase foregrounded}//
\endgl
\xe

\noindent The same four-way system extends to transitive and oblique content
questions. The wh-word type (agent, patient, location, instrument, and so on)
does not restrict the availability of any combination; all four remain possible
across the full inventory of wh-operators in Table~\ref{tab:nm-interrogatives}.

When Zone~\textsc{i} particles are present, they surface above the
wh-phrase (whether in situ or fronted) and above any overt \ird{by}, because
they occupy projections higher in the spine:

\pex
\a
\begingl
\gla Jena hlavdy Marek mulša?//
\glb where \textsc{infr} Marek.\Dir{} live-\Cont{}//
\glft \trsl{Where would Marek presumably be living?}\\
      \lingcontext{\ird{hlavdy} in \synhead{EvidP} above \synhead{PolP}; wh fronted}//
\endgl
\xe

\subsection{Multiple \textit{wh} and in-situ licensing}
\label{sec:th-multiple-wh}

When the clause contains more than one wh-phrase, the ordinary pattern is for
all wh-phrases to remain in situ. At most one wh-phrase may front to
\synhead{SpecFocP}, and only when that phrase is assigned specially marked focus
within the pair-list interpretation. The remaining wh-phrases are interpreted
as interrogative variables in their base positions, licensed by the
clause-level interrogative operator (overt \ird{by} or a null
\synhead{PolP} head).

\pex
\a
\begingl
\gla Jede ježe koupnik?//
\glb who what buy-\Pf{}//
\glft \trsl{Who bought what?}//
\endgl
\a
\begingl
\gla Jede jena mulša?//
\glb who where live-\Cont{}//
\glft \trsl{Who lives where?}//
\endgl
\xe

\noindent In these multiple-wh clauses both phrases remain in their
\synhead{IP}-internal base positions and are bound by the null interrogative operator in
\synhead{PolP}. The general licensing principle is: a wh-word within the scope
domain of the clause-level interrogative operator in \synhead{PolP} is interpreted as an
interrogative variable, regardless of its surface position. Overt movement to
\synhead{SpecFocP} marks the fronted phrase as the salient or foregrounded
variable in the pair-list; it does not itself establish the interrogative type
of the clause, which is always a function of \synhead{PolP}.

\subsection{Tag, echo, and indirect questions}
\label{sec:th-tag-echo}

Tag and echo questions, and indirect questions embedded under
\ird{poby}, are treated in §\,\ref{sec:questions} and
§\,\ref{sec:poby} respectively. For embedded content questions in
\ird{poby}-complements, the same distributional principle applies as in
matrix clauses: the embedded wh-phrase may remain in situ (the neutral
option) or move to the left periphery of the embedded clause for marked
focus. Wh-fronting within the embedded clause's own periphery is therefore
not obligatory; it signals the same kind of discourse prominence as fronting
in matrix questions. The \ird{poby} complementizer marks the embedded clause
as a non-assertive complement regardless of whether the internal wh-phrase
has fronted. The full analysis of \ird{poby} and its complement types is
in §\,\ref{sec:poby}.

\section{Topic--comment structure and information structure}\label{sec:topic-comment}

\subsection{The topic as the logical subject of predication}
\label{sec:tc-logical-subject}

The Iridian clause is organised around a \term{topic-comment bipartition}, and
the account of that bipartition must be precise enough to do real theoretical
work. The informal gloss --- topic is `what the sentence is about' --- is
insufficient, because it conflates two analytically separate relations: the
grammatical relation of subject, which is defined by morphosyntactic properties
such as case, agreement, and pivot behaviour, and the discourse relation of
topichood, which is defined by the semantic relation between a constituent and
the predication made about it. In the framework adopted here, following Kiss
(1995), the \term{topic} is the \term{logical subject of predication}: the
constituent with respect to which the rest of the clause --- the \term{comment}
--- constitutes a semantic predicate, expressing a property that is claimed to
hold of that individual. This is a discourse-semantic role, not a
morphosyntactic one, and it is independent of grammatical subjecthood in the
standard sense.

The independence of topichood from subjecthood is a theoretical necessity in
Iridian because the grammar does not recognise a language-wide grammatical
subject as its organising primitive. The stable privilege belongs instead to the
\term{absolutive pivot}: the direct-case core argument licensed by the alignment
system. In nominative-domain clauses --- the imperfective aspects --- the
absolutive pivot is typically the agent or sole participant. In
ergative-domain clauses --- the perfective and retrospective --- it is the
patient. The same argument position, determined by alignment, is the discourse
centre in both domains; but the semantic role of that centre changes with aspect
domain.

This matters for topichood in the following way. If `topic' were simply a
synonym for `grammatical subject', the Iridian facts would require two different
subjects: an agentive subject in nominative-domain clauses and a patient subject
in ergative-domain clauses. That is not a coherent analysis. Instead, the
grammar treats \term{topic} as a discourse role assigned to a structurally
identified position, Spec,\synhead{TopP}, and the absolutive pivot as the
default occupant of that position across aspect domains. The topic-comment
articulation is therefore prior to the agent--patient asymmetry, not derived
from it.

The distinction can be illustrated immediately with a minimal aspectual pair,
using the verb \ird{kup-} \trsl{buy}:

\pex\label{ex:tc-topic-subj}
\a
\begingl
\gla Marek tak tóme kupujime.//
\glb Marek.\Dir{} \Dem{}.\Prox{}.\Dir{} book.\Ind{} buy-\Cont{}//
\glft \trsl{Marek is buying this book.}
      \lingcontext{nominative domain; agent = absolutive pivot = topic}//
\endgl
\a
\begingl
\gla Tak tóm Marku koupnik.//
\glb \Dem{}.\Prox{}.\Dir{} book.\Dir{} Marek.\Trs{} buy-\Trs{}.\Pf{}//
\glft \trsl{This book, Marek bought.}
      \lingcontext{ergative domain; patient = absolutive pivot = topic}//
\endgl
\xe

\noindent The event participants are constant: Marek is the buyer and the book
is the thing bought. What changes is not the lexical event structure but the
alignment: the nominative-domain continuous places the agent in the direct case
and therefore makes the agent the absolutive pivot and clause topic; the
ergative-domain perfective places the patient in the direct case and makes the
patient the absolutive pivot and clause topic. A notion of `topic' that reduced
to `subject' could not handle this alternation without positing two grammatically
unrelated subject positions. The present analysis handles it as a single topic
mechanism --- the discourse prominence of the absolutive pivot --- interacting
with the aspectual determination of which participant receives the direct case.

The claim that topic is the logical subject of predication also entails that the
\term{comment} is a semantic predicate. In (\ref{ex:tc-topic-subj}a), the topic
is Marek and the comment is the property of being engaged in the act of buying
this book. In (\ref{ex:tc-topic-subj}b), the topic is the book and the comment
is the property of having been bought by Marek. The predication relation in
both cases is the same type, even though the discourse-prominent participant
differs. This is the sense in which the topic-comment structure is the primary
information-structural dimension of the Iridian clause: it organises the clause
into an individual and a property attributed to it, prior to any further
packaging by focus or modal operators.


\subsection{Absolutive pivot, alignment, and default topicality}
\label{sec:tc-abs-align}

The preceding subsection argued that topic and grammatical subject are
independent notions and that the absolutive pivot, not agenthood per se, is the
default discourse centre. This section formalises the connection between the
absolutive and topichood as a named mapping principle and explains its
interaction with aspect domain.

In neutral finite clauses, there is a systematic alignment between
morphological case marking and discourse structure. The direct-case
(absolutive) argument surfaces clause-initially in Spec,\synhead{TopP} and is
interpreted as the clause topic. This is not a semantic entailment --- the
direct-case argument is not logically required to be discourse-prominent ---
but the neutral or unmarked mapping between clause structure and discourse
packaging. The generalisation can be stated as follows:

\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
\item[\upshape Absolutive--Topic Alignment:] In unmarked finite clauses, the
  \synhead{DP} bearing the direct case is the default discourse topic and is
  merged or interpreted in Spec,\synhead{TopP}.
\end{description}

\noindent This principle is not a semantic law and not an inviolable syntactic
constraint. It is the neutral mapping: deviations from it are possible but
marked, and they trigger the resumptive, contrastive, and focus-related
interpretations discussed in §§\,\ref{sec:tc-marked-topics}--\ref{sec:tc-contrastive-topic}.
The principle is best understood as the Iridian instantiation of a more general
interface tendency: participants that are structurally prominent (absolutive
arguments, clause-initial position) tend also to be discourse-prominent.

The interaction with aspect is crucial. Aspect choice in Iridian is partly a
pragmatic decision about which participant is to be construed as the discourse
centre: the speaker selects the aspectual domain that places the intended topic
in absolutive position. The same lexical predicate can construe the same event
with different discourse centres:

\pex\label{ex:tc-aspect-topic}
\a
\begingl
\gla Marek travě piaštime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.}
      \lingcontext{nominative domain; agent = direct = topic}//
\endgl
\a
\begingl
\gla Travat Marku piaštnik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} eat-\Trs{}.\Pf{}//
\glft \trsl{Marek ate the bread.}
      \lingcontext{ergative domain; patient = direct = topic}//
\endgl
\xe

\noindent The alternation is not mere word-order variation. In
(\ref{ex:tc-aspect-topic}a), the aspectual morphology selects the nominative
domain and the agent surfaces in the direct case. The clause is organised around
Marek as its discourse centre: it is a clause about Marek's ongoing activity.
In (\ref{ex:tc-aspect-topic}b), the aspectual morphology selects the ergative
domain and the patient surfaces in the direct case. The clause is now about the
bread: specifically, about what happened to it. The ergative agent is present as
a core participant but is secondary from a discourse perspective --- it occupies
the transitive case and remains within the comment domain unless specially
promoted (§\,\ref{sec:tc-marked-topics}).

In the imperfective aspects (progressive, continuous, contemplative), agent and
topic therefore typically coincide; in the perfective and retrospective,
patient and topic typically coincide. This is one of the most theoretically
consequential properties of the Iridian grammar and is not merely a descriptive
observation about word order. It reflects the interaction of two independently
motivated systems --- the aspectual specification of which participant receives
the absolutive and the discourse mapping that places the absolutive in topic
position --- and it means that the topic-comment system in Iridian is
inextricably bound up with split-ergative alignment (see
§§\,\ref{sec:th-topic-comment-theory}, \ref{sec:th-wh-interrogation}, and
Chapter~\ref{ch:discourse}).


\subsection{The formal properties of ordinary topics}
\label{sec:tc-ordinary-topics}

Not every DP that might conceivably appear clause-initially is a topic, and
not every type of DP is equally available as a topic. Topics in the unmarked
case have specific formal properties, and distinguishing these properties from
those of other clause-initial material is essential for the analysis of the left
periphery as a whole.

An \term{ordinary topic} in Iridian is a \synhead{DP} that is
discourse-identifiable, referential or kind-denoting, and at least weakly
specific or recoverable in context. These properties are not independently
stipulated features; they follow from the semantic role of the topic as the
logical subject of predication. A predication requires an individual or kind
about which something is being said: predicating a property of an entity that
has not been introduced into, or cannot be recovered from, the discourse would
fail to connect the clause to its context of utterance. The ordinary topic is
therefore inherently discourse-anchored, though the nature and degree of that
anchoring may vary by topic type.

Four principal topic types can be distinguished within the class of ordinary
topics. \emph{Anchored referential \synhead{DP}s} are the paradigm case: DPs
bearing the anchoring suffix \ird{-to}, which signals that the referent is
tracked as a specific individual within the discourse. Anchored referential
topics may be discourse-anaphoric, situationally unique, or associated with a
contextually salient set. The anchoring suffix is by no means required on every
topic (§\,\ref{sec:tc-anchoring-interface}), but anchored topics are the most
prototypically discourse-linked type. \emph{Discourse-recoverable null topics}
arise when the topic referent is established by the prior discourse and
Spec,\synhead{TopP} is phonologically empty; these are treated in
§\,\ref{sec:tc-topic-drop}. \emph{Generic and kind-denoting \synhead{DP}s} may
appear as topics when interpreted as kinds or as generic individuals: the
predication is then a characterising statement about the kind, and the topic is
discourse-stable once introduced. \emph{Discourse-linked specific indefinites}
may appear as topics when the referent is being introduced as particular and
contextually distinguishable; in Iridian, the specifically indefinite particles
\ird{si} and \ird{druh} support this function
(§\,\ref{sec:nm-spec}).

The following examples illustrate each type:

\pex\label{ex:tc-topic-types}
\a\label{ex:tc-anch-topic}
\begingl
\gla Marko v zdravime.//
\glb Marek.\Dir{}.\Def{} already sleep-\Prog{}//
\glft \trsl{Marek is already sleeping.} \lingcontext{anchored referential topic}//
\endgl
\a\label{ex:tc-null-topic}
\begingl
\gla By v zdravime?//
\glb \textsc{q} already sleep-\Prog{}//
\glft \trsl{Is he already sleeping?} \lingcontext{null continuing topic; referent established}//
\endgl
\a\label{ex:tc-generic-topic}
\begingl
\gla Vlak drže na hlučnime.//
\glb wolf.\Dir{} night in howl-\Cont{}//
\glft \trsl{Wolves howl at night.} \lingcontext{generic topic; bare plural}//
\endgl
\a\label{ex:tc-spec-indef-topic}
\begingl
\gla druh člověk za dveřmi stojime.//
\glb \Indef{} person.\Dir{} behind door-\Pl{}.\Ind{} stand-\Cont{}//
\glft \trsl{A certain person is standing at the door.}
      \lingcontext{individuated indefinite topic; \ird{druh} marks individuation without anchoring}//
\endgl
\xe

\noindent In (\ref{ex:tc-spec-indef-topic}), the modifier \ird{druh} does not
make the referent discourse-recoverable in the anaphoric sense, but it
individuates the referent as a particular entity with contextual grounds for
its existence. This partial individuation is sufficient to support the
predication relation (§\,\ref{sec:nm-spec}). An unindividuated existential
predication --- the assertion that there is merely some entity, without any
individualising implication --- does not have a topic in this sense; that
construction is treated in §\,\ref{sec:tc-topicless}.


\subsection{Topic position in the clause spine}
\label{sec:tc-clause-spine}

The preceding subsections have presupposed that the topic occupies a designated
structural position. The topic is the \synhead{DP} in Spec,\synhead{TopP},
where \synhead{TopP} is the highest argument-hosting projection in the
articulated left periphery of the Iridian clause. The clause spine, set out in
Table~\ref{tab:th-clause-spine} and Figure~\ref{fig:th-left-periphery}, runs:

\begin{center}
\synhead{SpeechActP} $>$ \synhead{EvidP} $>$ \synhead{TopP} $>$
\synhead{PolP} $>$ \synhead{FocP} $>$ \synhead{IP} $>$
\synhead{PhasP} $>$ \synhead{PredP}
\end{center}

\noindent The topic is in Spec,\synhead{TopP} (projection 8). The comment
corresponds to the domain that \synhead{TopP} scopes over: everything from
\synhead{PolP} downward, including the focus position, the inflectional domain,
and the predicate field. \synhead{TopP} is dominated only by the evidential and
speech-act projections, which are not argumental and do not belong to the
comment.

This architecture has several immediate consequences. Evidential particles such
as \ird{oče} occupy \synhead{EvidP} outside and above the topic, not inside the
comment. The clause-typing particle \ird{by} is in Pol\textsuperscript{0},
below the topic and above the focus position; in polar questions, the topic
may remain in Spec,\synhead{TopP} while \ird{by} takes scope over the
predication within its domain. Focused constituents appear in
Spec,\synhead{FocP}, below the topic and within the comment domain, which
explains why a focused constituent and a topic may co-occur without structural
conflict (§\,\ref{sec:tc-focus-contrast}).

Whether the topic in Spec,\synhead{TopP} arrives there by movement from a
base-generated position within \synhead{IP}, or whether it is base-generated in
that position with its predication role established by a long-distance Agree
relation, is a question deliberately left open here. Both derivational routes
are architecturally consistent with the clause spine
(§\,\ref{sec:th-topic-comment-theory}); the present description is committal
about the surface position of the topic and about the topic-comment
interpretive division, while remaining agnostic about the precise derivational
mechanism pending further evidence from reconstruction and binding asymmetries.


\subsection{Topic versus focus}
\label{sec:tc-focus-contrast}

Topic and focus are distinct operations in the Iridian left periphery and must
not be conflated. The distinction is architectural: \synhead{TopP} is the topic
projection, and \synhead{FocP} --- the position immediately above \synhead{IP}
and below \synhead{PolP} --- is the focus projection. The two positions are not
interchangeable, and the constituents that occupy them have different discourse
functions.

A topic is discourse-given or at least discourse-linkable: it establishes the
individual about which the comment is a predication, and it typically
introduces no new information in itself. A focus, by contrast, is the
informationally or contrastively prominent element within the comment domain:
it identifies the relevant variable in an answer to a question under
discussion, or marks the constituent where the scalar or exhaustive
interpretation is concentrated. The focus is an operator whose scope is
internal to the comment; the topic is the frame within which that comment is
interpreted.

The two may co-occur, since topichood and focus are properties of different
projections:

\pex\label{ex:tc-top-foc}
\a\label{ex:tc-top-only}
\begingl
\gla Marek tak tóma obrojime.//
\glb Marek.\Dir{} \Dem{}.\Prox{}.\Dir{} book.\Ind{} read-\Cont{}//
\glft \trsl{Marek is reading this book.}
      \lingcontext{topic only; focus in situ at comment-final position}//
\endgl
\a\label{ex:tc-top-plus-foc}
\begingl
\gla Marko tak tóma obrojime.//
\glb Marek.\Dir{}.\Def{} \Dem{}.\Prox{}.\Dir{} book.\Dir{} read-\Cont{}//
\glft \trsl{As for Marek, it is this book he is reading.}
      \lingcontext{topic \ird{Marko}; \ird{tak tóm} fronted to Spec,\synhead{FocP}}//
\endgl
\xe

\noindent In (\ref{ex:tc-top-only}), Marek is the topic and the comment
identifies his ongoing activity; the focus is in situ at the comment-final
position (§\,\ref{sec:th-focus}). In (\ref{ex:tc-top-plus-foc}), Marek remains
the topic --- the clause is still \emph{about} Marek --- but \ird{tak tóm} has
been fronted to Spec,\synhead{FocP}, marking it as the focus: specifically this
book is what Marek is reading. The topic is the discourse anchor; the focus is
the marked operator within the comment.

Several further contrasts follow from this architecture. A topic must be
discourse-linkable; a focus need not be. A topic cannot by itself be the locus
of new, unpredicted information; a focus can. A topic takes scope over the
entire comment domain; a focus takes scope over the predicate field within that
domain. These distinctions are maintained throughout the grammar, and material
at the left periphery should always be identified as topic, focus, adverbial,
or clause-typing particle before any structural claim is made about it.


\subsection{Topic versus sentence adverbials and peripheral particles}
\label{sec:tc-adverbials}

One of the most practically important distinctions in the left periphery is
between true topics and other clause-peripheral material. It is tempting to
treat every clause-initial constituent as a topic simply because it precedes
the predicate. This is incorrect. A true topic is an argumental constituent
that stands in a predication relation with the comment: the comment predicates
a property of the topic, and the topic is the semantic individual about which
that predication holds. Sentence adverbials, evidential particles, and
clause-typing operators have entirely different grammatical and discourse
functions and should not be analysed as topics merely because they appear to
the left of the comment domain.

A \term{sentence adverbial} --- a temporal, locative, or modal adjunct that
scopes over the proposition --- is not a constituent of the argument structure
and is not the individual about which the comment predicates something.
\ird{Bych} \trsl{yesterday}, for instance, provides a temporal frame for the
event; the clause does not predicate a property \emph{of yesterday}. Similarly,
\term{evidential particles} such as \ird{oče} occupy \synhead{EvidP}, which is
above \synhead{TopP}: they contribute an epistemic or evidential stance on the
proposition as a whole. These material types are licensed by their own
projections in the left periphery and must be kept analytically separate from
topical material.

The following examples illustrate three distinct clause-initial configurations:

\pex\label{ex:tc-adv-contrast}
\a\label{ex:tc-adv-true-top}
\begingl
\gla Marek by v zdravime?//
\glb Marek.\Dir{} \textsc{q} already sleep-\Prog{}//
\glft \trsl{Is Marek sleeping yet?}
      \lingcontext{true topic; \ird{Marek} in Spec,\synhead{TopP}}//
\endgl
\a\label{ex:tc-adv-adv}
\begingl
\gla Bych by Marek v zdravime?//
\glb yesterday \textsc{q} Marek.\Dir{} already sleep-\Prog{}//
\glft \trsl{Was Marek already asleep yesterday?}
      \lingcontext{temporal adverbial \ird{bych} above \synhead{PolP}; \ird{Marek} is still the topic}//
\endgl
\a\label{ex:tc-adv-evid}
\begingl
\gla Oče Marko v zdravime.//
\glb \textsc{evid} Marek.\Dir{}.\Def{} already sleep-\Prog{}//
\glft \trsl{Apparently Marek is already sleeping.}
      \lingcontext{evidential \ird{oče} in Spec,\synhead{EvidP} above topic; \ird{Marko} is the topic}//
\endgl
\xe

\noindent In (\ref{ex:tc-adv-true-top}), \ird{Marek} is the topic: the clause
predicates a state of him. In (\ref{ex:tc-adv-adv}), \ird{bych} is a temporal
adverbial adjoined above \synhead{PolP}; it provides the temporal frame. The
clause still has a topic --- \ird{Marek} --- which follows the adverbial in the
linear string. In (\ref{ex:tc-adv-evid}), \ird{oče} is the evidential particle
in Spec,\synhead{EvidP}, scoping over the whole proposition including the topic.
The topic is still \ird{Marko}.

The practical consequence is significant: the left edge of an Iridian clause may
contain a stack of material from distinct projections --- a speech-act particle,
then an evidential, then a topic, then a clause-typing particle. Only the
constituent in Spec,\synhead{TopP} is the logical subject of predication. The
remainder should be analysed in their own designated positions.


\subsection{Multiple topics and the hierarchy of the left periphery}
\label{sec:tc-multiple-topics}

The architecture of \synhead{TopP} allows for the possibility that more than
one topic-level projection is active in a clause. Ordinary Iridian clauses have
a single topic in Spec,\synhead{TopP}, and iterated topic structure is not the
default. But the grammar appears to permit a second topic projection in specific
discourse environments, particularly where a \term{frame-setting topic}
establishes a domain of quantification or a contrastive partition, and a lower
\term{clause topic} identifies the individual about whom the comment predicates
something within that frame.

A frame-setting topic is a constituent that restricts the discourse domain in
which the comment is interpreted. It does not name the individual that the
comment is directly about; rather, it sets up a restricted context within which
the clause topic is identified and the comment evaluated. Frame-setting topics
are often formally marked by prosodic independence and, in some contexts, by a
resumptive element within the lower clause that co-refers with the frame:

\pex\label{ex:tc-frame-top}
\a\label{ex:tc-frame-town}
\begingl
\gla Tak dvor Marko v něm bydlime.//
\glb \Dem{}.\Prox{}.\Dir{} town.\Dir{} Marek.\Dir{}.\Def{} already in.it live-\Cont{}//
\glft \trsl{As for this town, Marek already lives in it.}
      \lingcontext{frame topic + clause topic; resumptive \ird{něm} co-refers with frame topic}//
\endgl
\a\label{ex:tc-frame-marek}
\begingl
\gla Marko tak tóm bych čitanik.//
\glb Marek.\Dir{}.\Def{} \Dem{}.\Prox{}.\Dir{} book.\Dir{} yesterday read-\Trs{}.\Pf{}//
\glft \trsl{As for Marek, this book he read yesterday.}
      \lingcontext{frame topic + ergative-domain absolutive topic}//
\endgl
\xe

\noindent In (\ref{ex:tc-frame-town}), \ird{tak dvor} \trsl{this town} is the
frame topic and \ird{Marko} is the clause topic; the resumptive pronoun
\ird{něm} inside the comment co-refers with the frame topic. In
(\ref{ex:tc-frame-marek}), \ird{Marko} is the frame topic and \ird{tak tóm}
is the patient-absolutive topic of the ergative clause, within the frame of
what Marek has done.

The distributional limits of iterated \synhead{TopP} await more detailed
investigation. The structural possibility is licensed by the cartographic
architecture: since \synhead{TopP} is a full projection with a specifier and a
complement, it may in principle be embedded within a higher \synhead{TopP}
whose specifier hosts the frame-setting topic. This analysis is consistent with
evidence from topic continuity and topic shift in extended discourse
(Chapter~\ref{ch:discourse}).


\subsection{Topic drop and null topics}
\label{sec:tc-topic-drop}

Topic drop --- the absence of an overt topic \synhead{DP} when the referent is
established by the prior discourse --- is the default strategy for topic
continuation in Iridian. It is not a separate construction or the result of a
dedicated deletion rule; it is the ordinary null spellout of Spec,\synhead{TopP}
under discourse-conditioned recoverability. The theoretical point is important:
topic drop does not reduce the clause to a structure without a topic projection.
Rather, Spec,\synhead{TopP} is syntactically present and semantically
interpreted --- its referential value is supplied by the discourse context ---
but it receives no phonological exponent.

The licensing condition is discourse recoverability. When the most recently
topicalised referent has not shifted and remains available as the intended
individual, Spec,\synhead{TopP} may be null. When the topic shifts, an overt
\synhead{DP} is typically required to signal the new discourse centre, though a
fronted pronoun with contrastive or resumptive force is also available. The
interaction with Zone~\textsc{i} particles follows directly: since those
particles occupy positions post-topic in the left periphery (§\,\ref{sec:th-topic}),
they appear at the left edge of the string when the topic is null, without any
separate movement being involved.

The following discourse sequence illustrates the pattern:

\pex\label{ex:tc-topdrop}
\a\label{ex:tc-topdrop-overt}
\begingl
\gla Marko domů chodime.//
\glb Marek.\Dir{}.\Def{} home go-\Cont{}//
\glft \trsl{Marek is going home.} \lingcontext{overt topic; referent introduced}//
\endgl
\a\label{ex:tc-topdrop-null1}
\begingl
\gla by v domě jest?//
\glb \textsc{q} already at.home be//
\glft \trsl{Is he home yet?}
      \lingcontext{null topic; referent continues from (\ref{ex:tc-topdrop-overt})}//
\endgl
\a\label{ex:tc-topdrop-null2}
\begingl
\gla ne, ještě cestujime.//
\glb no still travel-\Cont{}//
\glft \trsl{No, he is still travelling.}
      \lingcontext{null topic continues; negation particle \ird{ne}}//
\endgl
\xe

\noindent The null topic in (\ref{ex:tc-topdrop-null1}--\ref{ex:tc-topdrop-null2})
is semantically interpreted as co-referential with \ird{Marko} from
(\ref{ex:tc-topdrop-overt}). No referential shift has occurred, and no overt
\synhead{DP} is needed. The question particle \ird{by} and the phasal particle
\ird{v} appear at the left edge of the string because their structural positions
are post-topic, not because any movement has targeted them.

Topic drop should be distinguished from radical pro-drop in languages that lack
agreement morphology altogether. In Iridian, there is no verbal person
agreement; the recoverability of null topics therefore depends entirely on
discourse tracking and the prior absolutive pivot. The topic most naturally
continued by zero anaphora is the most recent absolutive argument, not simply
the most salient animate participant (§\,\ref{sec:th-topic};
Chapter~\ref{ch:discourse}). This makes the absolutive status of the referent
--- not its agentive or grammatical-subject status --- the crucial factor in
licensing null continuation.


\subsection{Apparent and genuinely topicless clauses}
\label{sec:tc-topicless}

Not every clause without an overt topic \synhead{DP} is structurally topicless.
The preceding section showed that null topics are the ordinary form of topic
continuation under discourse recoverability. A clause that lacks an overt topic
may therefore be either (a) a clause whose topic is syntactically present but
phonologically null, or (b) a clause that genuinely lacks a topic --- one in
which Spec,\synhead{TopP} is not projected, or is projected without an
argument. These two types must be distinguished because they have different
discourse properties and different structural analyses.

\term{Apparent topicless clauses} are clauses in which Spec,\synhead{TopP} is
phonologically null but semantically filled. The null topic has a referential
value recoverable from the discourse; it supports further topic continuation by
zero anaphora; and it behaves structurally like any overt topic with respect to
aspect selection and the ordering of comment-internal material.
Examples~(\ref{ex:tc-topdrop-null1}--\ref{ex:tc-topdrop-null2}) above are
apparent topicless clauses of this type.

\term{Genuinely topicless clauses} are clauses in which the predication does
not have a pre-established discourse individual as its logical subject. These
include existential and presentational constructions --- clauses whose function
is to introduce an entity or situation into the discourse rather than to
predicate a property of an already established individual. In such clauses there
is no prior referent to anchor a topic, and the clause as a whole asserts the
existence or occurrence of something. The existential construction with
\ird{jec-} is the primary example (§\,\ref{sec:existential-constructions}):

\pex\label{ex:tc-topicless}
\a\label{ex:tc-apparent-topicless}
\begingl
\gla by koupnik?//
\glb \textsc{q} buy-\Trs{}.\Pf{}//
\glft \trsl{Did he buy it?}
      \lingcontext{apparent topicless; null topic and null object, both recoverable}//
\endgl
\a\label{ex:tc-pres-topicless}
\begingl
\gla jec druh člověk za dveřmi.//
\glb \Exst{} \Indef{} person.\Dir{} behind door-\Pl{}.\Ind{}//
\glft \trsl{There is a certain person at the door.}
      \lingcontext{genuinely topicless presentational; \ird{druh} individuates; entity introduced into discourse}//
\endgl
\a\label{ex:tc-exist-topicless}
\begingl
\gla jec hluk v dvoře.//
\glb \Exst{} noise.\Dir{} in yard.\Ind{}//
\glft \trsl{There is noise in the yard.}
      \lingcontext{genuinely topicless existential-eventive; situation introduced}//
\endgl
\xe

\noindent In (\ref{ex:tc-apparent-topicless}), both the topic \synhead{DP} and
the object are null but both are recoverable: the buyer and the thing bought
have already been established in the discourse. Spec,\synhead{TopP} is present
but phonologically empty. In (\ref{ex:tc-pres-topicless}--\ref{ex:tc-exist-topicless}),
by contrast, the \ird{jec-} clauses introduce new referents into the discourse.
There is no prior individual about whom the clause predicates a property; the
person at the door and the noise in the yard are the entities being asserted to
exist, not the continuation of an established topic. These clauses do not
project an argumentally filled Spec,\synhead{TopP}.

Genuinely topicless clauses are not defective or marginal; they are a
grammatically well-defined type in Iridian, closely associated with the
\ird{jec-}/\ird{zad-} existential paradigm, with certain environmental
predicates, and with presentational discourse moves that introduce entities for
the first time. The overlap between the existential construction and the
genuinely topicless clause is one of the places where the topic-comment system
and the existential chapter (§\,\ref{sec:existential-constructions}) mutually
support one another rather than remaining separate descriptive domains.


\subsection{Marked topics: non-direct, resumptive, and promoted arguments}
\label{sec:tc-marked-topics}

The Absolutive--Topic Alignment principle (§\,\ref{sec:tc-abs-align})
identifies the direct-case pivot as the default topic. All departures from
this default are marked: they signal that the speaker has a specific discourse
motivation for presenting a non-absolutive participant as the clause's discourse
centre, rather than accepting the mapping that aspect domain provides. Marked
topics are therefore not free variants of ordinary topics; they carry additional
interpretive weight and are associated with specific discourse functions.

The nominal chapter establishes (§§\,\ref{sec:nm-topic-interface}--\ref{sec:nm-topic-oblique})
that anchoring on transitive-case and indirect-case \synhead{DP}s is always more
plainly expressive of contrastive, resumptive, or topic-shift function than
anchoring on direct-case \synhead{DP}s. The reason is structural: a
transitive-case agent is not the absolutive pivot and is not in
Spec,\synhead{TopP} under the neutral mapping; if such an agent bears the
anchoring suffix, the anchoring performs an additional function --- signalling
the agent's individuation or promotion --- beyond the baseline discourse-tracking
role that anchoring has on direct-case \synhead{DP}s.

In the ergative domain, the neutral clause has the patient as absolutive topic.
A speaker who wishes instead to make the agent the discourse centre may either
select a nominative-domain aspect (which makes the agent the absolutive pivot by
aspect choice), or produce a marked clause in which the agent is specially
fronted or anchored. The latter creates a marked agent topic. Compare:

\pex\label{ex:tc-marked-top}
\a\label{ex:tc-neutral-top}
\begingl
\gla Tak tóm Marku koupnik.//
\glb \Dem{}.\Prox{}.\Dir{} book.\Dir{} Marek.\Trs{} buy-\Trs{}.\Pf{}//
\glft \trsl{This book, Marek bought.}
      \lingcontext{neutral ergative clause; patient = absolutive = topic}//
\endgl
\a\label{ex:tc-marked-agent-top}
\begingl
\gla Markuto tak tóm koupnik.//
\glb Marek-\Def{}.\Trs{} \Dem{}.\Prox{}.\Dir{} book.\Dir{} buy-\Trs{}.\Pf{}//
\glft \trsl{As for Marek (himself), he bought this book.}
      \lingcontext{marked: anchored agent in transitive case promoted to topic field}//
\endgl
\xe

\noindent In (\ref{ex:tc-neutral-top}), the book is the absolutive pivot and the
clause topic; Marek occupies the transitive case in the comment domain. In
(\ref{ex:tc-marked-agent-top}), the anchored agent \ird{Markuto} has been
promoted to or interpreted in the topic field, while the book retains the direct
case. This is not the ordinary ergative mapping: the clause signals agent
resumption, contrast, or local topic shift. The interpretive effect is typically
one of emphasis or correction --- it was Marek, not someone else, who bought
this book --- or a switch of discourse tracking onto Marek as the continuing
topic.

A directly comparable construction involving an indirect-case argument is also
available when a recipient or beneficiary is being individuated as the most
discourse-prominent participant. Such clauses are marked by the same criteria
and are developed further in the discourse chapter (Chapter~\ref{ch:discourse}).


\subsection{Contrastive topic}
\label{sec:tc-contrastive-topic}

Contrastive topic is a distinct phenomenon that should not be reduced to
ordinary topic plus incidental emphasis, nor to focus. A \term{contrastive
topic} is a topic bearing a contrastive feature: it establishes the clause as
being about an individual X while simultaneously invoking a salient set of
alternatives to X, implying that the comment may not hold of all members of
that set. The structure is one of partial predications: the clause provides
information about X, and the contrastive framing signals that the corresponding
predication about the alternatives to X remains open or is differently answered.

This differs from focus in a fundamental respect. In a focused clause, the
focus constituent exhaustively identifies the relevant value within the comment
(it is \emph{this book} that Marek is reading, not some other): the
identification is within the predicate domain. In a contrastive topic
construction, the contrastive element is the subject of the predication, not the
identified value within the predicate: it is \emph{Marek} (among others) about
whom this comment holds. The alternatives are not alternatives to the complement
of the predicate but alternatives to the topic individual.

\pex\label{ex:tc-ctop}
\a\label{ex:tc-ctop-ordinary}
\begingl
\gla Marko tak tóme čitajime.//
\glb Marek.\Dir{}.\Def{} \Dem{}.\Prox{}.\Dir{} book.\Ind{} read-\Cont{}//
\glft \trsl{Marek is reading this book.}
      \lingcontext{ordinary topic; no contrastive alternatives invoked}//
\endgl
\a\label{ex:tc-ctop-contr}
\begingl
\gla Marko tak tóme čitajime, ale Janko ne.//
\glb Marek.\Dir{}.\Def{} \Dem{}.\Prox{}.\Dir{} book.\Ind{} read-\Cont{} but Janek.\Dir{}.\Def{} \Neg{}//
\glft \trsl{Marek is reading this book; Janek is not.}
      \lingcontext{contrastive topic \ird{Marko}; Marek contrasted with Janek}//
\endgl
\a\label{ex:tc-ctop-obj}
\begingl
\gla Tak tóm Marku koupnik, ale si tuto list ne.//
\glb \Dem{}.\Prox{}.\Dir{} book.\Dir{} Marek.\Trs{} buy-\Trs{}.\Pf{} but \Sanch{} \Dem{}.\Med{}.\Dir{} letter.\Dir{} \Neg{}//
\glft \trsl{This book, Marek bought; a certain particular letter, he did not.}
      \lingcontext{contrastive topic in ergative-domain clause; \ird{si} speaker-anchors the letter}//
\endgl
\xe

\noindent In (\ref{ex:tc-ctop-ordinary}), Marek is an ordinary topic: the clause
predicates reading of him, but no alternatives are invoked. In
(\ref{ex:tc-ctop-contr}), the contrastive feature on \ird{Marko} signals that
the addressee should consider the contrast with Janek: it is Marek (as opposed
to Janek) about whom this reading predication holds. In (\ref{ex:tc-ctop-obj}),
the contrastive topic is \ird{tak tóm} in an ergative-domain clause: this book
(as opposed to a particular letter) is the thing Marek bought. The contrastive
topic here is the patient-absolutive, the ordinary topic of an ergative clause,
bearing an additional contrastive feature.

In the phonological component of the grammar, contrastive topics typically
receive a distinct prosodic contour, the analysis of which is deferred to the
prosody chapter. What the present section establishes is that contrastive topics
are syntactically and discourse-structurally distinct from both ordinary topics
and from focus, and that Spec,\synhead{TopP} is the structural locus of both
ordinary and contrastive topics, with the contrastive feature as the
distinguishing property.


\subsection{The topic--anchoring interface}
\label{sec:tc-anchoring-interface}

The interaction between clause-level topichood and the nominal anchoring system
is treated in detail in Chapter~\ref{ch:nouns} (§\,\ref{sec:nm-topic-interface}),
where the \term{Topicality--Anchoring Harmony} preference is stated: a
\synhead{DP} bearing the feature [\textsc{+top}] in Spec,\synhead{TopP} is
preferentially realised with the anchoring suffix \ird{-to} in its nominal
functional hierarchy. The present section draws out the descriptive consequence
for the exposition of topic-comment structure.

The preference rests on a straightforward functional logic. Topics are
discourse-salient, discourse-trackable, and identified individuals. These are
precisely the conditions that the anchoring suffix encodes: \ird{-to} signals
that the referent is an anchored discourse token, tracked within the common
ground (§\,\ref{sec:nm-def-semantics}). The strongest co-occurrence of topic and
anchoring is therefore expected with ordinary referential topics. The preference
is weaker with generic and bare nominal topics, where the relevant entity is a
kind rather than a particular, and absent with null topics, which are by
definition recoverable without overt morphology.

A minimal contrast makes the distinction visible:

\pex\label{ex:tc-anch-contrast}
\a\label{ex:tc-anch-overt}
\begingl
\gla Marko v zdravime.//
\glb Marek.\Dir{}.\Def{} already sleep-\Prog{}//
\glft \trsl{Marek is already sleeping.}
      \lingcontext{anchored topic; explicit discourse tracking}//
\endgl
\a\label{ex:tc-anch-bare}
\begingl
\gla Marek v zdravime.//
\glb Marek.\Dir{} already sleep-\Prog{}//
\glft \trsl{Marek is already sleeping.}
      \lingcontext{unanchored topic; less explicit tracking}//
\endgl
\xe

\noindent The two clauses are truth-conditionally equivalent. The difference is
discourse-structural: (\ref{ex:tc-anch-overt}) presents Marek as a trackable
individual whose identity is being flagged in the common ground, whereas
(\ref{ex:tc-anch-bare}) presents him as the clause topic without that additional
marking. The anchoring suffix does not change the propositional content; it
adjusts the degree to which the topic is explicitly registered as a discourse
token. The Topicality--Anchoring Harmony preference is violable: bare nominal
topics, generic topics, and contrastive topics may all appear without overt
anchoring. The preference captures a statistical correlation with a structural
explanation, not a categorical requirement.


\subsection{Summary}
\label{sec:tc-summary}

The analysis of topic-comment structure developed in this section can be
summarised in the following claims:

\begin{enumerate}[wide]
\item The Iridian clause is organised primarily by a \term{topic-comment
  bipartition}. The topic is the logical subject of predication; the comment is
  the semantic predicate attributed to the topic. This articulation is
  independent of grammatical subjecthood.

\item The default topic of a neutral finite clause is the direct-case absolutive
  pivot occupying Spec,\synhead{TopP}: the \term{Absolutive--Topic Alignment}.
  In the imperfective aspects the absolutive is the agent; in the perfective and
  retrospective it is the patient. The topic-comment system is therefore
  inextricably tied to split-ergative alignment.

\item Ordinary topics are discourse-identifiable, referential or kind-denoting,
  and at least weakly specific or recoverable. The four principal types are
  anchored referential \synhead{DP}s, null continuing topics, generics, and
  discourse-linked specific indefinites.

\item \synhead{TopP} is the highest argument-hosting projection in the clause
  spine. The comment corresponds to the domain below \synhead{TopP}. Evidential
  and speech-act projections are above \synhead{TopP} and do not belong to the
  comment.

\item Topic and focus are distinct left-peripheral operations: the topic is the
  discourse anchor; the focus is the informationally or contrastively prominent
  element within the comment. The two may co-occur in their respective
  projections.

\item Sentence adverbials, evidential particles, and clause-typing particles are
  not topics; they occupy their own designated projections.

\item \synhead{TopP} may be iterated in frame-setting and resumptive contexts,
  yielding a frame-setting topic and a lower clause topic.

\item Topic drop is the ordinary null spellout of Spec,\synhead{TopP} under
  discourse recoverability, not the absence of topic structure.

\item A clause without an overt topic may be an apparent topicless clause (null
  topic present, recoverable from discourse) or a genuinely topicless clause (no
  topic projected). Genuinely topicless clauses are associated primarily with
  existential and presentational predicates.

\item Marked topics --- anchored non-direct \synhead{DP}s, promoted agents,
  individuated recipients --- are departures from the Absolutive--Topic
  Alignment signalling resumption, contrast, or switch-tracking.

\item Contrastive topics bear a contrastive feature, invoking alternatives to
  the topic individual; they differ from both ordinary topics and from focus.

\item Topics are preferentially anchored because anchoring and topicality are
  functionally correlated; this preference is strongest with referential topics
  and absent with null topics.
\end{enumerate}

\section{Word order}\label{sec:word-order}

The traditional descriptive shorthand for Iridian word order is \term{SOV}, but
that label should not be mistaken for the structural primitive of the clause. A
more faithful characterization is \term{topic-initial, verb-final}, or more
specifically \term{absolutive-first, verb-final}. The finite verb itself is
rigidly clause-final; the ordering of preverbal nominals is governed by
information structure and alignment rather than by a fixed subject-object
template.

In nominative-domain clauses this topic-initial pattern often happens to look
like SOV, because the agent is also the absolutive pivot. In ergative-domain
clauses, however, the same template yields a patient-topic order of the type
\term{absolutive $>$ ergative $>$ verb}. The deeper generalisation is therefore
not `subject before object' but `pivot-topic before comment, with the finite verb
last'.

\subsection{Lexical manner and frequency adverbs in the preverbal field}
\label{sec:lexical-adverb-position}

Lexical adverbs belong to the comment domain of the clause rather than to the
topic field. In the unmarked structure, derived manner adverbs in \ird{-ě} are
best analysed as low adjuncts adjoined to \synhead{vP} or to the outer \synhead{VP}. Because
Iridian is head-final, these adverbs surface in the preverbal field, below
topic and focus material and outside the fixed hierarchy of preverbal particles
described in Chapter~\ref{ch:function-words}. Their neutral placement is
therefore lower than that of clause-typing or stance particles but still to
the left of the finite verb.

\begin{center}
\term{[\synhead{TopP} Topic [\synhead{IP} \ldots [\synhead{vP} MANNER-ADV [\synhead{VP} Object V ]]]]}
\end{center}

\noindent The neutral surface order in an intransitive clause is accordingly
\term{Adverb $>$ Verb}:

\pex
\a
\begingl
\gla Skorně zmačime.//
\glb quickly run-\Prog{}//
\glft \trsl{I am running quickly.}//
\endgl
\xe

\noindent In a transitive clause, where the object remains in the low
complement domain, the neutral order is \term{Adverb $>$ Object $>$ Verb}:

\pex
\a
\begingl
\gla Skorně bylam piaštébime.//
\glb quickly child.\Dir{} eat-\Appl{}-\Prog{}//
\glft \trsl{I am feeding the child quickly.}//
\endgl
\xe

\noindent Frequency adverbs such as \ird{ohly} \trsl{sometimes},
\ird{roca} \trsl{often}, \ird{člíně} \trsl{infrequently}, and \ird{sabo}
\trsl{never} belong to the same broad lexical-adverb domain, though their
semantics are higher and more event-quantificational than those of pure manner
adverbs. Even so, they are not members of the rigid particle series. More
leftward placement of lexical adverbs is possible under discourse pressure, but
such ordering belongs to marked information structure and should not be taken
as the neutral pattern.

% [Examples: SOV default, pragmatic reordering, focus movement]

\section{The copula}\label{sec:copula}

Iridian distributes equative and classificatory predication across two
constructions that differ both structurally and semantically. The default
strategy is the \term{zero copula}: a null functional head that licenses a bare
equative predication without event structure. The second strategy is the overt
marked copular verb \ird{neh-}, which introduces an event of transition and
projects a full verbal spine. The two constructions are not merely stylistic
variants; they encode a principled semantic contrast between atemporal
classification on the one hand and entering a classification on the other.

Property predication --- the expression of adjectival meanings such as
\trsl{be beautiful} or \trsl{be good} --- belongs to the stative verb system and
is not treated here; see Chapters~\ref{ch:overview} and \ref{ch:verbs}. Existential
predication through the \ird{jec-}/\ird{zad-} pair is structurally distinct and
treated separately in §\,\ref{sec:existential-constructions}. The
equative strategies, their structural type, negation, and temporal range are
summarised in Table~\ref{tab:copula-overview}.

\begin{table}[ht]
  \footnotesize
  \caption{Equative and classificatory predication strategies.}\label{tab:copula-overview}
  \medskip
  \begin{tabularx}{\textwidth}{L L L L}
    \toprule
    \textsc{predication type} & \textsc{construction} &
    \textsc{negation} & \textsc{temporal range} \\
    \midrule
    Equative / class-membership & zero copula
      ($\varnothing$Cop\textsuperscript{0}) &
      \ird{staly} (free morpheme) & present / atemporal \\
    Transitional / inchoative & \ird{neh-} (overt verb) &
      \ird{zá-neh-} (prefix) & full aspect paradigm \\
    Property / stative & stative verb &
      \ird{zá-} + stative verb & full aspect paradigm \\
    \bottomrule
  \end{tabularx}
\end{table}

\subsection{Equative and class-membership clauses: the zero copula}%
\label{sec:zero-copula}

\subsubsection{The construction and its distribution}

The default pattern for equative clauses --- those asserting identity, role, or
class membership --- is zero copula. The topic \synhead{DP} occupies the initial position
and the predicative complement appears in the direct case in clause-final
position; no overt verb is present. This is the stylistically neutral, unmarked
construction.

\pex
\a
\begingl
\gla Marek doktor.//
\glb Marek.\Dir{} doctor.\Dir{}//
\glft \trsl{Marek is a doctor.}//
\endgl
\a
\begingl
\gla Bách němscvení kompozitor.//
\glb Bach.\Dir{} German.\textsc{attr} composer.\Dir{}//
\glft \trsl{Bach is a German composer.}//
\endgl
\xe

Inclusive and class-membership predications follow the same zero-copula pattern
and carry no inference beyond simple predication. The predicative complement
always appears in the direct case regardless of which case would otherwise be
assigned in the same structural position.

\subsubsection{Structural analysis: \synhead{EquP} and the absence of AspP}

The zero copula is not merely absent from the surface string; it is the
phonologically null exponent of a dedicated functional head,
Cop\textsuperscript{0}, which projects an \term{equative phrase} (\synhead{EquP}). \synhead{EquP}
is the functional layer that licenses the predication relation between the topic
\synhead{DP} and the predicative \synhead{DP}. Crucially, \synhead{EquP} is \emph{not} selected by
Asp\textsuperscript{0}: no aspectual projection is present in the
zero-copula clause.

The reason is structural. Aspect requires an eventuality --- a state, process,
or transition --- to scope over. Equative predications of the type \trsl{Marek
is a doctor} or \trsl{Bach is a composer} express a relation between two
individuals, not an eventuality in any aspectual sense. There is no event of
being-a-doctor that could be viewed as complete, ongoing, or prospective in the
way that eating or arriving can. In the absence of an eventuality, \synhead{AspP} cannot
be projected, and \synhead{EquP} is selected directly by \synhead{NegP} (when present) or by the
higher \synhead{IP} layer. The affirmative clause structure is:

\begin{center}
\begin{forest}
  syntax tree
  [\synhead{EquP}
    [{\synhead{DP}\\\textit{Marek}\\{\footnotesize\textsc{[dir]}}\\{\scriptsize topic}}]
    [{Eq$'$}
      [{\synhead{DP}\\\textit{doktor}\\{\footnotesize\textsc{[dir]}}\\{\scriptsize pred.}}]
      [{Eq$^{0}$\\{$\varnothing$}}]
    ]
  ]
\end{forest}
\end{center}

The two DPs bear the direct case as the default case assigned under \synhead{EquP}: both
the topic and the predicative complement are direct-case nominals within a
caseless, event-free predication domain.

\subsubsection{Temporal restriction}

The temporal corollary of the absence of \synhead{AspP} is that the zero-copula
construction is restricted to \term{present-time} and \term{atemporal} or
\term{generic} predications. A sentence like \ird{Marek doktor} is interpreted
as holding at the moment of utterance or as a classification that transcends any
particular time. It cannot, without additional context, be interpreted as a past
or future statement. This restriction follows directly from the lack of any
tense or aspect projection: there is no functional head in the clause that could
anchor the predication to a non-present time.

For past equative predications, \ird{neh-} in the retrospective (\ird{nac}) is
used; see §\,\ref{sec:neh-past}. For future-oriented predictions, the
contemplative of \ird{neh-} (\ird{nách}) is available. The zero copula's
restriction to present and generic contexts thus parallels the zero-copula
pattern found in Russian, Arabic, and Hebrew, where the copula is similarly
absent in the present indicative but required in other tenses --- though in
Iridian the structural basis is the absence of \synhead{EquP}'s eventuality rather than
the suppression of a tensed copula.

\subsubsection{Negation: \ird{staly} as the canonical Neg\textsuperscript{0} exponent}

When a zero-copula equative is negated, \ird{staly} appears in clause-final
position. The verbal negative prefix \ird{zá-} does not arise: it requires a
phonological host within the verbal complex, and no such host is present.

\pex
\a
\begingl
\gla Marek doktor staly.//
\glb Marek.\Dir{} doctor.\Dir{} \NegCop{}//
\glft \trsl{Marek is not a doctor.}//
\endgl
\a
\begingl
\gla Doktor Marek staly.//
\glb doctor.\Dir{} Marek.\Dir{} \NegCop{}//
\glft \trsl{Marek is not the doctor.}//
\endgl
\xe

\noindent The two examples differ in which \synhead{DP} occupies the topic position,
not in the position of \ird{staly}. The semantic contrast --- \trsl{not a
doctor} (predicational) vs.\ \trsl{not the doctor} (identificational) --- arises
from topicalization, not from any change in the negation structure.

Structurally, \ird{staly} is the phonological exponent of
Neg\textsuperscript{0} in equative contexts. It bears no aspect marking, has no
paradigmatic alternants, and cannot appear in any non-negative environment ---
all properties consistent with its status as a functional head rather than a
verbal root. Its clause-final position is not exceptional: as the head of \synhead{NegP}
in a head-final language, it follows its complement (the \synhead{EquP} predication). \synhead{NegP}
selects \synhead{EquP} directly, with no intervening \synhead{AspP}:

\begin{center}
\begin{forest}
  syntax tree
  [\synhead{NegP}
    [{\synhead{DP}\\\textit{Marek}\\{\footnotesize\textsc{[dir]}}\\{\scriptsize topic}}]
    [{Neg$'$}
      [{Neg$^{0}$\\\ird{staly}}]
      [\synhead{EquP}
        [{\synhead{DP}\\\textit{doktor}\\{\footnotesize\textsc{[dir]}}\\{\scriptsize pred.}}]
        [{Eq$'$}
          [{Eq$^{0}$\\{$\varnothing$}}]
        ]
      ]
    ]
  ]
\end{forest}
\end{center}

\noindent The topic \synhead{DP} occupies Spec,\synhead{NegP}; the predicative \synhead{DP} remains within
\synhead{EquP} as the complement of Eq\textsuperscript{0}.

This framing inverts the usual presentation of \ird{staly} as an `exceptional'
or `restricted' negation strategy. It is in fact the \emph{canonical} exponent
of Neg\textsuperscript{0}: it surfaces in exactly the position the clause spine
predicts for a head-final \synhead{NegP} head. The prefix \ird{zá-}, by contrast, is the
PF-conditioned variant --- it arises only when a verbal complex is present and
provides a cliticization host. When no such host exists, as in zero-copula
clauses, Neg\textsuperscript{0} surfaces freely in its default head-final
position as \ird{staly}. The same logic applies cross-linguistically: Turkish
\ird{değil} and Swahili \ird{si} occupy the same structural slot in their
respective zero-copula systems, surfacing as invariant clause-final (or
predicate-final) negative heads in the absence of a verbal host. See
§\,\ref{sec:clausal-negation} for the unified treatment across all predicate
types, and §\,\ref{sec:vm-neg-exponents} of Chapter~\ref{ch:verbs} for the
formal Distributed Morphology account.

When \ird{neh-} is present --- that is, when the copular predication is
expressed through the overt transitional verb rather than the zero copula ---
the construction has a verbal host, and negation therefore follows the standard
verbal pattern: \ird{zá-} procliticizes onto \ird{neh-}. \ird{Staly} does not
arise in this context not because it is `blocked' by \ird{neh-} but because
the condition for its appearance --- absence of a verbal host --- is not met.

\subsection{Identificational equatives and participant nominals}%
\label{sec:ident-equative}

When the predicative complement of a zero-copula equative clause is a
\term{participant nominal} (a nominalised verb form, §\,\ref{sec:nmlz-predicative}
of Chapter~\ref{ch:verbs}), the construction is \term{identificational}: the
topic \synhead{DP} is asserted to be the unique referent satisfying the description encoded
in the nominal. The schematic contrast with a plain verbal assertion is:

\pex
\a
\begingl
\gla Lobek Janku piaštnik.//
\glb apple.\Dir{} Janek.\Trs{} eat-\textsc{trs}-\Pf{}//
\glft \trsl{Janek ate the apple.} \lingcontext{verbal clause: propositional assertion}//
\endgl
\a
\begingl
\gla Lobek Jancí piaštnikou.//
\glb apple.\Dir{} Janek.\Ind{} eat-\textsc{trs}-\Pf{}-\Nmlz{}//
\glft \trsl{The apple is what Janek ate.} \lingcontext{identificational equative: patient nominal}//
\endgl
\xe

\noindent In (b), the apple is not merely a participant in the event but is
being identified as the unique entity satisfying the nominalised description. An
agent-nominal equative identifies the topic as the agent of the nominalised event:

\ex
\begingl
\gla Janek lubcí piaščikou.//
\glb Janek.\Dir{} apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{}//
\glft \trsl{Janek is the one who ate the apple.}//
\endgl
\xe

\noindent All internal arguments of the nominalised event appear in the indirect
case (\Ind{}) regardless of the case they would bear in the finite verbal clause:
the agent of the transitive verb (normally \Trs{}) and the patient of the
antipassive-derived agent nominal (normally suppressed in finite contexts) both
appear as \Ind{} arguments within the nominal predicate.

\paragraph{Negation of identificational equatives.}
When an identificational equative is negated with \ird{staly}, the result is
\term{identificational negation}: the topic \synhead{DP} is denied as the unique satisfier
of the description, with the strong implication that some other referent is the
actual satisfier. This contrasts systematically with verbal propositional negation:

\pex
\a
\begingl
\gla Lobek Janku zápiaštnik.//
\glb apple.\Dir{} Janek.\Trs{} \Neg{}-eat-\textsc{trs}-\Pf{}//
\glft \trsl{Janek did not eat the apple.} \lingcontext{propositional negation: denies that the event occurred}//
\endgl
\a
\begingl
\gla Lobek Jancí piaštnikou staly.//
\glb apple.\Dir{} Janek.\Ind{} eat-\textsc{trs}-\Pf{}-\Nmlz{} \NegCop{}//
\glft \trsl{The apple is not what Janek ate.} \lingcontext{identificational negation: event presupposed; identification denied}//
\endgl
\xe

\pex
\a
\begingl
\gla Janek lubcí piaščikou staly.//
\glb Janek.\Dir{} apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{} \NegCop{}//
\glft \trsl{Janek is not the one who ate the apple.} \lingcontext{agent-identification denied: it was someone else}//
\endgl
\xe

\noindent Example (a) of the first set is compatible with the eating not having
occurred at all. Examples using \ird{staly} presuppose that the eating did occur
and assert that the identified entity is the wrong one: \ird{piaštnikou staly}
says that the apple is misidentified as Janek's eating; \ird{piaščikou staly}
says that Janek is misidentified as the eater. In both cases the implication
that the correct referent exists --- that something else was eaten, or that
someone else ate the apple --- arises from the exhaustivity of the identificational
predication itself rather than from any special property of \ird{staly}.

\ird{Staly} is the same Neg\textsuperscript{0} head here as in plain equative
negation. The contrastive/focus force is a semantic consequence of negating an
identificational (exhaustive) predication, not a distinct grammatical function
of \ird{staly}. See §\,\ref{sec:clausal-negation} for the unified structural
account, §\,\ref{sec:vm-neg-exponents} of Chapter~\ref{ch:verbs} for the
formal DM analysis, and §\,\ref{sec:vm-ident-neg} of Chapter~\ref{ch:verbs}
for the full scope-theoretic analysis including the formal semantics of
identificational negation and the contrast table with propositional \ird{zá-}.

\subsection{The overt copular verb \ird{neh-}: transition and inchoative
predication}\label{sec:neh}

\subsubsection{Status and structural contrast with the zero copula}

The root \ird{neh-} is the overt copular verb of Iridian. Its most basic
function is to encode \term{transition}: the topic \synhead{DP} is asserted to be moving
from one classification or condition into another, or to have crossed a threshold
between them. This transitional meaning is not a pragmatic implication but a
semantic contribution of the root itself.

The structural difference between the zero copula and \ird{neh-} is not merely
that one is overt and the other silent. The zero copula projects \synhead{EquP}, a
non-eventive predication layer with no \synhead{AspP}. \ird{Neh-}, by contrast, projects a
proper verbal predicate: it introduces a \term{transition event} --- an
eventuality whose onset, duration, and completion can be aspectually
individuated. This eventuality licenses Asp\textsuperscript{0}, which selects
the \synhead{VP}, and the full verbal spine (\synhead{VP} $\subset$ \synhead{AspP} $\subset$ \synhead{NegP}) is
projected in the usual way. The two structures contrast as follows:

\begin{center}
\begin{minipage}[t]{0.42\textwidth}
\centering
\textit{Zero copula:}\\[6pt]
\begin{forest}
  syntax tree
  [\synhead{EquP}
    [{\synhead{DP}\\{\footnotesize topic}}]
    [{Eq$'$}
      [{\synhead{DP}\\{\footnotesize pred.}}]
      [{Eq$^{0}$\\{$\varnothing$}}]
    ]
  ]
\end{forest}
\end{minipage}%
\hfill
\begin{minipage}[t]{0.52\textwidth}
\centering
\textit{Overt copula:}\\[6pt]
\begin{forest}
  syntax tree
  [\synhead{AspP}
    [{\synhead{DP}\\{\footnotesize topic}}]
    [{Asp$'$}
      [{Asp$^{0}$}]
      [\synhead{VP}
        [{\synhead{DP}\\{\footnotesize pred.}}]
        [{\ird{neh-}}]
      ]
    ]
  ]
\end{forest}
\end{minipage}
\end{center}

\noindent The consequence for negation is immediate: since \ird{neh-} is a
verbal root embedded under \synhead{AspP}, the standard verbal negation strategy applies.
Neg\textsuperscript{0} is spelled out as the prefix \ird{zá-}, which
procliticizes to the left edge of the inflected form. \ird{Staly} does not
arise because the condition for its appearance --- the absence of a verbal host
--- is not satisfied.

\subsubsection{Paradigm}

The paradigm of \ird{neh-} is strongly suppletive and morphologically irregular;
principal forms are given in Table~\ref{tab:neh-paradigm}.

\begin{table}[ht]
  \footnotesize
  \caption{Principal forms of \ird{neh-} (overt copula).}\label{tab:neh-paradigm}
  \medskip
  \begin{tabularx}{0.84\textwidth}{L l l}
    \toprule
    {\sc form} & {\sc affirmative} & {\sc negative} \\
    \midrule
    Continuous          & \ird{neha}   & \ird{záneha} \\
    Progressive         & \ird{nejme}  & \ird{zánejme} \\
    Contemplative       & \ird{nách}   & \ird{zánách} \\
    Perfective          & \ird{nek}    & \ird{zánek} \\
    Retrospective       & \ird{nac}   & \ird{zánac}\\
    \bottomrule
  \end{tabularx}
\end{table}

\noindent The negative forms are entirely regular: \ird{zá-} (reduced to
\ird{za-} before unstressed syllables) procliticizes onto the corresponding
affirmative form. No suppletive negative root exists for \ird{neh-}; the
formation parallels that of all other finite verbs.

\subsubsection{Aspectual semantics}

\ird{Neh-} introduces a transition event, and its aspectual forms characterise
different perspectives on that event in the usual way.

The \textsc{continuous} (\ird{neha}) presents the transition as ongoing and
unbounded: the subject is in the process of moving into the new classification.
The inchoative flavour is inherent to \ird{neh-} itself, not to the continuous
aspect.

\pex
\a
\begingl
\gla Marek doktor neha.//
\glb Marek.\Dir{} doctor.\Dir{} \Cop{}-\Cont{}//
\glft \trsl{Marek is becoming a doctor.}//
\endgl
\a
\begingl
\gla Marek doktor záneha.//
\glb Marek.\Dir{} doctor.\Dir{} \Neg{}-\Cop{}-\Cont{}//
\glft \trsl{Marek is not becoming a doctor.}//
\endgl
\xe

The \textsc{progressive} (\ird{nejme}) adds an agentive or effortful dimension:
the subject is actively engaged in the process of transition. Where the
continuous reports a change in progress, the progressive foregrounds the subject's
directed activity toward the new state.

\ex
\begingl
\gla Marek doktor nejme.//
\glb Marek.\Dir{} doctor.\Dir{} \Cop{}-\Prog{}//
\glft \trsl{Marek is working toward becoming a doctor.}//
\endgl
\xe

The \Ctpl{} (\ird{nách}) locates the transition in prospective
time: the subject will enter the new classification. It is neutral between
imminent and more remote futures.

\ex
\begingl
\gla Marek doktor nách.//
\glb Marek.\Dir{} doctor.\Dir{} \Cop{}-\Ctpl{}//
\glft \trsl{Marek will become a doctor.}//
\endgl
\xe

The \textsc{perfective} (\ird{nek}) presents the transition as a completed
event: the subject has crossed the threshold and entered the new state. The
resulting state is typically understood as still holding unless contextually
cancelled.

\pex
\a
\begingl
\gla Tak byl kapitan nek.//
\glb \Dem{}.\Prox{} child.\Dir{} captain.\Dir{} \Cop{}-\Pf{}//
\glft \trsl{This child became captain.}//
\endgl
\a
\begingl
\gla Tak byl kapitan zánek.//
\glb \Dem{}.\Prox{} child.\Dir{} captain.\Dir{} \Neg{}-\Cop{}-\Pf{}//
\glft \trsl{This child did not become captain.}//
\endgl
\xe

\subsubsection{The retrospective and past equative predication}%
\label{sec:neh-past}

The \textsc{retrospective} (\ird{nac}) presents the transition as situated in
a past frame and viewed from the current moment. Its semantics are
aspectually neutral between the transitional reading (the subject entered the
state at some point in the past) and the stative reading (the subject was in
the state during the relevant past period).

\pex
\a
\begingl
\gla Marek doktor nac.//
\glb Marek.\Dir{} doctor.\Dir{} \Cop{}-\Ret{}//
\glft \trsl{Marek was a doctor. / Marek had been a doctor.}//
\endgl
\a
\begingl
\gla Marek doktor zánac.//
\glb Marek.\Dir{} doctor.\Dir{} \Neg{}-\Cop{}-\Ret{}//
\glft \trsl{Marek was not a doctor.}//
\endgl
\xe

This dual function of the retrospective is a direct consequence of the
temporal restriction on the zero copula. Since the zero copula is confined to
present-time and atemporal predications, past equative contexts require an overt
verb, and \ird{nac} is the form that fills this role. The transitional
implication that \ird{neh-} normally carries is therefore weakened or absent
in retrospective equatives: \ird{Marek doktor nac} does not necessarily imply
that Marek underwent a process of becoming a doctor, only that the doctor
classification held in the past. The implication is a pragmatic default, not a
semantic entailment of the retrospective form, and is readily cancelled by context.

Whether the past classification still holds is not encoded. \ird{Marek doktor
nac} is compatible both with Marek's remaining a doctor and with his having
ceased to be one; the current-state inference, if present, is a conversational
implicature from the choice of a past-framed form over the zero-copula present.

The sequential converb \ird{nesu} --- colloquially \ird{nu} --- is a suppletive
form established independently of the regular aspect paradigm. Its
grammaticalisation as a discourse particle is described in
Section~\ref{sec:seq-copula} of Chapter~\ref{ch:complex}.

\section{Existential constructions}
\label{sec:existential-constructions}

\subsection{Overview}
\label{sec:existential-overview}%
\label{sec:jec}% retained for cross-reference compatibility
\label{sec:zad}% retained for cross-reference compatibility
\label{sec:jec-zad}% retained for cross-reference compatibility

Iridian expresses existential predication through a pair of suppletive verbal
roots that together constitute the affirmative/negative alternation of a single
functional head, Exist\textsuperscript{0}. The affirmative root \ird{jec-}
asserts the existence or presence of a referent; the negative root \ird{zad-}
denies it. Both inflect for the full aspect system; both are morphologically
irregular and suppletive, with the perfective and retrospective built on
distinct suppletive stems. The two paradigms are given together in
Table~\ref{tab:jec-zad-paradigm}.

\begin{table}[ht]
  \footnotesize
  \caption{Principal forms of \ird{jec-} (\textsc{exst}) and \ird{zad-} (\textsc{neg.exst}).}\label{tab:jec-zad-paradigm}
  \medskip
  \begin{tabularx}{0.88\textwidth}{L l l}
    \toprule
    {\sc form} & \ird{jec-} (\textsc{exst}) & \ird{zad-} (\textsc{neg.exst}) \\
    \midrule
    Continuous          & \ird{jeca}   & \ird{zada} \\
    Progressive         & \ird{jecime} & \ird{zdíme} \\
    Contemplative       & \ird{jecach} & \ird{zadach} \\
    Perfective          & \ird{jak}    & \ird{zděk} \\
    Retrospective       & \ird{ján}   & \ird{znení} \\
    \bottomrule
  \end{tabularx}
\end{table}

The two roots are not independent lexical items but are best understood as
suppletive allomorphs of the same Exist\textsuperscript{0} head: \ird{jec-} is
the default exponent, inserted when \synhead{NegP} is absent from the derivation;
\ird{zad-} replaces it when \synhead{NegP} is present, absorbing the negative feature
entirely and leaving Neg\textsuperscript{0} with no independent PF reflex. The
structural and distributional consequences of this analysis are developed in
§\,\ref{sec:neg-exst-alternation} and §\,\ref{sec:vm-jec-zad} below. The
division of labour between \ird{jec-} and the zero-copula construction is
loosely parallel to the Japanese distinction between the existential verbs
\textit{iru} and \textit{aru} on the one hand and the equative
\textit{da}/\textit{desu} on the other: \ird{jec-} covers existence and
presence; zero copula covers identity and class membership.

Iridian has three main productive uses of the existential construction: to
assert existence or presence, including location in a particular place
(\S\,\ref{sec:existence-and-presence}), to express possession
(\S\,\ref{sec:existence-possession}), and certain idiomatic presentational
constructions (\S\,\ref{sec:existential-presentatives}).


\subsection{Existence and presence}\label{sec:existence-and-presence}

The primary use of \ird{jec-} is to assert that a referent exists or is present
at the moment of utterance or within a contextually given frame. The entity in
question stands in the direct case. \ird{Zad-} forms the direct negative
counterpart, asserting non-existence or absence under the same structural
conditions.

\pex
\a
\begingl
\gla Byl jeca.//
\glb child.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{There is a child. / A child is present.}//
\endgl
\a
\begingl
\gla Byl zada.//
\glb child.\Dir{} \NegExst{}-\Cont{}//
\glft \trsl{There is no child. / No child is present.}//
\endgl
\xe

\pex
\a
\begingl
\gla Kroumaště na pivo s jeca.//
\glb refrigerator.\Ind{} in beer.\Dir{} still \Exst{}-\Cont{}//
\glft \trsl{There is still (some) beer left in the refrigerator.}//
\endgl
\a
\begingl
\gla Kroumaště na pivo zada.//
\glb refrigerator.\Ind{} in beer.\Dir{} \NegExst{}-\Cont{}//
\glft \trsl{There is no beer in the refrigerator.}//
\endgl
\xe

Within the Minimalist framework, \ird{jec-} is the phonological exponent of a
dedicated functional head, \term{Exist}\textsuperscript{0}, which projects
\term{ExistP} as the innermost stratum of the clause spine --- the slot occupied
by root \synhead{VP} in ordinary verbal clauses. The sole argument it licenses is the
\term{existential pivot}: the \synhead{DP} asserted to exist, bearing the direct case as
the complement of Exist\textsuperscript{0}. ExistP is then selected by
Asp\textsuperscript{0}, which contributes the aspect suffix, and feeds the
higher functional sequence including \synhead{NegP}. When \synhead{NegP} is absent,
Exist\textsuperscript{0} is spelled out as \ird{jec-}; when \synhead{NegP} is present,
suppletive Vocabulary Insertion replaces \ird{jec-} with \ird{zad-}, leaving
Neg\textsuperscript{0} with no independent PF exponent. The affirmative and
negative structures of \ird{Byl jeca} and \ird{Byl zada} are shown below:

\begin{center}
\begin{forest}
  syntax tree
  [\synhead{AspP}
    [{\synhead{DP}\\\textit{byl}\\{\footnotesize child.\Dir{}}\\{\scriptsize pivot}}]
    [{Asp$'$}
      [{Asp\\\ird{-a}\\{\footnotesize\textsc{[cont]}}}]
      [ExistP
        [{Exist$'$}
          [{Exist\\\ird{jec-}}]
        ]
      ]
    ]
  ]
\end{forest}
\hspace{2em}
\begin{forest}
  syntax tree
  [\synhead{NegP}
    [{\synhead{DP}\\\textit{byl}\\{\footnotesize child.\Dir{}}\\{\scriptsize pivot}}]
    [{Neg$'$}
      [{Neg\\{\small($\varnothing$: absorbed by suppletive)}}]
      [\synhead{AspP}
        [{Asp$'$}
          [{Asp\\\ird{-a}\\{\footnotesize\textsc{[cont]}}}]
          [ExistP
            [{Exist$'$}
              [{Exist\\\ird{zad-}\\{\scriptsize suppletive}}]
            ]
          ]
        ]
      ]
    ]
  ]
\end{forest}
\end{center}

\noindent The pivot surfaces to the left as the absolutive topic constituent,
base-generated preverbal under head-finality; no movement is required. In the
affirmative it occupies Spec,\synhead{AspP}; in the negative it occupies Spec,\synhead{NegP}. The
Neg\textsuperscript{0} feature is discharged by conditioning the suppletive
readjustment \ird{jec-} $\rightarrow$ \ird{zad-} at Exist\textsuperscript{0},
and has no residual PF exponent of its own.

\paragraph{Location.} \ird{Jec-} is also used to locate an entity: the entity
appears in the direct case and the location is expressed by an indirect noun
phrase followed by a postposed clitic adposition. \ird{Zad-} negates location,
asserting that the referent is not present at the named location.

\pex
\a
\begingl
\gla Marek dumě na jeca.//
\glb Marek.\Dir{} house.\Ind{} in \Exst{}-\Cont{}//
\glft \trsl{Marek is at home.}//
\endgl
\a
\begingl
\gla Marek dumě na zada.//
\glb Marek.\Dir{} house.\Ind{} in \NegExst{}-\Cont{}//
\glft \trsl{Marek is not at home. / Marek is absent.}//
\endgl
\xe

\noindent The locative phrase is merged as a PP adjunct to ExistP, introducing
the spatial domain within which the existential relation holds, without altering
the valency of Exist\textsuperscript{0}. The pivot remains the sole direct
argument; the structures parallel those for bare existentials above, with the PP
adjoined at the ExistP periphery in both affirmative and negative derivations.


\subsection{Possession}\label{sec:existence-possession}

Iridian has no separate verb meaning \trsl{have}. What English presents as
\trsl{I have X} is in Iridian construed as \trsl{there is X at me}: the
possessor stands in the indirect case, the possessed thing in the direct, and
the construction is a special case of existential predication. \ird{Zad-}
produces the negative possessive, asserting that the possessed item does not
exist within the possessor's sphere.

The ordinary arrangement of the clause places the possessor phrase first, the
possessed noun next, and the existential verb last:

\begin{center}
[\emph{possessor}.\Ind{} \ \ird{na}] \ \ [\emph{possessee}.\Dir{}] \ \ \ird{jec-} / \ird{zad-}
\end{center}

\noindent The possessed noun is the thing whose existence is asserted or denied
relative to the possessor as reference point. In grammatical terms it is the
existential pivot --- the direct argument of Exist\textsuperscript{0} --- while
the possessor is an oblique locative domain, a PP adjunct to ExistP, within
which the existential relation holds.

\pex
\a
\begingl
\gla Bylam na tóm jeca.//
\glb child.\Ind{} in/on book.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{The child has a book.}//
\endgl
\a
\begingl
\gla Bylam na tóm zada.//
\glb child.\Ind{} in/on book.\Dir{} \NegExst{}-\Cont{}//
\glft \trsl{The child does not have a book.}//
\endgl
\xe

\pex
\a
\begingl
\gla Tóm bylam na jeca.//
\glb book.\Dir{} child.\Ind{} in/on \Exst{}-\Cont{}//
\glft \trsl{As for the book, it is with the child.}//
\endgl
\a
\begingl
\gla Tóm bylam na zada.//
\glb book.\Dir{} child.\Ind{} in/on \NegExst{}-\Cont{}//
\glft \trsl{As for the book, the child does not have it.}//
\endgl
\xe

\noindent The second pair illustrates that topic-initial order is fully
available with \ird{zad-} on the same terms as with \ird{jec-}: the alternation
belongs to the preverbal field and is independent of the affirmative/negative
opposition.

Following Freeze (1992) and the localist tradition, possession is treated here
as a structural subtype of location: \trsl{the book exists at the child}. The
possessor phrase [\textit{possessor}.\Ind{} \ird{na}] and a plain locative
phrase [\textit{location}.\Ind{} \ird{na}] are structurally parallel: both are
oblique PPs that modify ExistP without altering the valency of
Exist\textsuperscript{0}. The affirmative and negative possessive structures
are:

\begin{center}
\begin{forest}
  syntax tree
  [ExistP
    [{PP\\\ird{bylam na}\\{\footnotesize possessor}}]
    [ExistP
      [{\synhead{DP}\\\textit{tóm}\\{\footnotesize book.\Dir{}}\\{\scriptsize pivot}}]
      [{Exist$'$}
        [{Exist\\\ird{jec-}}]
      ]
    ]
  ]
\end{forest}
\hspace{2em}
\begin{forest}
  syntax tree
  [\synhead{NegP}
    [{\synhead{DP}\\\textit{tóm}\\{\footnotesize book.\Dir{}}\\{\scriptsize pivot}}]
    [{Neg$'$}
      [{Neg\\{\small($\varnothing$)}}]
      [\synhead{AspP}
        [{Asp$'$}
          [{Asp\\\ird{-a}\\{\footnotesize\textsc{[cont]}}}]
          [ExistP
            [{PP\\\ird{bylam na}\\{\footnotesize possessor}}]
            [ExistP
              [{Exist$'$}
                [{Exist\\\ird{zad-}\\{\scriptsize suppletive}}]
              ]
            ]
          ]
        ]
      ]
    ]
  ]
\end{forest}
\end{center}

\noindent The structural parallelism between locative and possessive
existentials is a direct consequence of this unified analysis: both
constructions merge a PP at the ExistP periphery, and the only grammatical
difference is the semantic content of the oblique phrase itself. The suppletive
readjustment applies identically in both frames; the PP adjunct is unaffected by
the presence or absence of \synhead{NegP}.

Since possession is expressed through the existential verb, possessive clauses
naturally inflect across the full aspectual system, and \ird{jec-} and
\ird{zad-} participate symmetrically in these aspectual contrasts. The
continuous is the common unmarked form: \ird{Bylam na tóm jeca} \trsl{the child
has a book}; \ird{Bylam na tóm zada} \trsl{the child does not have a book}. The
progressive conveys a relation in the course of arising: \ird{Bylam na tóm
jecime} \trsl{the child is getting a book}. The contemplative projects into
prospect: \ird{Bylam na tóm jecach} \trsl{the child will have a book}. The
perfective, applied to so stative a predicate, yields an inchoative sense:
\ird{Bylam na tóm jak} \trsl{the child got a book}. The retrospective places the
relation in a prior frame: \ird{Bylam na tóm žac} \trsl{the child had a book}.
In all these forms the construction remains fundamentally existential: the
possessed noun remains the direct pivot, the possessor remains oblique, and the
aspect alters only the speaker's view of the relation.

Since the possessed noun is the existential pivot, indefinite possessees are
especially natural. Definite possessees are not excluded, but with them the
clause more readily shades toward a meaning of location, availability, or
custody rather than settled ownership --- a consequence of the existential
source of the construction.


\subsection{Event-anchored existential pivots}\label{sec:existential-presentatives}

The existential predicate in Iridian consistently selects a nominal pivot: the entity
asserted to exist stands as the direct-case complement of Exist\textsuperscript{0}. The
constructional family defined by this selectional requirement is broader than the
overt-head presentational use treated in earlier descriptions. Three subtypes can be
distinguished according to the internal structure of the pivot \synhead{DP}:

\begin{enumerate}[wide]
\item a bare \synhead{NP} (plain existential; §\,\ref{sec:existence-and-presence}),
\item a complex \synhead{DP} with an overt lexical head modified by a
  participant-oriented attributive (overt-head presentational;
  §\,\ref{sec:exst-overt-head}),
\item a headless \synhead{DP} whose sole content is a participant nominal derived from
  verbal structure by Iridian's participant-nominal morphology (headless
  participant-nominal; §\,\ref{sec:exst-headless-nmlz}).
\end{enumerate}

\noindent The three subtypes share a single structural identity: all three occupy
Spec,\synhead{ExistP} as the nominal complement of Exist\textsuperscript{0}. They differ
not in the matrix construction but in the internal architecture of the pivot \synhead{DP}.

\subsubsection{Architectural decision: participant nominals, not silent nouns}
\label{sec:exst-pivot-arch}

The third subtype requires an explicit analytical decision. When the existential
pivot introduces an otherwise unidentified participant --- the function served by English
\trsl{somebody}, \trsl{something}, \trsl{nobody}, \trsl{nothing} --- a possible analysis
would posit an indefinite null lexical noun ($\varnothing$N) heading the pivot, with the
verbal description functioning as its attributive modifier. That analysis is explicitly
rejected for Iridian.

Iridian does not ordinarily lexicalize English-style \trsl{someone}, \trsl{something},
\trsl{no one}, and \trsl{nothing} as free pronominal items in clause structure, and it
does not build their equivalents by projecting a semantically bleached null noun.
Instead, the indefinite pivot is a bona fide participant nominal: an nP/DP derived
directly from verbal structure by the language's participant-nominal morphology
(§\,\ref{sec:nmlz-predicative} of Chapter~\ref{ch:verbs}). The nominal character of
the pivot is the output of n\textsuperscript{0}, whose categorial feature nominalises the
embedded verbal structure. No lexical noun is inserted or covertly present; the
nominalization is the whole pivot.

This treatment is principled within the grammar's DM architecture. Iridian already has a
robust system of participant nominals whose orientation is determined by voice morphology
(§\,\ref{sec:nmlz-predicative} of Chapter~\ref{ch:verbs}): antipassive stems yield
agent-oriented nominals, unmarked transitive stems yield patient-oriented ones, and
applicative stems yield nominals for the promoted participant. The indefinite existential
construction draws on precisely this system. The participant orientation of the
existential pivot is encoded by the same voice-sensitive stem morphology that governs
participant nominals throughout the grammar; no independent mechanism is required.

\subsubsection{Overt-head presentational pivots}
\label{sec:exst-overt-head}

The overt-head presentational construction is the most structurally elaborated subtype.
The pivot \synhead{DP} consists of a lexical head noun in the direct case modified by a
participant-oriented attributive. Its canonical shape is:

\begin{center}
[\textit{internal-arg.\Ind{}} \ \textit{participant-oriented-attributive} \ \textit{head-NP}.\Dir{}] \ \ \ird{jeca}
\end{center}

The modifier is canonically oriented toward the participant the embedded stem privileges:
the patient or theme of a transitive predicate, the recipient or beneficiary of an
applicative stem, or more generally the non-agentive participant that emerges from the
event as the discourse-relevant referent. The construction therefore introduces \trsl{a
letter that the woman wrote} or \trsl{a child someone fed} rather than asserting the
writing or the feeding for its own sake.

\ird{Jeca} is fixed in this use. Temporal and aspectual interpretation come entirely from
the embedded attributive, which may be perfective, retrospective, continuous, progressive,
or contemplative. The matrix \ird{jeca} signals only that the referent is being introduced
into the current discourse.

\pex
\a
\begingl
\gla Obrohnikovní senica jeca.//
\glb read-\textsc{trs}-\Pf{}-\textsc{attr} news.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{There is some news that (I) read.} \lingcontext{patient/theme introduced}//
\endgl
\a
\begingl
\gla Že dbornikovní kíl jeca.//
\glb \textsc{1sg}.\Ind{} hear-\textsc{trs}-\Pf{}-\textsc{attr} song.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{There is a song that I heard.} \lingcontext{stimulus introduced via perception event}//
\endgl
\a
\begingl
\gla Vednikovní ryc jeca.//
\glb see-\textsc{trs}-\Pf{}-\textsc{attr} cat.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{There is a cat that (I) saw.} \lingcontext{perceived referent introduced via completed event}//
\endgl
\a
\begingl
\gla Verce blětébovní byl jeca.//
\glb toy.\Ind{} give-\Appl{}-\Pf{}-\textsc{attr} child.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{There is a child that he gave a toy to.} \lingcontext{recipient introduced via transfer event}//
\endgl
\xe

The attributive is an adjunct within the pivot \synhead{DP}; it is not a predicate of
the matrix clause and contributes no independent propositional content at the matrix
level. The matrix Exist\textsuperscript{0} asserts only $\textsc{exist}(x)$ for the
complex, event-anchored pivot $x$. This decomposition directly accounts for the fixedness
of \ird{jeca}: because the existential head's semantic contribution is reduced to
$\textsc{exist}(x)$, the least-marked aspect (continuous, unbounded) is the appropriate
exponent; any other aspect would impose a viewpoint on the existential state itself,
conflicting with the temporal interpretation already contributed by the embedded
attributive.

\begin{center}
\begin{forest}
  syntax tree
  [\synhead{AspP}
    [{\synhead{DP}\\{\footnotesize pivot}}
      [{AttrP\\\ird{obrohnikovní}\\{\footnotesize read-\textsc{trs}-\Pf{}-\textsc{attr}}}]
      [{\synhead{NP}\\\ird{senica}\\{\footnotesize news.\Dir{}}}]
    ]
    [{Asp$'$}
      [{Asp\\\ird{jec-a}\\{\scriptsize Exist+Cont}}]
      [ExistP
        [{Exist$'$}
          [{Exist\\\ird{jec-}}]
        ]
      ]
    ]
  ]
\end{forest}
\end{center}

Because the function of this construction is to introduce a specific new referent into the
discourse model, non-agentive pivot heads are strongly favoured. Agents are the
prototypical topical and volitional participants; they are characteristically presupposed
or previously introduced, making them poor existential pivots. Patients, themes,
recipients, and other non-agentive participants are characteristically discourse-new and
are therefore the natural heads of a presentational pivot \synhead{DP}.

\ex
\begingl
\gla ?Senicěte obroždkovní chrás jeca.//
\glb news.\Ind{}.\Def{} read-\textsc{antip}-\Pf{}-\textsc{attr} person.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{There is a person who read the news.} \lingcontext{formally possible, but
  marked: agentive head disfavoured in overt-head presentational}//
\endgl
\xe

This discourse preference is, however, a property of the overt-head subtype specifically.
The headless participant-nominal construction (§\,\ref{sec:exst-headless-nmlz}) is the
default indefinite strategy across all participant roles and does not inherit any
preference for non-agentive pivots.

The negative counterpart of the overt-head presentational is formed with \ird{zada}.
\synhead{NegP} scopes over ExistP, and the construction asserts that no referent satisfying
the event-anchored description exists. As with \ird{jeca}, the continuous form \ird{zada}
is fixed; the aspectual interpretation of the embedded event comes entirely from the
attributive.

\pex
\a
\begingl
\gla Obrohnikovní senica zada.//
\glb read-\textsc{trs}-\Pf{}-\textsc{attr} news.\Dir{} \NegExst{}-\Cont{}//
\glft \trsl{There is no news that (I) read.} \lingcontext{negative overt-head presentational; event-anchored referent denied}//
\endgl
\a
\begingl
\gla Verce blětébovní byl zada.//
\glb toy.\Ind{} give-\Appl{}-\Pf{}-\textsc{attr} child.\Dir{} \NegExst{}-\Cont{}//
\glft \trsl{There is no child that he gave a toy to.} \lingcontext{negative presentational; recipient referent denied}//
\endgl
\xe

\begin{center}
\begin{forest}
  syntax tree
  [\synhead{NegP}
    [{\synhead{DP}\\{\footnotesize pivot}}
      [{AttrP\\\ird{obrohnikovní}\\{\footnotesize read-\textsc{trs}-\Pf{}-\textsc{attr}}}]
      [{\synhead{NP}\\\ird{senica}\\{\footnotesize news.\Dir{}}}]
    ]
    [{Neg$'$}
      [{Neg\\{\small($\varnothing$)}}]
      [\synhead{AspP}
        [{Asp$'$}
          [{Asp\\\ird{zad-a}\\{\scriptsize NegExst+Cont}}]
          [ExistP
            [{Exist$'$}
              [{Exist\\\ird{zad-}\\{\scriptsize suppletive}}]
            ]
          ]
        ]
      ]
    ]
  ]
\end{forest}
\end{center}

\noindent Exist\textsuperscript{0} now asserts $\neg\textsc{exist}(x)$ for the complex
pivot: no referent satisfying the event-anchored description exists. The attributive
remains an adjunct within the pivot \synhead{DP} with no matrix-level propositional
force. The disfavour of agentive pivot heads carries over to \ird{zada} for the same
reasons as in the affirmative.

The plural particle \ird{ne} cannot be combined with a \ird{zad-} clause; see
Section~\ref{sec:definiteness} of Chapter~\ref{ch:nouns}.

\subsubsection{Headless participant-nominal pivots}
\label{sec:exst-headless-nmlz}

The headless participant-nominal construction employs a pivot \synhead{DP} that contains
no lexical head noun. The pivot's entire content is a participant nominal derived from
verbal structure, whose voice morphology determines participant orientation. This
construction is the default means of expressing indefinite participant reference in
Iridian.

The participant nominal is built by the same morphological process that produces the
identificational predicates of §\,\ref{sec:nmlz-predicative} of
Chapter~\ref{ch:verbs}: n\textsuperscript{0} (spelled out as \ird{-ou}, with regular
stem-final contractions) applies to a verbal structure in which voice and aspect have
already been specified. The participant orientation follows from the voice of the embedded
stem:

\begin{itemize}
\item an antipassive-stem nominal denotes the agent of the event (\trsl{the one who Verb'd});
\item an unmarked transitive-stem nominal denotes the patient or theme (\trsl{that which was Verb'd});
\item an applicative-stem nominal denotes the promoted participant (\trsl{the one who was Verb'd-to}).
\end{itemize}

\noindent When such a nominal occupies the existential pivot position, Exist\textsuperscript{0}
asserts the existence of an individual satisfying the participant description. The event
description is internal to the nominalized pivot, not a contribution of the matrix
predicate. The matrix \ird{jec-} contributes only $\textsc{exist}(x)$; the verbal
material, together with its aspect and voice, supplies the property that identifies the
individual introduced.

\pex
\a
\begingl
\gla Ščenikou jeca.//
\glb arrive-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{There is someone who arrived.} \lingcontext{agentive indefinite: intransitive nominal, agent orientation}//
\endgl
\a
\begingl
\gla Lubcí piaščikou jeca.//
\glb apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{There is someone who ate an apple.} \lingcontext{agentive indefinite: antipassive agent-nominal}//
\endgl
\a
\begingl
\gla Vednikou jeca.//
\glb see-\textsc{trs}-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{There is something (I) saw.} \lingcontext{patient indefinite: transitive patient-nominal}//
\endgl
\a
\begingl
\gla Verce blětébovnikou jeca.//
\glb toy.\Ind{} give-\Appl{}-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{There is someone who was given a toy.} \lingcontext{goal indefinite: applicative recipient-nominal}//
\endgl
\xe

The formal structure of the headless participant-nominal existential differs from the
overt-head construction only in the internal architecture of the pivot. The nP directly
daughter-of-DP contains the Voice/Asp/Root complex; n\textsuperscript{0} introduces
nominal category; no lexical NP head is present.

\begin{center}
\begin{forest}
  syntax tree
  [\synhead{AspP}
    [{\synhead{DP}\\{\footnotesize pivot}}
      [{\synhead{nP}}
        [{\synhead{VoiceP}\\{Root + Voice/Asp}}]
        [{n\textsuperscript{0}\\\ird{-ou}\\{\scriptsize\Nmlz{}}}]
      ]
    ]
    [{Asp$'$}
      [{Asp\\\ird{jec-a}\\{\scriptsize Exist+Cont}}]
      [ExistP
        [{Exist$'$}
          [{Exist\\\ird{jec-}}]
        ]
      ]
    ]
  ]
\end{forest}
\end{center}

\noindent The semantics is parallel to the plain and overt-head cases. The nP denotes a
property of individuals, event-anchored by the verbal material it encodes;
Exist\textsuperscript{0} introduces a witness individual satisfying that property. The
eventive content is entirely internal to the pivot; no clausal embedding of the verbal
structure into the matrix occurs.

The contrast between the two constructional subtypes is already visible in the grammar's
existential question--answer paradigm:

\pex
\a
\begingl
\gla Vednikou by jeca?//
\glb see-\textsc{trs}-\Pf{}-\Nmlz{} \textsc{q} \Exst{}-\Cont{}//
\glft \trsl{Did you see anything?} \lingcontext{headless NMLZ pivot; referent as yet unidentified}//
\endgl
\a
\begingl
\gla Vednikovní ryc jeca.//
\glb see-\textsc{trs}-\Pf{}-\textsc{attr} cat.\Dir{} \Exst{}-\Cont{}//
\glft \trsl{There is a cat that I saw.} \lingcontext{overt-head pivot; specific referent introduced}//
\endgl
\xe

\noindent In (a), the pivot is a headless nominal (\ird{vednikou}), because no referent
has yet been identified: the question asks whether any individual satisfying the
existential description exists. In (b), the pivot is a complex \synhead{DP} with an overt
lexical head, because the cat is now a specific entity being introduced into the discourse
model. These are not competing strategies but complementary options: the headless
construction is the appropriate form when a participant is entirely unidentified; the
overt-head construction is the appropriate form when a referent is specific enough to
anchor to a lexical category and an event description.

\subsubsection{Indefinite participant introduction}
\label{sec:exst-indef-pos}

The headless participant-nominal construction is the primary means by which Iridian
expresses what English achieves with the free indefinite pronouns \trsl{somebody},
\trsl{someone}, \trsl{something}, and \trsl{anything}. Free pronominal items are not the
ordinary Iridian strategy; the existential pivot is instead a participant nominal whose
voice morphology encodes the relevant participant role.

\pex
\a
\begingl
\gla Ščenikou jeca.//
\glb arrive-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{Someone arrived.} \lingcontext{lit.\ there is one-who-arrived; agentive indefinite}//
\endgl
\a
\begingl
\gla Lubcí piaščikou jeca.//
\glb apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{Someone ate the apple.} \lingcontext{lit.\ there is one-who-ate-it; agentive transitive indefinite}//
\endgl
\a
\begingl
\gla Vednikou jeca.//
\glb see-\textsc{trs}-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{(I) saw something.} \lingcontext{lit.\ there is a seen-thing; patient indefinite}//
\endgl
\a
\begingl
\gla Verce blětébovnikou jeca.//
\glb toy.\Ind{} give-\Appl{}-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{Someone was given a toy.} \lingcontext{lit.\ there is one-who-was-given-a-toy; goal indefinite}//
\endgl
\xe

Agentive indefinites in (a) and (b) are fully natural. This is a significant difference
from the overt-head subtype, where agent-oriented pivot heads are disfavoured. The
discourse preference for non-agentive pivots is a property of the overt-head
presentational construction specifically, grounded in its referent-introduction function.
The headless construction asserts only that some participant of the relevant type exists;
because this function is indifferent to the discourse-topic properties that make agents
poor existential pivots in the overt-head type, agentive participant nominals are
unproblematic here.

Older descriptive treatments noted that Iridian systematically reframes English
indefinite-pronoun clauses as existential constructions built from nominalization, rather
than employing dedicated indefinite pronominal items. The present analysis gives that
observation formal content: the indefinite existential is not a construction containing a
silent lexical noun but one in which the participant nominal itself is the entire pivot,
with participant orientation encoded by voice morphology internal to n\textsuperscript{0}'s
complement.

\subsubsection{Negative existential indefinites}
\label{sec:exst-indef-neg}

Negative indefinites --- English \trsl{nobody}, \trsl{no one}, \trsl{nothing} --- are
expressed by the negative existential \ird{zad-} with a participant-nominal pivot. The
derivation is structurally parallel to the positive: \synhead{NegP} is projected,
Neg\textsuperscript{0} triggers suppletive Vocabulary Insertion at
Exist\textsuperscript{0}, and the resulting \ird{zad-} asserts $\neg\textsc{exist}(x)$
for the individual satisfying the nominal description.

\pex
\a
\begingl
\gla Ščenikou zada.//
\glb arrive-\Pf{}-\Nmlz{} \NegExst{}-\Cont{}//
\glft \trsl{Nobody arrived.} \lingcontext{lit.\ there is no one-who-arrived}//
\endgl
\a
\begingl
\gla Lubcí piaščikou zada.//
\glb apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{} \NegExst{}-\Cont{}//
\glft \trsl{Nobody ate the apple.} \lingcontext{lit.\ there is no one-who-ate-it}//
\endgl
\a
\begingl
\gla Vednikou zada.//
\glb see-\textsc{trs}-\Pf{}-\Nmlz{} \NegExst{}-\Cont{}//
\glft \trsl{Nothing was seen. / (I) saw nothing.} \lingcontext{lit.\ there is no seen-thing}//
\endgl
\xe

The negation is existential: no individual satisfying the participant description exists.
No presupposition is generated about whether the relevant event occurred. This property
distinguishes negative existential indefinites from identificational negation, treated in
the following subsection.

\subsubsection{Relation to identificational equatives}
\label{sec:exst-vs-ident}

The same participant nominals that serve as existential pivots with \ird{jec-}/\ird{zad-}
also serve as predicates in identificational equatives (§\,\ref{sec:ident-equative}).
The two constructions must be carefully distinguished, because they differ both in
structural position and in semantic scope.

In an identificational equative, the participant nominal is the predicative complement of
a zero-copula clause. The topic \synhead{DP} is identified as the unique entity satisfying
the description; the event is presupposed. Negation of this construction with \ird{staly}
yields identificational negation: the presupposed event remains intact, and only the
identity of the relevant participant is denied.

\pex
\a
\begingl
\gla Janek lubcí piaščikou.//
\glb Janek.\Dir{} apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{}//
\glft \trsl{Janek is the one who ate the apple.} \lingcontext{identificational equative: event presupposed; Janek identified as agent}//
\endgl
\a
\begingl
\gla Janek lubcí piaščikou staly.//
\glb Janek.\Dir{} apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{} \NegCop{}//
\glft \trsl{Janek is not the one who ate the apple.} \lingcontext{identificational negation: event still presupposed; Janek misidentified as agent}//
\endgl
\xe

\noindent The contrastive implication that someone else ate the apple arises because
identificational predications are exhaustive: negating an exhaustive identification
entails the existence of a different satisfier. This is the expected interpretation in
discourse (§\,\ref{sec:ident-equative}).

The existential construction with a participant-nominal pivot has a fundamentally
different semantic profile:

\pex
\a
\begingl
\gla Lubcí piaščikou jeca.//
\glb apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{} \Exst{}-\Cont{}//
\glft \trsl{Someone ate the apple.} \lingcontext{existential: asserts existence of an agent; no event presupposition}//
\endgl
\a
\begingl
\gla Lubcí piaščikou zada.//
\glb apple.\Ind{} eat-\Antip{}-\Pf{}-\Nmlz{} \NegExst{}-\Cont{}//
\glft \trsl{Nobody ate the apple.} \lingcontext{negative existential: no agent exists; event not presupposed}//
\endgl
\xe

\noindent In the existential, the participant nominal is an argument (the pivot) in
Spec,\synhead{ExistP}; \ird{zad-} scopes over Exist\textsuperscript{0}, denying the
existence of any individual satisfying the description. No event presupposition survives.
The contrast in structural position --- nominal as predicate vs.\ nominal as argument ---
drives the contrast in presuppositional profile.

Ordinary negative indefinites (\trsl{nobody}, \trsl{nothing}) therefore belong to the
existential system with \ird{zad-}, not to the identificational equative with \ird{staly}.
The identificational construction is the appropriate choice only when the discourse
specifically presupposes the event and contests the identity of its participant. Deploying
\ird{staly} + participant nominal in a context where no such presupposition is intended
would produce an unintended exhaustivity implication that the event occurred and some
other participant was responsible.

A small set of items --- including \ird{niho} and \ird{zide} --- appear to function as
negative-polarity or quantificational operators in certain registers and may interact with
both the existential and identificational systems. Their full characterization remains an
open question and is not pursued here.

\subsubsection{Discourse profile of the two constructional subtypes}
\label{sec:exst-discourse}

The overt-head and headless participant-nominal constructions have complementary discourse
profiles. They are not competing strategies for the same communicative function but are
specialized for different referential tasks.

The overt-head presentational construction is the natural means of introducing a specific,
contentfully described new referent into the discourse model. The lexical head noun
provides a category for the referent; the attributive anchors it to an event in which it
is the salient participant. This construction is especially apt when the speaker wishes to
present an entity rich enough in description to support further reference --- to introduce
a cat, a song, a piece of news, or a child given a toy. Because referent introduction
presupposes that the referent is hearer-new and non-presupposed, non-agentive participants
are the most natural pivot heads; their discourse-new status is the norm, while agents are
characteristically already in play.

The headless participant-nominal construction performs a different function: it asserts
the existence of an otherwise unidentified participant without assigning that participant a
lexical category. It is the appropriate form when the speaker wishes only to assert that
some individual, as yet unnamed and uncharacterized, bears the relevant participant
relation to an event. This construction is indifferent to participant role: agentive,
patient, and goal pivots are all equally regular.

The presentational construction is not simply a periphrastic variant of the corresponding
eventive clause. \ird{Vednikovní ryc jeca} and the corresponding plain verbal clause share
the same truth conditions --- both are true when the speaker saw a cat --- but they differ
in discourse function. The plain clause reports an event of seeing; the overt-head
presentational performs a distinct discourse act: the introduction of the cat as a new
participant into the discourse model, with the seeing event supplying the identifying
anchor. The distinction is sharpest in context. After the existential question \ird{Vednikou
by jeca?} (\trsl{Did you see anything?}), the overt-head answer \ird{Vednikovní ryc jeca}
is highly felicitous because the question has already made the entity --- not the event ---
the relevant discourse target. After a question targeting the agent (\trsl{Who saw the
cat?}), the overt-head presentational is a poor response: the construction is designed
to bring a non-agentive participant onto the discourse stage, and the question has made the
agent the target instead. The headless construction, by contrast, is the default form for
asserting the existence of any participant type, agentive or otherwise, when no richer
description is yet warranted.


\subsection{Suppletive exclusivity: \ird{jec-}, \ird{zad-}, and the \synhead{NegP} alternation}\label{sec:neg-exst-alternation}%
\label{sec:vm-jec-zad}%
\label{sec:vm-jec}

The account in this section is the Exist\textsuperscript{0}-local unpacking of
the Neg\textsuperscript{0} allomorphy analysis in
§\,\ref{sec:vm-neg-exponents} of Chapter~\ref{ch:verbs}. That analysis
characterises \ird{zad-} as the existential exponent of Neg\textsuperscript{0}:
\synhead{NegP} is projected, and its Neg\textsuperscript{0} feature is discharged by
conditioning suppletive Vocabulary Insertion at Exist\textsuperscript{0},
leaving Neg\textsuperscript{0} with no independent PF reflex. The present
section redescribes the same facts from the perspective of
Exist\textsuperscript{0}: \ird{jec-} and \ird{zad-} are suppletive allomorphs
of the same functional head, with \synhead{NegP} presence as the conditioning environment.
The two descriptions are notational variants of one analysis, not competing
accounts.

\ird{Jec-} and \ird{zad-} are in strict complementary distribution conditioned
by the presence or absence of \synhead{NegP} in the derivation. Wherever \synhead{NegP} is absent,
Exist\textsuperscript{0} is spelled out as \ird{jec-}; wherever \synhead{NegP} is present,
it is spelled out as \ird{zad-} and \ird{jec-} is excluded. This is an instance
of the same conditioned allomorphy that governs Neg\textsuperscript{0}'s
exponence across all Iridian clause types: \ird{zá-} (prefix) in verbal clauses,
\ird{staly} (free morpheme) in equative clauses, and suppletive root replacement
in existential clauses. The three are not separate lexical negation strategies
but conditioned phonological reflexes of a single \synhead{NegP} head.

The semantic content of the alternation is straightforwardly compositional.
\ird{Jec-} contributes $\textsc{exist}(x)$; \ird{zad-} contributes
$\neg\textsc{exist}(x)$, where the negation is the propositional negation
licensed by \synhead{NegP}. The same opposition is preserved across all construction types
--- bare existential, locative, possessive, and presentational --- and there is
no grammatical asymmetry between the two roots in terms of which construction
types each permits.

The sole constructional asymmetry is pragmatic. In the presentational use, both
\ird{jeca} and \ird{zada} are fixed continuous forms, but the positive
presentational functions as a discourse-introducing device (it asserts and
presents a new referent), whereas the negative presentational presupposes the
availability of the event-anchored description in order to deny a referent
matching it. This follows from general properties of existential negation and
requires no grammatical stipulation.

\paragraph{The impossibility of \ird{*zájec-}.} A natural question is whether
the NEG prefix \ird{zá-} can be used with \ird{jec-} to form a negative
existential. It cannot. Forms like \ird{*zájeca} and \ird{*zájecime} are
ungrammatical across all existential construction types; negative existential
predication is always and exclusively expressed by \ird{zad-}.

\pex
\a
\begingl
\gla Bylam na tóm zada.//
\glb child.\Ind{} in/on book.\Dir{} \NegExst{}-\Cont{}//
\glft \trsl{The child does not have a book.} \lingcontext{\ird{zad-} obligatory in the negative possessive}//
\endgl
\a
\begingl
\gla {*}Bylam na tóm zájeca.//
\glb child.\Ind{} in/on book.\Dir{} \Neg{}-\Exst{}-\Cont{}//
\glft \trsl{(intended: the child does not have a book)} \lingcontext{\ird{zá-} cannot prefix \ird{jec-}}//
\endgl
\xe

\pex
\a
\begingl
\gla Obrohnikovní senica zada.//
\glb read-\textsc{trs}-\Pf{}-\textsc{attr} news.\Dir{} \NegExst{}-\Cont{}//
\glft \trsl{There is no news that (I) read.} \lingcontext{\ird{zad-} obligatory in the negative presentational}//
\endgl
\a
\begingl
\gla {*}Obrohnikovní senica zájeca.//
\glb read-\textsc{trs}-\Pf{}-\textsc{attr} news.\Dir{} \Neg{}-\Exst{}-\Cont{}//
\glft \trsl{(intended: there is no news that (I) read)} \lingcontext{\ird{zá-} cannot prefix \ird{jec-}}//
\endgl
\xe

The structural basis is straightforward. When Neg\textsuperscript{0} is present
in an existential clause, it triggers suppletive Vocabulary Insertion at the
Exist\textsuperscript{0} node, replacing \ird{jec-} with \ird{zad-}. The
Neg\textsuperscript{0} feature is thereby discharged structurally in full; no
\synhead{NegP} head remains to be spelled out as \ird{zá-}. The prefix \ird{zá-} is the PF
reflex of Neg\textsuperscript{0} only when Neg\textsuperscript{0} has not
already been discharged through suppletive readjustment. In verbal clauses,
where no suppletive alternation is available at the predicate root, \ird{zá-} is
the sole exponent; in existential clauses, the suppletive root exhausts the
Neg\textsuperscript{0} feature and \ird{zá-} accordingly never surfaces.

The form \ird{*zájeca} would require Neg\textsuperscript{0} to be doubly
exponed: once as the suppletive trigger replacing \ird{jec-} at
Exist\textsuperscript{0}, and again as a prefix attaching to the surviving
positive root. Since a single syntactic head cannot discharge twice, this
configuration is structurally ill-formed. The same logic excludes \ird{*zázada}:
stacking \ird{zá-} onto the already-negative \ird{zad-} would require two
independent Neg\textsuperscript{0} heads in the same clause, violating the
uniqueness of the \synhead{NegP} projection. The prohibition on \ird{zá-} in existential
clauses and the prohibition on \ird{*zázad-} are therefore the same prohibition,
both following from the single-\synhead{NegP} constraint on the clause spine.
\section{Questions: descriptive recap}\label{sec:questions}

The syntactic analysis of Iridian interrogatives is developed in full in
§\,\ref{sec:th-left-periphery}--§\,\ref{sec:th-wh-interrogation}. The
present section provides a brief descriptive summary for cross-reference and
points forward to the relevant complement-clause and discourse chapters.

\subsection{Polar questions}

Polar (yes-no) questions are formed by the overt particle \ird{by} in
Pol\textsuperscript{0} (§\,\ref{sec:th-polar-questions}). In colloquial speech
a rising intonation pattern may suffice without overt \ird{by}; this is a
prosodic reduction of the same clause type, not a distinct syntactic
construction. The Class~1 phasal particles \ird{v} and \ird{s} may co-occur
with \ird{by} to specify the phasal profile of the question (§\,\ref{sec:class1-particles}).

\subsection{Content questions}

Content questions are formed by the combination of a wh-operator (see
Table~\ref{tab:nm-interrogatives} in Chapter~\ref{ch:nouns}) and a
Pol\textsuperscript{0} head, which may be phonologically null or spelled out
as overt \ird{by}; both options are standardly available. The wh-operator may
remain in situ or move to the left-peripheral focus position for discourse
prominence; both orders are grammatical in standard Iridian. The in-situ and
fronted variants differ in the discourse salience attributed to the wh-variable,
not in interrogative well-formedness. A content question with overt \ird{by}
is more explicitly typed as a direct question, while one with null
Pol\textsuperscript{0} is the lighter, more neutral form; the choice is
a matter of discourse profile, not grammatical markedness. The full structural
analysis and four-combination paradigm are in §\,\ref{sec:th-wh-interrogation}.

\subsection{Tag, echo, and indirect questions}

Tag questions append a polar question fragment after a declarative clause,
repeating the relevant predicate or using a proform. Echo questions repeat a
clause with rising intonation, often with \ird{by} added or with fronting of a
wh-phrase in place of the queried element. Indirect questions are embedded
under the non-assertive complementizer \ird{poby}; they may be of the polar
or wh-content type (§\,\ref{sec:poby}).

\section{Negation in clausal context}\label{sec:clausal-negation}

At the clausal level, negation in Iridian is expressed by a single functional
projection, \synhead{NegP}, whose Neg\textsuperscript{0} head occupies position 4 in the
clause spine (above \synhead{AspP}, below \synhead{IP}/\synhead{MoodP}; see Table~\ref{tab:th-clause-spine} in
§\,\ref{sec:th-clause-structure}). The surface form of Neg\textsuperscript{0}
varies by predicate type, yielding three distinct phonological reflexes that
together cover all clause types. This section provides a descriptive account of
the distribution; the formal DM analysis of Neg\textsuperscript{0} allomorphy
(including the prefix paradox and the unified \synhead{NegP} argument) is developed in
§\,\ref{sec:vm-neg} and §\,\ref{sec:vm-neg-exponents} of
Chapter~\ref{ch:verbs}; Figure~\ref{fig:neg0-comparative} there provides
the comparative structural summary of the three exponents in a single
diagram.

\subsection{Verbal clauses: \ird{zá-}}

In finite verbal clauses, Neg\textsuperscript{0} is spelled out as the prefix
\ird{zá-} (reduced to \ird{za-} in unstressed environments). Because the
verbal complex is clause-final under head-finality, and because \ird{zá-}
has no independent prosodic weight, it procliticizes to the left edge of the
verbal complex. The \synhead{NegP} head is syntactically above \synhead{AspP}, but PF
cliticization places the prefix before the root --- a phonological rather
than a syntactic inversion. See §\,\ref{sec:negation} of Chapter~\ref{ch:verbs}
and §\,\ref{sec:vm-neg} of Chapter~\ref{ch:verbs} for the full analysis.

\subsection{Equative clauses: \ird{staly}}

In zero-copula equative and class-membership clauses, no verbal complex is
present, and \ird{zá-} therefore has no host to cliticize onto.
Neg\textsuperscript{0} is instead spelled out as the free morpheme \ird{staly},
which surfaces in the canonical head-final position of a \synhead{NegP} head:
clause-finally, after the topic \synhead{DP} and the predicative complement \synhead{DP}. This is
the position the clause spine directly predicts. \ird{Staly} is morphologically
invariant; it bears no aspect, mood, or person marking of any kind, consistent
with its status as a functional head rather than a verbal root. See
§\,\ref{sec:zero-copula} for exemplification of predicational equative negation
(\ird{Marek doktor staly}) and §\,\ref{sec:ident-equative} for identificational
equative negation (\ird{Lobek Jancí piaštnikou staly}).

\subsection{Identificational negation: scope contrast with \ird{zá-}}

The availability of participant nominal predicates (§\,\ref{sec:ident-equative})
introduces a structurally motivated contrast between two types of clausal
negation that share the same morphological means but differ in semantic scope:

\begin{description}[leftmargin=2em, labelwidth=!, labelindent=0em]
\item[\upshape Propositional negation (\ird{zá-}).] The prefix \ird{zá-} negates
  the event description expressed by the finite verb. The negated clause denies
  that the event occurred (or was performed by the relevant participant). No
  presupposition is generated about whether the event occurred in some other
  form.
\item[\upshape Identificational negation (\ird{staly} + participant nominal).]
  The combination of a participant nominal predicate and \ird{staly} negates the
  identification: the topic \synhead{DP} is denied as the unique satisfier of the
  nominalised description. The event is presupposed to have occurred; only the
  identity of the participant is in question.
\end{description}

\pex
\a
\begingl
\gla Lobek Janku zápiaštnik.//
\glb apple.\Dir{} Janek.\Trs{} \Neg{}-eat-\textsc{trs}-\Pf{}//
\glft \trsl{Janek did not eat the apple.}
      \lingcontext{propositional: the eating may not have occurred at all}//
\endgl
\a
\begingl
\gla Lobek Jancí piaštnikou staly.//
\glb apple.\Dir{} Janek.\Ind{} eat-\textsc{trs}-\Pf{}-\textsc{nmlz} \NegCop{}//
\glft \trsl{The apple is not what Janek ate.}
      \lingcontext{identificational: eating presupposed; the apple is the wrong identification}//
\endgl
\xe

\noindent The contrastive implication (`it was something else') in (b) arises
because identificational predications are exhaustive: asserting that $x$ is the
entity satisfying the description entails that no other entity does. Negating
an exhaustive assertion therefore entails the existence of a different satisfier.
This implication is pragmatically cancellable in principle but is the strongly
expected interpretation in discourse. For the formal scope analysis --- including
the semantic formula, the contrast with propositional \ird{zá-}, and the
derivation of the contrastive implication from exhaustivity --- see
§\,\ref{sec:vm-ident-neg} of Chapter~\ref{ch:verbs}.

\subsection{Existential clauses: \ird{zad-}}

In existential clauses, the presence of Neg\textsuperscript{0} in the derivation
triggers suppletive root replacement: the existential root \ird{jec-} is replaced
by \ird{zad-}, which carries the full aspect paradigm of a finite verb.
Neg\textsuperscript{0} itself has no independent PF exponent in this
configuration; it is discharged structurally by licensing the suppletive Vocabulary
Item. The prefix \ird{zá-} does not appear because there is no
Neg\textsuperscript{0} head remaining to be spelled out separately. See
§\,\ref{sec:zad} for paradigm and exemplification.

\subsection{Mood-internal negation: prohibitive and negative hortative}

The prohibitive \ird{-é}, the negative hortative \ird{-ku}, and the negative
irrealis \ird{-ir} each carry [+\textsc{neg}] as a lexically specified feature
of the \synhead{MoodP} Vocabulary Item. When a \synhead{MoodP} head bears this feature, \synhead{NegP} is
absent from the derivation: the clause is built with the negative \synhead{MoodP} merging
directly above \synhead{AspP}, and \ird{zá-} does not surface because there is no
Neg\textsuperscript{0} node for it to spell out. The [+\textsc{neg}] of these
forms is directive-scoped (qualifying a command, suggestion, or hypothetical),
not propositional, and therefore not equivalent to the truth-conditional negation
of \synhead{NegP}. For the full \synhead{MoodP} analysis, including the paradigm of affirmative and
negative Vocabulary Items and the restricted palatalizing behaviour of \ird{-ir}, see
§\,\ref{sec:vm-moodp} of Chapter~\ref{ch:verbs}.

\subsection{Summary}

The distribution is summarised in the following table:

\begin{table}[ht]
  \footnotesize
  \caption{Negation strategies by predicate/mood type.}\label{tab:neg-exponents}
  \medskip
  \begin{tabularx}{\textwidth}{L L L L}
    \toprule
    \textsc{predicate/mood type} & \textsc{affirmative} & \textsc{negative} & \textsc{negation scope} \\
    \midrule
    Verbal (indicative)           & verbal complex                & \ird{zá-} prefix; \synhead{NegP} projected          & propositional \\
    Equative / zero copula        & $\varnothing$ copula            & \ird{staly} (free morpheme, clause-final) & predicational \\
    Identificational (part.\ nom.) & part.\ nom.\ + $\varnothing$  & part.\ nom.\ + \ird{staly}                & identificational \\
    Existential                   & \ird{jec-}                    & \ird{zad-} (suppletive root; Neg\textsuperscript{0} absorbed, no independent PF exponent); \synhead{NegP} projected & propositional \\
    \midrule
    \multicolumn{4}{l}{\textit{Mood-internal negation: \synhead{NegP} not projected; [+\textsc{neg}] on \synhead{MoodP} head\rlap{\textsuperscript{†}}}} \\[2pt]
    Command (imperative)          & \ird{-ím}                     & \ird{-é} (prohibitive)                    & directive scope \\
    Suggestion (hortative)        & \ird{-ká}                     & \ird{-ku} (neg.\ hortative)               & directive scope \\
    Hypothetical (irrealis)       & \ird{-ěv}                     & \ird{-ir} (neg.\ irrealis)                & modal/hypothetical scope \\
    \bottomrule
    \multicolumn{4}{p{\textwidth}}{\textsuperscript{†}\footnotesize
      Directive scope (prohibitive, neg.\ hortative) and modal/hypothetical
      scope (neg.\ irrealis) are both instances of \synhead{MoodP}-internal
      [+\textsc{neg}]; neither involves \synhead{NegP} projection nor truth-conditional
      propositional negation. See §\,\ref{sec:vm-moodp} of
      Chapter~\ref{ch:verbs} for the unified formal account and the semantic
      distinction between the two sub-types.} \\
  \end{tabularx}
\end{table}

\noindent The formal analysis --- including the DM account of the prefix paradox
(\ird{zá-} as a procliticized \synhead{NegP} head in head-final context), the unified \synhead{NegP}
argument across all predicate types, and the \synhead{MoodP} paradigm of affirmative and
negative Vocabulary Items --- is in §\,\ref{sec:vm-neg} and
§\,\ref{sec:vm-neg-exponents} and §\,\ref{sec:vm-moodp} of Chapter~\ref{ch:verbs}.

\section{Comparative and superlative constructions}\label{sec:comparison}

Comparison in Iridian is expressed analytically using particles that combine
with stative verbs. The subject of comparison stands in the direct case; the
standard of comparison takes the indirect case. Two particles are
distinguished: \ird{stá} for direct comparison (\textit{more\ldots than},
\textit{-er than}) and \ird{zrc} for equative comparison (\textit{as\ldots
as}).

\subsection{Direct comparison}

Direct comparison is formed with the particle \ird{stá} placed between the
standard of comparison and the stative verb. The stative verb inflects for the
full range of aspects, giving the usual range of meanings --- ongoing state,
entry into the state, past state, and so on --- within the comparative frame:

\pex
\a
\begingl
\gla Janek Marcí stá četna.//
\glb Janek.\Dir{} Marek.\Ind{} \textsc{comp} be.tall-\Cont{}//
\glft \trsl{Janek is taller than Marek.}//
\endgl
\a
\begingl
\gla Janek Marcí stá četnime.//
\glb Janek.\Dir{} Marek.\Ind{} \textsc{comp} be.tall-\Prog{}//
\glft \trsl{Janek is becoming taller than Marek.}//
\endgl
\a
\begingl
\gla Janek Marcí stá četnik.//
\glb Janek.\Dir{} Marek.\Ind{} \textsc{comp} be.tall-\Pf{}//
\glft \trsl{Janek became taller than Marek.}//
\endgl
\xe

\noindent where \ird{četn-} is \trsl{be tall}. The particle \ird{stá} may also
combine with the \ird{-á} attributive form (§\ref{sec:stative-attr}) to form
an attributive comparative. The particle precedes the \ird{-á} form
prenominally:

\pex
\a \ird{stá četná pres} \trsl{the taller man}
\a \ird{Marcí stá četná pres} \trsl{the man who is taller than Marek}
\xe

\subsection{Equative comparison}

Equative comparison is formed with the particle \ird{zrc}. The structure is
otherwise identical to the direct comparative, with the subject in the direct
case and the standard in the indirect. Negation of the equative falls on the
stative verb itself, which takes the regular negative prefix \ird{zá-}
(§\ref{sec:negation}):

\pex
\a
\begingl
\gla Janek Marcí zrc četna.//
\glb Janek.\Dir{} Marek.\Ind{} \textsc{eq} be.tall-\Cont{}//
\glft \trsl{Janek is as tall as Marek.}//
\endgl
\a
\begingl
\gla Janek Marcí zrc záčetna.//
\glb Janek.\Dir{} Marek.\Ind{} \textsc{eq} \Neg{}-be.tall-\Cont{}//
\glft \trsl{Janek is not as tall as Marek.}//
\endgl
\xe
