\chapter{Discourse}\label{ch:discourse}

% ============================================================
% STATUS: SCAFFOLD — largely new; no direct predecessor in 2020 archive.
% Some material in archive/2020/chapters/08-semantics-and-usage.tex may be relevant.
% This chapter is theoretically the most ambitious of the rewrite:
% it reframes the alignment system as a discourse-level phenomenon and
% treats the converb chain as the basic unit of narrative structure.
% ============================================================

\section{Sentence-chains as discourse units}\label{sec:discourse-chains}

% ============================================================
% Iridian discourse is organized in "sentence chains": sequences of
% converb clauses capped by a finite verb. Each sentence chain functions
% as a single discourse unit. The chain has:
%   - Converb clauses = background, supporting, or subordinate events
%   - Finite verb = the foregrounded, topically relevant event
% This parallels the foreground/background distinction found in many
% verb-final or SOV languages with converb systems.
% ============================================================

\section{Topic continuity and topic shift}\label{sec:topic-continuity}

% The absolutive participant tends to persist as topic across sentence
% chains. When the topic shifts, this is typically signaled by:
%   - Lexical re-introduction of the new topic
%   - A demonstrative or proximate/obviative distinction
% The obviative/proximate system of demonstratives (ša vs. dní) is the
% primary cross-clause topic-tracking device.

\subsection{Reference management without morphological cross-reference}
\label{sec:ref-management}

Iridian manages reference tracking without any pronominal clitic series on
the verb or noun. The devices available are: (a)~overt demonstratives,
which encode the proximate/obviative distinction and serve as the primary
third-person referential series; (b)~overt first- and second-person
pronouns, employed when required by contrast, emphasis, disambiguation, or
low discourse recoverability; (c)~the reflexive anaphor, which forces local
coreference within the clausal domain; and (d)~zero anaphora, the unmarked
strategy in contexts of stable topic continuity, where the absolutive
argument of the preceding clause is recoverable without any overt expression.

Pro-drop in Iridian is accordingly a purely discourse-driven null-argument
system. The verb does not morphologically cross-reference its arguments
through a clitic or agreement layer: argument drop is licensed entirely
by discourse recoverability and topic continuity, and the burden of
disambiguation falls on syntax and pragmatics rather than on right-edge
person morphology.

\section{Definiteness in discourse context}\label{sec:discourse-definiteness}

% The definiteness marker -to on a noun signals that the referent is
% identifiable to the hearer. In discourse, first mention is typically
% bare (no definiteness marker); subsequent mention may use the marker,
% the demonstrative, or drop the noun entirely (pro-drop).
% The interaction of definiteness, topic, and absolutive alignment
% is the key thread here.

\subsection{Anchoring and the neutral discourse center}
\label{sec:disc-def}

The anchoring suffix \ird{-to} is one of the clearest nominal diagnostics for
how discourse prominence is distributed in Iridian. Anchoring on a direct-case
\synhead{DP} is the neutral way to maintain or reactivate a referent already established
as the clause's discourse center. By contrast, anchoring on transitive or
indirect DPs is markedly more contrastive, resumptive, or re-identifying
(\S\,\ref{sec:nm-def-nondirect}). The nominal morphology therefore supports the
clausal analysis: the unmarked discourse center is the absolutive/direct pivot,
not the agent as such.

This matters especially in ergative-domain transitives. If such clauses were
fundamentally subject-topic clauses in disguise, one would expect the ergative
agent to show the least marked topic behavior. Instead the grammar points the
other way. A direct-case patient is the neutral anchored referent and the
natural target for zero continuation, while an anchored ergative agent carries
added discourse force: contrast, resumption, or a deliberate topic shift. The
anchoring system is thus an overt nominal reflex of the absolutive--topic
connection rather than a free-floating label of `definiteness'.

\section{Aspect choice as narrative strategy}\label{sec:aspect-narrative}

% The split-ergative alignment means that aspect choice simultaneously
% encodes:
%   (a) the temporal/aspectual shape of the event
%   (b) which participant is the discourse topic
% In narrative, speakers use the perfective/retrospective to bring the
% patient into topic/absolutive position (foregrounding the event outcome)
% and use nominative aspects to keep the agent as topic (foregrounding
% the agent's activity).
% This is the core pragmatic principle behind the aspect system.

In extended discourse, aspect choice is inseparable from participant tracking. The perfective and retrospective place the patient in the absolutive and are therefore preferred when the discourse is about what happened \emph{to} a participant: they are the natural vehicle for narrative foreground, event culmination, and topic shift to an affected referent. The nominative aspects, by contrast, preserve the agent or sole participant as absolutive and are accordingly favoured where the continuity of an actor matters more than the bounded result of the event.

The same principle explains the discourse value of the antipassive. Speakers choose the antipassive when a transitive event is relevant chiefly as part of an agent's ongoing activity and not because the patient itself needs to be established, identified, or tracked. The construction keeps the agent in absolutive position, suppresses any patient from the grammatical core, and thereby functions as a topic-maintenance device. It is especially natural when the patient is generic, incorporated into the event type, inferable from context, or simply of no further narrative consequence; for the morphological details see Chapter~\ref{ch:verbs}.

The causative serves a different discourse purpose. By adding a causer, it allows the speaker to recast an event as one of inducement, delegated action, permission, or coercion. This is a useful viewpoint-management device in narrative: a simple predicate presents an event as directly performed, whereas a causative can foreground authority, distribute responsibility across two participants, or distance the discourse topic from the event's direct execution. In perfective contexts this often yields a strongly interventionist reading; in retrospective contexts it more readily supports permissive, indirect, or weak-control interpretations. The formal argument structure of the causative is described in Chapter~\ref{ch:verbs}; what matters here is that the construction lets the speaker decide whether the discourse should centre on the doing, the causing, or the managed consequences of the action.

\subsection{Mini-texts: continuity and re-centering}

Short discourse sequences make the same point more vividly than single clauses
in isolation.

\pex
\a
\begingl
\gla Marek travě piastime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.}//
\endgl
\a
\begingl
\gla Zmacime.//
\glb run-\Prog{}//
\glft \trsl{[He] is running.} \lingcontext{agent continuity in the nominative domain}//
\endgl
\xe

\noindent Here the null topic of the second clause is naturally interpreted as
Marek, because the first clause placed Marek in the absolutive as its pivot.
This is the familiar agent-continuity pattern of the nominative domain.

\pex
\a
\begingl
\gla Travat Marku piaštnik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} eat-\textsc{trs}-\Pf{}//
\glft \trsl{Marek ate the bread.}//
\endgl
\a
\begingl
\gla Piastnime.//
\glb eat-\textsc{pass}-\Prog{}//
\glft \trsl{[It] is being eaten.} \lingcontext{patient continuity after an ergative-domain clause}//
\endgl
\xe

\noindent In this sequence the null topic of the second clause is most naturally
the prior absolutive \ird{travat} \trsl{the bread}, not the agent \ird{Marku}.
Continuing the agent instead requires overt reintroduction or a more marked
discourse configuration, because the perfective clause has already centered the
patient as pivot.

\pex
\a
\begingl
\gla Travat Marku piaštnik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} eat-\textsc{trs}-\Pf{}//
\glft \trsl{Marek ate the bread.}//
\endgl
\a
\begingl
\gla Marek travě piastime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.} \lingcontext{re-centering from patient-topic narration to agent-topic description}//
\endgl
\xe

\noindent The event type is comparable across the two clauses, but the discourse
center is not. The perfective clause is about what happened to the bread; the
progressive clause is about Marek's ongoing activity. A shift from perfective
narration to progressive description therefore often coincides with a shift in
which participant the discourse tracks most naturally.

\section{Tail-head linkage}\label{sec:tail-head}

% Tail-head linkage is a discourse-structuring device where the final
% clause of one sentence chain is repeated (in whole or in abbreviated
% form) at the start of the next chain. This is attested in the sample
% texts and is a common device in narrative Iridian.

\section{Discourse particles}\label{sec:discourse-particles}

% ============================================================
% STATUS: STUB — to be documented
% Categories:
%   - Topic-management particles (topic shift, topic continuation)
%   - Stance particles (mirative, dubitative, concessive)
%   - Connectives (now, then, well, anyway)
% Some of these were described as "evidential" in the 2020 archive
% (under the QUOT/Infer/Rep labels). They should be reconciled with
% the evidential slot in the morphological template.
% ============================================================

\section{Register variation}\label{sec:register-discourse}

% Literary vs. colloquial clause-linking preferences:
%   - Literary: tends toward full converb chains, explicit nominalization
%   - Colloquial: nu as discourse particle, purposive-future, shorter chains
%   - The increasing use of finite coordinated clauses under Slavic influence


\section{Discourse-level word order}\label{sec:discourse-word-order}

\subsection{Clause-final discourse particles}
\label{sec:disc-final-particles}

Iridian permits a small class of clause-final discourse particles,
placed after the final prosodic constituent of the clause, most
typically after the finite verbal complex. These particles do not
belong to the ordinary verbal inflectional template and do not affect
argument structure, alignment, or truth-conditional content. Their
function is instead pragmatic: they mark the speaker's stance toward
the proposition and toward the addressee at the level of the utterance
as a whole.

Synchronically, these particles are best treated as utterance-edge
clitics or weak particles rather than verbal suffixes. Semantically,
they express such values as appeal to shared knowledge, assertive
presentation, corrective insistence, or emotive intensification. The
core system is intentionally small.

The principal clause-final particles are the following.

\paragraph{\ird{len}}
The particle \ird{len} is confirmatory and shared-ground oriented. It
presents the proposition as accessible, recognisable, or available for
the addressee's ratification, with values such as ‘you know’, ‘as we
both know’, or ‘is that not so’. It is the least forceful member of the
set and is well suited to mild prompting, reassurance, and alignment of
speaker and addressee.

\paragraph{\ird{prav}}
The particle \ird{prav} is assertive and presentational. It marks the
utterance as one the speaker puts forward with particular commitment,
roughly ‘indeed’, ‘truly’, or ‘mark this’. Diachronically, it may be
understood as historically related to a truth-word, but synchronically
it functions as a discourse particle rather than a lexical predicate.
In clause-final position it signals strengthened assertion without
necessarily implying contrast.

\paragraph{\ird{však}}
The particle \ird{však} is corrective or contrastive. It indicates that
the utterance stands against an assumption, implication, or line of
expectation in the discourse, with values such as ‘after all’, ‘but in
fact’, ‘mind you’, or ‘contrary to that’. It is especially natural in
rebuttal, clarification, argumentative discourse, and narrative
correction.

\paragraph{\ird{blěc}}
The particle \ird{blěc} is a marked colloquial and low-register
emotive particle. Unlike \ird{len}, \ird{prav}, and \ird{však}, it
does not belong to the neutral core discourse system, but adds rough
speaker attitude such as irritation, incredulity, impatient insistence,
or coarse emphasis. Its contribution is primarily expressive and
register-sensitive. In formal or careful style it is normally avoided;
in colloquial speech it may reinforce commands, corrections, or
exasperated assertions.

Taken together, these particles form a small utterance-final system in
which \ird{len} marks common-ground appeal, \ird{prav} assertive
presentation, \ird{však} corrective contrast, and \ird{blěc} emotive
or vulgar intensification. Further examples and distributional
restrictions are discussed below.
