\chapter{Discourse, Register, and Spoken Iridian}
\label{ch:discourse}\label{ch:spoken}\label{ch:semantics}

% ============================================================
% This chapter brings together three domains that belong to a single
% descriptive continuum: how Iridian organises information across clauses
% and texts; how register and social context shape grammatical choices;
% and how the spoken and colloquial variety departs from the written
% standard. The three domains interact: aspect choice as a narrative
% device is also a discourse-register choice; many spoken morphemes are
% formally integrated into the same inflectional machinery that the rest
% of the grammar describes; and politeness strategies depend on the same
% anchoring and topic mechanisms analysed in earlier chapters.
% ============================================================

\section{Introduction}\label{sec:discourse-intro}

This chapter deals with Iridian grammar above the level of the individual
clause. The earlier chapters described the internal structure of the verb and
the noun phrase, the organisation of the simple sentence, and the mechanics of
clause chaining. What those chapters largely set aside is the question of how
speakers use those structures in continuous discourse, how register and social
context constrain grammatical choice, and how the spoken and colloquial variety
of the language differs from the written standard. Those questions are the
subject of the present chapter.

The three sections are not independent. The absolutive-pivot analysis of
split-ergativity, developed in Chapters~\ref{ch:verbs} and~\ref{ch:simple}, is
most fully vindicated at the discourse level, where aspect choice proves to be
simultaneously a choice of event shape and a choice of which participant to
track as topic. The preverbal particle system of Chapter~\ref{ch:function-words}
has a clause-peripheral counterpart in the utterance-final particles described
here. And many of the spoken morphemes discussed in the final section are
formally unremarkable from the perspective of Iridian morphology: what marks
them as colloquial is their origin and register, not their integration into the
grammatical system.

%
% ================================================================
%  PART I: DISCOURSE STRUCTURE
% ================================================================
%

\section{Sentence chains as discourse units}\label{sec:discourse-chains}

% ============================================================
% Iridian discourse is organised in "sentence chains": sequences of
% converb clauses capped by a finite verb. Each sentence chain functions
% as a single discourse unit. The chain has:
%   - Converb clauses = background, supporting, or subordinate events
%   - Finite verb = the foregrounded, topically relevant event
% This parallels the foreground/background distinction found in many
% verb-final or SOV languages with converb systems.
% ============================================================

\section{Topic continuity and topic shift}\label{sec:topic-continuity}

% The absolutive participant tends to persist as topic across sentence
% chains. When the topic shifts, this is typically signalled by:
%   - Lexical re-introduction of the new topic
%   - A demonstrative or proximate/obviative distinction
% The obviative/proximate system of demonstratives is the
% primary cross-clause topic-tracking device.

\subsection{Reference management without morphological cross-reference}
\label{sec:ref-management}

Iridian manages reference tracking without any pronominal clitic series on the
verb or noun. The devices available are: (a)~overt demonstratives, which encode
the proximate/obviative distinction and serve as the primary third-person
referential series; (b)~overt first- and second-person pronouns, employed when
required by contrast, emphasis, disambiguation, or low discourse recoverability;
(c)~the reflexive anaphor, which forces local coreference within the clausal
domain; and (d)~zero anaphora, the unmarked strategy in contexts of stable topic
continuity, where the absolutive argument of the preceding clause is recoverable
without any overt expression.

Pro-drop in Iridian is accordingly a purely discourse-driven null-argument
system. The verb does not morphologically cross-reference its arguments through
a clitic or agreement layer: argument drop is licensed entirely by discourse
recoverability and topic continuity, and the burden of disambiguation falls on
syntax and pragmatics rather than on right-edge person morphology.

\section{Anchoring in discourse context}\label{sec:discourse-definiteness}

% The anchoring suffix -to on a noun signals that the referent is
% identifiable to the hearer. In discourse, first mention is typically
% bare (no anchoring marker); subsequent mention may use the marker,
% a demonstrative, or drop the noun entirely (zero anaphora).
% The interaction of anchoring, topic, and absolutive alignment
% is the key thread here.

\subsection{The neutral discourse centre in direct-case position}
\label{sec:disc-def}

The anchoring suffix \ird{-to} is one of the clearest nominal diagnostics for
how discourse prominence is distributed in Iridian. Anchoring on a direct-case
\synhead{DP} is the neutral way to maintain or reactivate a referent already
established as the clause's discourse centre. By contrast, anchoring on
transitive or indirect DPs is markedly more contrastive, resumptive, or
re-identifying (\S\,\ref{sec:nm-def-nondirect}). The nominal morphology
therefore supports the clausal analysis: the unmarked discourse centre is the
absolutive/direct pivot, not the agent as such.

This matters especially in ergative-domain transitives. If such clauses were
fundamentally subject-topic clauses in disguise, one would expect the ergative
agent to show the least marked topic behaviour. Instead the grammar points the
other way. A direct-case patient is the neutral anchored referent and the
natural target for zero continuation, while an anchored ergative agent carries
added discourse force: contrast, resumption, or a deliberate topic shift. The
anchoring system is thus an overt nominal reflex of the absolutive--topic
connection rather than a free-floating label of `definiteness'.

\section{Aspect choice as narrative strategy}\label{sec:aspect-narrative}

% The split-ergative alignment means that aspect choice simultaneously
% encodes:
%   (a) the temporal/aspectual shape of the event
%   (b) which participant is the discourse topic
% In narrative, speakers use the perfective/retrospective to bring the
% patient into topic/absolutive position (foregrounding the event outcome)
% and use nominative aspects to keep the agent as topic (foregrounding
% the agent's activity).
% This is the core pragmatic principle behind the aspect system.

In extended discourse, aspect choice is inseparable from participant tracking.
The perfective and retrospective place the patient in the absolutive and are
therefore preferred when the discourse is about what happened \emph{to} a
participant: they are the natural vehicle for narrative foreground, event
culmination, and topic shift to an affected referent. The nominative aspects, by
contrast, preserve the agent or sole participant as absolutive and are
accordingly favoured where the continuity of an actor matters more than the
bounded result of the event.

The same principle explains the discourse value of the antipassive. Speakers
choose the antipassive when a transitive event is relevant chiefly as part of an
agent's ongoing activity and not because the patient itself needs to be
established, identified, or tracked. The construction keeps the agent in
absolutive position, suppresses any patient from the grammatical core, and
thereby functions as a topic-maintenance device. It is especially natural when
the patient is generic, incorporated into the event type, inferable from
context, or simply of no further narrative consequence; for the morphological
details see Chapter~\ref{ch:verbs}.

The causative serves a different discourse purpose. By adding a causer, it
allows the speaker to recast an event as one of inducement, delegated action,
permission, or coercion. This is a useful viewpoint-management device in
narrative: a simple predicate presents an event as directly performed, whereas a
causative can foreground authority, distribute responsibility across two
participants, or distance the discourse topic from the event's direct execution.
In perfective contexts this often yields a strongly interventionist reading; in
retrospective contexts it more readily supports permissive, indirect, or
weak-control interpretations. The formal argument structure of the causative is
described in Chapter~\ref{ch:verbs}; what matters here is that the construction
lets the speaker decide whether the discourse should centre on the doing, the
causing, or the managed consequences of the action.

\subsection{Mini-texts: continuity and re-centering}

Short discourse sequences make the same point more vividly than single clauses
in isolation.

\pex
\a
\begingl
\gla Marek travě piastime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.}//
\endgl
\a
\begingl
\gla Zmacime.//
\glb run-\Prog{}//
\glft \trsl{[He] is running.} \lingcontext{agent continuity in the nominative domain}//
\endgl
\xe

\noindent Here the null topic of the second clause is naturally interpreted as
Marek, because the first clause placed Marek in the absolutive as its pivot.
This is the familiar agent-continuity pattern of the nominative domain.

\pex
\a
\begingl
\gla Travat Marku piaštnik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} eat-\textsc{trs}-\Pf{}//
\glft \trsl{Marek ate the bread.}//
\endgl
\a
\begingl
\gla Piastnime.//
\glb eat-\textsc{pass}-\Prog{}//
\glft \trsl{[It] is being eaten.} \lingcontext{patient continuity after an ergative-domain clause}//
\endgl
\xe

\noindent In this sequence the null topic of the second clause is most naturally
the prior absolutive \ird{travat} \trsl{the bread}, not the agent \ird{Marku}.
Continuing the agent instead requires overt reintroduction or a more marked
discourse configuration, because the perfective clause has already centred the
patient as pivot.

\pex
\a
\begingl
\gla Travat Marku piaštnik.//
\glb bread.\Dir{}.\Def{} Marek.\Trs{} eat-\textsc{trs}-\Pf{}//
\glft \trsl{Marek ate the bread.}//
\endgl
\a
\begingl
\gla Marek travě piastime.//
\glb Marek.\Dir{} bread.\Ind{} eat-\Prog{}//
\glft \trsl{Marek is eating bread.} \lingcontext{re-centering from patient-topic narration to agent-topic description}//
\endgl
\xe

\noindent The event type is comparable across the two clauses, but the discourse
centre is not. The perfective clause is about what happened to the bread; the
progressive clause is about Marek's ongoing activity. A shift from perfective
narration to progressive description therefore often coincides with a shift in
which participant the discourse tracks most naturally.

\section{Tail-head linkage}\label{sec:tail-head}

% Tail-head linkage is a discourse-structuring device where the final
% clause of one sentence chain is repeated (in whole or in abbreviated
% form) at the start of the next chain. This is attested in the sample
% texts and is a common device in narrative Iridian.

\section{Discourse particles}\label{sec:discourse-particles}\label{sec:discourse-word-order}

% ============================================================
% STATUS: Partly developed (clause-final particles below);
% preverbal discourse particles remain a stub.
% Categories to document:
%   - Topic-management particles (topic shift, topic continuation)
%   - Stance particles (mirative, dubitative, concessive)
%   - Connectives (now, then, well, anyway)
% Some of these were described as "evidential" in earlier literature
% (under the QUOT/Infer/Rep labels). They should be reconciled with
% the evidential slot in the morphological template (Ch. 3).
% ============================================================

\subsection{Clause-final discourse particles}\label{sec:disc-final-particles}

Iridian permits a small class of clause-final discourse particles, placed after
the final prosodic constituent of the clause, most typically after the finite
verbal complex. These particles do not belong to the ordinary verbal
inflectional template and do not affect argument structure, alignment, or
truth-conditional content. Their function is instead pragmatic: they mark the
speaker's stance toward the proposition and toward the addressee at the level of
the utterance as a whole.

Synchronically, these particles are best treated as utterance-edge clitics or
weak particles rather than verbal suffixes. Semantically, they express such
values as appeal to shared knowledge, assertive presentation, corrective
insistence, or emotive intensification. The core system is intentionally small.

The principal clause-final particles are the following.

\paragraph{\ird{len}} The particle \ird{len} is confirmatory and shared-ground
oriented. It presents the proposition as accessible, recognisable, or available
for the addressee's ratification, with values such as `you know', `as we both
know', or `is that not so'. It is the least forceful member of the set and is
well suited to mild prompting, reassurance, and alignment of speaker and
addressee.

\paragraph{\ird{prav}} The particle \ird{prav} is assertive and presentational.
It marks the utterance as one the speaker puts forward with particular
commitment, roughly `indeed', `truly', or `mark this'. Diachronically, it may be
understood as historically related to a truth-word, but synchronically it
functions as a discourse particle rather than a lexical predicate. In
clause-final position it signals strengthened assertion without necessarily
implying contrast.

\paragraph{\ird{však}} The particle \ird{však} is corrective or contrastive. It
indicates that the utterance stands against an assumption, implication, or line
of expectation in the discourse, with values such as `after all', `but in fact',
`mind you', or `contrary to that'. It is especially natural in rebuttal,
clarification, argumentative discourse, and narrative correction.

\paragraph{\ird{blěc}} The particle \ird{blěc} is a marked colloquial and
low-register emotive particle. Unlike \ird{len}, \ird{prav}, and \ird{však}, it
does not belong to the neutral core discourse system, but adds rough speaker
attitude such as irritation, incredulity, impatient insistence, or coarse
emphasis. Its contribution is primarily expressive and register-sensitive. In
formal or careful style it is normally avoided; in colloquial speech it may
reinforce commands, corrections, or exasperated assertions.

Taken together, these particles form a small utterance-final system in which
\ird{len} marks common-ground appeal, \ird{prav} assertive presentation,
\ird{však} corrective contrast, and \ird{blěc} emotive or vulgar
intensification. Further examples and distributional restrictions are discussed
below.

%
% ================================================================
%  PART II: REGISTER AND PRAGMATICS
% ================================================================
%

\section{Register and formality}\label{sec:register}\label{sec:register-discourse}

% ============================================================
% Bring together the Ch.10 register/formality stub and the
% Ch.8 literary-vs-colloquial stub. Key contrasts:
%   - Literary: full converb chains, explicit nominalisation
%   - Colloquial: nu as discourse particle, purposive-future, shorter chains
%   - Increasing use of finite coordinated clauses under Slavic influence
% Port and update from 2020 archive; update case terminology in examples.
% ============================================================

\section{Politeness and the T-V distinction}\label{sec:politeness}

% ============================================================
% The 2020 archive had a detailed treatment of the T-V distinction:
%   - 2pl used as polite singular
%   - 1pl used for speaker humility in formal contexts
%   - Pro-drop as a politeness strategy (avoiding pronouns entirely)
%   - Use of names and titles as quasi-honorifics
% The plural marker ne/ně as an honorific with proper names and
% kinship terms (ne Stám Kovárž, ne mlazka) should be discussed here.
% ============================================================

\section{Honorifics}\label{sec:honorifics}

% [Develop from 2020 archive notes on ně/ne as honorific particle]

\section{Phatic expressions}\label{sec:phatic}

% [Greetings, leave-takings, thanks, apologies — cross-reference with
%  §\,\ref{sec:fixed-converbs} in Chapter~\ref{ch:complex}]

\section{Idiomatic expressions}\label{sec:idioms}

% [Selected idioms from the 2020 archive, updated glosses]

%
% ================================================================
%  PART III: SPOKEN AND COLLOQUIAL IRIDIAN
% ================================================================
%

\section{Spoken Iridian: an overview}\label{sec:spoken-overview}

This section sketches a small part of the spoken and urban colloquial register
of present-day Iridian. The written standard remains the reference variety of
this grammar, but it would be misleading to treat everyday spoken usage as
merely a noisier version of that standard. In ordinary conversation, speakers
make frequent use of reduced connectives, clause-peripheral hedges, semantically
bleached nouns, and other high-frequency colloquial forms whose contribution is
partly lexical and partly interactional.

The treatment here is explicitly synchronic and non-exhaustive. Several spoken
phenomena have already been noted elsewhere in the grammar, for example the
contraction of sequential \ird{nesu} to colloquial \ird{nu} in
Chapter~\ref{ch:complex} and the intentional-future use of the purposive converb
in the same chapter. What follows does not attempt a complete grammar of casual
speech. It instead documents a small set of common colloquial morphemes whose
behaviour is especially revealing for the interaction of lexical bleaching,
discourse management, and ordinary Iridian morphosyntax.

\section{Common colloquial morphemes}\label{sec:spoken-morphemes}

\subsection{The connective \ird{nu}}\label{sec:spoken-nu}

The colloquial connective \ird{nu} has already been introduced in
Section~\ref{sec:seq-copula} of Chapter~\ref{ch:complex}. Synchronically it is
best treated as the reduced spoken continuation of sequential \ird{nesu}, but
its discourse function is now wider than simple sequential clause-linking. In
rapid speech it often serves as a light connective between clauses, turns, or
discourse steps, with meanings ranging from loose \trsl{and then} to weaker
\trsl{so}, \trsl{well then}, or simple transition management.

What matters for the present section is that \ird{nu} belongs to colloquial
discourse management, not to the rigid preverbal particle hierarchy of
Chapter~\ref{ch:function-words}. It sits at the edge of clauses and utterances,
helps the speaker move the discourse along, and shows how readily spoken Iridian
recruits reduced forms into connective work.

\subsection{\ird{Žema} and reduced \ird{žma}}\label{sec:spoken-zema}

Among colloquial noun-like items, \ird{žema} is especially revealing. In
present-day spoken Iridian it is best analysed as a semantically bleached
colloquial lexeme with three closely related functions. First, it may serve as
a placeholder noun meaning roughly \trsl{thingy}, \trsl{that thing}, or
\trsl{what-do-you-call-it}. Secondly, it may function as a dismissive or
evaluative noun meaning \trsl{nonsense}, \trsl{excuses}, or \trsl{all that
fuss}. Thirdly, it may appear at the edge of an utterance as a playful hedge,
roughly \trsl{not really}, \trsl{just kidding}, or \trsl{I am only saying it
lightly}.

Alongside the fuller form \ird{žema}, colloquial speech also shows a more
reduced variant \ird{žma}. The reduced form is faster, more interactional, and
more typical of relaxed or intimate conversation, but it is not a different
lexeme with different core meanings. Rather, it is a spoken reduction of the
same colloquial item. Speakers may therefore choose \ird{žema} in clearer or
less reduced casual speech, while \ird{žma} is especially natural in rapid,
playful, or dismissive turns.

\subsubsection{Category and core values}

The basic grammatical point is that \ird{žema} behaves as a noun-like base in
ordinary Iridian syntax. In its placeholder use it fills a nominal slot without
contributing a stable lexical class of its own:

\pex
\a
\begingl
\gla Tak žema Janku koupnik.//
\glb this thingy Janek.\Trs{} buy-\Trs{}-\Pf{}//
\glft \trsl{Janek bought that thing / that whatchamacallit.}//
\endgl
\a
\begingl
\gla Tak žema.//
\glb this nonsense//
\glft \trsl{That is nonsense / that is just an excuse.}//
\endgl
\xe

In the first example, \ird{žema} is simply a placeholder noun occupying the
patient slot of an ordinary transitive clause. In the second, the same form is
interpreted dismissively: the speaker rejects what has just been said or done as
empty, unserious, or merely performative. The lexical content is therefore weak
in both uses, but the syntactic behaviour remains nominal.

\subsubsection{Plurality and expressive reduplication}

The nominal behaviour of the item becomes still clearer when it combines with
the plural clitic \ird{ne}. As elsewhere in the noun phrase
(§\,\ref{sec:nm-num}), \ird{ne} does not force a reading of neatly countable
plurality. With \ird{žema}, it usually denotes a loose aggregate: \trsl{all
those things}, \trsl{all that stuff}, or, in a more evaluative context,
\trsl{all that nonsense}.

\pex
\a
\begingl
\gla Ne žema Marek koupnik.//
\glb \Pl{} thingy Marek.\Trs{} buy-\Trs{}-\Pf{}//
\glft \trsl{Marek bought all that stuff.}//
\endgl
\a
\begingl
\gla Ne žemažema Marek koupnik.//
\glb \Pl{} thingy-thingy Marek.\Trs{} buy-\Trs{}-\Pf{}//
\glft \trsl{Marek bought all that random junk / all that nonsense stuff.}//
\endgl
\xe

The reduplicated form \ird{žemažema} adds an expressive nuance of vagueness,
theatricality, or dismissal. It is more overtly colloquial than plain \ird{žema}
and is especially natural when the speaker wishes to downplay the importance of
the referents or treat them as an undifferentiated heap of extras, details, or
irritations. In rapid speech the same pattern may surface as reduced
\ird{žmažma}, but the fuller reduplicated form is the clearest citation shape.

A further compressed form \ird{žežma} appears in particularly rapid or intimate
speech, resulting from syncope of the medial syllable of the full reduplicated
shape: \ird{žemažema} contracts to \ird{žežma}. This mirrors the reduction of
the simplex \ird{žema} to \ird{žma} and has a comparable sociolinguistic
distribution---casual, fast-paced, or interactionally marked registers---without
any change in meaning. The three members \ird{žemažema}, \ird{žmažma}, and
\ird{žežma} are therefore best treated as stylistic rather than semantic
variants of the reduplicated base.

Informal writing and colloquial lexicographic listings often record the plural
combinations as single orthographic words: \ird{nežma} (for \ird{ne} +
\ird{žma}) and \ird{nežežma} (for \ird{ne} + \ird{žežma}). These spellings
reflect the phonological cohesion of the plural clitic with the base in fast
speech; grammatically, \ird{ne} retains its ordinary clitic status. The
gradient of expressivity noted above is preserved in the plural: \ird{nežma}
is the more neutral aggregate plural (\trsl{all that stuff / all those things}),
while \ird{nežežma} is more overtly evaluative (\trsl{all that performative
nonsense / all that random fussing}), carrying into the plural the heightened
dismissiveness of the reduplicated base.

\subsubsection{Clause-peripheral hedge use}

The same colloquial base may also appear clause-peripherally, especially in the
reduced form \ird{žma}. In this use it no longer behaves like an ordinary noun
inside the clause, but as a parenthetical hedge by which the speaker playfully
withdraws, softens, or reframes what has just been said. Even so, it should not
be treated as a member of the formal particle hierarchy. Its contribution is
interactional and parenthetical rather than structural in the narrow sense.

\pex
\begingl
\gla Janek v ščenik, žma.//
\glb Janek.\Dir{} already arrive-\Pf{} hedge//
\glft \trsl{Janek has already arrived, not really / just kidding.}//
\endgl
\xe

This utterance-edge use follows naturally from the same semantic bleaching seen
in the nominal uses. The speaker introduces a proposition and then immediately
marks it as playful, non-committal, or not to be taken at full face value.
Synchronically, then, \ird{žema} and \ird{žma} form a small colloquial system:
an underspecified noun in the syntax, an evaluative noun in stance-marking
predication, and a parenthetical hedge in clause-peripheral discourse use.

\subsubsection{Verbal use and productive
recategorisation}\label{sec:spoken-zema-verbal}

The most grammatically productive extension of the \ird{žema} system is the
recruitment of the nominal base as a colloquial verbal stem. These verbal uses
are best understood not as a single fixed lexical entry but as an instance of
productive recategorisation: the nominal base is assigned a thematic class and
then set to work with ordinary Iridian valency and aspect-alignment morphology.
This is a familiar process in colloquial Iridian, where emotionally vivid or
semantically bleached bases are frequently pressed into verbal service without
any formal derivational affix mediating the transition.

The colloquial verbal stem is \ird{žemat-}, assigned to thematic class~I
(thematic consonant \ird{-t}), with antipassive stem \ird{žemač-}. The
assignment to class~I reflects a well-attested tendency for recently verbalised
colloquial bases to adopt the most common and unmarked thematic alternation. As
a verbal stem, \ird{žemat-} takes the full complement of valency and
aspect-alignment morphology described in Chapter~\ref{ch:verbs}. What remains
unusual is not the morphological behaviour---it is entirely transparent---but
the semantics: the lexical content of the stem is no more determinate than that
of the nominal base from which it derives.

In its actor-focused (antipassive) use, the stem describes the agent as engaging
in vague, performative, evasive, or unserious activity. The core meaning is most
readily glossed as \trsl{to mess around}, \trsl{to make excuses}, \trsl{to put
on a show}, or \trsl{to fuss about without achieving anything}, though the exact
nuance depends on context in the same way as the nominal base does. The
antipassive perfective \ird{žemačék} reports a completed episode of such
activity; the antipassive retrospective \ird{žemačac} presents it with
continuing present relevance and a characteristically evaluative charge---the
speaker frames the prior activity as having amounted to nothing or as having
been mere performance; the continuous \ird{žemata} describes it as an ongoing
state or habitual disposition:

\pex
\a
\begingl
\gla Anka žemačék.//
\glb Anka.\Dir{} žemat-\Antip{}-\Pf{}//
\glft \trsl{Anka was just messing around / making excuses.}//
\endgl
\a
\begingl
\gla Anka žemačac.//
\glb Anka.\Dir{} žemat-\Antip{}-\Ret{}//
\glft \trsl{Anka has been making nothing but excuses.}//
\endgl
\a
\begingl
\gla Anka žemata.//
\glb Anka.\Dir{} žemat-\Cont{}//
\glft \trsl{Anka is always fussing / is the type to make excuses.}//
\endgl
\xe

In its patient-focused (ergative) use, the stem \ird{žemat-} takes the
transitive marker \ird{-n-} and acquires a closely related but distinct reading:
\trsl{to handle carelessly}, \trsl{to give only a vague or makeshift treatment
to}, \trsl{to fudge}, or \trsl{to treat as a trivial matter}. In the ergative
aspects the agent appears in the transitive case and the patient in the
directional:

\pex
\a
\begingl
\gla Anka tak žematnik.//
\glb Anka.\Trs{} that.\Dir{} žemat-\Trs{}-\Pf{}//
\glft \trsl{Anka fudged that / handled that carelessly.}//
\endgl
\a
\begingl
\gla Anka tak žematnac.//
\glb Anka.\Trs{} that.\Dir{} žemat-\Trs{}-\Ret{}//
\glft \trsl{Anka has been treating that carelessly the whole time.}//
\endgl
\xe

The same verbal suffixes attach to the reduplicated base \ird{žemažema-}
(antipassive stem \ird{žemažemač-}) or to its compressed variant \ird{žežma-}
(antipassive stem \ird{žežmač-}). The reduplicated verbal is semantically
parallel to the reduplicated nominal: it is more overtly expressive and more
dismissive than the simple verbal form, conveying not merely an isolated episode
of vague activity but a sustained, layered, or theatrically amplified pattern of
performative moves:

\pex
\a
\begingl
\gla Anka žemažemačék.//
\glb Anka.\Dir{} \textsc{rdp}-žemat-\Antip{}-\Pf{}//
\glft \trsl{Anka went through all her usual vague fussing.}//
\endgl
\a
\begingl
\gla Anka žežmata.//
\glb Anka.\Dir{} \textsc{rdp}-žemat-\Cont{}//
\glft \trsl{Anka is always going through all that performative nonsense.}//
\endgl
\xe

Semantically, all verbal uses share a stable core even as contextual readings
vary widely. The predicate never names a concrete or specific action; it instead
attributes to the agent a \emph{stance}---vagueness, evasiveness, theatricality,
or dismissal---and that stance is precisely what the nominal base also
contributes. \ird{Žemačék} in one context means \trsl{she was just messing
around}; in another, \trsl{he was making excuses}; in a third, \trsl{they were
fussing over trivial extras}. What remains constant is the evaluative frame: the
speaker characterises the activity not by what was done but by how it was done,
and the \textit{how} is always some variant of unserious, unfocused, or merely
performative.

It should further be noted that the verbal morphology is entirely transparent.
The forms produced from \ird{žemat-} follow all ordinary Iridian verbal rules:
class~I mutation of the thematic consonant under the antipassive (\ird{-t} →
\ird{-č}), ergative-absolutive alignment in the perfective and retrospective,
nominative-accusative alignment in the continuous and progressive, and the
regular fusion of aspect and alignment into a single suffix. The colloquial
character of the item lies entirely in its origin and semantics; its
morphological behaviour is unremarkable. This is the hallmark of productive
recategorisation: a semantically bleached colloquial noun has entered verbal
morphology without requiring any special accommodation from the language's
ordinary morphological machinery.

\section{The \ird{ty}-supplement in spoken interaction}
\label{sec:spoken-ty}

The appositional particle \ird{ty} and the supplement construction it heads
are introduced in the nominal chapter (\S\,\ref{sec:nm-appos}). The present
section treats the specifically interactional properties of the
right-peripheral variant, which is substantially more productive in spoken
than in written language and which serves a distinct set of discourse functions
in everyday speech.

\subsection{Referent-first ordering as a spoken strategy}
\label{sec:spoken-ty-strategy}

Spoken language imposes real-time production constraints that written language
does not. A speaker who launches a noun phrase cannot always plan its full
modifier structure before the head noun must be uttered; equally, elaborating
descriptions may become relevant only as the discourse develops, after the
referent has already been introduced. The right-peripheral \ird{ty}-phrase
directly serves both situations. The speaker commits to a head noun,
establishes the referent, and then appends the clarification or specification
via \ird{ty} as a subsequent informational unit:

\pex
\a \ird{pres ty že vednikový i}
   \trsl{the man---that is, the one I saw}
\a \ird{nuž ty mrocná i}
   \trsl{the city---the beautiful one}
\a \ird{Karel ty mlazka}
   \trsl{Karel---my brother, you know}
\xe

\noindent This ordering reflects the referential logic of spoken discourse: the
head noun anchors the referent in the common ground and the \ird{ty}-phrase
adds identifying or descriptive content that the hearer can use to resolve or
enrich the reference. The construction is particularly valuable when the
describing material is long or syntactically complex, since a heavy prenominal
modifier stack---a nominalised relative clause, a multi-word adjectival phrase,
a postpositional elaboration---requires planning the entire modifier before
uttering the head noun. Right-peripheral \ird{ty} makes it possible to deliver
the head noun immediately and defer the specifying material.

\subsection{Repair and afterthought uses}
\label{sec:spoken-ty-repair}

The most distinctively spoken use of the \ird{ty}-supplement is as a repair
and afterthought device. A speaker may introduce a referent, continue the
utterance, and then append a \ird{ty}-phrase retroactively to specify it more
precisely:

\pex
\ird{pres vednik, ty dumě na i}
   \trsl{I saw the man---the one at the house}
\xe

% Add how to use with Not for disambiguation

In such cases the \ird{ty}-phrase is structurally detached from the preceding
clause or \synhead{NP}. It is not embedded inside the \synhead{NP} it
elaborates; it is appended to the right edge of the utterance as a freestanding
supplement, anchored to the earlier referent anaphorically via \ird{i}. The
prosodic signature is a short break before \ird{ty}---a slight pause or the
falling-then-rising contour typical of repair and self-correction---followed by
the supplement at a somewhat lower pitch level.

This repair-like use extends to cases where the \ird{ty}-phrase follows a
complete finite clause and reaches back to a referent introduced earlier in the
utterance:

\pex
\a \ird{Vednik, ty dumi na i}
   \trsl{I saw [him]---the one at the house}
\xe

Here \ird{ty} is not repairing an error but elaborating a
discourse referent that was already recoverable from context (via zero anaphora
in the main clause), offering the hearer additional identifying information as
an aside.

As described in \S\,\ref{sec:nm-appos-right}, the anaphoric pronoun \ird{i}
appears inside the \ird{ty}-phrase whenever the supplement's descriptive content
is non-nominal and requires a pronominal head to attach to. In spoken use this
is the normal case: spoken \ird{ty}-supplements typically contain adjectival or
relative-clause material, and \ird{i} serves as the pronominal host that allows
this material to form a well-headed nominal phrase while observing
head-finality. The resulting pattern, \ird{ty} \textsc{[modifier i]}, is highly
productive and has a fixed prosodic shape: \ird{ty} carries a mild stress, the
modifier is pronounced at its normal weight, and \ird{i} is typically reduced or
unstressed at the right edge of the supplement.

Right-peripheral \ird{ty}-phrases are unmarked in spoken interaction. Their
frequency, their use in repair contexts, and their role as a strategy for
deferring heavy modifier material all reflect production-driven motivations that
are native to spontaneous speech. In formal and written language the
construction remains available but is marked: a right-peripheral \ird{ty}-phrase
in writing carries a deliberate elaborative implication, foregrounding the
supplement as a rhetorically distinct informational act rather than presenting
it as an integrated modifier. Heavy adnominal material---nominalised relative
clauses that would be unwieldy prenominally---can motivate the construction even
in formal prose without special effect; a short adjective or a simple nominal
going postnominal in writing, by contrast, is interpretable as emphatic or
contrastive. The contrast with the unmarked prenominal strategy is always
available to the writer as a pragmatic resource.
