\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}
\markboth{abbreviations}{abbreviations}
\section*{Glossing Abbreviations}
\begin{longtabu} {>{\scshape}YY[2.5]}
	1		& first person\\
	2		& second person\\
	3		& third person\\
	4		& fourth person\\
	abl		& abilitative mood\\
	ade		& adessive\\
	agt		& agent\\
	anim	& animate\\
	att		& attributive\\
	ben		& benefactive focus\\
	caus	& causative\\
	comp	& comparative\\
	cond	& conditional\\
	ctpv	& contemplative aspect\\
	cv		& converb\\
	dem		& demonstrative\\
	dist	& distal\\
	excl	& exclusive\\
	expl	& expletive\\
	g		& generic number\\
	gen 	& genitive\\
	ger		& gerund\\
	hort	& hortative mood\\
	inan	& inanimate\\
	incl	& inclusive\\
	incp	& inceptive\\
	inf		& infinitive\\
	inst	& instrumental case\\
	ipf		& imperfective aspect\\
	iv		& instrumental focus\\
	loc		& locative\\
	lv		& locative focus\\
	med		& medial\\
	neg		& negative\\
	nz		& nominalizer\\
	opt		& optative mood\\
	pat		& patient\\
	perm	& permissive mood\\
	pf		& perfective aspect\\
	pl		& plural\\
	prosp	& prospective aspect\\
	proh	& prohibitive mood\\
	prox	& proximal\\
	pv		& patientive focus\\
	q			& question particle\\
	quot	& quotative mood\\
	ref		& reflexive focus\\
	ret		& retrospective aspect\\
	rz		& relativizer\\
	s		& singular\\
	sbj		& subjunctive mood\\
	str		& strong form\\
	sup		& supine\\
	wk		& weak form\\
\end{longtabu}

\clearpage

\section*{Abbreviation of Language Names}
\begin{longtabu} {YY[2.5]}
	Cz.		& Czech\\
	Eng.	& English\\
	Fr.		& French\\
	Ger.	& German\\
	Gk.		& Greek\\
	Hu.		& Hungarian\\
	It.		& Italian\\
	Lat.	& Latin\\
	OCS		& Old Church Slavonic\\
	Pol.	& Polish\\
	Rus.	& Russian\\
	Sl.		& Common Slavic/Slavonic\\
	Slk.	& Slovakian\\
	Uk.		& Ukrainian\\
\end{longtabu}


\section*{Other Symbols}
\begin{longtabu} {YY[2.5]}
	C 		& consonant\\
	C\sx{j}	& palatalized consonant\\
	D 		& voiced stop\\
	N 		& nasal consonant\\
	P 		& stop\\
	T 		& unvoiced stop\\
	V		& vowel\\
	Ṽ	 	& nasalized vowel\\
	V\tss{u}& unstable vowel\\
	\orth{~}& orthographic representation\\
	/~/		& phonemic transcription\\
	\nt{ }	& phonetic transcription\\
	\sim	& alternates with\\
\end{longtabu}
