\chapter{Semantics and Usage}

\section{Register}
\section{Politeness and forms of address}\index{form of address}\index{politeness}\label{sec:politaddr}

\subsection{Politeness and formality in Iridian}\index{politeness}\index{formality}

Although not as complex and as pervasive as the politeness/formality system found in Japanese\index{Japanese} or Korean\index{Korean}, Iridian formally encodes\index{markedness} more sociolinguistic information than its neigbouring languages such as Czech\index{Czech} or Hungarian\index{Hungarian}.

Broadly speaking, Iridian distinguishes between three levels of speech\index{level of speech}:\footnote{
	The English names are of course imperfect. It would perhaps be more correct,---if not more illustrative of their differences,---to call the polite speech level \emph{formal} and the formal speech level \emph{honorific}. What we call above as polite is more close to what linguists would call `formal' mainly because the strategy is one of distance and not deference. Moreover, although the formal speech level may be used to signal respect and shows a strong tendency to use honorifics and titles, the main usage remains that of showing an even greater detachment on the part of the speaker than would have otherwise have been possible when using the polite speech level.
} (1)~\emph{polite} speech, which serves more or less as the ``default'' level of politeness, as this is the speech level most often used by, say, strangers when talking to each other; (2)~\emph{formal} speech, which is used in more formal settings, where the speaker wants to distance themself from the listener or explicitly signal their politeness, such as in a conversation among business associates or when talking to a divinity; and (3)~\emph{casual} speech, which is used between close friends and family members, or to or among children. These levels of speech are not definite, of course, and politeness is more properly viewed as a spectrum (cf., e.g., \cite{hansonjap}) as speakers would often switch from one level of speech to another even when speaking to the same person, or within a single conversation.

The distinction between politeness\index{politeness} (which for the purpose of this grammar we can define as the psychological or social distance between speakers) on the one hand, and formality (which we can define as situational distance) on the other, is not always one made (or kept) in Iridian. Indeed, more often than not, these categories are often viewed by most speakers as essentially being the same. This is further complicated by the fact that the distinction between the various speech levels is not morphologically marked\index{markedness} but is facilitated instead by the preference for certain constructions and forms of address\index{form of address}.


The choice of which speech level to use with which speaker and in which scenarios is influenced by a lot of factors. It would be helpful, however, to analyse these factors as being influenced by two main considerations: the relationship---more specifically, the familiarity,---between the speakers, and the social setting in which the conversation or interaction is taking place.\footnote{
	One could take a look as well at the dimensions (or `semantics,' to use the authors' term) that influence the formality/politeness distinctions made in a language, proposed by \textcite{browngilman} in their study on the development of second-person pronouns and address forms. Although on the surface, the politeness distinction in Iridian is not dual, we see (as discussed \emph{infra}) that we can in fact classify the speech levels as either familiar (T) and distant (V). Where most Indo-European languages, however, predicate this distinction on the power semantic (i.e., the T-V distinction is made initially when a speaker of one power group speaks to a member of another), Iridian bases this initially on the solidarity semantics, thus creating a T-V distinction first when there is no solidarity (perceived or otherwise) between speakers, and only secondarily on the basis of the power semantic.
}

The first consideration, the relationship between speakers, divides the levels of speech into two groups: {\sc familiar}\index{familiar speech} speech, which consists of the casual speech level and {\sc distant} speech\index{distant speech} which consists of both the formal and polite speech levels. This distinction is perhaps of greater actual importance than that introduced earlier between the levels of speech, as the differences between familiar speech and distant speech are more pronounced than the differences between formal speech and polite speech, which are often more subtler. Distant speech is characterized by a preference to indirect speech acts where possible. For example, direct imperatives or prohibitives are virtually unused in distant speech, replaced instead with hortative constructions, or in more formal situations with questions or optative constructions. Consider for example the following:

\pex
\a Imperative in familiar speech:\\
\ird{Mina n\'avilastn\'im.} \trsl{Open the door!}
\a Alternative constructions in distant speech:
\beginsubsub
  \b{-} {Neutral, using the hortative:\\
  	\ird{Mina n\'avilastnik\'a.} \trsl{Please open the door.}}
  \b{-} {More polite, using \ird{am luhnink\'a}:\\
  	\ird{Mina se n\'avilastu am luhnink\'a}\\\trsl{May (you) be thanked because the door was closed.}}
  \b{-} {More formal and more polite, using a question:\\
  	\ird{Mina n\'avilastn\'i\v{s} to mu\v{z}nali\v{t}?}\\
  	\trsl{Is it possible that the door will be closed?}
  	}
\endsubsub
\xe

Perhaps a direct consequence of this preference for indirect speech acts over direct ones is the strategy of {\sc pronoun avoidance} so heavily employed in distant speech.


\subsection{Forms of address, titles, and honorifics}\index{honorific}\index{terms of courtesy|see{honorific}}\index{courtesy|see{honorific}}\index{form of address}

A {\sc honorific} is a form of address used to indicate respect or courtesy. The most common honorifics in
Iridian are the masculine \ird{St\'am} equivalent to the English\index{English} \trsl{Sir} and the feminine
\ird{Nau} equivalent to the English \trsl{Madame/Ma'am.} When addressing a person of an unknown
gender\index{gender}, the term \irdp{Obečne}{mercy/grace} is used.

Both \ird{St\'am} and \ird{Nau} may be followed by the addressee's last name. They should never be used with
the first name as it would be considered sarcastic or rude. In writing, these are abbreviated as \ird{S.}
and \ird{N.}, respectively. If the name of the person being addressed is not
known, the placeholders \irdp{vieda}{man} and \irdp{huzak}{woman} are used, thereby producing \ird{St\'am Vieda}
and \ird{Nau Huzak}. When writing, these are often abbreviated to {\sc s.v.} and {\sc n.h.}, respectively.
The usage of \ird{St\'am Vieda} and \ird{Nau Huzak} is similar to how the third person may sometimes be used
in English\index{English} to politely address someone (e.g., saying, \trsl{Will the gentleman yield?}) but while
it may sometimes appear dated or overly formal in English\index{English}, this practice is still commonly observed
in Iridian, especially when addressing strangers.

Other common titles include \ird{Doktor} used when addressing physicians, \ird{Majestet} or \ird{Kopi\v{z}n\'ast}
when addressing members of the royalty (with the latter reserved for reigning monarchs), \ird{Eselenc} when
addressing certain high-ranking officials such as senators, governors, and ambassadors, \ird{Eminenc} when addressing
cardinals of the Catholic Church, \ird{Obe\v{c}ne} or \ird{Prac} when addressing judges and magistrates, and
\ird{Tieho\v{z}n\'ast} or \ird{Hilda\v{z}n\'ast} or \ird{Hilden\'i T\'a\v{t}}\footnote{This form of address, meaning
\trsl{Holy Father} or more commonly its abbreviation {\sc h.t.}, is used in writing when referring to the Pope in the
third person.} when addressing the Pope or the religious leaders from other traditions.

When addressing or referring to multiple individuals the term \ird{maše} (originally meaning \trsl{crowd}
but now exclusively employed as a honorific) is used. This is often preceded, both in the written and spoken forms,
by the non-nominal supine\index{supine} \irdp{prehoda\v{s}ce}{esteemed/praiseworthy.}

\subsection{Salutations and valedictions in the written language}\index{salutation}\index{valediction}\index{written correspondence}

The general salutation in most formal correspondence uses the honorific \irdp{St\'am}{Sir} or \irdp{Nau}{Madame}.
The last name of the addressee may also follow, although more often than not, the simple honorific\index{honorific} 
should suffice. When addressing a collegiate entity or a collection of people, the term \irdp{Maše}{crowd} or
\irdp{Prehodašce maše}{Esteemed/praiseworthy crowd} is used instead.

If the addressee holds a specific title, the title is included in the salutation. In some cases, the wife of the
title-holder may be addressed using \ird{Nau} followed by the title, although this practice is slowly falling out
of use, except in most diplomatic correspondence, where it is still considered standard. Below are some examples:


\begin{itemize}[nosep]
	\item \irdp{St\'am/Nau Prezident}{Mister/Madame President}
	\item \irdp{St\'am/Nau Brac}{Mister/Madame Member of the Parliament}
	\item \irdp{St\'am/Nau Kanclár}{Mister/Madame Chancellor}
	\item \irdp{St\'am/Nau Holva}{Mister/Madame Chairman/Chairwoman}
	\item \irdp{St\'am/Nau Provízor}{Mister/Madame Professor}
\end{itemize}

Where the addressees are multiple individuals who hold specific titles, the honorific \ird{St\'am} or \ird{Nau} is replaced with \irdp{prehodašce}{esteemed, praiseworthy}. When used this way, the title is normally not capitalised. Note also that \ird{prehodašce} will only be used in a salutation when there are multiple addressees.

\begin{itemize}[nosep]
	\item \irdp{Prehodašce brac}{Esteemed members of the Parliament}
	\item \irdp{Prehodašce provízor}{Esteemed members of the faculty}
\end{itemize}

When the addressee is a medical doctor, the salutation \irdp{Doktor}{doctor} is used. When writing to members of the clergy, it is customary to use \irdp{Pápka}{My father} or \irdp{Mlazka}{My brother.}

It is considered rude to use a person's first name by itself in the salutation. A more common way is to add the suffix \irdp{-óm}{our} or \irdp{-(e)m}{my} to the name or the diminutive form of the name. Alternatively the terms \irdp{kamarád}{colleague, comrade} or \irdp{naž}{friend} or their diminutives may also be used. This approach is particularly common in e-mail correspondence between work colleagues.

Standard valedictions used in formal written correspondence\index{correspondence|see{written correspondence}} in Iridian tend to be more complex than the ones used in English. Below is 

\begin{itemize}[nosep]
	\item \irdp{(St\'am/Nau) oblostnení mavac/respekt akceptirniká}{Sir/Madame, please accept my sincerest regards (\emph{lit.}, wishes)/respect.}
	\item \irdp{Dá zespodení/spietnení pokár\'i biležit}{I will remain your most humble/loyal servant.}
	\item \irdp{Dá zespodení/spietnení byl\'i biležit}{I will remain your most humble child.}\footnote{This is often used among religious people when writing to members of the clergy.}
	\item \irdp{Oblostnení mavacu/respektu še hroznik.}{With the sincerest regards/respect has this letter been sent.}

\end{itemize}

Increasingly, especially in e-mail\index{e-mails} correspondence, it has become more common to use the following valedictions instead:

\begin{itemize}[nosep]
	\item \irdp{Mavac/\v{S}e mavacu}{Regards/with wishes/regards.}
	\item \irdp{Oblostnení}{Most sincere}
\end{itemize}

In more informal situations, such as between close friends and family, the following are used:

\begin{itemize}[nosep]
	\item \irdp{Dá}{I/me}
	\item \irdp{Bes/Mach bes/Nic bes}{Hug/Two hundred hugs/A thousand hugs}
	\item \irdp{Beska/Mach beska/Nic beska}{Little hug/Two hundred little hugs/A thousand little hugs}
	\item \irdp{\v{S}e hloubu/Hloubževí}{With love/Loving}
	\item \irdp{\v{Z}už/Mach žuž/Nic žuž}{Kiss/Two hundred kisses/A thousand kisses}
\end{itemize}

As mentioned earlier, specific examples of written correspondence in Iridian can be found in \S\,\ref{sec:writcorr}.


\section{Phatic Expressions and Social Formulas}

\section{Idiomatic Expressions}\index{idiomatic expressions}

\section{Punctuation}