\chapter{Sample Texts}

\section{The \emph{Pater Noster}}

\section{Milan Kundera, \ttla{A Kidnapped West or the Tragedy of Central Europe}}

{\small
% NOTE ON THE TRANSLATION
The translation is based on the French text of Kundera's essay \ttla{Un Occident kidnapp\'e: ou la trag\'edie de l'Europe centrale} first published in \ttlb{Le D\'ebat} in 1983. The full text is available online at various websites, with the link I used in the references. Due to copyright considerations, a translation has not been provided, although interlineal glosses and explanatory notes have been added where  I believe they are needed, in addition to the lexicon at the end. The text itself contains its own footnotes however and to distinguish Kundera's notes from those I have added, I have included included his name at their end.
}

\begin{center}1.\end{center}
1956 svem\'i Septembru Ma\v{z}arevn\'i Znova Byr\'ov\'i direktor\'am, byr\'o
nastol\'am jedno\v{c}nil ko obien\'i vniho minutu, ru\v{s}\v{c}evnie ur\'a\v{z}
po Budape\v{s}ta \v{s}el\v{c}ice to-\v{z}e t\'elexu laska mieta kudn\'i expedica
pashval\'ebik. Expedice to nie neitu uho\v{z}nek: >> M\'e za Ma\v{z}ar\'oma a
za Evropa shra\v{z}ach<<.

Nie neite je\v{z}e-no pr\'ones\v{c}ev\'i? Ma\v{z}ar\'oma a \v{s}e laska Evropu ru\v{s}\v{c}evn\'i \v{s}ar\'am zbavujinalu to \v{z}votu pr\'ones\v{c}ev\'i. Ma Evropa zbavujinale to je\v{z}e pr\'onestu?

Ma \v{z}en\'a --- >>za bl\'aha a za Evropa shra\v{z}\'a<< --- to \v{z}e Leningrada \v{z}e Mu\v{s}h\'ova z\'avadn\'etev\'i to neite, ma \v{c}e je Budape\v{s}ta, \v{c}e je Var\v{s}\'ava.

\begin{center}2.\end{center}
Vade, Evropa-te je\v{z}e-no za ona ma\v{z}arevna, ona \v{c}e\v{s}\v{c}evna, ona pol\v{s}\v{c}evna?
